% see localcommands for redefinition of \lsCollectionMetadataToBibliography

\title{Head-Driven Phrase Structure Grammar}
%\subtitle{The handbook\newlineCover\textit{\LARGE {\spaceskip=3.5pt Second revised edition} }}
%\subtitle{The handbook\newlineCover{} \mysmall{{\itshape Second revised edition}}}
\subtitle{The handbook\newlineCover{} {\itshape Second revised edition}}
%\subtitle{The handbook\newlineCover{} {Second revised edition}} 
\renewcommand{\lsImpressumCitationText}{%
Stefan Müller, Anne Abeillé, Robert D. Borsley \& Jean-Pierre Koenig (eds.).
2024. Head-Driven Phrase Structure Grammar: The handbook, 2nd revised edn.
(Empirically Oriented Theoretical Morphology and Syntax 9). Berlin: Language
Science Press.}

\BackBody{Head-Driven Phrase Structure Grammar (HPSG) is a linguistic framework that models   linguistic knowledge on all descriptive levels (phonology, morphology, syntax, semantics,   pragmatics) by using feature value pairs, structure sharing, and relational constraints. This   volume summarizes work that has been done since the mid 80s. Various chapters discus formal foundations and   basic assumptions, describe the evolution of the framework and go into the details of various   syntactic phenomena. Separate chapters are devoted to non-syntactic levels of description. The book also handles related fields and research areas (gesture, sign languages, computational   linguistics) and has a part in which HPSG is compared to other frameworks (Lexical Functional Grammar, Categorial Grammar, Construction Grammar, Dependency Grammar, and Minimalism).
 }
%\dedication{Change dedication in localmetadata.tex}
\typesetter{Stefan Müller, Elizabeth Pankratz}
\proofreader{Elizabeth Pankratz}
\author{Stefan Müller and Anne Abeillé and Robert D. Borsley and Jean-​Pierre Koenig}
\BookDOI{10.5281/zenodo.13637708}
\renewcommand{\lsISBNdigital}{978-3-96110-482-6}
\renewcommand{\lsISBNhardcover}{978-3-98554-111-9}
\renewcommand{\lsSeries}{eotms}
\renewcommand{\lsSeriesNumber}{9}
\renewcommand{\lsID}{478}
\renewcommand{\paperhivetext}{Errata:~\href{https://paperhive.org/documents/remote?type=langsci&id=259}{paperhive.org/documents/remote?type=langsci\&id=259}}
\renewcommand{\githubtext}{Source code available from \href{https://www.github.com/langsci/259}{www.github.com/langsci/259}}


\renewcommand{\lsCoverTitleFont}[1]{\sffamily\addfontfeatures{Scale=MatchUppercase}\fontsize{45.5pt}{15.1mm}\selectfont #1}

%\newcommand{\mysmall}[1]{{\small #1}}

%      <!-- Local IspellDict: en_US-w_accents -->
