%% -*- coding:utf-8 -*-

\documentclass[output=paper
                ,modfonts
                ,nonflat
	        ,collection
	        ,collectionchapter
	        ,collectiontoclongg
 	        ,biblatex
                ,babelshorthands
                ,newtxmath
                ,draftmode
                ,colorlinks, citecolor=brown
]{./langsci/langscibook}

\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
  \input{../localpackages}
  \input{../localcommands}

  \togglepaper[9]
}{}

\author{%
	Stephen Wechsler\affiliation{The University of Texas}%
	\and Jean-Pierre Koenig\affiliation{University at Buffalo}%
	\lastand Anthony Davis\affiliation{Southern Oregon University}%
}
\title{Argument structure and linking}

% \chapterDOI{} %will be filled in at production

%\epigram{}%We are up against one of the great sources of philosophical bewilderment: we must try to find a substance for a substantive.}
\abstract{In this chapter, we discuss the nature and purpose of argument structure in HPSG, focusing on the problems that theories of argument structure are intended to solve, including: (1) the relationship between semantic arguments of predicates and their syntactic realizations, (2) the fact that lexical items can occur in more than one syntactic frame (so-called valence or diathesis alternations), and (3) argument structure as the locus of binding principles. We also discuss cases where the argument structure of a verb includes more than predicted from the meaning of the verb, as well as rationales for a lexical approach to argument structure.}


\begin{document}
\maketitle
\label{chap:argumentstr}\label{chap-argumentstr}\label{chap-arg-st}

\section{Introduction}


For a verb or other predicator to compose with the phrases or pronominal affixes expressing its semantic arguments, the grammar must specify the mapping between the semantic participant roles \is{participant roles|see{thematic role, semantic role, proto-role}} and syntactic dependents of that verb.  For example, the grammar of English indicates that the subject of \word{eat} fills the eater role and the object of \word{eat}  fills the role of the thing eaten.  In HPSG, this mapping is usually broken down into two simpler mappings by positing an intermediate representation called \argst (`argument structure').  The first mapping connects the participant roles within the semantic \content with the elements of the value of the \argst feature; here we will call the theory of this mapping \emph{linking theory} (see Section \ref{linking-sec}).  The second mapping connects those \argst list elements to the elements of the  \val (`valence') lists, namely \comps (`complements') and \subj (`subject'; or \spr, `specifier'); we will refer to this second mapping as \emph{argument realization} (see Section \ref{sec:arg-st}).\footnote{Some linguists, such as \citet{LevinandRappaport2005}, use the term ``argument realization'' more broadly, to encompass linking as well.}  These two mappings \is{mapping|see{linking, argument realization}} are illustrated with the simplified lexical sign for the verb \word{eat} in (\ref{eat}) (for ease of presentation, we use a standard predicate-calculus representation of the value of \content in (\ref{eat}) rather than the attribute-value representation we introduce later on). 
%JPK 1-13-20 Just explained the notation 


\begin{exe} 
	\label{eat}
\ex	Lexical sign for the verb \word{eat}\\
{\avmoptions{center}
\begin{avm}
\[phon & \< $\text{\rm eat}$ \> \\
%TD 17 July: maybe /i:t/, or orth rather than phon
val & \[ subj \ \ & \<  {\@1} \> \\ 
comps \ \ & \< {\@2}  \> \] \\ 
arg-st & \< {\@1}\textsc{np}$_i$ , {\@2}np$_j$ \> \\	
content  \ \ & \textit{eat}($i, j$)\] 
\end{avm}}
\end{exe}

	
\noindent
In (\ref{eat}), ``NP'' abbreviates a feature structure representing syntactic and semantic information about a nominal phrase.  The variables $i$ and 
$j$ are the referential indices for the eater and eaten arguments, respectively, of the \textit{eat} relation.  The semantic information in 
NP$_i$ semantically restricts the value or referent of $i$. 

The \argst feature plays an important  
role in HPSG grammatical theory.  In addition to regulating the mapping from semantic arguments to grammatical relations, \argst is the locus of the theories of anaphoric binding and other construal relations such as control and raising.  (This chapter focuses on the function of \argst  in semantic mapping, with some discussion of binding and other construal relations only insofar as they interact with that mapping.  A more detailed look at binding is presented in \crossrefchapteralt{binding}.)   

In HPSG, verb diathesis alternations, voice alternations, and derivational processes such as category conversions are all captured within the lexicon (see Section~\ref{alternations} and \crossrefchapteralt{lexicon}).  The different variants of a word are grammatically related either through lexical rules or by means of the lexical type hierarchy.  HPSG grammars explicitly capture paradigmatic relations between word variants, making HPSG a \textit{lexical approach to argument structure}, in the sense of \citet{MWArgSt}.
This fundamental property of lexicalist theories contrasts with many transformational approaches, where such relationships are treated as syntagmatically related through operations on phrasal structures representing sentences and other syntactic constituents.  Arguments for the lexical approach are reviewed in Section \ref{lexicalapproach}.  

Within the HPSG framework presented here, we will formulate and address a number of empirical and theoretical questions: 

\begin{itemize}
\item We know that a verb's meaning influences its valence requirements (via the \argst list, on this theory). 
 What are the principles governing the mapping from \content to \argst ?  Are some aspects of \argst idiosyncratically stipulated for individual verbs?  Which aspects of the semantic \content  bear on the value of \argst, and which aspects do not?  (For example, what is the role of modality?)  
\item How are argument alternations defined with respect to our formal system?  For each alternation we may ask which of the following it involves: a shuffling of the \argst list;  a change in the mapping from \argst to \val; or  a change in the \content, with a concomitant change in the \argst?  
\end{itemize}

\noindent
These questions will be addressed below in the course of presenting the theory.  We begin by considering \argst itself (Section \ref{sec:arg-st}), followed by the mapping from \argst to \val (Section \ref{argst-valence-sec}), and the mapping from \content to \argst (Section \ref{linking-sec}).  \is{mapping|see{linking, argument realization}}
The remaining sections address further issues relating to argument structure: the nature of argument alternations, extending the \argst attribute to include additional elements, whether \argst is a universal feature of languages, and a comparison of the lexicalist view of argument structure presented here with phrasal approaches.


\section{The representation of argument structure in HPSG}

\label{sec:arg-st}

In the earliest versions of HPSG, the selection of dependent phrases was specified in the \subcat feature \is{SUBCAT} of the head word (\citealt{pollard+sag:1987}, \citealt[Chapters~1--8]{pollard+sag:1994}).  The value of \subcat is a list of items, each of which corresponds to the \synsem value of a complement or subject.  The following are \subcat features for an intransitive verb, a transitive verb, and a transitive verb with obligatory PP complement:


\begin{exe} 
\ex \label{subcats}
\begin{xlist}
\ex	\word{laugh}:  $[$ \subcat $\langle$ \textsc{np} $\rangle ]$
\ex    \word{eat}:  $[$ \subcat $\langle$ \textsc{np}, \textsc{np} $\rangle ]$
\ex    \word{put}:  $[$ \subcat $\langle$ \textsc{np}, \textsc{np}, \textsc{pp} $\rangle ]$
\end{xlist}
\end{exe}

\noindent
Phrase structure rules in the form of immediate dominance schemata \is{immediate dominance schema} identify a certain daughter node as the head daughter (\textsc{head-dtr}) and others, including subjects, as complement daughters (\textsc{comp-dtrs}).  In keeping with the \emph{Subcategorization Principle}, \is{Subcategorization Principle} here paraphrased from \cite[34]{pollard+sag:1994}, list items are effectively ``cancelled'' from the \subcat list as complement phrases, including the subject, are joined with the selecting head:

\begin{exe}
\ex Subcategorization Principle: In a headed phrase, the \subcat value of the \headdtr (`head daughter') is the concatenation of the phrase's \subcat list with the list of \synsem values of the \compsdtrs (`complement daughters').
\end{exe}

\noindent
Phrasal positions are distinguished by their saturation level: ``VP'' is defined as a verbal projection whose \subcat list contains a single item, corresponding to the subject, and ``S'' is defined as a verbal projection whose  \subcat list is empty. 

\is{ARG-ST!as locus of binding}
The ``subject'' of a verb, a distinguished dependent with respect to construal processes such as binding, control, and raising, was then defined as the first item in the  \subcat list, hence the last item with which the verb combines.   However, defining ``subject'' as the last item to combine with the head proved inadequate \citep[Chapter~9]{pollard+sag:1994}.  There are many cases where the dependent displaying subject properties need not be the last item added to the head projection.  For example, in German the subject is a nominal in nominative case \citep{Reis82}, but the language allows subjectless clauses containing only a dative or genitive non-subject NP.  If that oblique \is{oblique argument} NP is the only NP dependent to combine with the verb, then it is \emph{ipso facto} the last NP to combine, yet such obliques lack the construal properties of subjects in German.  

Consequently, the \subcat list was split into two valence lists, a \subj list of length zero or one for subjects, and a \comps list for complements.  Nonetheless, certain grammatical phenomena, such as binding and other construal processes, must still be defined on a single list comprising both subject and complements \citep{Manning+Sag:1999}. Additionally, some syntactic arguments are unexpressed or realized by affixal pronouns, rather than as subject or complement phrases.  
The new list containing all the syntactic arguments of a predicator was named \textsc{arg-st} (`argument structure').  

In clauses without implicit or affixal arguments, the \textsc{arg-st} is the concatenation of  \textsc{subj} \isfeat{SUBJ} and \textsc{comps} \isfeat{COMPS} respectively.  For example, the \subcat list for \word{put} in (\ref{subcats}c) is replaced with the following:

\begin{exe} 
	\label{put}
\ex	%\textit{put}: \\
\begin{avm}
[ phon & < \rm put > \\
val & [ subj \ \ & <  \ {@1} \ > \\ 
comps \ \ & < \  {@2}, {@3} \  > ] \\ 
arg-st & < \ {@1}np , {@2}np, {@3}pp \  > ] 
\end{avm}
\end{exe}

\noindent
The idealization according to which \argst is the concatenation of \subj and \comps is canonized as the \emph{Argument Realization Principle} (ARP) \is{Argument Realization Principle} \citep[494]{SWB2003a}.  Systematic exceptions to the ARP, that is, dissociations between \textsc{valence} and \argst, are discussed in Section \ref{argst-sec} below.  

A predicator's \textsc{valence} lists indicate its requirements for syntactic concatenation with phrasal dependents (Section \ref{argst-valence-sec}). 
\argst, meanwhile, provides syntactic information about the expression of semantic roles and is related, via linking theory, to the lexical semantics of the word (Section \ref{argst-sec}).  
The \argst list contains specifications for the union of the verb's local phrasal dependents (the subject and complements, whether they are semantic arguments, raised phrases, or expletives) and its arguments that are not realized locally, whether they are unbounded dependents, affixes, or unexpressed arguments.\is{VALENCE!as distinct from ARG-ST} 
\is{ARG-ST!as distinct from VALENCE} 

Figure \ref{fig:over} provides a schematic representation of %``typical'' 
linking and argument realization in HPSG,  illustrated with the verb \textit{donate}, as in \textit{Mary donated her books to the library}.   Linking principles govern the mapping of participant roles in a predicator's \content to %direct 
\is{mapping|see{linking, argument realization}}
elements of the \argst list.   Argument realization is shown in this figure only for mapping to \val, which represents locally realized phrasal dependents; affixal and null arguments are not depicted (but are discussed below).
%Here, the semantic roles are just arbitrary labels, but we discuss in Section (\ref{linking-sec}) how they can be systematically related to lexical entailments of predicators.
The \argst and \val lists in this figure contain only arguments linked to participant roles, \is{participant roles|see{thematic role, semantic role, proto-role}}  but in Section \ref{sec:extended-arg-st} we discuss proposals for extending \argst to include additional elements.
In Section \ref{argst-valence-sec}, we examine cases where the relationship between \argst and \val
violates the ARP.\todostefan{Figure~\ref{fig:over} has to be redone.}

\begin{figure}[htbp!]\footnotesize
% this really does not work with externaliaztion 02.05.2020
\tikzexternaldisable
\begin{tabular}{p{3cm}p{4.5cm}p{4cm}}
	Semantics (\attrib{content}) & {\avmoptions{center}\begin{avm}
 \[\asort{$\textit{donate-rel}$}
 donor	& $i$ \\
 recipient & $j$ \\
 theme & $k$   \]
 \end{avm}} & Semantic relation denoted by a verb\\
  & \Large{\begin{tikzcd}
 {} \arrow[leftrightarrow]{d}{\,\,\,\text{\textit{Linking Principles}}}
               \\ 
               {}
\end{tikzcd} } &  \\
& \\
	Argument Structure (\attrib{arg-st}) & 
	{\avmoptions{center}\begin{avm}
	\[arg-st &  \< \@1\textsc{np}$_i$, \@2\textsc{np}$_{k}$, \@3\textsc{pp}[$to$]$_{j}$ \> \]	
 \end{avm}} & List of syntactic arguments of the verb \\
 & \Large{\begin{tikzcd}
 {} \arrow[leftrightarrow]{d}{\,\,\,\text{\textit{Argument Realization Principles}}}
               \\ 
               {}
\end{tikzcd} } & \\
	Syntax (\attrib{valence}) & {\avmoptions{center}\begin{avm}
 \[ subj & \< \@1\textsc{np}$_{i}$ \> \\
 comps & \< \@2\textsc{np}$_{k}$, \@3\textsc{pp}[$to$]$_{j}$ \> \] 	
 \end{avm}}   
 & Lists of locally realized phrasal dependents
\end{tabular}
\caption{\label{fig:over}Linking and argument realization in HPSG, illustrated with the verb \textit{donate}} 
\end{figure}



\section{Argument realization: The mapping between \argst and \val lists}
\label{argst-valence-sec}

\subsection{Variation in the expression of arguments}
\label{express-sec}

The \val feature is responsible for composing a verb with its phrasal dependents, but this is just one of the ways that semantic arguments of a verb are expressed in natural language.  Semantic arguments can be expressed in various linguistic forms: as local syntactic dependents (\subj and \comps), as affixes, or displaced in unbounded dependency constructions (\textsc{slash}). 

Affixal arguments can be illustrated with the first person singular Spanish verb \textit{hablo} `speak.\textsc{1sg}', as in \ref{hablo}.


\begin{exe} 
\ex	\label{hablo}
\begin{xlist}
\ex 		\gll Habl-o espa\~{n}ol.  \\
		speak-\textsc{1sg} Spanish  \\
		\glt `I speak Spanish.'
\ex \textit{hablo} `speak.\textsc{1sg}': \\
{\avmoptions{center}
\begin{avm}
\[ phon & \< $\text{\rm ablo}$ \> \\
val & \[ subj & \<  \ \  \> \\ 
comps & \< \@2   \> \] \\ 
arg-st & \< np:\[$ppro$ \\ index \[pers & $1st$ \\ num & $sg$ \] \]  , \@2np  \>  \]
\end{avm}}
\end{xlist}
\end{exe}

\noindent
The \textit{-o} suffix contributes the first person singular pronominal subject content to the verb form (the morphological process is not shown here; see \crossrefchapteralt{morphology}).  The pronominal subject  appears on the \argst list and hence is subject to the binding theory.  But it does not appear in \subj , if no subject NP appears in construction with the verb.   

A lexical sign whose \argst list is just the concatenation of its \subj and \comps lists conforms to the Argument Realization Principle \is{Argument Realization Principle} (ARP); such signs are called \word{canonical signs}\is{canonical sign} by \citet{Boumaetal2001}.  Non-canonical signs, which violate the ARP, have been approached in two ways.  In one approach, a lexical rule takes as input a canonical entry and derives a non-canonical one by removing items from the \val lists, while adding an affix or designating an item as an unbounded dependent by placement on the \textsc{slash} list.  \is{VALENCE!as distinct from ARG-ST} \is{ARG-ST!as distinct from VALENCE} 
In the other approach, a feature of each \argst list item specifies whether the item is subject to the ARP (hence mapped to a \val list), or ignored by it (hence expressed in some other way).  
See the chapter on the lexicon for more detail and \citet{MillerandSag1997} for a treatment of French clitics as affixes. 

A final case to consider is null anaphora, in which a semantic argument is simply left unexpressed and receives a definite pronoun-like interpretation.  Japanese \textit{mi-} `see' is transitive but the object NP can be omitted as in (\ref{jap}).

\begin{exe}
	\ex\label{jap}
		\gll Naoki-ga mi-ta.  \\
		Naoki-\ig{nom} see-\ig{pst}  \\
		\glt `Naoki saw it/him/her/*himself.'
\end{exe} 

\noindent
Null anaphors of this kind typically arise in discourse contexts similar to those that license ordinary weak pronouns, and the unexpressed object often has the obviation effects characteristic of overt pronouns, \is{obviation} as shown in (\ref{jap}).  HPSG eschews the use of silent formatives like `small \textit{pro}' when there is no evidence for such items, such as local interactions with the phrase structure.  Instead, null anaphors of this kind are present in \argst but absent from \val lists.  \argst is directly linked to the semantic \content and is the locus of binding theory, so the presence of a syntactic argument on the \argst list but not a \val list
 accounts for null anaphora.\is{ARG-ST}  To account for obviation, the \argst list item, when unexpressed, receives the binding feature of ordinary (non-reflexive) pronouns, usually \textit{ppro}.  This language-specific option can be captured in a general way by \val and \argst defaults in the lexical hierarchy for verbs.   

\subsection{The syntax of \argst and its relation to \textsc{valence}}
\label{argst-sec}


The ordering of members of the \argst list represents a preliminary syntactic structuring of the set of argument roles.  In that sense, \argst functions as an interface between the lexical semantics of the verb and the expressions of dependents as described in Section \ref{argst-valence-sec}.  Its role thus bears some relation to the initial stratum in Relational Grammar \citep{PerlmutterandPostal1984}, \is{Relational Grammar} \textit{argument structure} (including intrinsic classifications) in LFG Lexical Mapping Theory \is{Lexical Functional Grammar} \citep{Bresnan+etal:2015}, macroroles in Role and Reference Grammar \citep{VanValinandLapolla1997}, D-structure in Government/Binding theory, and the Merge positions of arguments in Minimalism, \is{Minimalism} assuming in the last two cases the Uniform Thematic Alignment Hypothesis \citep{Baker1988} or something similar.  However, it also differs from all of those in important ways.  


Semantic constraints on \argst are explored in Section \ref{linking-sec} below.  But \argst represents not only
% is  structured not only by    
semantic distinctions between the arguments, but also  %by 
syntactic  ones.  Specifically, the list ordering represents relative syntactic \textit{obliqueness} of arguments.   The least \isi{oblique argument} is the subject (\subj), followed by the complements (\comps).  Following \citet{Manning1996}, term arguments (direct arguments, i.e., subjects and objects) are  assumed to be less oblique than ``oblique'' arguments (adpositional and oblique case marked phrases), followed finally by predicate and clausal complements.  The transitive ordering relation on the \argst list is called \textit{o-command} (`obliqueness command'):\is{o-command|see{anaphoric binding}} \is{anaphora|(} the list item that corresponds to the subject o-commands those corresponding to complements; a list item corresponding to an object o-commands \is{o-command|see{anaphoric binding}} those corresponding to any obliques; and so on (see \crossrefchapteralt{binding} for details).
 
% TD 18 Jan. 20: I think the following paragraph and examples could just be deleted.  They presuppose the non-canonical linking analysis we're skeptical of later, and they don't really seem to usefully illustrate or advocate anything.
% \is{alternation}
%Voice alternations like the passive \is{passive}  are sometimes defined on the \argst list and illustrate the ordering of terms before obliques on the \argst list.   Passivization, in some analyses that go back to \citet{pollard+sag:1987}, alters the syntactic properties of \argst list items: the initial item of the active, normally mapped to \subj of the active, is an oblique (\textit{by} phrase) or unexpressed argument in the passive.  Given that terms precede obliques in the list order, any term arguments must o-command the passive oblique, so passive effectively reorders the initial item in \argst to a list position following any terms.   
%JPK 1-15-20 I changed the wording in the paragraph above to reflect our more complex story about the passive

%\begin{exe}
%\ex \label{passive}
%\begin{xlist}
%\ex Susan gave Mary a book.
%\ex Mary was given a book by Susan.
% \end{xlist}
% \end{exe}

%\begin{exe}
%\ex \label{pasargst}
%\begin{xlist}
%\ex \textit{give} (active): $[ \textsc{arg-st}  < \textsc{np}_i, \textsc{np}_j, \textsc{np}_k > ]$
%\ex \textit{given} (passive): $[ \textsc{arg-st}  < \textsc{np}_j, \textsc{np}_k, \textsc{pp}[by]_i > ]$
% \end{xlist}
% \end{exe}

Relative obliqueness conditions a number of syntactic processes and phenomena, including anaphoric binding.  The o-command relation replaces the c"=command in the Principles A, B, and C of Chomsky's \citeyearpar{Chomsky:1981} configurational theory of binding.  For example, HPSG's Principle B states that an ordinary pronoun cannot be o-commanded by its coargument antecedent, which accounts for the pronoun obviation observed in the English sentence \textit{Naoki$_i$ saw him$_{*i/j}$}, and also accounts for obviation in the Japanese sentence (\ref{jap}) above.  

Relative obliqueness also conditions the \isi{accessibility hierarchy} of \citet{KeenanandComrie1977}, according to which a language allowing relativization of some type of dependent also allows relativization of any less oblique than it.  Hence if a language has relative clauses at all, it has subject relatives; if it allows obliques to relativize, then it also allows subject and object relatives; and so on.  
Similar implicational universals apply to verb agreement with subjects, objects, and obliques \is{oblique argument} \citep{greenberg:1966}.  \is{anaphora|)}

Returning now to argument realization, we saw above that the rules for the selection of the subject from among the verb's arguments are also stated in terms of the \argst list.  In a canonical\ realization the subject is the first list item, o-commanding all of its coarguments.
% TD 10 Jan. 20: added the following to transition to the next section
In various non-canonical circumstances, such as those we noted above, o-command relations do not correspond to ordering on the valence lists, and this can be reflected in phenomena such as anaphoric binding.
In the following section we examine another kind of non-canonical relationship between \argst and \val in more detail: syntactic ergativity, exemplified by Balinese.

\subsection{Syntactic ergativity}
\label{ergativity}\label{arg-st-sec-ergativity}
\is{ergativity|(}
The autonomy of \argst from the \val lists is further illustrated by cross"=linguistic variation in the mapping between them.  As just noted, in English and many other languages, the initial item in \argst maps to the subject.  However,  languages with so-called \emph{syntactically ergative} clauses have been analyzed as following a different mapping rule.  Crucially, the \argst ordering in those languages is still supported by independent evidence from properties such as binding and NP versus PP categorial status of arguments.
Balinese (Austronesian), as analyzed by \citet{Wechsler+Arka:1998}, is such a language.  In the morphologically unmarked, and most common voice, called \emph{Objective Voice}\is{voice!objective} (\attrib{ov}), the subject is any term \textit{except} the \argst-initial one.   

\il{Balinese|(} Balinese canonically has SVO order, regardless of the verb's voice form \citep{Artawa1994, Wechsler+Arka:1998}.  The preverbal NPs in  (\ref{bal1}) are the surface subjects and the postverbal ones are complements.  When the verb appears in the unmarked objective voice (\attrib{ov}), a non-initial term is the subject, as in (\ref{bal1-a}).    But verbs in the \emph{Agentive Voice}\is{voice!agentive} (\attrib{av}) select as their subject the \argst{}-initial item, as in (\ref{bal1-b}).  

\begin{exe}
	\ex\label{bal1}
\begin{xlist}
\ex \label{bal1-a}	\gll Bawi adol ida.  \\
		pig \textsc{ov}.sell \textsc{3sg}   \\
		\glt `He/She sold a pig.'
\ex	\label{bal1-b}\gll Ida ng-adol bawi.  \\
		\textsc{3sg} \textsc{av}-sell pig   \\
		\glt `He/She sold a pig.'
\end{xlist}
\end{exe} 

\noindent
A ditransitive verb, \is{ditransitive} such as the benefactive applied form of \textit{beli} `buy' in (\ref{bal2}), has three term arguments on its \argst list.  The subject can be either term that is non-initial in \argst{}:

\begin{exe}
	\ex\label{bal2}
\begin{xlist}
\ex 	\gll Potlote ento beli-ang=a I Wayan.  \\
		pencil-\textsc{def} that \textsc{ov}.buy-\textsc{appl}=3 \textsc{art} Wayan   \\
		\glt `(s)he bought Wayan the pencil.'
\ex 	\gll I Wayan beli-ang=a potlote ento.   \\
		\textsc{art} Wayan \textsc{ov}.buy-\textsc{appl}=3 pencil-\textsc{def} that   \\
		\glt `(s)he bought Wayan the pencil.'
\end{xlist}
\end{exe} 

\noindent
Wechsler and Arka argue that Balinese voice alternations \is{alternation} do not affect \argst list order. 
Thus the agent argument can bind a coargument reflexive pronoun (but not vice versa), regardless of whether the verb is in OV or AV form:

\begin{exe}
	\ex\label{bal3}
\begin{xlist}
\ex 	\gll Ida ny-ingakin ragan idane. \\
		\textsc{3sg} \textsc{av}-see self\\
		\glt ‘(s)he saw himself/herself’
\ex 	\gll {Ragan idane} cingakin ida. \\
		self \textsc{ov}.see \textsc{3sg} \\
		\glt ‘(s)he saw himself/herself’
\end{xlist}
\end{exe} 

\noindent
The `seer' argument o-commands \is{o-command|see{anaphoric binding}} the `seen', with the AV versus OV voice forms regulating subject selection:\is{anaphora}

\begin{exe} 
	\label{avsee}
\ex	Agentive Voice form of `see': \\
{\avmoptions{center}
\begin{avm}
\[ phon & \< $nyinkagin$ \> \\
val & \[ subj & \<  \@1  \> \\ 
comps & \< \@2   \> \] \\ 
arg-st & \< \@1np$_{i}$  , \@2np$_j$ \> \\
content  & \[\asort{see-rel}   seer & i \\ seen & j \]
  \] 
\end{avm}}
\end{exe}

\begin{exe} 
	\label{ovsee}
\ex	Objective Voice form of `see': \\
{\avmoptions{center}
\begin{avm}
\[ phon & \< $cinkagin$ \> \\
val & \[ subj  & \<  \@2 \> \\ 
comps & \< \@1 \> \] \\ 
arg-st & \< \@1np$_{i}$  , \@2np$_j$ \> \\
content  & \[ \asort{see-rel}   seer & i \\ seen & j \]
  \] 
\end{avm}}
\end{exe}

\noindent
Languages like Balinese illustrate the autonomy of \argst .  
Although the agent binds the patient in both (\ref{bal3}a) and (\ref{bal3}b),  the binding conditions cannot be stated directly on the \is{thematic hierarchy|see{thematic roles}} thematic hierarchy.  For example, in HPSG a raised argument appears on the \argst list of the raising verb, even though that verb assigns no thematic role \is{thematic role|see{semantic role, participant role, thematic hierarchy}} to that list item.  But a raised subject can bind a coargument reflexive in Balinese (this is comparable to English \textit{John seems to himself to be ugly}).  Anaphoric binding in Balinese raising constructions thus behaves as predicted by the \argst based theory \citep{Wechsler1999}.  
In conclusion, neither \val nor \content provides the right representation for defining binding conditions, but \argst fits the bill.  


Syntactically ergative languages besides Balinese that have been analyzed as using an alternative mapping between \argst and \val include Tagalog, Inuit, some Mayan languages, Chukchi, Toba Batak, Tsimshian languages, and Nad{\"e}b \citep{Manning1996,Manning+Sag:1999}.  

Interestingly, while the GB/Minimalist configurational binding theory may be defined on analogues of \val or \content,  those theories lack any analogue of \argst.  This leads to special problems for such theories in accounting for binding in many Austronesian languages like Balinese.  In transformational theories since \citet{Chomsky:1981}, anaphoric binding conditions are usually stated with respect to the A-positions (`argument positions').  A-positions are analogous to HPSG \val list items, with relative c"=command in the configurational structure corresponding to relative list ordering in HPSG, in the simplest cases.  Meanwhile, to account for data similar to (\ref{bal3}), where agents asymmetrically bind patients, Austronesian languages like Balinese were said to define binding on the ``thematic structure'' encoded in d-structure or Merge positions, where agents asymmetrically c"=command patients regardless of their surface positions \citep{Guilfoyle+etal:1992}.  But the interaction with raising shows that neither of those levels is appropriate as the locus of binding theory \citep{Wechsler1999}.\footnote{To account for (\ref{bal3}b) under the configurational binding theory, the subject position must be an A-bar position, but to account for binding by a raised subject, it must be an A-position.  See \citet{Wechsler1999}. } 

\il{Balinese|)}

\is{ergativity|)}

\subsection{Symmetrical objects}
\is{symmetrical object}
We have thus far tacitly assumed a total ordering of elements on the \argst list, but \citet{AMM2013a}, \citet{Ackermanetal2017} propose a partial ordering
\is{ARG-ST!partial ordering on}
 for certain so-called \emph{symmetrical object languages}.  In \ili{Moro} (Kordofanian), the two term complements of a ditransitive \is{ditransitive} verb have exactly the same object properties.  Relative linear order of the theme and goal arguments is free, as shown by the two translations of (\ref{moro}) (from \citealt[9]{Ackermanetal2017}; \attrib{cl} `noun class'; \attrib{sm} `subject marker').


\begin{exe}
	\ex\label{moro}
\gll   \'{e}-g-a-nat\textipa{S}-\'{o} \'{o}r\'{a}\textipa{N}  \textipa{N}e\textipa{R}\'{a}  \\
        1\textsc{sg.sm-cl}g-\textsc{main}-give-\textsc{pfv}    \textsc{cl}g.man \textsc{cl}\textipa{N}.girl \\
\glt `I gave the girl to the man.’ / `I gave the man to the girl.’
\end{exe} 

\noindent
More generally, the two objects have identical object properties with respect to occurrence in post-predicate position, case marking, realization by an object marker, and ability to undergo passivization \citep[9]{Ackermanetal2017}.

\citet{Ackermanetal2017} propose that the two objects are unordered on the \argst list.  This allows for two different mappings to the \comps list, as shown here:

\begin{exe} 
\ex		\label{moro-avm1}
\begin{xlist}
\ex Goal argument as primary object: \\
{\avmoptions{center}
\begin{avm}
\[ val & \[ subj & \<  \@1 \> \\ 
comps & \< \@2 , \@3  \> \] \\ 
arg-st & \<  \@1np$_{i}$  ,\{ \@2np$_{j}$ , \@3np$_{k}$ \}   \> \\
content  & \[ \tp{give-rel}  \\ agent & i \\ goal & j \\ theme & k \]
  \] 
\end{avm}}
\ex Theme argument as primary object: \\
{\avmoptions{center}
\begin{avm}
\[ val & \[ subj & \<  \@1 \> \\ 
comps & \< \@3 , \@2 \> \] \\ 
arg-st & \< \@1np$_{i}$  , \{ \@2np$_{j}$ , \@3np$_{k}$  \} \> \\
content  & \[ \tp{give-rel}  \\ agent & i \\ goal & j \\ theme & k \]
  \] 
\end{avm}}
\end{xlist}
\end{exe}

\noindent
The primary object properties, which are associated with the initial term argument of \comps, can go with either the goal or theme argument. 

To summarize this section, while the relationship between \argst, \subj, and \comps lists was originally conceived as a straightforward one, enabling binding principles to maintain their simple form by defining \argst as the concatenation of the other two, the relationship was soon loosened.
%Non-canonical 
Looser
relationships between \argst and the \val lists are invoked in accounts of several core syntactic phenomena.
Arguments not realized overtly in their canonical positions 
due to extraction, cliticization, or pro-drop (null anaphora) \is{anaphor!null|see{obviation}}appear on \argst but not in any \val list.  
Accounts of syntactic ergativity in HPSG involve variations in the mapping between \argst and \val lists; in particular, the element of \subj is not, in such languages, the first element of \argst.  
Modifications of \argst play a role in some treatments of passivization, where its expected first element is suppressed, and in languages with multiple, symmetric objects, where a partial rather than total ordering of \argst elements has been postulated (see Section \ref{passives} for details on the analysis of passives in HPSG).
Thus \argst has now acquired an autonomous % independent 
status within HPSG, and is not merely a predictable rearrangement of information present in the valence lists.  
%present elsewhere in lexical entries.


\section{Linking: the mapping between semantics and \argst}
\label{linking-sec}
\is{ARG-ST!as locus of linking|see{linking}|(}

\subsection{HPSG approaches to linking}

The term \textit{linking} refers to the mapping specified in a lexical entry between participant roles \is{participant roles|see{thematic role, semantic role, proto-role}} in the semantics and their syntactic representations on the \argst list.  
Early HPSG grammars stipulated the linking of each verb:  semantic \content values with predicator-specific attributes like \textsc{devourer} and \textsc{devoured} were mapped to the subject and object, respectively, of the verb \textit{devour}.  But linking follows general patterns across verbs, and across languages; e.g., if one argument of a transitive verb in active voice has an agentive role, it will map to the subject, not the object, 
except in syntactically ergative languages described in Section \ref{ergativity} above, and in those languages the linking is just as regularly reversed.
Those early HPSG grammars did not capture the regularities across verbs.  
%Thus these early accounts were regarded as unsatisfying, because they lead to purely stipulative accounts of linking, specified verb by verb.  
%  ALERT: ok to remove this sentence?  I'd rather not be so negative.  also I actually think the 'unsatisfying' account is right.  :)  i.e. I think innovation involves a clustering algorithm that leads to patterns; no grammar rules are needed to explain those patterns.  
% TD 28 Jan 20: I have reworded the sentence to accommodate that viewpoint, but I think it should remain in some form as an explanation of why more general constraints were sought. 
%JPK 1-29-20 I agree with Tony's rewording.
%SW 1-29-20  How about the new sentence above?  It still says why more general constraints were sought, but doesn't involve a statement that I disagree with, in the sense that "purely stipulative" is a pejorative in our field.  I don't agree with the view that these regularities should be captured in the competence grammar, so to me those 'stipulations' are good.  I'm not insisting you agree with me on that, just that we say things in a neutral way.  

To capture those regularities, HPSG researchers beginning with \citet{Wechsler1995b} and \citet{Davis1996} formulated linking principles stated on more general semantic properties that hold across verbs.  

Within the history of linguistics, there have been three general approaches to modeling the lexico-semantic side of linking: thematic role types \is{thematic role|see{semantic role, participant role, thematic hierarchy}}
(P\={a}\d{n}ini ca.\ 400 B.C., \citealt{Fillmore1968}); lexical decomposition \is{lexical decomposition} \citep{FoleyandvanValin1984,RappaportandLevin1998}; and the proto-roles approach \is{proto-role|see{thematic role, semantic role, participant role}} \citep{Dowty1991}.   In developing linking theories within the HPSG framework, \citet{Wechsler1995b} and \citet{Davis1996} employed a kind of lexical decomposition that also incorporated some elements of the proto-roles approach.  The reasons for preferring this over the alternatives are discussed in Section \ref{thetaroles} below.  

\is{linking!and meaning|(}
Wechsler's (\citeyear{Wechsler1995b}) linking theory constrains the relative order of pairs of arguments on the \argst list according to semantic relations entailed between them.  For example, his \emph{notion rule}\is{Notion Rule}\is{experiencer predicates} states that if one participant in an event is entailed to have a mental notion of another, then the first must precede the second on the \argst list.  The \textit{conceive-pred} type is defined by the following type declaration (based on \citealt[127]{Wechsler1995b}, with formal details adjusted for consistency with current usage):

\begin{exe}
	\ex\label{conceive}
	\textit{conceive-pred:}  
	{\avmoptions{center}
	\begin{avm} 
		\[arg-st  &  \<  \textsc{np}$_i$, \textsc{np}$_j$ \> \\
		content  & \[\asort{conceive-rel}  
		conceiver & $i$ \\
		conceived & $j$ \] 
		\]
	\end{avm}
	}
\end{exe}

This accounts for a host of linking facts in verbs as varied as \word{like}, \word{enjoy}, \word{invent}, \word{claim}, and \word{murder}, assuming these verbs belong to the type \textit{conceive-pred}.  
It explains the well-known contrast between experiencer-subject \word{fear} and experiencer-object \word{frighten} verbs:  \word{fear} entails that its subject has some notion of its object, so \word{The tourists feared the lumberjacks} entails that the tourists are aware of the lumberjacks.  But the object of \word{frighten} need not have a notion of its subject: in \word{The lumberjacks frightened the tourists (by cutting down a large tree that crashed right in front of them)}, the tourists may not be aware of the lumberjacks' existence.  

Two other linking rules appear in \citet{Wechsler1995b}.  One states that ``affected themes'', that is, participants that are entailed to undergo a change, map to the object, rather than subject, of a transitive verb.  Another states that when stative transitive verbs entail a part-whole relation between the two participants, the whole maps to the subject and the part to the object: for example,  \textit{X includes Y} and \textit{X contains Y} each entail that \textit{Y} is a part of \textit{X}.    

These linking constraints do not rely on a total ordering of thematic roles, nor on an exhaustive asssignment of thematic role types to every semantic role in a predicator. Instead, a small set of partial orderings of semantic roles, based on lexical entailments,\is{entailments!lexical} suffices to account for the linking patterns of a wide range of verbs. 
This insight was adopted in a slightly different guise in work by \citet{Davis1996}, \citet{Davis2001}, and \cite{DavisandKoenig2000b}, who develop a more elaborated representation of lexical semantics, with which simple linking constraints can be stated.
The essence of this approach is to posit a small number of dyadic semantic relations such as \textit{act-und-rel} (`actor-undergoer relation') with attributes  \attrib{act(or)} \isfeat{ACT(OR)} and \attrib{und(ergoer)} \isfeat{UNDER(GOER)} that serve as intermediaries between semantic roles and syntactic arguments (akin to the notion of Generalized Semantic Roles discussed in \citealt{VanValin1999}).  

What are the truth conditions of \textit{act-und-rel}?  
Following \citet{Fillmore1977}, \citet{Dowty1991}, and \citet{Wechsler1995b}, Davis and Koenig note that many of the pertinent lexical entailments come in related pairs.
For instance, one of Dowty's entailments is that one participant causally affects another, and of course the other is entailed to be causally affected.
Another involves the entailments in Wechsler's notion rule (\ref{conceive}); one participant is entailed to have a notion of another. 
These entailments of paired participant types characterize classes of verbs (or other predicators), and can then be naturally represented as dyadic relations in \attrib{content}.  Collecting those entailments, we arrive at a disjunctive statement of truth conditions:

\ea
\label{def-act-und-rel}
\textit{act-und-rel}($x,y$) is true iff $x$ causes a change in $y$, or $x$ has a notion of $y$, or \ldots
\z

\noindent
We can designate the $x$ participant  in the pair as the value of \attrib{actor} (or \attrib{act}) and $y$ as the value of \attrib{undergoer} (or \attrib{und}), in a relation of type \type{act-und-rel}.   Semantic arguments that are \attrib{actor} or \attrib{undergoer} will then bear at least one of the entailments characteristic of \attrib{actor}s or \attrib{undergoer}s \citep[72]{DavisandKoenig2000b}. This then simplifies the statement of linking constraints for all of these paired participant types.
\citet{Davis1996} and \citet{KoenigandDavis2001} argue that this obviates counting the relative number of proto-agent and proto-patient entailments, which is what  \citet{Dowty1991} had advocated.

The linking constraints (\ref{act-vb-linking}) and (\ref{und-vb-linking}) state that 
a verb whose semantic \content is of type \emph{act-und-rel} will be constrained to link the \attrib{act} participant to the the first element of the verb's \argst list (its subject), and the \attrib{und} participant to the second element of the verb's \argst list (this is analogous to Wechsler's constraints based on partial orderings).  The attribute \attrib{key} \is{KEY} selects one predication as relevant for linking, among a set of predications included in a lexical item's \attrib{content}; we furnish more details below.
%JPK 1-13-2020 I added a sentence to briefly defined KEY.
% TD 13 Jan. 20: modified a bit for clarity

These linking constraints can be viewed as parts of the definition of lexical types, as in \citet{Davis2001}, where each of the constraints in (\ref{act-vb-linking})--(\ref{emb-act-vb-linking}) defines a particular class of lexemes (or words).\footnote{Alternatively, (\ref{act-vb-linking}) (and other linking constraints) can be recast as implicational constraints on lexemes or words 
\citep{KoenigandDavis2003}.    (\ref{act-vb-linking-alt}) is an implicational constraint indicating that a word whose semantic content includes an \textsc{actor} role must map that role to the initial item in the \argst list. 

\begin{exe}
	\ex\label{act-vb-linking-alt}
	{\avmoptions{center}
	\begin{avm}
		\[content$|$key & \[act & \@1 \] 
		\]\,
		$\Rightarrow$
		\[ 
		arg-st & \<\textsc{np}$_{\@1}$,  \ldots \>
		\]
	\end{avm}
	}
\end{exe} }   

\begin{exe}
	\ex\label{act-vb-linking}
	{\avmoptions{center}
	\begin{avm}
		\[content$|$key & \[act & \@1 \] \\
		arg-st & \<\textsc{np}$_{\@1}$,  \ldots \>
		\]
	\end{avm}
	}
\end{exe}

\begin{exe}
	\ex\label{und-vb-linking}
	{\avmoptions{center}
	\begin{avm}
		\[content$|$key & \[und & \@2 \] \\
		arg-st & \<\ldots, \textsc{np}$_{\@2}$,  \ldots \>
		\]
	\end{avm}
	}
\end{exe}

\begin{exe}
	\ex\label{emb-act-vb-linking}
	{\avmoptions{center}
	\begin{avm}
		\[content$|$key & \[\asort{cause-possess-rel} 
									soa & \[act & \@3 \] \] \\
		arg-st & \<\rm \textit{synsem}\> \,$\oplus$ \<\textsc{np}$_{\@3}$,  \ldots \>
		\]
	\end{avm}
	}
\end{exe}


\noindent
The first constraint, in (\ref{act-vb-linking}), links the value of \attrib{act} (when not embedded within another attribute) to the first element of \argst.
The second, in (\ref{und-vb-linking}), merely links the value of \attrib{und} (again, when not embedded within another attribute) to some NP on \argst.
Given this understanding of how the values of \attrib{act} and \attrib{und} are determined, these constraints cover the linking patterns of a wide range of transitive verbs: \word{throw} (\attrib{act} causes motion of \attrib{und}), \word{slice} (\attrib{act} causes change of state in \attrib{und}), \word{frighten} (\attrib{act} causes emotion in \attrib{und}), \word{imagine} (\attrib{act} has a notion of \attrib{und}), \word{traverse} (\attrib{act} ``measures out'' \attrib{und} as an incremental theme), and \word{outnumber} (\attrib{act} is superior to \attrib{und} on a scale).

The third constraint, in (\ref{emb-act-vb-linking}), links the value of an \attrib{act} attribute embedded within a \attrib{soa} (`state of affairs') \isfeat{SOA (State of Affairs)} attribute to an NP that is second on \argst.
This constraint accounts for the linking of the (primary) object of ditransitives.\is{ditransitive}
In English, these verbs (\word{give}, \word{hand}, \word{send}, \word{earn}, \word{owe}, etc.) involve (prospective) causing of possession \citep{Pinker1989,Goldberg1995}, and the possessor is represented as the value of the embedded \attrib{act} in (\ref{emb-act-vb-linking}).
There could be additional constraints of a similar form in languages with a wider range of ditransitive constructions; conversely, such a constraint might be absent in languages that lack ditransitives entirely.
As mentioned earlier in this section, the range of subcategorization options varies somewhat from one language to another.

The \attrib{key} attribute in (\ref{act-vb-linking})--(\ref{emb-act-vb-linking}) also requires further explanation.\is{KEY}
The formulation of linking constraints here employs the architecture used in \citet{KoenigandDavis2006}, in which the semantics represented in \content values is expressed as a set of \emph{elementary predications}, formalized within Minimal Recursion Semantics \citep{Copestakeetal2001,Copestakeetal2005}.
Each elementary predication is a simple relation, but the relationships among them may be left unspecified.
For linking, one of the elementary predications is designated the \attrib{key}, and it serves as the locus of linking.
This allows us to %sidestep thorny questions of lexical semantics and linking, such as 
indicate the linking of participants that play multiple roles in the denoted situation. 
%, or whether all aspects of a particpant's involvement in an situation type are properly represented in \content.
The \attrib{key} selects one relation as the ``focal point,'' and the other elementary predications are then irrelevant as far as linking is concerned. 
The choice of \attrib{key} then becomes an issue demanding consideration; we will see in the discussion of argument alternations \is{alternation} in Section \ref{alternations} how this choice might account for some alternation phenomena.
\is{linking!and meaning|)}

%Note too that these linking constraints are treated as constraints on classes in the lexical hierarchy
These linking constraints apply to word classes in the lexical hierarchy
 (see \crossrefchapteralt{lexicon}).
One consequence of this fact merits brief mention.
Constraint (\ref{und-vb-linking}), which links the value of \attrib{und} to some NP on \argst, is a specification of one class of verbs.
Not all verbs (and certainly not all other predicators, such as nominalizations) with a \content value containing an \attrib{und} value realize it as an NP.
Verbs obeying this constraint include the transitive verbs noted above, and intransitive ``unaccusative'' verbs such as \word{fall} and \word{persist}.
But some verbs with both \attrib{act} and \attrib{und} attributes in their \content are intransitive, such as \word{impinge (on)}, \word{prevail (on)}, and \word{tinker (with)}.
Interactions with other constraints, such as the requirement that verbs (in English, at least) have an NP subject, determine the range of observed linking patterns.

These linking constraints also assume that the proto-role \is{proto-role|see{thematic role, semantic role, participant role}} attributes \attrib{actor}, \attrib{undergoer}, and \attrib{soa} are appropriately matched to entailments, as described above.
Other formulations are possible, such as that of \citet{KoenigandDavis2003}, where the participant roles \is{participant roles|see{thematic role, semantic role, proto-role}} pertinent to each lexical entailment are represented in \content by corresponding, distinct attributes.

In addition to the linking constraints, there may be some very general well-formedness conditions on linking.  We rarely find verbs that obligatorily map one semantic role to two distinct members of the \argst list, both expressed overtly.  A verb meaning `eat', but with that disallowed property, could appear in a ditransitive sentence like (\ref{rah-a}), with the meaning that Pat ate dinner, and his dinner was a large steak.  

\begin{exe}
\ex[*]{\label{rah-a} 
Pat ate dinner a large steak.
}
\end{exe}

\noindent
Typically, semantic arguments map to at most one (overtly expressed) \argst list item \citep[262--268]{Davis2001}.

% TD 10 Jan. 20: added the following:
% JPK 13 Jan.2020: slightly reworded
Having set out some general principles of linking and their implementation in HPSG, we now briefly discuss linking of oblique arguments.\is{oblique argument}
We also return in the remainder of this section to issues relating to lexical semantic representations as they pertain to linking.
To what extent are the elements of \argst determined by lexical semantics?
Do HPSG lexical semantic representations require thematic roles?
And how does other information in these representations, such as modality and modifier scope, affect linking?
\is{ARG-ST!as locus of linking|see{linking}|)}

\subsection{Linking oblique arguments}
\is{oblique argument}
%We have not yet discussed 
In this section we discuss linking of oblique arguments, that is, PPs and oblique case marked NPs.
In some instances, a verb's selection of a particular preposition appears at least partly arbitrary;
it is hard to explain why English speakers accept \word{hanker after} and \word{yearn for}, but  not \word{*yearn after}.
In these cases, the choice of preposition may be stipulated by the individual lexical entry.
But as \citet{Gawron1986} and \citet{Wechsler1995} have shown, many prepositions selected by a verb have semantic content.
\word{For} in the above-mentioned cases, and in \word{look for}, \word{wait for}, and \word{aim for}, is surely not a lexical accident.
And in  cases like \word{cut with}, \word{with} is used in an instrumental sense, denoting a \type{use-rel} relation, as with verbs that either allow (\word{eat}) or require (\word{cut}) an instrument \citep{KoenigandDavis2006}.
\citet{Davis1996} and \citet{Davis2001} adopt the position of Gawron and Wechsler in his treatment of linking to PPs.
As an example of this kind of account, the linking type in (\ref{with-linking}) characterizes a verb selecting a \word{with}-PP. 
The PP argument is linked from the \rels list \isfeat{RELS} rather than the \attrib{key}. \is{KEY}

\begin{exe}\ex\label{with-linking}
{\avmoptions{center}\begin{avm}
\[content & \[key & \@1 \\
              rels & \<\@1, 
                                   \@2\[\asort{use-rel} \\
                                    act & a \\
                                    und & u  \\
                                    soa & s \], 
                                 \ldots \> \] \\
    arg-st & \< \ldots, \textsc{pp}$_{with}$:\@2 \ldots  \>                          
                           \]
                                             \end{avm} }
\end{exe} 


Apart from the details of individual linking constraints, we have endeavored here to describe how linking can be modeled in HPSG using the same kinds of constraints used ubiquitously in the framework.
Within the hierarchical lexicon (see \crossrefchapteralt{lexicon}), constraints between semantically defined classes and syntactically defined ones can furnish an account of linking patterns, and there is no resort to additional mechanisms such as a thematic hierarchy \is{thematic hierarchy|see{thematic roles}} or numerical comparison of entailments.\is{entailments!lexical}



\subsection{To what extent does meaning predict linking?}
\is{linking!and meaning|(}

The framework outlined above allows us to address the following question: how much of linking is strictly determined by semantic factors, and how much is left open to 
lexically arbitrary subcategorization specifications, or perhaps subject to other factors?

Subcategorization --~the position and nature of \argst elements, in HPSG terms~-- is evidently driven to a great extent by semantics,
but debate continues about how much, and which components of semantics are involved.
%Taking a viewpoint on one end of the decompositional spectrum, there are many elements of lexical semantics that could appear on \argst.
%For example, denominal verbs like \word{pocket}, \word{chisel}, and \word{saddle}; verbs of sound emission like \word{laugh} and \word{rumble}, and verbs with entailed participants like \word{spit}, \word{bleed}, and \word{urinate} all involve participant types that typically are syntactically unrealized.
%Reviving syntactic representations from Generative Semantics, some  Minimalist accounts have advocated exactly that (see the chapter on HPSG and Minimalism for more discussion) and so do linking theories based on the kind of conceptual structure developed in \citet{Jackendoff1990}, to some extent .
%Within HPSG too, there has been an increasing tendency to include ``non-core'' semantic participants--- adjuncts, adverbials, secondary predications, and implicit arguments such as null anaphors--- on the \argst list.
%We will examine this in more detail in Section \ref{extended-arg-st}.
%Even with regard to core, syntactically realized, arguments, we can ask how closely subcategorization mirrors lexical semantics.
%Here as well, 
Views have ranged from the strict, highly constrained relationship in which lexical semantics essentially determines syntactic argument structure to a looser one in which some elements of subcategorization may be stipulated.
Among  the first camp are those who espouse the \isi{Uniformity of Theta Assignment Hypothesis}  proposed in \citet[46]{Baker1988} which maintains that ``identical thematic relationships between items are represented by identical structural relationships'' in the syntax (see also \citealt{Baker1997}).
With regard to the source of \is{alternation!diathesis} diathesis alternations, \citet[12--13]{Levin1993} notes that ``studies of these properties suggest that argument structures might in turn be derivable to a large extent from the meaning of words'', and accordingly ``pursues the hypothesis of semantic determinism seriously to see just how far it can be taken''.

Others, including \citet[Section 5.3]{pollard+sag:1987} and \citet[Section 5.1]{Davis2001}, have expressed caution, pointing out cases where subcategorization and diathesis alternations \is{alternation!diathesis} seem to be at least partly arbitrary.
Pollard and Sag note contrasts like these:

\begin{exe}
\ex \label{ps-subcat-ex}
\begin{xlist}
\ex    Sandy spared/*deprived Kim a second helping.
\ex    Sandy *spared/deprived Kim of a second helping.
\citep[ex.\ 214--215]{pollard+sag:1987}
\end{xlist}
\end{exe}

\noindent
And Davis provides these pairs of semantically similar verbs with differing subcategorization requirements:

\begin{exe}
\ex \label{ard-subcat-ex}
\begin{xlist}
\ex    Few passengers waited for/awaited the train.
\ex    Homer opted for/chose a chocolate frosted donut.
\ex    The music grated on/irritated the critics.
\citep[ex.\ 5.4]{Davis2001}
\end{xlist}
\end{exe}

Other cases where argument structure seems not to mirror semantics precisely include raising constructions, in which one of a verb's direct arguments bears no semantic role to it at all.
Similarly, overt expletive arguments cannot be seen as deriving from some participant role in a predicator's semantics.
Like the examples above, these phenomena suggest that some aspects of subcategorization are specified independently of semantics.

Another point against strict semantic determination of argument structure comes from cross"=linguistic observations of subcategorization possibilities.
It is evident, for example, that not all languages display the same range of direct argument mappings.
Some lack ditransitive \is{ditransitive} constructions entirely (Halkomelem), some allow them across a limited semantic range (English), some quite generally (Georgian), and a few permit tritransitives (Kinyarwanda and Moro).
\citet{Gerdts1992} surveys about twenty languages and describes consistent patterns like these.
The range of phenomena such as causative and applicative formation in a language is constrained by what she terms its ``relational profile''; this includes, in HPSG terms, the number of direct NP arguments permitted on its \argst lists.
Again, it is unclear that underlying semantic differences across languages in the semantics of verbs meaning \word{give} or \word{write} would be responsible for these general patterns.

%Summarizing, there is much evidence tempting us to derive the contents of \argst solely from lexical semantics.
%If this ultimately proves feasible, then \argst serves more as a convenient interface notion with little possibility of independently expressing strictly syntactic aspects of subcategorization.
%This view, however satisfying it might be, does not accord with our current best understanding of the syntactic and semantic evidence.
%In the following sections we delve into some of the nuances that make linking more than a simple rendering of lexical semantics.
%We begin by noting a point on which HPSG accounts of linking differ from many others --- the absence of traditional thematic roles.
%ALERT:   I don't understand this paragraph, so I provisionally commented it out.   
% TD 28 Jan. 20: OK. I'm not strongly attached to this paragraph and I suspect it's there partly as a transition to what's next.  If you don't feel that's needed, let's junk it.
%JPK 1-29-20 OK with junking the paragraph. 

\is{linking!and meaning|)}
\subsection{HPSG and thematic roles}
\label{thetaroles}

\is{thematic role|see{semantic role, participant role, thematic hierarchy}|(}

The \textsc{arg-st} list constitutes the syntactic side of the mapping between semantic roles and syntactic dependents.  As \argst is merely an ordered list of arguments, without any semantic ``labels'', it contains no counterparts to thematic role types, such as \textsc{agent}, \textsc{patient}, \textsc{theme}, or \textsc{goal}.  Thematic roles like these, however, have been a mainstay of linking in generative grammar since \citet{Fillmore1968} and have antecedents going back to P\={a}\d{n}ini.
Ranking them in a \is{thematic hierarchy|see{thematic roles}} \emph{thematic hierarchy}, and labeling each of a predicator's semantic roles with a unique thematic role, then yields an ordering of roles analogous to the ordering on the \argst list.  Indeed, it would not be difficult to import this kind of system into HPSG, as a means of determining the order of elements on the \argst list.  However, HPSG researchers have generally avoided using a thematic hierarchy, for reasons we now briefly set out.

\citet{Fillmore1968} and many others thereafter have posited a small set of disjoint thematic roles, with each semantic role of a predicator assigned exactly one thematic role.
Thematic hierarchies depend on these properties for a consistent linking theory, but they do not hold up well to formal scrutiny.
\citet{Jackendoff1987} and \citet{Dowty1991} note (from somewhat different perspectives) that numerous verbs have arguments not easily assigned a thematic role from the typically posited inventory (e.g., the objects of \word{risk}, \word{blame}, and \word{avoid}), that more than one argument might sensibly be assigned the same role (e.g., the subjects and objects of \word{resemble}, \word{border}, and some alternants of commercial transaction verbs), and that multiple roles can be sensibly assigned to a single argument (the subjects of verbs of volitional motion are like both an \textsc{agent} and a \textsc{theme}).
In addition, consensus on the inventory of thematic roles has proven elusive, and some, notoriously \textsc{theme}, have resisted clear definition.
Work in formal semantics, including \citet{LadusawandDowty1988}, \citet{Dowty1989}, \citet{Landman2000}, and \cite{Schein2002}, casts doubt on the prospects of assigning formally defined thematic roles to all of a predicator's arguments, at least in a manner that would allow them to play a crucial part in linking.
Thematic role types seem to pose problems, and there are alternatives that avoid those problems.  As \citet{Carlson1998} notes about thematic roles: ``It is easy to conceive of how to write a lexicon, a syntax, a morphology, a semantics, or a pragmatics without them''.

%
%Moreover, it is not clear how an ordered hierarchy of thematic roles could be modeled elegantly within HPSG's typed feature structure formalism, though it is certainly not impossible to achieve in some fashion.  Equally unclear is how to compute and compare total numbers of proto-agent and proto-patient entailments for each of a predicator's semantic roles, as \citet{Dowty1991} suggests, because no mechanisms are available for numerical calculations. As a constraint-based formalism, HPSG has led researchers to seek other ways of incorporating insights from these approaches into the framework.

\is{thematic role|see{semantic role, participant role, thematic hierarchy}|)}

\subsection{\content decomposition and \argst}
\is{lexical decomposition|(}
%ok, I see.  I agree, this should be discussed.  So, how does this sound as a list of criteria for inclusion in the decomposition?  Include something in the decomposition (in CONTENT) only if needed in order to:
%
%-- account for sublexical scope, e.g. result states (the Dowty 1979 stuff)
%-- account in a Pinker / Davis & Koenig way for argument alternations  
%-- account for linking in our system
%
%Basically this is everything the grammar needs to see.  Anything else?     
%
%So I'll say this, and I'll briefly contrast it with Hale & Keyser, Jackendoff.  Sound ok?  

Instead of thematic role types, lexical decomposition is typically used in HPSG to model the semantic side of the linking relation.  The word meaning represented by the \content value is decomposed into elementary predications that share arguments, as described in Section \ref{linking-sec} above.  Lexical decompositions cannot be directly observed, but the decompositions are justified indirectly by the roles they play in the grammar.  Decompositions play a role in at least the following processes:

\begin{itemize}
\item  \textit{Linking.}  As described in Section \ref{linking-sec}, linking constraints are stated on semantic relations like \textit{act-und-rel} (`actor-undergoer relation'), so those relations must be called out in the \content field.
\item \textit{Sublexical scope.}  Certain modifiers can scope over a part of the situation denoted by a verb \citep{Dowty:1979a}.  
\end{itemize}

Consider sentence (\ref{again}).

\begin{exe}
\ex \label{again} John sold the car, and then he bought it again.
\end{exe}

In this sentence, the adverb \textit{again} either adds the presupposition that John bought it before, or, in the more probable interpretation, it adds the presupposition that \textit{the result of buying the car} obtained previously.  The result of buying a car is owning it, so this sentence presupposes that John previously owned the car. Thus the decomposition of the verb \textit{buy} includes a \textit{possess-rel} (`possession relation') holding between the buyer and the goods.  This is available for modification by adverbials like \textit{again}.

\begin{itemize}
\item \textit{Argument alternations.}  Some argument alternations can be modeled as the highlighting of different portions of a single lexical decomposition.  See Section \ref{alternations}.  \is{alternation}
\end{itemize}  

\noindent
In general, sublexical decompositions are included in the \content field only insofar as they are visible to the grammar for processes like these.  

% TD 19 Jan 20: I eedited this paragraph a bit
The \argst list lies on the syntax side of linking.  Just as the roles and predicates within \content must be motivated by (linguistic) semantic considerations, the presence of elements on \argst is primarily motivated by their syntactic visibility.  Many \argst list items are obviously justified, being explicitly expressed as subject and complement phrases, or as affixal pronouns.  In addition, certain implicit arguments should appear on \argst if, for instance, they are subject to the binding theory constraints that apply to \argst, as discussed in Section \ref{express-sec} above.

Some implicit arguments can also participate in the syntax, for example, by acting as controllers of adjunct clauses.  This could plausibly be viewed as evidence that such arguments are present on the \argst list.
English rationale clauses,\is{rational clause} like the infinitival phrase in (\ref{mka}), are controlled \is{controlled argument} by the agent argument in the clause, \textit{the hunter} in this example.
The implicit agent of a short passive \is{passive} can likewise control the rationale clause as shown in (\ref{mkb}).
But control is not possible in the middle construction (\ref{mkc}) even though loading a gun requires some agent.
This contrast was observed by \citet{KeyserandRoeper1984} and confirmed in experimental work by \citet{MaunerandKoenig2000}.  

\begin{exe}
\ex\label{mk}
\begin{xlist}
\ex[]{
\label{mka}
The shotgun was loaded quietly by the hunter to avoid the possibility of frightening off the deer.
}
\ex[]{\label{mkb}
The shotgun was loaded quietly to avoid the possibility of frightening off the deer.
}
\ex[*]{
\label{mkc}
The shotgun had loaded quietly to avoid the possibility of frightening off the deer.
}
\end{xlist}
\end{exe}



\noindent
If the syntax of control  is specified such that the controller of the rationale clause is an (agent) argument on the \argst list of the verb, then this contrast is captured by assuming that the agent appears on the \argst list of the passive verb but not the middle.


\subsection{Modal transparency}
\is{modal transparency|(}
Another observation concerning lexical entailments \is{entailments!lexical} and linking was developed by \citet{KoenigandDavis2001}, who point out that linking appears to ignore modal elements of lexical semantics, even when those elements invalidate entailments (expanding on an observation implicit in \citealt{Goldberg1995}).
For instance, there are various English verbs that display linking patterns like the ditransitive \is{ditransitive} verbs of possession transfer \word{give} and \word{hand}, but which denote situations in which the transfer need not, or does not, take place.
Thus, \word{offer} describes a situation where the transferrer is willing to effect the transfer, \word{owe} one in which the transferor should effect the transfer but has not yet, \word{promise} describes a situation where the transferor commits to effect the transfer, and \word{deny} one in which the transferor does not effect the contemplated transfer. 
Koenig and Davis argue that modal elements should be clearly separated in \content values from the representations of predicators and their arguments.  (\ref{promise-sem}) exemplifies this factoring out of sublexical modal information from core situational information. This pattern of linking functioning independently of sublexical modal information applies not only to these ditransitive cases, but also to verbs involving possession (cf.\ \word{own} and \word{obtain} vs.\ \word{lack}, \word{covet}, and \word{lose}), perception (\word{see} vs.\ \word{ignore} and \word{overlook}), and carrying out an action (\word{manage} vs.\ \word{fail} and \word{try}).  Whatever the role of lexical entailments in linking, then, the modal components should be factored out, since the entailments that determine, e.g., the ditransitive linking patterns of verbs like \word{give} and \word{hand} do not hold for \word{offer}, \word{owe}, or \word{deny}, which display the same linking patterns. The constraints in (\ref{act-vb-linking})--(\ref{emb-act-vb-linking}) need only be minimally altered to target the value of \attrib{sit-core}, representing the `situational core' of a relation.

\begin{exe}
\ex\label{promise-sem} The lexical semantic representation of \word{promise} 
\citep[101]{KoenigandDavis2001} \\
{\avmoptions{center}
\begin{avm} 
\[\asort{$\textit{promise-sem}$ $\wedge$ $\textit{cause-possess-sem}$}
  sit-core &  \@3\[\asort{$\textit{cause-possess-rel}$}
                act & \@1 \\
               und & \@2 \\
                soa & \[sit-core & \@5\[\asort{$\textit{have-rel}$}
                                       act & \@2 \\
                                       und & \@4\]\]\]\\
   modal-base & \< \[\asort{$\textit{deontic-mb}$ $\wedge$ $\textit{condit-satis-mb}$}
                soa  & \@3 \] \>\] \end{avm}
}
 \end{exe}
 
% TD 10 Jan. 20: addded the following, including the summary subsection
%JPK 13 Jan 20: rewordinf the very end of the sentence to make it crisper.
This kind of semantic decomposition preserves the simplicity of linking constraints, while representing the differences between verbs that straightforwardly entail the relation between the arguments in the situational core and verbs for which those entailments do not hold, because their meaning contains a modal component restricting those entailments to a subset of possible worlds.
\is{modal transparency|)}

\is{lexical decomposition|)}

\subsection{Summary of linking}

In this section we have examined HPSG approaches to linking.
HPSG constrains the mapping between participant roles \is{participant roles|see{thematic role, semantic role, proto-role}}  in \attrib{content} and their syntactic realizations on \argst based on entailments of the semantic relations in \attrib{content}.
These constraints do not require a set of thematic roles arranged in a hierarchy.
Nor do they require a numerical comparison of entailments holding for each participant role, which has been an influential alternative to a thematic hierarchy.
Rather, they reference the types of relations wthin a lexical entry's \attrib{content}, and the subcategorization requirements of its \argst.
Information from both is necessary because, although semantics is a strong determinant of argument realization, independent stipulations of subcategorization appear to be needed, too.
Finally, we have examined the role of modal information in lexical semantics, which seem not to interact much with linking, and described mechanisms proposed within HPSG that separate this information from the situational core that drives linking.

In the remainder of this chapter, we will examine the relationship of argument structure to argument alternations, including passives, as well as broader questions concerning the addition of other elements like modifiers to \argst, the universality of \argst across languages, and whether \argst is best regarded as solely a lexical attribute or one that should also apply to phrases or constructions.


\section{The semantics and linking of argument alternations}
\label{alternations}
\is{alternation|(}
A single verb can often alternate between
%occur 
%in varied syntactic contexts, 
 various alternative patterns of dependent phrases, situations called either 
%which are termed 
\emph{argument alternations},
\emph{valence alternations}, or \emph{diathesis alternations}.
\citet{Levin1993} lists around 50 kinds of alternations in English, and English is not untypical in this regard.

%TD 14 Jan 20: I suggest removing the following, which is not referenced later and has nothing in particular to do with HPSG
%and there are still more, including the alternation illustrated in (\ref{chair}).  
%
%\begin{exe}
%	\ex\label{chair}
%	\begin{xlist}
%	\ex\label{chair-a} I found that the chair was comfortable
%	\ex\label{chair-b} I found the chair to be comfortable
%	\ex\label{chair-c} I found the chair comfortable	
%	\end{xlist}
%\end{exe}

How has argument structure in HPSG been used to account for alternations?
Many alternations exhibit (often subtle) meaning differences between the two 
alternants.
%Most alternations have been associated with subtle differences between the meanings of the two alternants, though they may be suble.
%Consequently, alternants with different argument structures and syntactic contexts owe this difference to different \attrib{content} values.
We first discuss alternations due to these differences in meaning, showing how their differing \argst lists arise from differences in \attrib{content}.  
We then examine some alternations where meaning differences are less apparent.
Although the \attrib{content} values of the two alternants in such cases may not differ, we can analyze the alternation in terms of a different choice of \attrib{key} predicate in each.
Lastly, we consider active-passive voice alternations, which are distinct from other alternations in important ways.
% TD 28 Jan 20: what was the motivation for commenting out the following?
%It is doubtful that meaning differences between passives and corresponding actives can furnish an explanation for their syntactic differences.
%Indeed, because passivization applies so productively in many languages, to verbs that range so widely in meaning, it would be difficult to model it as a semantic alternation or as a change of \attrib{key} predicate.
%Accounts of passivization in HPSG have therefore resorted to two kinds of non-canonical lexical entries.
%One relaxes linking constraints, so that the active form's subject is not the first element of the passive form's \argst.
%The other relaxes the mapping between elements of \argst and elements of the \val list, in a manner similar to the treatment of Balinese voice alternations sketched in section \ref{ergativity}.

\subsection{Meaning-based argument alternations}

One well-studied alternation, the locative alternation, is exemplified by the two uses of \word{spray} in (\ref{spray}).

\begin{exe}
\ex \label{spray}
\begin{xlist}
\ex \label{spraya} \textit{spray$_{\text{loc}}$}: Joan sprayed the paint onto the statue.
\ex \label{sprayb} \textit{spray$_{\text{with}}$}: Joan sprayed the statue with paint.
 \end{xlist}
 \end{exe}


It is typically assumed that these two different uses of \textit{spray} in (\ref{spray}) have slightly different meanings, with the statue being in some sense more affected in the \word{with} alternant.
This exemplifies the ``holistic'' effect of direct objecthood, \is{holistic effect of direct objecthood} which we will return to.
Here, we will examine how semantic differences between alternants relate to their linking patterns.
The semantic side of linking has often been devised with an eye to syntax (e.g., \citealt{Pinker1989}, and see \citealt{KoenigandDavis2006} for more examples).
There is a risk of stipulation here, without independent evidence for these semantic differences.
In the case of locative alternations, though, the meaning difference between (\ref{spraya}) and (\ref{sprayb}) is easily stated (and Pinker's intuition seems correct), as (\ref{sprayb}) entails (\ref{spraya}), but not conversely.
Informally, (\ref{spraya}) describes a particular kind of caused motion situation, while  (\ref{sprayb}) describes a situation in which this kind of caused motion additionally results in a caused change of state.
The difference is depicted in the two structures in (\ref{spray-sem}).

\begin{exe}\ex\label{spray-sem}
\begin{xlist}
\ex \label{spray-sema} \textsc{cause (Joan, go (paint, to (statue)))}
\ex \label{spray-semb} \textsc{act-on (Joan, statue, by (cause (Joan, go (paint, to (statue)))))}
\end{xlist}
\end{exe}

This description of the semantic difference between sentences (\ref{spraya}) and (\ref{sprayb}) provides a strong basis for predicting their different argument structures.
But we still need to explain how linking principles give rise to this difference.
Pinker's account rests on semantic structures like (\ref{spray-sem}), in which depth of embedding reflects sequence of causation, with ordering on \argst stemming from depth of semantic embedding, a strategy adopted in \citet{Davis1996} and \citet{Davis2001}.
This is one reasonable alternative, although the resulting complexity of some of the semantic representations raises valid questions about what independent evidence supports them.
An alternative appears in \citet{KoenigandDavis2006}, who borrow from \isi{Minimal Recursion Semantics} (see \crossrefchapteralt{semantics} for an introduction to MRS).
MRS ``flattens'' semantic relations, rather than embedding them in one another, so the configuration of these \emph{elementary predications} \is{elementary predication} with respect to one another is of less import.
It posits a \attrib{relations} (or \rels) \isfeat{RELS} attribute that collects a set of elementary predications, each representing some part of the predicator's semantics.
In Koenig and Davis' analysis, a \attrib{key} attribute specifies a particular member of \rels as the relevant one for linking (of direct syntactic arguments). \is{KEY}
In the case of (\ref{sprayb}), the \attrib{key} is the caused change of state description.
These MRS-style representations of the two alternants of \word{spray}, with different \attrib{key} values, are shown in (\ref{spray-on}) and (\ref{spray-with}).

\begin{exe}
\ex\label{spray-on}
{\avmoptions{center}\begin{avm}\[key & \@5 \\
                    rels & \<\@5\[\asort{$\textit{spray-ch-of-loc-rel}$} 
                                    act & \@1 \\
                                    und & \@4 \\
                                    soa & \[\asort{\textit{ch-of-loc-rel}} 
                                                figure & \@4\]\] \>\]
                  \end{avm}}
\end{exe}

\begin{exe}\ex\label{spray-with}
{\avmoptions{center}\begin{avm}
\[key & \@3\[\asort{$\textit{spray-ch-of-st-rel}$}
                                    act & \@1 \\
                                    und & \@2  \\
                                    soa & \[\asort{ch-of-st-rel} 
                                                und & \@2\] 
                      \] \\
                   rels & \<\@3, 
                                   \[\asort{$\textit{use-rel}$}
                                    act & \@1 \\
                                    und & \@4  \\
                                    soa & \@3 \], 
                                 \[\asort{$\textit{spray-ch-of-loc-rel}$} 
                                    act & \@1 \\
                                    und & \@4 \\
                                    soa & \[\asort{ch-of-loc-rel} 
                                                figure & \@4\]\] \>
                           \]
                                             \end{avm} }
\end{exe}                  
                  
Generalizing from this example, one possible characterization of valence alternations, implicit in \citet{KoenigandDavis2006}, is as systematic relations between two sets of lexical entries in which the \rels of any pair of related entries are in a subset/superset relation (a weaker version of that definition would merely require an overlap between the \rels values of the two entries). 
Consider another case; (\ref{caus-inch}) illustrates the causative-inchoative alternation, where the intransitive alternant describes only the change of state, while the transitive one ascribes a explicit causing agent.

\begin{exe}
\ex\label{caus-inch}
\begin{xlist}
	\ex\label{caus-inch-a}John broke the window.
	\ex\label{caus-inch-b}The window broke.
\end{xlist}	
\end{exe}

Under an MRS representation, the change of state relation is a separate member of \rels; it is also included in the \rels of the transitive alternant, which contains a cause relation as well.
Again, the \rels value of one member of each pair of related entries is a subset of the \rels value of the other.

Many other alternations involve one argument shifting from direct to oblique.\is{oblique argument}
Some English examples include conative, locative preposition drop, and \word{with} preposition drop alternations, as shown in (\ref{altex}):

\begin{exe}\ex\label{altex}
\begin{xlist}
\ex \label{altexa} Rover clawed (at) Spot. 
\ex \label{altexb} Bill hiked (along/on) the Appalachian Trail.
\ex \label{altexc} Burns debated (with) Smithers.
\end{xlist}
\end{exe}

\noindent
The direct object argument in (\ref{altexa}) is interpreted as more ``affected'' than its oblique counterparts:  if Rover clawed Spot, we infer that Spot was subjected to direct contact with Rover's claws and may have been injured by them, while if Rover merely clawed \word{at} Spot, no such inference can be made.
Similarly, to say that one has hiked the Appalachian Trail as in the transitive variant of (\ref{altexb}) suggests that one has hiked its entire length, while the prepositional variants merely suggest one hiked along some portion of it.  In still other cases like (\ref{altexc}), the two variants seem to differ very little in meaning.  

%How might these varying intuitions related to ``affectedness'' relate to lexical semantic representations like those in the (\ref{spray-on}) and (\ref{spray-with})?  

\citet{Beavers2010} observes the following generalization over direct-oblique alternations:  the direct variant entails the oblique one, and can have an additional entailment that the oblique variant lacks.  
His \emph{Morphosyntactic Alignment Principle} \is{Morphosyntactic Alignment Principle} states this generalization in terms of ``L-thematic roles'', which are defined as sets of entailments associated with individual thematic roles:\is{thematic role|see{semantic role, participant role, thematic hierarchy}}   

\begin{exe}
\ex\label{beavers-map}
When participant $x$ may be realized as either a direct or \isi{oblique argument} of verb V, it bears L-thematic role $R$ as a direct argument and L-thematic role $Q\subseteq_{M}R$ as an oblique.
\citep[848]{Beavers2010}
\end {exe}

\noindent
Here, $Q$ and $R$ are roles, defined as sets of individual entailments, and  $Q\subseteq_{M}R$ means that set $Q$ is a subset of $R$ that is minimally different from $R$, differing in at most one entailment.
%a ``minimally weaker'' role than $R$; in other words, there is no role
%$P$ in the predicator such that $Q⊂P⊂R$.
Thus, the substantive claim is essentially that the MAP rules out ``verbs where the alternating participant has \textsc{MORE} lexical entailments as an oblique than the corresponding object realization'' \citep[849]{Beavers2010}.
% \citet{beavers:2005} presents this generalization in the HPSG framework, using MRS semantic representations similar to those employed above.  
% TD 31 Jan. 20: I've now looked at Beavers (2005) -- thanks for tracking that down -- and I think there are some substantive differences worth noting briefly here, so I've revived some of the stuff that was commented out below, putting it here in a revised form.  
The notion of a stronger role in Beavers' analysis has a rough analogue in terms of whether a particular elementary predication is present in the semantics of a particular alternant.
\citet{Beavers2005} describes a version of the Morphosyntactic Alignment Principle implemented in HPSG, which posits a separate \attrib{roles} attribute within \attrib{content}, containing a list of labeled roles.
The roles are ordered on the \attrib{roles} list, determined at least partly by direction of causality, although this is not fully worked out.
Each role can be regarded as a bundle of entailments.
The bundle of entailments varies slightly between different alternants of verbs like those in (\ref{altex}), and the Morphosyntactic Alignment Principle comes into play, comparing the sets of entailments constituting each role.
Assessing which of two roles is stronger, according to this principle, requires some additional mechanisms within HPSG that are not spelled out.

Beavers notes the resemblance between his account and numerical comparison approaches such as those of \citet{Dowty1991} and \citet{AckermanandMoore2001}.
He points out that the direct object bears an additional entailment in each alternant. However, the specific entailment involved depends on the verb; the entailments involved in each of the examples in (\ref{altex}), for instance, are all different.  Thus, comparing the numbers of entailments holding for a verb's arguments in each alternant is crucial.

%The entailments Beavers employs differ somewhat from those we have discussed here, involving quantized change, nonquantized change, potential for change (where change can refer to change in location, possession, state, or something more abstract), furnishing the clear ordering by strength that is central to his proposal.
%But they do resemble entailments of semantic relations we have represented as elementary predications, such as incremental theme, change of state, and possession, along with the modal effects described in \citet{KoenigandDavis2001}.
%Thus the notion of a stronger role in Beavers' analysis has a rough analog in terms of whether a particular elementary predication is present in the semantics of a particular alternant.
%And only if an elementary predication is present, can it be designated as the \attrib{key}, and its roles linked directly.
%For example, in (\ref{spray-on}), there is nothing representing affectedness of the location, while in (\ref{spray-with}), there is, and it is designated as the \attrib{key}.
%As noted earlier, the semantics in (\ref{spray-with}) represents this additional entailment borne by the location argument.
%However, we are not aware of any simple, general way to represent Beavers' MAP within the EP-based model of \citet{KoenigandDavis2006}.  
%ALERT:  Why not?  can't you associate the transitive V with an extra entailment?  Then when you replace the object with a PP that entailment goes away.  
% TD 28 Jan 20: You might be able to do that, if there's a consistent set of entailments that go with direct objecthood.  But the constraint as presented is about numbers of entailments.  Let's think about it a bit.
% JPK 1-29-20 The problem is that there is no consistent entailment associated with objecthood according to John's approach. It is a meta-semantic statement: whatever the meaning of the "input", strenghten it (clearly, a simplified description). I can't see how to model that in the way HPSG is supposed to work. We maybe could stipulate an additional semantic type strengthened_affectedness that would be vacuous for some objects and have a different model-theoretic effect for different classes of verbs alternating between obliques and direct objects, but that would be a hack. Am I missing something?
%Indeed, there is an aspect of Beavers' view that seems more in accord with numerical comparison approaches such as those of \citet{Dowty1991} and \citet{AckermanandMoore2001}, in that role strength is determined by the number of entailments that hold of it relative to others.  

\subsection{Relationships between alternants}

Having outlined the semantic basis of the different linking patterns of alternating verbs, we briefly take up two other issues.
First is the question of how the alternants are related to one another.
Second is how \attrib{key} selection has been used to account not just for alternants of the same verb, but for (nearly) synonymous verbs whose semantics contain the same set of elementary predications.

The hypothesis pursued in \citet{Davis1996} and \citet{Davis2001}  is that 
most alternations are the consequence of classes of lexical entries having
two related meanings. This follows researchers such as \citet{Pinker1989} and \citet{Levin1993} in modeling subcategorization alternations as underlyingly meaning alternations. 
This change in meaning is crucial to the \citet{KoenigandDavis2006} \attrib{key} shifts as well. In some cases, the value of the \rels attribute of the two valence alternates differ (as in the two alternates of \word{spray} in the so-called \word{spray/load} alternation we discussed earlier).
In some cases, the alternation might be different construals of the same event for some verbs, but not others, as \citet{RappaportandLevin2008} claim for the English ditransitive \is{ditransitive} alternations, which adds the meaning of transfer for verbs like \word{send}, but not for verbs like \word{promise}; a \attrib{key} change would be involved (with the addition of a \type{cause-possess-rel}) for the first verb only. But \attrib{key} shifts and diathesis alternations do not always involve a change in meaning. The same elementary predications can be present in the \attrib{content} values of two alternants, with each alternant designating a different elementary predication as the \attrib{key}. \is{KEY}

Koenig and Davis propose this not only for cases in which there is no obvious meaning difference between two alternants of a single verb, but also for different verbs that appear to be truth-conditionally equivalent.
% TD 31 Jan 20: revisions made throughout this paragraph to checnge from buy/sell to substitute/replace
The verbs \word{substitute} and \word{replace} are one such pair.
The two sentences in (\ref{subrep}) illustrate this equivalence.

\begin{exe}
\ex\label{subrep}
\begin{xlist}
\ex\label{subrepa}They substituted an LED for the burnt-out incandescent bulb.
\ex\label{subrepb}They replaced the burnt-out incandescent bulb with an LED.
\end{xlist}
\end{exe}

These two verbs denote a type of event in which a new entity takes the place of an old one, through (typically intentional) causal action.
%Koenig \& Davis suggest that both the removal of the old entity and the placement of the new one are represented as elementary predications in the \attrib{content} values of these verbs.
Koenig and Davis decompose both verb meanings into two simpler actions of removal and placement: `x removes y (from g)' and `x places z (at g)', each represented as an elementary predication in the \attrib{content} values of these verbs.  


% one famous example being the verbs of commercial exchange \word{buy} and \word{sell} (but see \citet[387-388]{VanValin1999}, \citet[20]{LevinandRappaport2005}, and \citet{Wechsler:2005} for arguments that \word{buy} and \word{sell} are not equivalent). 
% ??   if they were equivalent then 'John bought it' and 'John sold it' would have the same meaning, which nobody believes.  Not sure what you mean here.  ALERT
% Koenig and Davis argue that a commercial event involves two reciprocal actions, a transfer
%an exchange [exchange = 'an act of giving one thing and receiving another (especially of the same type or value) in return.'  So a sale involves one exchange, not two.] of goods (which involves giving goods and obtaining goods) and 
%an exchange a transfer
% of money (which involves giving money and obtaining money). Individual verbs might select  one or the other these four relations, thus accounting for the differences in subject and object selection. As shown in (\ref{buy-lex}) and (\ref{sell-lex}), each of these verbs contains four elementary predications: one \type{exch-give-rel} and one \type{obtain-rel} for the transfer of goods, and one of each for the counter-transfer of money. 
%ALERT: there are 2, not 4 EPs in the AVM for 'sell', right?  Also I don't understand this analysis.  why are there 4 EPs for 'buy'?   The seller doesn't have to own the item he sells; cp. salesmen.  Nor does the buyer necessarily come to own it; cp. purchasing agents.  See my little v paper.   More generally, what is the basis for including all this structure?  a word usually has an infinite set of lexical entailments, so accounting for entailments can't be sufficient.  
% TD 28 Jan 20: I think we could avoid some of these issues (which obscure the main point here anyway) by using an example other than 'buy'/'sell', such as 'substitute' (X for Y)/'replace' (Y with X). I agree that 'sell' differs in meaning from 'buy' (a plausible suggestion is that it means ``cause to buy''), so it's not a great example to use, even though historically it has been sometimes held up as one.. 
%JPK 1-29-20 Fine with the substitution (so to speak!). For whatever it is worth, I disagree witht the claim that it's clear that "sell" and "buy" are not truth conditionally equivalent. I could imagine making the claim that BOX (sell <-> buy) and that examples typically cited that suggest it's false are different uses of "buy" (different "micro-meanings"). But that's for another day.  
%SW 1-29-20.  You think that 'JP bought a horse' and 'JP sold a horse' have the same meaning?  
%JPK 1-31-20 No I mean that "Somebody sold a horse to JP" and "JP bought a horse from somebody" have the same meaning. At least, one can make the claim that they mutually entail each other, and some people's view that's having the same meaning :).
%\word{Buy} designates the \type{obtain-rel} representing the transfer of goods as the \attrib{key}, while \word{sell} designates the \type{exch-give-rel} representing the transfer of goods as the \attrib{key}.  Other verbs, such as \word{pay} or \word{charge}, choose elementary predications representing the counter-transfer as the \attrib{key}.  
% The relevant portions of the entries for \word{buy} and \word{sell} in (\ref{buy-lex}) and (\ref{sell-lex}) below illustrate: critically, the \attrib{key} relation for \word{buy} is not the same as that for \word{sell}.


%\begin{exe}
%\ex\label{buy-lex}
%A representation of the relevant parts of the lexical entry for \word{buy}: \\
%{\avmoptions{center}
%\begin{avm}\[content & \[key &\@7 \\
%                         relations & \<\@7\[act & \@1 (buyer) \\
%                          und & \@2 \\
%                          soa & \@5\[\asort{$\textit{possess-rel}$} 
%                                    act & \@1 \\
%                                    und & \@2 (goods)\]\],
%                        \[act & \@1 (buyer) \\
%                          und & \@4 \\
%                    soa & \@6\[\asort{$\textit{possess-rel}$}
%                                    act & \@3 (seller) \\ 
%                                    und & \@4 (money)\]\] \\
%                              \[\asort{$\textit{obtain-rel}$}
%                   act & \@3 (seller) \\
%                          und & \@4 \\
%                    soa & \@6\[\asort{$\textit{possess-rel}$} 
%                                    act & \@3 (seller) \\ 
%                                    und & \@4 (money)\]\], 
%\@8\[\asort{$\textit{exch-give-rel}$}
%                    act & \@3 (seller) \\
%                          und & \@2 \\
%                    soa & \@5\[\asort{$\textit{possess-rel}$} 
%                                    act & \@1 \\ 
%                                    und & \@2 (goods)\]\]
%                                                         \> \] \\
%             arg-st & \<NP:\@1, NP:\@2, PP(from):\@8\>\]
%\end{avm}}	
%\end{exe}

%\begin{exe}
%\ex\label{sell-lex}	
%A representation of the relevant parts of the lexical entry for \word{sell}: \\
%{\avmoptions{center}
%\begin{avm}\[content & \[key &\@7 \\
%                         relations & \<\@7\[act & \@1 (seller) \\
%                                          und & \@2 \\
%                    soa & \@6\[\avmspan{\textit{possess-rel}} \\
%                                    act & \@3(buyer) \\ 
%                                    und & \@2 (goods)\]\],
%                             \[act & \@1 (seller) \\
%                          und & \@4 \\
%                    soa & \@6\[\avmspan{\textit{possess-rel}} \\
%                                    act & \@1 \\ 
%                                    und & \@4 (money)\]\]\> \] \\
%             arg-st & \<NP:\@1, NP:\@2, PP(to):\@7\>\]
%\end{avm}}
%\end{exe}



%\pagebreak

% TD 31 Jan. 20: We might consider simplifying and clarifying these AVMs a bit.  We might dispense with all the SIT-CORE and modal aspects that aren't pertinent to the point we're making here, and distinguish the removal and placement relations with different names.

\begin{exe}
\ex \label{fig:place} Representation of `x places y (at g)' \\
{\avmoptions{center} 
\begin{avm} \@a $=$  \[\avmspan{\rm \textit{place-rel}} \\
                                                         act & \@1 (x) \\
                                                         und & \@2 (y)  \\
                                                         soa & \[{\rm\textit{motion-sem} }\\
                                                                      sit-core & \[{\rm\textit{motion-rel}} \\
                                                                                            fig & \@2 \\
                                                                                           grnd & g\]\\
                                                                      modal-base & \< \> \] \]                                       
                  \end{avm}
}
\end{exe}


\begin{exe}
\ex \label{fig:remove} Representation of `x removes z (from g)' \\
{
\avmoptions{center} 
\begin{avm}  \@b $=$ \[\avmspan{\rm \textit{remove-rel}} \\
                                    act & \@1 (x) \\
                                    und & \@3 (z) \\
                                    soa & \[\textit{\rm \textit{motion-sem}} \\
                                                  sit-core & \@4\[{\rm\textit{motion-rel}} \\
                                                                     fig & \@3 \\
                                                                      grnd & g\]\\
                                                  modal-base & \<\[{\rm\textit{neg-rel}} \\
                                                                                  soa & \@4\] \> \] \]                                                                                   
                  \end{avm}
}
\end{exe}

Either one can be selected as the \attrib{key}.
In the lexical entry of \word{replace}, the removal predication is the value of \attrib{key}, while in the lexical entry of \word{substitute}, the placement of the new object is the value of \attrib{key}.
(\ref{fig:replace}) and (\ref{fig:subst}) show the \attrib{content} values of these two verbs under this account, where \fbox{a} and \fbox{b} abbreviate the structures in (\ref{fig:place}) and (\ref{fig:remove}).
In both cases, the same linking constraints apply between the  \attrib{key} and the \argst list, but the two verbs have different argument realizations because their \attrib{key} values differ, even though their semantics are equivalent.


\begin{exe}
\ex \label{fig:replace} \textsc{content} value of \word{substitute for} \\
{
\avmoptions{center} 
\begin{avm}\[\textsc{key} &  \@a \\
                   \textsc{rels} & \< \@a, \@b \> \]
                  \end{avm}
}
\end{exe}


\begin{exe}
\ex \label{fig:subst} \textsc{content} value of \word{replace with} \\
{
\avmoptions{center} 
\begin{avm}\[\textsc{key} &  \@b \\
                   \textsc{rels} & \< \@a, \@b \> \]
                  \end{avm}
}
\end{exe}

As a final example of the effect of alternations on fine-grained aspects of verb meaning, we consider the source-final product alternation exemplified in (\ref{carve}), where the direct object can be either the final product or the material source of the final product. 

\begin{exe}
\ex\label{carve}
\begin{xlist}
	\ex\label{carve-a} Kim made/carved/sculpted/crafted a toy (out of the wood).
	\ex\label{carve-b} Kim made/carved/sculpted/crafted the wood into a toy.
\end{xlist}
\end{exe}

Davis proposes that the (\ref{carve-a}) sentences involve an alternation between the two meanings represented in (\ref{carve-sem}), each associated with a distinct entry. We adapt \citet{Davis2001} to make it consistent with \citet{KoenigandDavis2006} and also treat the alternation as an alternation of \emph{entries} with distinct meanings. 
%Note that in the alternation described in (\ref{carve-sem}), we use, informally, a double-headed arrow. 
Lexical rules are a frequent analytical tool used to model to alternations between two related meanings of a single entry illustrated in (\ref{carve-sem}). One of the potential drawbacks of a lexical rule approach to valence alternations is that it requires selecting one alternant as basic and the other as derived. This is not always an easy decision, as \citet{Goldberg1995} or \citet{LevinandRappaport1994} have pointed out (e.g., is the inchoative or the causative basic?). Sometimes, morphology provides a clue, although in different languages the clues may point in different directions.  French, and other Romance languages, use a ``reflexive'' clitic as a detransitivizing affix.  In English, though, there is no obvious ``basic'' form or directionality. It is to avoid committing ourselves to a directionality in the relation between the semantic contents described in (\ref{carve-sem}) that we eschew treating it as a lexical rule.

% TD 10 Jan. 20: the arrow in this figure is pretty ugly; can we change it, perhaps to the sort of arrows that are in figure \ref{fig:over}?  We might consider orienting it vertically as well, rather than horizontally
%JPK 1-13-20: I removed the arrows and reworded the last sentence in the preceding paragraph.



\begin{exe}
	\ex\label{carve-sem}
	\begin{xlist}
	\ex \label{carve-sem-a}
{\avmoptions{center}
\begin{avm}
\[
content \[key & \@3\[\asort{affect-incr-th-rel} 
act & \@1 \\
soa & \@2(final product)\] \\
rels & \< \@3, \@4 \> \] \]\end{avm}}
\ex \label{carve-sem-b}
{\avmoptions{center}
\begin{avm}
\[
content \[key &
\@4\[\asort{affect-incr-th-rel}
act & \@1 \\
und & (source material) \\
soa & \[\asort{affect-incr-th-rel} 
		act & \@1 \\
		soa & \@2(final product)\] \] \\
rels & \< \@3, \@4 \>
\] \]
\end{avm}		
}
\end{xlist}
\end{exe}
 
% TD 14 Jan. 20: new subsection for passives
\subsection{The problem of passives}
\label{passives}
\is{passive(}
\is{alternation!voice|see{passive}}
% TD 14 Jan. 20: revised this introductory material a bit, to fit with the new stuff
Although most diathesis alternations can be modeled as alternations in meaning and as \attrib{key} shifts, some arguably cannot. 
One prominent example is the active/passive alternation.
Other widely attested constructions, such as raising constructions, similarly involve no change in meaning, but we will examine only passives here. 

The semantics of actives and corresponding long passives, as in (\ref{act-pass}), are practically identical and the difference between the two alternants is pragmatic in nature. 

\begin{exe}
	\ex\label{act-pass}
	\begin{xlist}
		\ex\label{act-pass-a}Fido dug a couple of holes.
		\ex\label{act-pass-b}A couple of holes were dug by Fido.
	\end{xlist}
\end{exe}

%We might regard the relationship between actives and passives as a degenerate case of the subset relationship between \rels attributes, where both the \rels and \attrib{key} values of the two entries are identical.
% TD 16 Jan. 20: I don't quite follow this; is it referencing Beavers's stuff?  Maybe we explain a bit more, or just delete this first sentence.
%SW: Same here.  I deleted it.  
%But we must still account for the demotion of the active subject in the corresponding passive, and, in the context of this chapter, what the \argst list of a passive looks like.
In this section, we outline two possible approaches to 
%this question.
the passive.  
Both of them treat the crucial characteristic of passivization as subject demotion (see \citealt{Blevins2003} for a thorough exposition of this characterization), rather than object advancement, as proposed, e.g., in \isi{Relational Grammar} \citep{PerlmutterandPostal1983b}.
As we will see, there are various options for implementing this general idea of demotion within HPSG.

% TD 14 Jan. 20: new material inserted here
The first approach, which goes back to \citet{pollard+sag:1987}, assumes that passivization targets the first member of a \attrib{subcat} list and either removes it or optionally puts it last on the list, but as a PP.
This approach is illustrated in (\ref{pass-lr}), a possible formulation of a lexical rule for transitive verbs adapted to a theory that replaces \attrib{subcat} with \argst, as discussed in \citet[67]{Manning+Sag:1999}. 
See \citew{Mueller2003e} for a  more refined formulation of the passive lexical rule for German that accounts for impersonal passives, and \citet{Blevins2003} for a similar analysis.
The first NP is demoted and either does not appear on the output's \argst or is a PP coindexed with the input's first NP's index (see \crossrefchapteralt{binding} 
%and \crossrefchapteralt{diachronic} 
for scholars who assume the latter view within this handbook).
Linking in passives thus violates the constraints in (\ref{act-vb-linking})--(\ref{emb-act-vb-linking}), specifically (\ref{act-vb-linking}), which links the value of \attrib{act} to the first element of \argst. (We use one possible feature-based representation for lexical rules to help comparing approaches to passives.) See \citet{Meurers2001} and \crossrefchaptert{lexicon} for a discussion of various approaches to lexical rules.
Note that we use the attribute \attrib{lex-dtr} rather than the \attrib{in(put)} attribute used in the representation of lexical rules in \citet[76]{Meurers2001} to avoid any procedural implications; nothing substantial hinges on this labeling change.

%\avmsetup{columnsep=.3ex}

 \ea
\label{pass-lr}
 %% {\avmoptions{center}\begin{avm}
 
 %% \[\asort{passive-verb}
 %% arg-st & \@2\textup{(}$\oplus$\<PP[\type{by}$]$$_{i}$\>\textup{)} \\
 %% lex-dtr & \[\asort{stem}
 %%             head & \type{verb}\\
 %%             arg-st & \<\@1 $_{i}$\> $\oplus$ \@2\<np, \ldots\>
 %%           \]
 %% \]
 %% \end{avm}}
\avmtmp{
[\type*{passive-verb}
 arg-st  & \2 ( \+ < ! PP[\type{by}]$_i$ ! > )\\
 lex-dtr & [\type*{stem}
            head   & verb\\
            arg-st & < \1$_i$ > \+ \2 < NP, \ldots > ]]  
}
\avmtmp[columnsep=.3ex]{
[\multicolumn{2}{@{}l@{}}{\raisebox{0.7ex}{\type{passive-verb}}}\\
 arg-st  & \2 ! \upshape ~( $\oplus$ \sliste{ PP[\type{by}]$_i$  } ) !\smallskip\\
 lex-dtr & [\raisebox{1.1ex}{\type{stem}}\\
            head   & verb\\
            arg-st & \sliste{ \1$_i$ } \+ \2 \sliste{ NP, \ldots } ]]  
}

\z
% ALERT:  should the attribute be INPUT instead of IN?
% TD 28 Jan. 20: I'm fine with either
%JPK 1-29-20: I used "IN" because this is what Meurers (2001) uses and we cite in the lexicon chapter, i.e. for consistency across chapters. That said, if you both want INPUT, fine with me.
We will refer to this approach as the non-canonical linking analysis of passives. This kind of analysis invites at least three questions.\is{linking!and passives}
First, as already noted, the constraint linking the value of \attrib{act} to the first element of \argst is violated.
If passives --- widespread and hardly exotic constructions --- violate canonical linking constraints, how strong an account of linking can be maintained?
Second, what other predictions, such as changes in binding behavior, control constructions, and discourse availability, arise from the altered \argst of passives?
Third, what is the status of the \word{by}-phrase in long passives, and how is it represented on the \argst list?

Another approach maintains the \argst list of the active verb in its passive counterpart, thereby preserving linking constraints.
Passives differ from actives under this account in their non-canonical mapping from \argst to \val lists; the subject is not the first element of the \argst list. This analysis bears some resemblance to the distinction between macro-roles and syntactic pivots in Role and Reference Grammar, with passives having a marked mapping from macro-roles to syntactic pivot \citep{VanValinandLapolla1997}.  
In this kind of approach, the passive subject might be the second element of the \argst list, as in a typical personal passive, or an expletive element, as in impersonal passives.
In a long passive, the first element of \argst is coindexed with a PP on the \attrib{comps} list or an adjunct.
This analysis is reminiscent of the account of Balinese objective voice presented in Section \ref{ergativity} in that the account of both phenomena uses a non-canonical mapping between \argst and \val lists.
% ALERT:  We argued in some detail that passive differs from OV and can't work this way.  cp. binding "They don't seem to mind being insulted by each other."    Balinese would be:  "Each other insulted-OV them."
% TD 28 Jan. 20: Yes, I don't wish to imply that passive and OV are the same, just that the non-canonical mapping between arg-st abd valence in this treatment of passives resembles the account of OV presented earlier.  If that isn't clear, we should reword this so that it is.
%JPK 1-29-20 You are right, Steve, that our wording was unclear. I added something to the end of the sentence to make the parallel clearer. Hopefully, that clarifies things.
%SW 1-29-20.  Sorry, but it doesn't help. I'll email about it.
A version of this view is proposed in \citet[241]{Davis2001}, who proposes the representation in (\ref{pass-lxm}) for passive lexemes (as before, we substitute the attribute name \attrib{lex-dtr} for \attrib{in}).

\begin{exe}
\ex\label{pass-lxm}
{\avmoptions{center}
\begin{avm}
\[\asort{passive-verb}
subj & \@2  \\
comps & \@3 \\
arg-st & \@1$($\<xp\>$\textup{)}\oplus$\@2$\oplus$\@3 \\
lex-dtr & \[\asort{trans-stem}
			arg-st & \@1 \\
			content & \@4 \]\\
content & \@4
\]
\end{avm}}
\end{exe}

% TD 15 Jan. 20: There are some incidental differences between \ref{pass-lr} and \ref{pass-lxm} that potentially obscure the comparison we wish to make.  The first is presented as a lexical rule with an arrow between input and output, while the second is shown as a morpholigically complex lexical entry.  But that difference is irrelevant here, so we should present them in parallel fashion.  I don't have a strong opinion about which format to use, though I would say that if we ever want to refer to information in the stem, we might find it simpler to use the morpholigically complex lexical entry option.  We can say that we've modified the originals to make them easily comparable, and perhaps try to be consistent with what we say about lexical rules in the Lexicon chapter (e.g., change \attrib{stem} to \attrib{in}, as I've done just above).
% TD 15 Jan. 20: Also, it looks like the brackets around the outer \argst in \ref{pass-lxm} are a bit messed up.

We will refer to this as the non-canonical argument realization analysis of passives. Again, at least three 
issues must be addressed.
%JPK 1-29-20 I changed the phrasing of this sentence to remove an implicature of "questions" that was not meant.
%questions immediately present themselves.
First, the standard mapping between the elements of \argst and those of the \val lists is violated.
If passives violate these canonical mapping constraints, how strong an account of the relationship between \argst and \val can be maintained?
Second, as with the non-canonical linking analysis, what predictions, such as changes in binding behavior, control constructions, and discourse availability, arise from the non-canonical valence values in passives?
Third, what is the status of the \word{by}-phrase in long passives, and how is it represented on the \argst list?  If the logical subject remains the first element of a passive verb's \argst list, does it appear as an additional oblique element on \argst as well? 

The implications of weakening canonical constraints under each of these analyses have not been thoroughly addressed, to our knowledge.
%ALERT.  Wayan Arka and I addressed it. maybe not thoroughly enough  
% TD 28 Jan. 20: Sorry; in that case this should be changed to reflect that.
%JPK 1-29-20: Agree with Tony, feel free to edit the way you want, Steve.
We are unaware, for example, of proposals that limit non-canonical linking in HPSG to only the kind observed in passives.
One might begin by stipulating that linking concerns only NP (i.e., ``direct'') arguments on \argst, but the implications of this have not yet been well explored.
With respect to the non-canonical argument realization analysis, the required variation in \argst to \val mappings has been investigated somewhat more (see Section \ref{argst-valence-sec} for some details), especially in connection with ergativity and voice alternations, and also in analyses of pro-drop, cliticization, and extraction \citep{MillerandSag1997, Manning+Sag:1999, Boumaetal2001}.
Thus, there are some independent motivations for positing non-canonical mappings between \argst and \val lists.
But we will leave matters here in regard to the general advantages and drawbacks of non-canonical linking versus non-canonical argument realization.

As for passives in particular, the two analyses make different predictions regarding binding and control by the ``logical subject'' (the subject of the corresponding actives).
Under the non-canonical linking analysis, it is not present on the \argst (and \val) lists of short passives, so it is predicted to be unavailable to any syntactic process that depends on elements of \argst.
Binding and varieties of control that reference these elements therefore cannot involve the logical subject.
Under the non-canonical argument realization analysis, the logical subject is present on the \argst lists of short passives, so it is predicted to play much the same role in binding as it does in corresponding actives.
However, we can see that, at least when unexpressed, this is not the case, as in (\ref{no-bind}).
%ALERT:  Then why present this as an open question above?  oh ok... but still.
% TD 28 Jan. 20: We can reword that so it doesn't read like an open question; where is that, exactly?
\begin{exe}
\ex[*]{\label{no-bind}
The money was donated to himself. (\word{himself} intended to refer to the donor)
}
\end{exe}

Certain control constructions also illustrate this point.
While the unexpressed logical subject can control rationale clauses in English, as exemplified above in (\ref{mkb}), not all cases of control exhibit parallel behavior.
The Italian consecutive \word{da} + infinitive construction \citep{Perlmutter1984, Sanfilippo1998} appears to be controlled by the surface subject, as shown in (\ref{italian-da}).\il{Italian}

% TD 18 Jan. 20: these examples are from Sanfilippo (1998), (49a) and  (51)

\begin{exe}
\ex     \label{italian-da}
\begin{xlist}
\ex
\gll Gino ha  rimproverato Eva tante   volte da    arrabbiarsi.  \\
     Gino has scolded      Eva so.many times so.as to.get.angry  \\
\glt `Gino scolded Eva so many times that he/*she got angry.'
\ex
\gll Eva fu  rimproverata da Gino tante   volte da    arrabbiarsi.  \\
     Eva was scolded      by Gino so.many times so.as to.get.angry  \\
\glt `Eva was scolded by Gino so many times that *he/she got angry.'
\end{xlist}
\end{exe}

Although there are other factors involved in the choice of controller of consecutive \word{da} infinitive constructions, it is clear that the logical subject in the passivized main clause cannot control the infinitive.
Thus, even if it remains the initial element of the passive verb's \argst, it must be blocked as a controller.
Sanfilippo argues from these kinds of examples that the passive \word{by}-phrase should be regarded as a ``thematically bound'' (i.e., linked) adjunct that does
not appear on the passive verb's \argst list, but on the \attrib{slash} list.
However, this would require some additional mechanism to explain the involvement of the \word{by}-phrase in binding, noted below, and possibly with respect to other evidence for including adjuncts on \argst, as discussed in the Section \ref{sec:extended-arg-st}.

In addition, the implicit agent of short passives is ``inert'' in discourse, as discussed in \citet{KoenigandMauner1999}.
It cannot serve as an antecedent of cross"=sentential pronouns without additional inferences, as shown in (\ref{a-def}), where the referent of \word{he} cannot without additional inference be tied to the logical subject argument of \word{killed}, i.e., the killer.

\begin{exe}
\ex\label{a-def}
\# The president was killed. He$_{i}$ was from Iowa.
\end{exe}

Note that the discourse inertness of the implicit agent in (\ref{a-def}) does not follow from its being unexpressed, as shown by the indefinite use of the subject pronoun \word{on} in \ili{French} \citep[241--244]{Koenig1998c} or \ili{Hungarian} bare singular objects \citep[89--108]{FarkasandDeSwart2003}.
These, though syntactically expressed, do not introduce discourse referents either.
In such cases, as well as in passives under the non-canonical argument realization analysis, the first member of the \argst list must therefore be distinguished from indices that introduce discourse referents.
% TD 17 Jan. 20: Here's a bit that might be added to the paragraph above, with a brief mention of how ``inertness'' might be handled, and maybe Kathol's idea of two different kinds of indices (cross-ref to Agreement chapter for that as well).

These facts would seem to favor the non-canonical linking analysis.
However, there are options for representing the inertness of the logical subject under the non-canonical argument realization analysis.
One possibility is to introduce a special subtype of the type \type{index}, which we could call \type{inert} or \type{null}; by stipulation, it could not correspond to a discourse referent.
This is also one way to treat the inertness of expletive pronouns, so it has some plausible independent motivation.
Unlike expletive pronouns, the logical subject of passives is linked to an index in \attrib{content}.
Its person and number features therefore cannot be assigned as defaults (e.g., third person singular \word{it} and \word{there} in English), but must correspond to those of the entity playing the relevant semantic role in \attrib{content}.
\citet[251--253]{Davis2001} offers a slightly different alternative using the dual indices \attrib{index} and \attrib{a-index}, following the distinction between \attrib{agr} and \attrib{index} used in \citet[240--250]{Kathol1999b} to model different varieties of agreement.
The \attrib{a-index} of a passive verb's logical subject is of type \type{null}, which, by stipulation, can neither o-command other members of \argst nor appear on \val lists.
%This would be the case for both personal and impersonal passives.
In both impersonal and short personal passives, the logical subject is coindexed with a role in \attrib{content} representing an unspecified human (or animate).
These analyses of logical subject inertness have not been pursued, however.

% TD 17 Jan. 20: I'm not sure how this would work, actually.  For linking, the logical subject needs to be coindexed with the relevant role in \attrib{content}.  So that index can't be \type{null}.  Attempting to use Wechsler's \attrib{index} vs. \attrib{concord} distinction, we'd then heed to say that the logical subject's \attrib{concord} is the \type{null} element.  But my understanding of Wechsler's distinction is that a \type{null} value of \attrib{concord} wouldn't block binding.  Binding is still stated in terms of \attrib{index}.  I think Kathol's use of \attrib{a-index} was a bit different, but I haven't checked.


% TD 16 Jan. 20: I suggest deleting the following, now paraphrased elsewhere.
%Such an analysis, though, seems to run afoul of data that suggest that unexpressed agents of passives can bind reflexives or control \is{controlled argument} the subject of rationale clauses \is{rationale clause} we presented above and already discussed by Manning and Sag. For such reasons as well as the view that voice phenomena of the kind we presented in Section \ref{ergativity} for Balinese are better treated as different kinds of relations between the \argst list and the valence lists, a frequent approach these days seems to treat passive as a non-canonical relation between the \argst list and the valence lists (see \crossrefchapteralt{binding} and \crossrefchapteralt{complex-predicates} for scholars who assume such a view within this handbook). One possible version of this view is proposed in \citet[241]{Davis2001} who proposes the representation in (\ref{pass-lxm}) for passive lexemes.
%
%\begin{exe}
%\ex\label{pass-lxm}
%{\avmoptions{center}
%\begin{avm}
%\[\asort{passive-verb}
%subj & \@2  \\
%comps & \@3 \\
%arg-st & \@1(\<XP\>$\oplus$\@2$\oplus$\@3 \\
%stem & \[\asort{trans-stem}
%			arg-st & \@1 \\
%			content & \@4 \]\\
%content & \@4
%\]
%\end{avm}}
%\end{exe}

%The first member of the \argst list is not mapped onto a valence list, but the rest of the \argst list is mapped onto the valence lists as per the Argument Realization Principle, i.e. the second member of the \argst list is the value of the \attrib{subj} \isfeat{SUBJ} attribute and the rest of the \argst list is the value of the \attrib{comps} \isfeat{SUBJ} attribute.  \citep{Manning+Sag:1999} proposes a different version of this non-canonical alignment between \argst and valence lists approach to passives. Such differences are not critical in the context of this chapter; not is the representation of stems tentatively assumed by Davis. What matters is that, in contrast to the ``cycling'' argument structure approach to passives illustrated in (\ref{pass-lr}), the \argst list remains unchanged, what is changed is how that list relates to valence lists.

%Several issues remain to be addressed in the non-canonical alignment approach to passives. We simply mention them here and leave a full treatment to some other venue. 

% TD 17 Jan. 20: we need to tie what we say about by-phrases here to what we say above, which will require a bit of editing and maybe another example or two (the Russian ones from Perlmutter are clear in showing that the by-phrase --- actually an instrumental case NP --- can bind another argument.

% TD 19 Jan. 20: Slightly revised and augmented text on by-phrases
Finally, we turn to \word{by}-phrases in long passives.
In languages like English, a \word{by}-phrase can express the lexeme's logical subject.
Under both the non-canonical linking and non-canonical argument realization analyses, this might be represented as an optional oblique complement on \argst, as indicated in (\ref{pass-lr}) and (\ref{pass-lxm}), respectively.
As noted, the non-canonical argument realization analysis would then posit that the \argst of passives includes two members that correspond to the same argument, which again shows the need for an inert first element of \argst.
Another possibility is to treat such \word{by}-phrases as adjuncts (and therefore not part of the \argst list), see \citet[Chapter 7]{Hoehle78a} and \citet[292--294]{Mueller2003e} for German and \citet[180]{Jackendoff1990} for English. 
%or not targeting the logical subject per se \citep{Jackendoff1990}.
There is evidence, however, that \word{by}-phrases can serve as antecedents of anaphors in at least some languages.
\citet[111]{Collins2005} cites sentences like (\ref{by-bind}), which suggest that the complement of \word{by}-phrases can bind a reciprocal.

\begin{exe}
\ex\label{by-bind}The packages were sent by the children to each other.
\end{exe} 

Acceptability judgements of this and similar examples vary, but they are certainly not outright unacceptable.  Likewise, \citet[10]{Perlmutter1984} furnishes \ili{Russian} examples in which the logical subject (realized as an instrumental case NP) binds a reflexive (note that the English translation of it is also fairly acceptable).

\begin{exe}
\ex     \label{arg-st:russian-pass}
\gll Eta kniga byla kuplena Boris-om dlja sebja.  \\
     this book was bought Boris-\textsc{ins} for self  \\
\glt `This book was bought by Boris$_{i}$ for himself$_{i}$.'
\end{exe}

Given that binding is a relation between members of the \argst list, such data would seem problematic for an approach that does not include \word{by}-phrases on the \argst list.
Interestingly, Perlmutter also argues that Russian \word{sebja} is subject-oriented (see \crossrefchapteralt[Section~\ref{binding-sec-passive}]{binding}).
The intrumental NP \word{Borisom} can bind \word{sebja}, only because it corresponds to the subject of active \word{kupit'}, `buy'.
Assuming that is correct, an HPSG account of Russian passives would need some means of representing the logical subjecthood of these instrumental NPs; this might involve some way of accessing their active counterpart's \subj value, or of referencing the first element of the passive verb's \argst list, despite its inertness.

The interaction of binding and control with passivization across languages appears to be varied, and as we have noted, we are not aware of systematic investigations into this variation and possible accounts of it within HPSG.
Here, we have surveyed these phenomena and two possible approaches, while noting that some key issues remain unresolved.
Notably, both of these approaches introduce non-canonical lexical items, violating either linking or argument realization constraints that otherwise have strong support. 
Further work is required to assure that these can be preserved in a meaningful way, as opposed to allowing non-canonical structures to appear freely in the lexicon.
% TD 16 Jan. 20: Reworded the following short paragraph to make it more of a summary of the entire section
%Several possibilities suggest themselves here, but, as our goal, was to point out some of the difficulties that remain with theories of passives in HPSG, we will not pursue the matter further in this chapter.

\subsection{Summary}

We have examined in this section several approaches to argument alternations in HPSG and their implications for \argst.
For alternations based on semantic differences, different alternants will have different \attrib{content} values, and linking principles like those we outlined in the previous section account for their syntactic differences.
Even where such meaning differences are small, there are differing semantic entailments that can affect linking.
For some cases where there seems to be no discernible meaning difference between alternants, it is still possible for linking principles to yield syntactic differences, if the alternants select different \attrib{key} predications in \attrib{content}.
The active/passive alternation, however, cannot be accounted for in such a fashion, as it applies to verbs with widely varying \attrib{content} values.
HPSG accounts of passives therefore resort to lexical items that are non-canonical, either in their linking or in their mapping between \argst and \val.
Both of these are ways of modeling the demotion of the logical subject.
But there is as yet no consensus within the HPSG community on the correct analysis of passives.

%TD 14 Jan. 20: this paragraph will now need to be modified a bit and integrated with the new material
% TD 16 Jan. 20: At this point, I believes the following has already been covered, so I suggest deleting it
%One typical HPSG method for modeling valence alternations like passives is through lexical rules (see \crossrefchapteralt{lexicon}) with one alternant serving as input and the other as output; the main effect of the lexical rule in such an approach is to alter the \argst of the input, going from the \argst of \word{give} to that of \word{given} in (\ref{pasargst}). Critically, we must assume that the output cannot be subject to linking constraint (\ref{act-vb-linking}), since the actor argument is not linked to the first member of the \argst list.
%A simplified representation of what such a rule would look like is provided in (\ref{pass-lr}) where we assume that the input to the rule must be transitive ($\oplus$ indicates list concatenation).

%\begin{exe}
%\ex\label{pass-lr}
%{\avmoptions{center}
%\begin{avm}
%	\[\asort{$\textit{trans-vb}$}
%	head & $\textit{verb}$ \\
%		arg-st & \< \textsc{NP}$_{\@1}$, NP$_ {\@2}$\> $\oplus$ \@3$\textit{list}$
%	\] \,
%	$\mapsto$
%	\[head & \[vform & pass \] \\
%	arg-st & \<NP$_{\@2}$\> $\oplus$\@3 $\oplus$
%	\<PP[$_{by}$]$_{\@1}$\>
%	\]
%	\end{avm}
%}

%\end{exe}


%To sum up, in contrast to most meaning-driven alternations, valence alternations like the active/passive are  modeled through the use of lexical rules that alter the \argst of ``base'' entries. Which alternations pattern with active/passive and require positing lexical rules that alter a ``base'' entry's \argst list is, as of yet, not settled. 
%We turn now to roles that are putatively present semantically, but not realized syntactically at all. 

%JPK 1-18-20: I suggesting removin the above paragraph.
% TD: agreed

\is{passive)}
\is{alternation|(}
\section{Extended \argst}
\label{sec:extended-arg-st}
\is{ARG-ST!extended|see{resultative, clause union}|(}

Most of this chapter focuses on cases where semantic roles linked to the \argst list are arguments of the verb's core meaning. But in quite a few cases, complements (or even subjects) of a verb are not part of this basic meaning; consequently, the \argst list must be extended to include elements beyond the basic meaning. We consider three cases here, illustrated in (\ref{fish})--(\ref{adj}).  

Resultatives, \is{resultative} illustrated in (\ref{fish}), express an effect, which is caused by an action of the type denoted by the basic meaning of the verb. The verb \textit{fischen} `to fish' is a simple intransitive verb (\ref{fish}a) that does not entail that any fish were caught, or any other specific effect of the fishing (see \citealt[219--220]{Mueller2002b}).  
\il{German}

\begin{exe}
\ex\label{fish}
\begin{xlist}
\ex
\gll dass er  fischt\\
     that he  fishes\\
\glt `that he is fishing'
\ex 
\gll dass er ihn leer fischt\\
     that he it empty fishes\\
\glt `that he is fishing it empty'
\ex 
\gll wegen der Leerfischung der Nordsee \\
     because.of the empty.fishing of.the North.See.\textsc{gen} \\
\glt `because of the North Sea being fished empty' (\word{die tageszeitung}, 06.20.1996, p.\,6)
\end{xlist}
\end{exe}

\noindent
 In (\ref{fish}b) we see a resultative construction, with an object NP and a secondary predicate AP.  The meaning is that he is fishing, causing it (the body of water) to become empty of fish.  \citet[241]{Mueller2002b} posits a German lexical rule applying to the verb that augments the \argst list with an NP and AP, and adds the causal semantics to the \content (see \citealt{Wechsler2005result} for a similar analysis of English resultatives).     The existence of deverbal nouns like \textit{Leerfischung} `fishing empty', which takes the body of water as an argument in genitive case (see \ref{fish}c) confirms that the addition of the object is a lexical process, as noted by \citet{Mueller2002b}.  

%any effect or change of state due to the hammering. Since there is evidence th
Romance clause-union structures as in (\ref{faire}) have long been analyzed as cases where the complements of the complements of a clause-union verb (\word{faire} in (\ref{faire})) are complements of the clause-union verb itself \citep{Aissen1979}.
\il{French} 

\begin{exe}
\ex \label{faire}
\gll Johanna a fait manger les enfants. \\
Johanna has made eat the children \\
\glt `Johanna had the children eat.'
\end{exe}

\noindent
Within HPSG, the ``union'' of the two verbs' dependents is modeled via the composition of \argst lists of the \isi{clause union} verb, following \citet{HinrichsandNakazawa1994} (this is a slight simplification; see  \crossrefchapteralt{complex-predicates} for details). 



%JPK 1-14-2020: The following paragraph has been changed

\citet{AbeilleandGodard1997} have argued that many adverbs \is{adverbs as complements} including \word{souvent} in (\ref{adj}) and negative adverbs and negation in French are complements of the verb, and \citet{KimandSag2002} extended that view to some uses of negation in English. Such analyses hypothesize that some semantic modifiers are realized as complements, and thus should be added as members of \argst (or members of the \deps list, if one countenances such an additional list; see below). In contrast to resultatives, \is{resultative} which affect the meaning of the verb, or to clause union, where one verb co-opts the argument structure of another verb, what is added to the \argst list in these cases is typically considered a semantic adjunct and a modifier in HPSG (thus it selects the verb or VP via the \attrib{mod} attribute). 
\il{French}

\begin{exe}
\ex\label{adj}
 \gll
	Mes amis m’ ont souvent aidé. \\
	my friends me have often helped \\
	\glt `My friends often helped me.'
\end{exe}


Another case of an adjunct that behaves like a complement is found in (\ref{mourir}).  The clitic \word{en} expressing the cause of death   is not normally an argument of the verb \textit{mourir} `die', but rather an adjunct \citep{KoenigandDavis2006}: % discuss examples such as (\ref{mourir}).

\begin{exe}
	\ex\label{mourir}
	\gll Il en est mort. \\
	 he of.it is dead.\textsc{pfv.pst} \\
	 \glt `He died of it.' \citep[ex.\ 12a]{KoenigandDavis2006}
\end{exe}

\noindent
On the widespread assumption (at least within HPSG) that pronominal clitics are verbal affixes \citep{MillerandSag1997}, the adjunct cause of the verb \word{mourir} must be represented within the entry for \word{mourir}, so as to trigger affixation by \word{en}. \citet{Boumaetal2001} discuss cases where ``adverbials'', as they call them, can be part of a verb's lexical entry. To avoid mixing those adverbials with the argument structure list (and having to address their relative obliqueness with syntactic arguments of verbs), they introduce  an additional list, the dependents list (abbreviated as \deps) which includes the \argst list but also a list of adverbials. Each adverbial selects for the verb on whose \deps list it appears as an argument, as shown in (\ref{deps}). But, of course, not all verb modifiers can be part of the \deps list, and Bouma, Malouf, and Sag discuss at length some of the differences between the two kinds of ``adverbials''.

\begin{exe}
\ex\label{deps}	\type{verb} $\Rightarrow$
{\avmoptions{center}
\begin{avm}
	\[cont$|$key & \@2 \\
	  head & \@3 \\
	  deps & \@1 $\oplus$ $\textit{list}$(\[\textup{\textsc{mod}} & \[\textup{\textsc{head}} & \@3 \\
	  									 \textsc{key} & \@2\]\]) \\
	  arg-st & \@1
	\]
\end{avm}}
\end{exe}

Although the three cases we have outlined result in an extended \argst, the ways in which this extension arises differ. In the case of resultatives, \is{resultative} the extension results partly or wholly from changing the meaning in a way similar to \citet{RappaportandLevin1998}: by adding a causal relation, the effect argument of this causal relation is added to the membership in the base \argst list (see Section \ref{alternations} for a definition of the attributes \attrib{key} and \attrib{rels}; here it suffices to note that a \type{cause-rel} is added to the list of relations that are the input of the rule). 
%% ALERT:  is this about 51? if so then I don't quite follow how 51 works.  
%JPK 1-29-20: The causal relation is added to the list of RELS. It's flat (MRS), but isn't that what is needed for resultatives? 51 is oversimplified in many ways,  but it seems to me to get the idea across. Am I missing something.
%SW 1-29-20:   51 seems a little too vague.  e.g. it could represent resultative or causative. 
%JPK 1-31-20: I agree, but I am not sure it is worth going into more details. You think we should? 
The entries of the \isi{clause union} verbs are simply stipulated to include on their \argst lists the syntatic arguments of their (lexical) verbal arguments in (\ref{faire-ent}).
%TD 13 Jan. 20: reworded the definition of shuffle
%SW: jan 28. sorry, I rereworded it.  they dont have to be 'intermixed' do they?  ALERT
% TD 28 Jan. 20: Thanks; that's better
The symbol $\bigcirc$ \is{$bigcirc$|see{shuffle}} in this rule is known as ``shuffle''\is{shuffle}; it represents any list 
% in which the elements of the two lists are intermixed
containing the combined elements of the two lists, but with the relative ordering of elements on each list preserved.
In (\ref{faire-ent}), shuffling the members of the verbal complement's \argst list with those of the main verb's \argst list allows us represent a possible reordering of them (for example in passivization -- depending on one's approach to passivization in HPSG; see \crossrefchaptert{complex-predicates}).
%JPK 13 Jan 2020: Added a definition of the shuffle operation
Finally, (negative) adverbs that select for a verb (VP) are added to the \argst of the verb they select. A simplified representation of all three processes is provided in (\ref{res-lr})--(\ref{neg-lr}). 


\begin{exe}
	\ex\label{res-lr}
	{\avmoptions{center}
	\begin{avm}
		\[key & \@2 \\ rels & \@1\<\ldots \@2 \ldots \>\]
	\end{avm}
	$\mapsto$
	\begin{avm}
		\[key & \@3$\textit{cause-rel}$ \\ rels & \@1 $\oplus$ \@3\]
	\end{avm}
	}
\end{exe}

\begin{exe}
\ex\label{faire-ent}
	{\avmoptions{center}
	\begin{avm}
		\[arg-st & \<\ldots, \[head & $\textit{verb}$ \\
								arg-st & \@1\> \] $\bigcirc$ \@1\]
	\end{avm}
	}
\end{exe} 

\begin{exe}
	\ex\label{neg-lr}
	{\avmoptions{center}
	\begin{avm}
		\[arg-st & \@1\]
	\end{avm}
	$\mapsto$
	\begin{avm}
		\[arg-st & \@1 $\bigcirc$ \<\textsc{adv}$_{neg}$\>\]
	\end{avm}
	}
\end{exe}

\is{ARG-ST!extended|see{resultative, clause union}|)}
 

\section{Is \argst universal?}

% TD 31 Jan. 20: I have not examined the new version of this section in detail, but I suggest something like the following approach to this material (which might enable us to make it more concise):
% TD 31 Jan. 20: start with somthing like the following (commented out at the moment:
% We have examined several functions of \argst thus far, motivated by evidence drawn from arange of languages.
% It is an ordered list of arguments, indexed both to roles in \attrib{content} and to elements of \val lists.
% Oredering on the \argst list is determined both by a predicator's subcategorization requirements and by the semantic properties (entailments) of its roles.
% But as extraction, pro-drop, null anaphora, and voice alternations show, the relationship between semantic roles and \val lists is complex and \argst provides the needed interface.
% In articular, binding principles are stated with respect to ordering on \argst.
%
% If there are languages in which none of the foregoing kinds of evidence are to be found, the question of whether \argst should be posited in such language arise.
% This is in large part a methodological question, the answer to which is driven by one's stance on the relative importance of positing universal features across languages, even when there is no evidence supporting them in a particular language.
% Here, we review one case of this nature.

% TD 31 Jan. 20: Then, very briefly explain why there's no vidence supporting these functions of \argst in Oneida.  I'd say it's less important to introduce IMFL-STR than to present the idea that \argst isn't motivated.  I don't know if this will shorten the presentation or not, and I'm suggesting this just to rethink the presentation a bit, so if it doesn't strike you as the right way to go, just forget what I've writen here.

HPSG's \argst attribute does not seem to be a universal property of natural
language grammars.  The \argst feature is the intermediary between, on
the one hand, a semantic representation of an event or state in which
participants fill specific roles, and on the other, their syntactic and
morphological expression.  \argst is defined as a list of \type{synsem}
objects in the entry for a verb lexeme, and is used to model the following grammatical regularities of particular predicators or sets of predicators:

\begin{itemize}
\item The verb selects grammatical features such as part of speech
category, case marking, and preposition forms of its dependent
phrases.
\item \argst list items:
\begin{enumerate}
\item are unified with \val list items
representing grammatical properties of phrasal dependents (subject and
complements),
\item determine verbal morphology, or
\item are left unexpressed.
\end{enumerate}
\item The different contexts of occurrence of a verb correspond to distinct \argst values.
\item The inflected verb can indicate agreement features on both phrasal
and affixal arguments.
\item Binding conditions on arguments are defined on \argst, making
crucial use of the ordering relation (``obliqueness'').
\end{itemize}

\citet{KoenigandMichelson2014,KoenigandMichelson2015a,KoenigandMichelson2015b} argue that the grammatical encoding of semantic arguments in Oneida (Northern Iroquoian) \il{Oneida} does not display any of these properties.  In fact, the only function of the
corresponding intermediate representation in Oneida is to distinguish
the arguments of a verb for the purpose of determining verbal
prefixes indicating semantic person, number, and gender
features of animate arguments.  
For example,  the prefix \word{lak-} occurs if a third-person singular masculine proto-agent argument is acting on a first-person singular proto-patient argument as in \word{lak-hlo·lí-he\textipa{P}}  `he tells me’ (habitual aspect), whereas  the prefix \word{li-} occurs if a first singular proto-agent argument is acting on a third masculine singular argument, as in \word{li-hlo·lí-he\textipa{P}} `I tell him’ (habitual aspect).
 As there is no syntactic agreement, these verbal prefixes encode purely semantic features.
\type{Synsem} objects are therefore not appropriate for this intermediate
representation; all that is needed are semantic argument indices to
distinguish between (a maximum of two) animate co-arguments
distinguished for fixed argument roles for each verb.
Koenig and Michelson use the attribute \attrib{infl-str},\isfeat{INFL-STR} which is a list of referential indices for animate arguments, within the morphological information associated with each verb (within the value of \attrib{morph}; see \crossrefchapteralt{morphology}) for this highly restricted
Oneida feature that replaces \argst. If Koenig and Michelson are correct, the \argst list may thus not be a universal attribute of words, though present in the overwhelming majority of languages.
Linking, understood as constraints between semantic roles and members of the \argst list, is then but one possibility; constraints that relate semantic roles to an \attrib{infl-str} list of semantic indices is also an option.
In languages that exclusively exploit that latter possibility, syntax is indeed simpler.



%In this section, we briefly consider the question of whether something akin to Figure \ref{fig:over} offers a satisfying account of the grammatical encoding of semantic arguments  across all languages.
%Because of its role in accounting for the syntax of basic clauses, the presence of an \argst list on lexical entries comes with expectations about the syntactic realization of semantic roles.
%In recent work, \citet{KoenigandMichelson2015a} \il{Oneida} argue these expectations are not universally borne out, based on data from Oneida (Northern Iroquoian). The only grammatical reflex of semantic arguments in Oneida, they argue, is inflectional: the referencing of semantic arguments by so-called pronominal prefixes, which are better thought of as agreement markers \`a la \citet{Evans2002}. 
%If a language like \ili{Oneida} (Northern Iroquoian) only includes an ordering of semantic indices for inflectional purposes and no \argst list, a number of predictions follow, which Koenig and Michelson claim are borne out. Briefly summarizing their evidence, the relation between semantic arguments and external phrases, when they occur, is not necessarily one of co-indexing, no binding constraints exist between external phrases (e.g., condition C violations can be found), there are no valence alternations, and no syntactic constraints on extraction. In other words, Oneida contains none of the evidence supporting the presence of an \argst list and an ordering of syntactic arguments along an obliqueness hierarchy we have discussed in this chapter.


%Koenig and Michelson distinguish between grammatical and syntactic arguments (whether syntactically realized or unexpressed).
%Grammatical arguments include not only syntactic arguments (that is, those on \argst) but also inflectional referencing of semantic roles. 
% ALERT.  Help!  I'm confused.  I thought affixal arguments were on ARG-ST.  as needed for binding etc.  anyway that's what we said in sec. 3.1
%JPK 1-29-20: Hmm, we spend a few papers explaining the issue and I would be thrilled to include that here, but wouldn't that be leading us a little off topic? See my next point for the gist of the argument.
%SW 1-29-20:  the words "syntactic arguments (that is, those on \argst)" imply to me that only syntactic arguments are on \argst.  Is that what you mean? That's not what I have come to believe and not what we say in Sec. 3.1.  It;s not just syntactic arguments but also (some) null arguments and affixal arguments that appear on \argst.  That's how binding of null and affixal arguments works in HPSG-- binding is on ARG-ST.  Syntactic arguments correspond to VALENCE list items, not ARG-ST.  So this just seems wrong, as far as I can tell.
%JPK 1-31-20: Got it. The expression "syntactic arguments" is explained in the paper referenced here and unexpressed members of arg-st are syntactic arguments as per this definition (they play a role in control, e.g., or whatever). In contrast to things that "disappear" from the arg-st list. I added a parenthesis to clarify. Hope it does.
%Some ordering analogous to linking in other languages is present in Oneida, because the semantic roles are not arbitrarily associated with agreement morphemes, but this can be captured in an ordered list of semantic indices, called \attrib{infl-str} \isfeat{INFL-STR} in their proposal whose value it also a member of the Morphosyntactic feature Set (\attrib{ms}) that regulates exponence in Information-Based Morphology, see \citealt{CrysmannandBonami2016} and \crossrefchapteralt{morphology}, and \citet{Diazetal2019} for its application to Oneida prefixal verbal inflection in Oneida in particular. \attrib{infl-str} is part of the morphological information relevant to word-internal inflectional processes, what  \citet{Anderson1992} calls `Morphosyntactic Representation' in his treatment of Georgian agreement markers. The ordering of semantic indices on \attrib{infl-str} ensures that the predicator is properly inflected. For example,  the prefix \word{lak-} occurs if a third singular masculine proto-agent argument is acting on a first singular proto-patient argument as in \word{lak-hlo·lí-he\textipa{P}}  `he tells me’ (habitual aspect), whereas  the prefix \word{li-} occurs if a first singular proto-agent argument is acting on a third masculine singular argument, as in \word{li-hlo·lí-he\textipa{P}} `I tell him’ (habitual aspect). 

%It is tempting to think of the \attrib{infl-str} list as similar to the \argst list and some of the constraints ordering members of \attrib{infl-str} resemble linking constraints proposed in various HPSG approaches to linking. Compare (\ref{KD-03}) from \citet{KoenigandDavis2003} for ordering \type{synsem} members of \argst and and (\ref{KM-ex}) from \citet{KoenigandMichelson2015a} for ordering semantic indices on \attrib{infl-str}.

%\begin{exe}
	%\ex\label{KD-03}
%	{\avmoptions{center}
%	\begin{avm}\[content & \[\asort{$\textit{cause-rel}$}
%												causer & \@1\] \\
%                   arg-st & \<np, \ldots \>\]\end{avm}
%$\Rightarrow$ \begin{avm}\[
   %                        arg-st & \<np:\@1, \ldots\>\]
%\end{avm} 
%	}
%\end{exe}
%\begin{exe}
%\ex\label{KM-ex}	
%{\avmoptions{center}\begin{avm}\[content & \[\asort{$\textit{cause-rel}$}
%                                              causer & \@1anim\] \]\end{avm}
%$\Rightarrow$ \begin{avm}\[
 %                          infl-str & \<\@1, \ldots\>\]
%\end{avm}}
%\end{exe}


%But, there are critical differences that suggest the two lists have very different functions. We alluded above to the fact that none of the functions of \argst we discussed in this chapter apply to Oneida (e.g., binding, valence alternations, argument realization). 
%But, \citet{KoenigandMichelson2014,KoenigandMichelson2015b} provide additional evidence for the difference between semantic indices referenced morphologically through inflection and members of \argst. We briefly summarize some of their arguments here, First, members of \argst are of type \type{synsem} and include not only semantic indices, but semantic content, and, crucially, syntactic information about a verb's dependents, thus allowing the syntactic selection by a verb of its dependents. 
%In contrast, members of \attrib{infl-str} are merely semantic indices: asking for which syntactic features members of \attrib{infl-str} have would be a category mistake. Morphological referencing of semantic arguments in Oneida thus differs from morphological referencing in other polysynthetic languages where morphological referencing does not prevent, say, case marking on the corresponding \emph{syntactic} arguments (as in Eskimo languages). It differs also from morphological referencing of arguments by so-called object affixes in Romance languages where syntactic properties such as case or obliqueness affects morphological referencing (see \citealt{MillerandSag1997} for French and \citealt{Monachesi1999} for Italian within HPSG).
%Second, members of \argst are, canonically, mapped onto valence lists: \argst regulates (potiental) argument realization, to borrow a term from \citet{LevinandRappaport2005}. Members of \attrib{infl-str}, on the other hand, are never realized and, therefore mapped onto valence lists and plays no in the syntax of the language. Third, membership in \attrib{infl-str} is strictly determined morphologically and is not a ``simple'' reflection of semantic arguments (including syntactically unexpressed arguments as we discuss, for example, in Section \ref{passives}). As argued in detail in \citet{KoenigandMichelson2015}, only \emph{distinct},  \emph{animate} semantic arguments are members of \attrib{infl-str}. Thus, reflexive/reciprocal stems in Iroquoian bear a so-called intransitive prefix, i.e. mark only one argument, and verb forms that describe situations with three animate participants are only possible if the stem is reflexive/reciprocal, i.e. if only one member of \attrib{inf-str} corresponds to these two semantic arguments. Furthermore, and more importantly, the index of inanimate semantic arguments is never included on the \attrib{infl-str} list; thus no proposition or situation argument is ever part of the \attrib{infl-str} list and any semantically dyadic verb one of whose argument is inanimate behaves as a monadic verb: the inanimate argument is invisible grammatically. In other words, membership in \attrib{infl-str} is determined on a very different basis than membership in \argst. Finally, the relationship between the \attrib{infl-str} and morphological form is different from the relation between \argst and valence (and ultimately, syntactic realization). In the latter case, the expectation, at least as a default, is that distinct members of \argst have distinct syntactic realization. But any two-membered \attrib{infl-str} list is always expounded by a single prefix in Oneida. In other words, the exponence of \attrib{infl-str} feature values treats these values as single complex morphosyntactic features rather than a list of two distinct indices whose exponence is independent of each other. The previous arguments simply attest of the fact that inflection is not syntax and further corroborate that there is nothing like \argst in Oneida (see \citealt{KoenigandMichelson2015b} for further evidence of the morphological rather than syntactic nature of \attrib{infl-str}).



% ALERT.  I don't understand how Oneida differs from other polysynthetic  languages such that it requires its own special snowflake list.  why not use ARG-ST?  as in sec. 3.1, ex. 5.  The inflections could be agreement, a la Austin and Bresnan. 
%JPK 1-29-20: contrary to popular opinions (:)), not all polysynthetic languages are created equal and work the same way. It happens that there is no evidence for an arg-st list in Oneida and having it forces one to answer questions that have no empirical facts bearing on them (what's the order, ...). I am not sure I understand "snow-flake list", do you mean INFL-STR? If that's correct, this is not that weird fare from a morphologist point of view. My suspicion is that a worked out morphological analysis of Australian languages would basically have the same kind of analysis FOR THE MORPHOLOGY. The only thing that's unusual is the ordering you need, but that's something you need for languaages that have mnore than agreement marker (see Anderson 1992 for Georgian) (of course you can have a diacritic that amounts to the same as an ordered list). The only unusual thing about Oneida is that the pronominal prefixes are portmanteau-like in that you have one exponent for the combinations of members of INFL-STR.
%SW 1-29-20  Suppose you take your analysis, JP, but replace the attribute name INFL-STR with "ARG-ST" (but do everything else exactly as you have done it).  What aspect of that use of ARG-ST would make it unlike other uses of ARG-ST in HPSG work?   I.e. what would be un-ARG-ST-like about it?    Conversely, what ARG-ST-like properties are missing from Oneida?  Unless you answer that (in this paper) then it just looks like you renamed a feature.  
%JPK 1-31-20: I greatly expanded this section to answer your comment. Hopefully that clarifies things. 

\section{The lexical approach to argument structure}
\label{lexicalapproach}

We end this chapter with a necessarily brief comparison between the approach to argument structure we have described here and other approaches to argument structure that have developed since the 1990s.
This chapter describes a \textit{lexical approach to argument structure}, which is typical of research in HPSG.  The basic tenet of such approaches is that lexical items include argument structures, which represent essential information about potential argument selection and expression, but
abstract away from the actual local phrasal structure.  In contrast, \emph{phrasal approaches}, which
are common both in Construction Grammar \is{Construction Grammar} and in transformational approaches such as Distributed Morphology,\is{Distributed Morphology} reject such lexical argument structures.   Let us briefly review the reasons for preferring a lexical approach. (This section is drawn from \citealt{MWArgSt}, which may be consulted for more detailed and extensive argumentation.) 

In phrasal approaches \is{argument realization!phrasal approaches to|(} to argument structure, components of a verb's apparent meaning are actually ``constructional meaning'' contributed directly by the phrasal structure.  The linking constraints of the sort discussed above are then said to arise from the interaction of the verb meaning with the constructional meaning.  For example, agentive arguments tend to be realized as subjects, not objects, of transitive verbs.  On the theory presented above, that generalization is captured by the linking constraint (\ref{act-vb-linking}), which states that the \textsc{actor} argument of an \textit{act-und-rel} (`actor-undergoer relation') is mapped to the initial item in the \argst list.  In a phrasal approach, the agentive semantics is directly associated with the subject position in the phrase structure.  In transformational theories, a silent ``light verb'' (usually called ``little \textit{v}'') heads a projection in the phrase structure and assigns the agent role to its specifier (the subject).  In constructional theories, the phrase structure itself assigns the agent role.  In either type of phrasal approach, the agentive component of the verb meaning is actually expressed by the phrasal structure into which the verb is inserted.  


The lexicalist’s approach to argument structure provides essential information for a verb's potential
combination with argument phrases.   If a given lexical entry could only  combine with the particular set of phrases specified in a single \val feature, then the lexical and phrasal approaches would be difficult to distinguish: whatever information the lexicalist specifies for each \val list item could, on the phrasal view, be specified instead for the phrases realizing those list items.  
But crucially, the verb need not immediately combine with its specified
arguments.  Alternatively, it can meet other fates: it can serve as the input to a lexical rule; it
can combine first with a modifier in an adjunction structure; it can be coordinated with another
word with the same predicate argument structure; instead of being realized locally, one or more of
its arguments can be effectively transferred to another head’s valence feature (raising or argument
transfer); or arguments can be saved for expression in some other syntactic position (partial
fronting).   Here we consider two of these, lexical rules and coordination.  
 
The lexically encoded argument structure is abstract: it does not directly encode the phrase structure or
precedence relations between this verb and its arguments. This abstraction captures the commonality
across different syntactic expressions of the arguments of a given root.

\begin{exe}
\ex \label{nibble}
\begin{xlist}
\ex  The rabbits were nibbling the carrots.  
\ex  The carrots were being nibbled (by the rabbits).
\ex  a large, partly nibbled, orange carrot 
\ex  the quiet, nibbling, old rabbits
\ex  the rabbit's nibbling of the carrots
\ex  The rabbit gave the carrot a nibble.  
\ex  The rabbit wants a nibble (on the carrot).  
\ex  The rabbit nibbled the carrot smooth.
\end{xlist}
\end{exe}

\noindent
Verbs undergo  morpholexical operations like passive
(\ref{nibble}b), as well as antipassive, causative, and applicative in other languages.  They have cognates in
other parts of speech such as adjectives  (\ref{nibble}c, d) and nouns  (\ref{nibble}e, f, g).  
Verbs have been argued to form complex predicates with resultative secondary predicates (\ref{nibble}h), and with serial verbs in other languages.\is{resultative}   

The same root lexical entry \emph{nibble}, with the same meaning, appears in all of these contexts.
The effects of lexical rules together with the rules of syntax dictate the proper argument
expression in each context.  For example, if we call the first two arguments in an \argst list
(such as the one in (\ref{nibble}) above) Arg1 and Arg2 (or \attrib{act} or \attrib{und}), respectively, then in an active
transitive sentence Arg1 is the subject and Arg2 the object; in the passive, Arg2 is the subject and
the referential index of Arg1 is optionally assigned to a \emph{by}-phrase.  The same rules of
syntax dictate the position of the subject, whether the verb is active or passive.  When adjectives
are derived from verbal participles, whether active (\emph{a nibbling rabbit}) or passive (\emph{a
  nibbled carrot}), the rule is that whichever role would have been expressed as the subject of the
verb is assigned by the participial adjective to the referent of the noun that it modifies; see \citet{Bresnan:1982passive} 
and \citet[Chapter~3]{Bresnan+etal:2015}.  
The phrasal approach, in which the agent role is assigned to the subject position, is too rigid.  

Nor could this be solved by associating each syntactic environment with a different meaningful phrasal construction: an active construction with agent role in the subject position, a passive construction with agent in the \textit{by}-phrase position, etc.  The problem for that view is that one lexical rule can feed another.  In the example above, the output of the verbal passive rule (see (\ref{nibble}d)) feeds the adjective formation rule (see (\ref{nibble}e)).  
 
A verb can also be coordinated with another verb with the same valence requirements.  The two verbs then share their dependents.  This causes problems for the phrasal view, especially when a given dependent receives different semantic roles from the two verbs.  For example, in an influential phrasal analysis, \citet{hale+keyser:1993}
derived denominal verbs like \textit{to saddle} through noun incorporation out of a structure akin to
[PUT a saddle ON x].  Verbs with this putative derivation routinely coordinate and share
dependents with verbs of other types: 

\begin{exe}
\ex Realizing the dire results of such a capture and that he was the only one to prevent it, he quickly
[saddled and mounted] his trusted horse and with a grim determination began a journey that would
become legendary.\footnote{\url{http://www.jouetthouse.org/index.php?option=com_content&view=article&id=56&Itemid=63},
  21.07.2012}  
\end{exe}

\noindent
Under the phrasal analysis, the two verbs place contradictory demands on a single phrase structure.  But on the lexical analysis, this is simple V$^0$ coordination.  \is{argument realization!phrasal approaches to|)} 
 
To summarize, a lexical argument structure is an abstraction or generalization over various occurrences of the verb in syntactic contexts. To be sure, one key use of that argument structure is simply to indicate what sort of phrases the verb must (or can) combine with, and the result of semantic composition; if that were the whole story, then the phrasal theory would be viable. But it is not. As it turns out, lexically-encoded valence structure, once abstracted, can alternatively be used in other ways: among other possibilities, the verb (crucially including its valence structure) can be coordinated with other verbs that have a similar valence structure, or it can serve as the input to lexical rules specifying a new word bearing a systematic relation to the input word.   The phrasal approach prematurely commits to a single phrasal position for the realization of 
a semantic argument.  In contrast, a lexical argument structure gives a word the appropriate flexibility to account for the full range of expressions found in natural language.   
 

%\section*{Abbreviations}
%\section*{Acknowledgements}

%
{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this] 
}
\end{document}
