%% -*- coding:utf-8 -*-
\chapter{Foreword of the second edition}

%\largerpage
The second edition comes with a lot of small improvements: the index has been improved, typos have
been fixed, reference were added, and ORCIDs were added to authors and are displayed on the title pages of the papers now.

%1 properties
The type in the example (\ref{ex:prop38})
on p.\,\pageref{ex:prop38} was changed from \type{phrase} to \type{example-type}. As noted by
Philipp Trapp in 2022, the presence of the feature \textsc{head-daughter} would entail that the type
of the AVM is \type{headed-phrase} and since the type implication in (\ref{ex:prop38}) applies to all
structures of type \type{phrase} this would mean that all linguistics objects of type \type{phrase} have to be
of type \type{headed-phrase}, which would result in contradictions for all subtypes of \emph{phrase}
that are not of type \type{headed-phrase}. A schema for head-adjunct phrases (p.\,\pageref{ex-hd-adj-ph}) and a schema for
specifier-head phrases (p.\,\pageref{ex-head-spr-ph}) was added.

%2 evolution
The LILOG system is now mentioned in Chapter~\ref{chap-evolution} about the evolution of HPSG
(Section~\ref{sec-LILOG}). Links to historical sources were updated.

%4 lexicon
The argument realization principle in (\ref{wd-bouma}) on p.\,\pageref{wd-bouma} was fixed. It
contained too many brackets in the specification of the \depsl. Footnote~\ref{fn-subject-ARP} was
added to explain how constraints on the length of the \subjl can be enforced.

% 2023-11-21 SYNSEM|LOCAL in MORPH in (34) in lexicon.tex
%


%5 understudied languages
Many references were added to Chapter~\ref{understudied-languages}, the chapter on understudied languages. 

%6 agreement
% 04.09.2004
Example (\ref{ex-gender-agreement-historical-development}) on
p.\,\pageref{ex-gender-agreement-historical-development} was modified so that the respective
examples for historical stages of development of gender agreement are more similar.

%7 case
The example (\ref{ex:syn:gfr}) on p.\,\pageref{ex:syn:gfr} was changed to another example excluding the possibility that we
are dealing with an instance of non-matching free relative clauses.

%
% 31.01.2024 added Abbreviations for case.tex since illative is not in the Leipzig Glossing Rules.


%8 nominal structures
%9 arg-st
Footnote~\ref{fn-mapsto-lexical-rules} was added on p.\,\pageref{fn-mapsto-lexical-rules} to explain
the lexical rule notation using $\mapsto$ that was employed later in the chapter. A footnote
mentioning the mapping of non-canonical objects to valence features was added (p.\,\pageref{fn-non-cannonical-arg-st}). 
The type \type{ppro} used in (\ref{hablo}) on p.\,\pageref{hablo}, was explained.

%10 order
The relation \texttt{synsems2signs} was explained by adding (\ref{ex-schema-hc-flat-synsem-sign}),
which is an expansion of (\ref{schema-hc-flat}) on p.\,\pageref{schema-hc-flat}. The relation that
is used in the chapter on complex predicates has the same name now (see
(\ref{CP-ex-head-complements-phrase}) on p.\,\pageref{CP-ex-head-complements-phrase}). The
footnote~\ref{fn-order-lexical-Uszkoreit} was added. It discusses a lexical account of constituent
order assuming a separate lexical item for each ordering variant.

% order: 04.01.22
% Gray -> gray
% der Frau -> dem Kind
% added \ref{ex-schema-hc-flat-synsem-sign}

%11 complex predicates

% complex-predicates 04.01.22
% unified synsems2signs. The relation has the same name now in order.tex and complex-predicates.tex

% 17.01.2023 unified spelling of reduced-verb and basic-verb, added dot to example in
% complex-predicates-include.tex


%12 control and raising

%13 udc.tex
Some explanation about the lexical item for the empty element that is used in some analyses of
extraction was added to Section~\ref{sec:UDC:MoreOnGaps} of Chapter~\ref{chap-udc} and the 
caption of Figure~\ref{fig:UDC:Tree:ThatHeMightBeWrong} on
p.\,\pageref{fig:UDC:Tree:ThatHeMightBeWrong} was changed to reflect the fact that the analysis
involves a pseudo-filler rather than a so-called ``dishonest gap''.

% 25.01.22 Footnote~\ref{fn-hf-schema} was missing. % in udc

% 24.01.2023  removed space udc.tex, added Section to reference of Ross67 regarding ATB
% added crossref to island chapter.


%14 relative clauses
Relative pronouns are NPs. The respective representation in (\ref{x:rc-18}) on p.\,\pageref{x:rc-18}
was fixed. Valence features were added to the lexical item in (\ref{x:rc-17}).

% relative-clauses.tex 05.01.22
% \trace -> \trace{}
% glosses aligned in {x:rc-129}
% added language tag
% fixed index entry for Bavarian German

% 2023-12-12 Fixed NP for relative pronouns rather than N'.
% added SPR and COMPS for lexical item for relative pronoun
% Changed PP [3] into [LOC [3]] in figure in relative clause chapter
% Fixed PP[4], which should have been [3] in footnote in relclause chapter.


%15 islands
% 2024-09-07
A reference to recent work was added in Chapter~\ref{chap-islands}.

%16 coordination
\emph{Mary} is of category NP rather than N in Figure~\ref{coordphr} on p.\,\pageref{coordphr}. A
paragraph explaining (\ref{ex-binary-coordination}) was added on p.\,\pageref{ex-binary-coordination}.
%
% Bosque was checked. Data is not from there but from Anne.
% 2024-10-10

% 30.03.22 Figure 4, Mary is NP not N
%



%17 idioms
There was an NP too many in the \compsl in the lexical item in (\ref{le-idiomatic-spill}) on
p.\,\pageref{le-idiomatic-spill}. 
% 2024-09-07
References to recent publications were added to Chapter~\ref{chap-idioms}.

% 03.02.22 Idioms: NP in (8) too much, REL bad feature name, ref to Krenn&Erbach added



%18 negation
%19 ellipsis
%20 anaphoric binding
%21 morphology
%22 semantics
%23 information structure 

% 08.02.22 Information structure: added page numbers for Bildhauer & Cook 2010
%          fixed layout issue with Head-Dislocation Schema for Catalan
% 09.02.22 Added sentence about diff-list and reference to copestake2002.

%Diff-lists are now explained A bit of explanation and a reference was added to


%24 Processing
%2024-09-10
A note on psycholinguistic arguments for a lexical analysis of complex predicates was added to
Section~\ref{processing:sec-lexicalism} of Chapter~\ref{chap-processing}. 

%
% 14.02.22 Added glosses to helfen in chapter on processing
%






% 18.01.22 added language info for German examples


% 26.10.22 (38a) used to be phrase => but since the constraint referred to HD-DTR this would cause a
% conflict for unheaded phrases. Noticed by student Philipp Trapp.
% The left-hand daughter in (38b) must be SYNSEM X, noted by St.Mü.

% 01.11.22
% Daughter in head-filler-phrase must have SYNSEM|LOCAL instead of LOCAL. St. Mü.
%
% 22.11.22
% added comma in 14b in lexicon.tex
%
% 01.12.2022 index entries for \ominus
%
% 10.01.2023 Added comma in np.tex
%
%
%
% 2023-08-23 Kim Sells appeared in 2015 not in 2014, we missed this despite the check.
%
% 2023-09-26 The LILOG system is now mentioned in evolution.tex
%



%      <!-- Local IspellDict: en_US-w_accents -->
