\documentclass[output=paper]{langsci/langscibook} 
\author{%
	Joanna Nykiel\affiliation{Kyung Hee University, Seoul}%
	\lastand Jong-Bok Kim\affiliation{Kyung Hee University, Seoul}%
}
\title{Ellipsis}

% \chapterDOI{} %will be filled in at production

%\epigram{Change epigram in chapters/03.tex or remove it there }

\abstract{This chapter provides an overview of HPSG analyses of ellipsis. The structure of the chapter follows three types of ellipsis, nonsentential utterances, predicate ellipsis, and nonconstituent coordination, with three types of analyses applied to them. The analyses characteristically don't admit silent syntactic material for any ellipsis phenomena with the exception of certain types of nonconstituent coordination.}

\maketitle

\begin{document}
\label{chap-ellipsis}

{\avmoptions{center}

\section{Introduction}
         Ellipsis is a phenomenon that involves a noncanonical mapping between syntax and semantics. What appears to be a syntactically incomplete utterance still receives a semantically complete representation, based on the features of the surrounding context, be the context linguistic or nonlinguistic. The goal of syntactic theory is thus to account for how the complete semantics can be reconciled with the apparently incomplete syntax. One of the key questions here relates to the structure of the ellipsis site, that is, whether or not we should assume the presence of invisible syntactic material. This chapter begins by introducing three types of ellipsis (nonsentential utterances, predicate ellipsis, and nonconstituent coordination) that have attracted considerable attention and received treatment within HPSG. We next overview existing evidence for and against the so-called WYSIWYG (`What You See Is What You Get') approach to ellipsis, where no invisible material is posited at the ellipsis site. Finally, we walk the reader through three types of HPSG analyses applied to the three types of ellipsis presented in section 2.


\section{Three types of ellipsis}
         Depending on the type of analysis by means of which HPSG handles them, elliptical phenomena can be broadly divided into three types:
         nonsentential utterances, predicate ellipsis, and nonconstituent coordination.
          We overview the key features of these types here before discussing in greater detail how they have been brought to bear on the question of whether there is invisible syntactic structure at the ellipsis site or not. We begin with stranded XPs, which HPSG treats as nonsentential utterances, and then move on to predicate and argument ellipsis, followed by phenomena known as nonconstituent coordination.

\subsection{Nonsentential utterances}
This section introduces utterances smaller than a sentence, which we refer to as nonsentential utterances (NUs). These range from Bare Argument Ellipsis (BAE) (\ref{1}), including fragment answers (\ref{2}), to direct or embedded sluicing (\ref{3})-(\ref{4}). Sluicing hosts stranded wh-phrases and has the function of an interrogative clause, while BAE hosts XPs representing various syntactic categories and typically has the function of a declarative clause.\footnote{Several subtypes of nonsentential utterances can be distinguished, based on their contextual functions, which we don't discuss here (for a recent taxonomy, see Ginzburg 2011).}



\ea A: You were angry with them.\\ B: Yeah, angry with them and angry with the situation.\label{1}\z

\ea A: Where are we? \\B: In Central Park.\label{2}\z

\ea A: So what did you think about that?\\ B: About what? \label{3}\z

\ea A: There's someone at the door. \\B: Who?/I wonder who. \label{4}\z
The theoretical question NUs raise is whether they are parts of larger sentential structures or rather nonsentential structures whose semantic and morphosyntactic features are licensed by the surrounding context. To adjudicate between these views researchers have looked for evidence that NUs in fact behave as if they were fragments of sentences. As we will see in section 3, there is evidence to support both of these views. However, HPSG doesn't assume that NUs are underlyingly sentential structures.

\subsection{Predicate ellipsis and argument ellipsis}
The section looks at three constructions whose syntax includes null, hence noncanonical, elements. They are Verb Phrase Ellipsis (VPE), Null Complement Anaphora (NCA), and argument drop (or pro drop). VPE features stranded auxiliary verbs (\ref{5}) and NCA is characterized by omission of complements to some lexical verbs (\ref{6}). Argument drop refers to omission of a pronominal subject or an object argument, as illustrated in (\ref{7}) for Polish.

\ea A: I didn't ask George to invite you.\\B: Then who did?\label{5}\z

\ea Some mornings you can't get hot water in the shower, but nobody complains. \label{6} \z

\ea
\gll Pia p\'{o}\'{z}no wr\'{o}ci\l a do domu. {Od razu} posz\l a spa\'{c}.\\
\textsc{Pia} \textsc{late} \textsc{got} \textsc{to} \textsc{home} \textsc{right away} \textsc{went} \textsc{sleep}\\
\glt `Anna got home late. She went straight to bed.'
\label{7}
\z
Here the question is whether these null elements should be assumed to be underlyingly present in the syntax of theses constructions, and the answer is no.

\subsection{Nonconstituent coordination}
We focus on two instances of nonconstituent coordination, Right Node Raising (RNR) and gapping (Ross 1967), illustrated in (\ref{8}) and (\ref{9}) respectively. In RNR, a single constituent located in the right-peripheral position is associated with both conjuncts. In gapping, a finite verb is associated with both (or more) conjuncts but only present in the leftmost one. This results in what appears to be coordination of standard constituents and elements not normally defined as constituents (a stranded transitive verb in (\ref{8}) and a cluster of subject NP and object NP in (\ref{9})).
 \ea Ethan sold and Rasmus gave away all his CDs. \label{8}\z

 \ea Ethan gave away his CDs and Rasmus his old guitar. \label{9}\z
 To handle such constructions the grammar must either be permitted to coordinate noncanonical constituents, to generate coordinated constituents parts of which can undergo deletion, or to coordinate nonsentential utterances. As we will see, HPSG makes use of the latter two options.

\section{Evidence for and against invisible material at the ellipsis site}
This section is concerned with NUs and VPE since this is where the contentious issues arise of where ellipsis is licensed (sections 3.3. and 3.4) and whether there is invisible syntactic material in an ellipsis site (sections 3.1 and 3.2). Below we consider evidence for and against invisible structure found in the ellipsis literature; the evidence is based not only on intuitive judgments, but also experimental and corpus data.

\subsection{Connectivity effects}
Connectivity effects refer to parallels between NUs and their counterparts in sentential structures, thus speaking in favor of the existence of silent sentential structure. We focus on two kinds here: case-matching effects and preposition-stranding effects (for other examples of connectivity effects, see Ginzburg \& Miller 2018). It's been known since Ross (1969) that NUs exhibit case-matching effects, that is, they are typically marked for the same case that is marked on their counterparts in sentential structures. (\ref{10}) illustrates this for German, where case matching is seen between a wh-phrase functioning as an NU and its counterpart in the antecedent.

\ea
\gll Er will jemandem schmeicheln, aber sie wissen nicht wem/*wen.\\
\textsc{he} \textsc{will} \textsc{someone.DAT} \textsc{flatter}, \textsc{but} \textsc{they} \textsc{know} \textsc{not} \textsc{who.DAT/*who.ACC}\\
\glt `He wants to flatter someone, but they don't know whom.'\label{10}\z

Case-matching effects are crosslinguistically robust in that they are found in the great majority of languages with overt case marking systems, and therefore, they have been taken as strong evidence for the reality of silent structure. The argument is that the pattern of case matching follows straightforwardly from the silent structure that embeds an NU and matches the structure of the antecedent clause. However, a language like Hungarian poses a problem for this reasoning (see Jacobson 2016). Hungarian has verbs that assign one of two cases to their object NPs with no meaning difference, but case matching is still required between an NU and its counterpart, whichever case is marked on the counterpart. To see this, consider (\ref{11}) from Jacobson (2016: 356). The verb {\it hasonlit} assigns either sublative (SUBL) or allative (ALL) case to its object, but if SUBL is selected for an NU's counterpart, the NU must match this case.

\ea
A: \gll Ki-re hasonlit P\'{e}ter?\\
 \textsc{who.SUBL} \textsc{resembles} \textsc{Peter}\\
 \glt A: `Who does Peter resemble?'\\

B: \gll J\'{a}nos-ra/*J\'{a}nos-hoz.\\
\textsc{J\'{a}nos.SUBL/*J\'{a}nos.ALL}\\
\glt B: `J\'{a}nos.'\label{11}\z
Jacobson (2016) notes that there is some speaker variation regarding the (un)ac\-cepta\-bi\-li\-ty of case mismatch here at the same time that all speakers agree that either case is fine in a corresponding nonelliptical response to (\ref{11}A). This last point is important, because it shows that the requirement of---or at least a preference for---matching case features applies to NUs to a greater extent than it does to their nonelliptical equivalents, challenging connectivity effects.

Similarly problematic for case-based parallels between NUs and their sentential counterparts are some Korean data. Korean NUs can drop case markers more freely than their counterparts in nonelliptical clauses can, a point made in Morgan (1989). Observe the following exchanges:

  \ea
A: \gll Nwukwu-ka ku  chaek-ul  sa-ass-ni?\\
\textsc{who.NOM} \textsc{the} \textsc{book.ACC} \textsc{buy.PST.QUE}\\
\glt A: `Who bought the book?'\\

B: \gll Yongsu-ka/Yongsu/*Yongsu-lul.\\
\textsc{Yongsu.NOM/Yongsu/*Yongsu.ACC}\\
\glt B: `Yongsu.'

B$'$: \gll Yongsu-ka/*Yongsu  ku  chaek-ul  sa-ass-e\\
\textsc{Yongsu.NOM/*Yongsu} \textsc{the} \textsc{book.ACC} \textsc{buy.PST.DECL}\\
\glt B$'$: `Yongsu bought the book'\\
\label{12}\z
%
When an NU corresponds to a nominative subject in the antecedent (as in \ref{12}B), it can be either marked for nominative or caseless.
However, replacing the same NU with a full sentential answer, as in (\ref{12}B$'$), rules out case drop from the subject. This strongly suggests that the case-marked and caseless NUs couldn't have identical source sentences if they were to derive via PF-deletion.\footnote{Nominative differs in this respect from three other structural cases, dative, accusative and genitive, in that the latter may also be dropped from nonelliptical clauses (see Morgan 1989, Lee 2016, Kim 2016).}  Data like these led Morgan (1989) to propose that not all NUs have a sentential derivation, an idea later picked up in Barton (1998).

The same pattern is associated with semantic case. That is, in (\ref{13}), an NU can optionally be marked for comitative like its counterpart in the A-sentence or be caseless. But being caseless is not an option for the NU's counterpart.

\ea
A: \gll Nwukwu-wa/*nwukwu  hapsek-ul ha-yess-e?\\
\textsc{who.COM/*who}  \textsc{sitting.together.ACC} \textsc{do.PST.QUE}\\
\glt A: `With whom did you sit together?'\\

B: \gll Mimi-wa/*Mimi.\\
\textsc{Mimi.SRC/Mimi}\\
\glt B: `With Mimi/*Mimi.' \label{13}\z
The generalization for Korean is then that NUs may be optionally realized as caseless but may never be marked for a different case than is marked on their counterparts.

Overall, case-marking facts show that there is some morphosyntactic identity between NUs and their antecedents, though not to the extent that NUs have exactly the features that they would have if they were constituents embedded in sentential structures. The Hungarian facts also suggest that aspects of the argument structure of the antecedent relating to case licensing are relevant for an analysis of NUs.\footnote{Hungarian and Korean are in fact not the only problematic languages; for a list, see Vicente (2015).}

The second kind of connectivity effects goes back to Merchant (2001, 2004) and highlights apparent links between wh- and focus movement and the features of NUs. The idea is that prepositions behave the same under wh- and focus movement as they do under clausal ellipsis, that is, they pied-pipe or strand in the same environments. If a language (e.g., English) permits preposition stranding under wh- and focus movement ({\it With what did Harvey paint the wall?} vs {\it What did Harvey paint the wall with?}), then NUs may surface with or without prepositions, as illustrated in (\ref{14}) for sluicing and BAE.

\ea A: I know what Harvey painted the wall with.\\B: (With) what?/(With) primer.\label{14}\z
If there indeed was a link between between preposition stranding and NUs, then we would not expect prepositionless NUs in languages without preposition stranding. This expectation is disconfirmed by an ever-growing list of nonpreposition-stranding languages that do feature prepositionless NUs: Brazilian Portuguese (Almeida \& Yoshida 2007), Spanish and French (Rodrigues et al. 2009), Greek (Molimpakis 2018), Bahasa Indonesia (Fortin 2007), Emirati Arabic (Leung 2014), Russian (Philippova 2014), Polish (Szczegielniak 2008, Nykiel 2013, Sag \& Nykiel 2011), Czech (Caha 2011), Bulgarian (Abels 2017), and Serbo-Croatian (Stjepanovi\'{c} 2008, 2012). A few of these studies have presented experimental evidence that prepositionless NUs are acceptable, though --- for reasons still poorly understood --- they typically don't reach the same level of acceptability as their variants with prepositions do (see Nykiel 2013 for Polish and Molimpakis 2018 for Greek).

 It is evident from this research that there is no grammatical constraint on NUs that keeps track of what preposition-stranding possibilities exist in any given language. On the other hand, it doesn't seem sufficient to assume that NUs can freely drop prepositions, given examples of sprouting like (\ref{15}), in which prepositions are not omissible (see Chung 2006, 2013 on the inomissibility of prepositions under sprouting). The difference between (\ref{14}) and (\ref{15}) is that there is an explicit phrase the NU corresponds to (in the HPSG literature this phrase is termed a Salient Utterance (Ginzburg \& Sag 2000) or a Focus-Establishing Constituent (Ginzburg 2012)) in the former but not in the latter.

 \ea A: I know Harvey painted the wall.\\B: *(With) what?/Yeah, *(with) primer.\label{15}\z
This issue has not received much attention in the HPSG literature, though see Kim (2015).


\subsection{Island effects}
One of the predictions of the view that NUs are underlyingly sentential is that they should respect island constraints on long-distance movement. But as illustrated below, NUs (both sluicing and BAE) exhibit island-violating behavior.\footnote{Merchant (2004) argued that BAE, unlike sluicing, does respect island constraints, an argument that was later challenged (see e.g, Culicover \& Jackendoff 2005, Griffiths \& Lipt\'{a}k 2014). However, Merchant (2004) focused specifically on pairs of wh-questions and answers like (\ref{2}) and ran into the difficulty of testing for island-violating behavior, since a well-formed antecedent couldn't be constructed.}

\ea A: Harriet drinks scotch that comes from a very special part of Scotland.\\B: Where? (Culicover \& Jackendoff 2005: 245) \label{16}\z

\ea A: The administration has issued a statement that it is willing to meet with one of the student groups.\\B: Yeah, right---the Gay Rifle Club. (Culicover \& Jackendoff 2005: 245) \label{17}\z

Among Culicover \& Jackendoff's (2005: 245) examples of well-formed island-violating NUs are also sprouted NUs (those that correspond to implicit Salient Utterances) like (\ref{18})-(\ref{19}).

\ea A: John met a woman who speaks French.\\B: With an English accent?\label{18}\z

\ea A: For John to flirt with at the party would be scandalous. \\B: Even with his wife?\label{19}\z
Other research assumes that sprouted NUs are one of the two kinds of NUs that respect island constraints, the other kind being contrastive NUs, illustrated in (\ref{20}) (Chung,
Ladusaw \& McCloskey 1995, Merchant 2001, Griffiths \& Lipt\'{a}k 2014).

\ea A: Does Abby speak the same Balkan language that Ben speaks?\\B: *No, Charlie. (Merchant 2001)  \label{20}\z
 Schmeh et al. (2015) explore the acceptability of NUs like (\ref{20}) compared to NUs introduced by the particle {\it Yes} depicted in (\ref{21}). (\ref{20}) differs from (\ref{21}) in terms of discourse function in that it corrects rather than supplement the antecedent, which is signaled by a different response particle.

\ea A: John met a guy who speaks a very unusual language. \\B: Yes, Albanian. (Culicover \& Jackendoff 2005: 245) \label{21}\z
Schmeh et al. (2015) find that corrections lower acceptability ratings compared to supplementations and propose that this follows from the fact that corrections induce greater processing difficulty than supplementations do, and hence the acceptability difference between (\ref{20}) and (\ref{21}). This finding makes it plausible that the perceived degradation of island-violating NUs could ultimately be attributed to nonsyntactic factors, e.g., the difficulty of successfully computing a meaning for them.

In contrast to NUs, many instances of VPE appear to obey island constraints, as would be expected if there was unpronounced structure from which material was extracted. An example is depicted in (\ref{22}) (note that the corresponding sluicing NU is fine).

\ea *They want to hire someone who speaks a Balkan language, but I don't remember which they do [\sout{want to hire someone who speaks t}]. (Merchant
2001: 6)\label{22}\z
(\ref{22}) contrasts with well-formed examples like (\ref{23}) and (\ref{24}), from Ginzburg \& Miller (2018).

\ea He managed to find someone who speaks a Romance language, but a Germanic language, he didn't [\sout{manage to find someone who speaks t}].\label{23}\z
\ea He was able to find a bakery where they make good baguette, but croissants, he couldn't [\sout{find a bakery where they make good t}].\label{24}\z
As Ginzburg \& Miller (2018) rightly point out, we don't yet have a complete understanding of when or why island effects show up in VPE. Its behavior is at best inconsistent, failing to provide convincing evidence for silent structure.


\subsection{Structural mismatches}
 Because structural mismatches are not permitted by NUs (see Merchant 2005, 2013),\footnote{Ginzburg \& Miller (2018) cite examples---originally from Beecher (2008)---of sprouting NUs with nominal, hence mismatched, antecedents, e.g., (i).
 \ea We're on to the semi-finals, though I don't know who against.\z
 Somewhat similar examples, where NUs appear to take APs as antecedents, appear in COCA:
 \ea  A: Well, it's a defense mechanism. B: Defense against what?\z
 \ea Our Book of Mormon talks about the day of the Lamanite, when the church would make a special effort to build and reclaim a fallen people. And some people will say, Well, fallen from what? \z
The NUs in (ii)--(iii) repeat the lexical heads whose complements are being sprouted ({\it defense} and {\it fallen}), that is, they contain more material than is usual for NUs (cf. (i)). It seems that without this additional material it would be difficult to integrate the NUs into the propositions provided by the antecedents and hence to arrive at the intended interpretations.
} this section focuses on VPE and developments surrounding the question of which contexts license it. In a seminal study of anaphora, Hankamer and Sag (1976) classified VPE as a surface anaphor with syntactic features closely matching those of an antecedent present in the linguistic context. They argued in particular that VPE is not licensed if it mismatches its antecedent in voice. Compare (\ref{25}) and (\ref{26}) from Sag \& Hankamer (1984: 327).
\ea[*]{
The children asked to be squirted with the hose, so we did.  \label{25} 
}
\z
\ea[]{
The children asked to be squirted with the hose, so they were. \label{26}
}
\z
This proposal places tighter structural constraints on VPE than on other verbal anaphors (e.g., \emph{do it/that}) in terms of identity between an ellipsis site and its antecedent and has prompted extensive evaluation in a number of corpus and experimental studies in the decades following Hankamer \& Sag (1976). Below are examples of acceptable structural mismatches reported in the literature, ranging from voice mismatch (\ref{27}) to nominal antecedents (\ref{28}) to split antecedents (\ref{29}).

\ea This information could have been released by Gorbachev, but he chose not to [release it]. (Hardt 1993: 37) \label{27}\z

\ea Mubarak's survival is impossible to predict and, even if he does [survive], his plan to make his son his heir apparent is now in serious jeopardy. (Miller \& Hemforth 2014: )\label{28}\z

\ea Wendy is eager to sail around the world and Bruce is eager to climb Mt. Kilimanjaro, but neither of them can [do the things they want], because money is too tight. (Webber 1978: )\label{29}\z

There are two opposing views that have emerged from the empirical work. The first view takes mismatches to be grammatical and connects degradation in acceptability to violation of certain independent discourse (Kehler 2002, Miller 2011, Kertz 2013, Miller \& Hemforth 2014, Miller \& Pullum 2014) or processing constraints (Kim et al. 2011). Two types of VPE have been identified on this view through corpus work---auxiliary choice VPE and subject choice VPE---each with different discourse requirements with respect to the antecedent (Miller 2011, Miller \& Hemforth 2014, Miller \& Pullum 2014). The second view assumes that there is a grammatical ban on structural mismatch but violations thereof may be repaired under certain conditions; repairs are associated with differential processing costs compared to matching ellipses and antecedents (Arregui et al. 2006, Grant et al. 2012). If we follow the first view, it is perhaps unexpected that voice mismatch should consistently incur a greater penalty under VPE than when no ellipsis is involved, as recently reported in Kim \& Runner (2017). Kim \& Runner (2017) stop short of drawing firm conclusions regarding the grammaticality of structural mismatches, but one possibility is that the observed mismatch effects reflect a construction-specific constraint on VPE. HPSG analyses take structurally mismatched instances of VPE to be unproblematic and fully grammatical, while also recognizing construction-specific constraints: discourse or processing constraints formulated for VPE may or may not extend to other elliptical constructions, such as NUs (see Ginzburg \& Miller 2018 for this point).


\subsection{Nonlinguistic antecedents}
Like structural mismatches, the availability of nonlinguistic antecedents for an ellipsis points to the fact that it needn't be interpreted by reference to and licensed by a structurally identical antecedent. Although this option is somewhat limited, VPE does tolerate nonlinguistic antecedents, as shown in (\ref{30})--(\ref{31}) (see also Hankamer \& Sag 1976, Schachter 1977).
\ea Mabel shoved a plate into Tate's hands before heading for the sisters' favorite table in the shop. ``You shouldn't have.'' She meant it. The sisters had to pool their limited resources
just to get by. (Pullum \& Miller 2014, ex. 23)\label{30}\z

\ea Once in my room, I took the pills out. ``Should I?'' I asked myself. (Pullum \& Miller 2014, ex. 22a)\label{31}\z
Pullum \& Miller (2014) provide an extensive critique of the earlier work on the ability of VPE to take nonlinguistic antecedents, arguing for a streamlined discourse"=based explanation that neatly captures the attested examples as well as examples of structural mismatch like those discussed in section 3.3. The important point here is again that VPE is subject to construction-specific constraints which limit its use with nonlinguistic antecedents.

NUs appear in various nonlinguistic contexts as well. Ginzburg \& Miller (2018) distinguish three classes of such NUs: sluices (\ref{32}), exclamative sluices (\ref{33}), and declarative fragments (\ref{34}).

\ea (In an elevator) What floor? (Ginzburg \& Sag 2000: 298)\label{32}\z

\ea It makes people ``easy to control and easy to handle,'' he said, ``but, God forbid, at what a cost!'' (Ginzburg \& Miller To appear, ex. 34a)\label{33}\z

\ea BOBADILLA turns, gestures to one of the other men, who comes forward and gives him a roll of parchment, bearing the royal seal. ``My letters of appointment.'' (COCA)\label{34}\z
In addition to being problematic from the licensing point of view, NUs like these have been put forward as evidence against the idea that they are underlyingly sentential, because it is unclear what the structure that underlies them would be (see Ginzburg \& Sag 2000, Culicover \& Jackendoff 2005, Stainton 2006).\footnote{This is not to say that a sentential analysis of fragments without linguistic antecedents hasn't been attempted. For details of a proposal involving a `limited ellipsis' strategy, see Merchant (2004, 2010).}


\section{Analyses of NUs}
It is worth noting at the outset that the analyses of NUs within the framework of HPSG are based on an elaborate theory of dialog (Ginzburg 1994, Ginzburg \& Cooper 2004, 2014, Larsson 2002, Purver 2006, Fern\'{a}ndez 2006, Fern\'{a}ndez \& Ginzburg 2002, Fern\'{a}ndez et al. 2007, Ginzburg \& Fern\'{a}ndez 2010, Ginzburg et al. 2014, Ginzburg 2012, 2013) and on a wider range of data than is common practice in the ellipsis literature. Existing analyses of NUs go back to Ginzburg \& Sag (2000), who recognize declarative fragments (\ref{34a}) and two kinds of sluicing NUs, direct sluices (\ref{35}) and reprise sluices (\ref{36}) (the relevant fragments are bolded). The difference between direct and reprise sluices lies in the fact that the latter are requests for clarification of any part of the antecedent. For instance, in (\ref{36}) the referent of {\it that} is unclear to the interlocutor.

\ea ``I was wrong.'' Her brown eyes twinkled. ``Wrong about what?'' ``\textbf{That night}.'' (COCA) \label{34a}\z

\ea ``You're waiting,'' she said softly. ``\textbf{For what?}'' (COCA) \label{35} \z

\ea ``Can we please not say a lot about that?'' ``\textbf{About what?}'' (COCA) \label{36} \z

Ginzburg and Sag (2000: 304) make use of the constraint shown in (\ref{hf-cx}) (we have added information about the MAX-QUD) to generate NUs.




                                 \ea{\label{hf-cx}
\item[] Head-Fragment Construction
\item[] \small
\begin{avm}
\[CAT &S\\ %\[HEAD v\]\\
  %CONT &\[NUCL \@1\]\\
  CTXT & \[MAX-QUD $\lambda$\{$\pi$$^{i}$\}\\
  SAL-UTT \{  \[CAT \@2\\
                         CONT\ \[IND & {\it i}\\
                                  \]\]\}\]\]

                    \ \ $\Rightarrow$\ \
\[  CAT &  \@2\\
   CONT  &\[IND & {\it i}\]
\]
\end{avm}}\z
Let us see how this constructional constraint allows us to 
license NUs and capture their properties, including the connectivity effects we discussed in section 3.1. Note first that any phrasal category
can function as an NU, that is, can be mapped onto a sentential utterance as long as it corresponds to a Salient Utterance (SAL-UTT). This means that 
the head daughter's syntactic category must match that of a SAL-UTT, which is an attribute supplied by the surrounding context as a (sub)utterance of another contextual attribute---the Maximal Question under Discussion (MAX-QUD). The two contextual attributes SAL-UTT and MAX-QUD are introduced specifically for the purpose of analyzing NUs. The context gets updated with every new question-under-discussion, and MAX-QUD represents the most recent question-under-discussion, while SAL-UTT is the (sub)utterance with the widest scope within MAX-QUD. To put it informally, SAL-UTT represents a (sub)utterance of a MAX-QUD that has not been resolved yet. Its feature CAT supplies information relevant for establishing morphosyntactic identity with an NU, that is, syntactic category and case information, and (\ref{hf-cx}) requires that an NU match this information. Meanwhile, MAX-QUD provides the propositional semantics for an NU and is, typically, a unary question. The content of MAX-QUD can be supplied by linguistic or nonlinguistic context. In the prototypical case, MAX-QUD arises from the most recent wh-question uttered in a given context (\ref{37}), but can also arise (via accommodation) from other forms found in the context, such as constituents bearing focal accent (\ref{38}) and constituents in need of clarification (\ref{39}), or from a nonlinguistic context (\ref{40}).\footnote{Ginzburg (2012) uses the notion of the Dialog Game Board (DGB) to keep track of all information relating to the common ground between interlocutors. The DGB is also the locus of contextual updates arising from each new question-under-discussion that is introduced.}
  \ea A: What did Barry break? \\B: The mike.\label{37}\z

  \ea A: Barry broke the MIKE. \\B: Yes, the only one we had.\label{38}\z

  \ea A: Barry broke the mike. \\B: Who?\label{39}\z

  \ea (Cab driver to passenger on the way to airport) A: Which airline?\label{40}\z

The existing analyses of NUs (Ginzburg 2012, Sag \& Nykiel 2011, Kim 2015, Abeill\'{e} et al. 2014, Abeill\'{e} \& Hassamal 2017) are based on Ginzburg and Sag's (2000) constraint. Below we illustrate how it is applied to the declarative fragment in (\ref{37}).

\ea\label{41}
\begin{forest}
[\begin{avm}\avml \hfil S\\
\[CAT &\[HEAD v\]\\
  %CONT & \[NUCL \@1\]\\
 CTXT & \[MAX-QUD $\lambda$\{$\pi$$^{i}$\}\[{\it break}({\it b,i})\]\\
  SAL-UTT \{\[CAT\ \@2\\
                          CONT & \[IND & {\it i}\\
                                 \] \]\}\]\]  \avmr \end{avm}
[\begin{avm}\avml \hfil NP\\
             \[CAT  \@2\\
  CONT & \[IND & {\it i}%\\
   %\PARAMS & \{$\pi$$^{i}$\}
   \]\]\avmr\end{avm}
 [The mike]]]
\end{forest}
\z  

This construction-based analysis, in which dialogue updating plays
a key role in the licensing of NUs, can also offer a simple account of
sprouting examples like (\ref{35}).\footnote{Ginzburg and Sag (2000: 330) suggest a way of analyzing sprouted NUs, such as (i). The implied direct object of {\it eat} functioning as SAL-UTT here would appear as a noncanonical synsem on the verb's ARG-ST list, but not on the COMPS list, and thereby be able to provide appropriate morphosyntactic identity information.

\ea  A: And what did you do then? \\B: I ate. \\A: What?\z


} As discussed in Kim (2015), we could take an unrealized oblique argument of
the verb {\it wait} as an instance of indefinite null instantiation (INI) (see Ruppenhofer and Michaelis 2014):

\ea Lexical entry for {\it wait}:\\
 \begin{avm}
 \[FORM \q<{\it wait}\q>\\
   ARG-ST \<NP\jbsub{{\it i}}, PP\jbsub{{\it x}}\>\\
   SYN\[SUBJ \<NP[{\it overt}]\>\\
        COMPS \<PP[{\it ini}]\>\]\\
   SEM {\it wait}({\it i, x})\]
   \end{avm}\label{wait}\z
%
The lexical information specifies that the second argument of {\it wait} can be an unrealized indefinite NP while the first argument needs to be an overt one. Now consider the dialogue in (\ref{35}). Uttering
the sentence {\it You're waiting?} would then update the DGB with a SAL-UTT represented by the unrealized NP:
%
\ea
\begin{avm}
\[%{\it dgb}\\
 DGB \[SAT-UTT \[SYN  PP\[{\it ini}\\
                          PFORM {\it for}\\
                        \IND\ {\it  x}\]\\
                 SEM {\it wait.for}({\it i,x})\]\]\] \end{avm}\z
%
The fragment {\it for what?}, matching this SAL-UTT,
projects a well-formed NU in accordance with the Head-Fragment Construction.\footnote{See the detailed analysis of such sprouting examples in Kim (2015).}

The advantages of the nonsentential analyses sketched here follow from their ability to capture limited morphosyntactic parallelism between NUs and SAL-UTT without having to account for why NUs don't behave like constituents of sentential structures. The island-violating behavior of NUs is unsurprising on this analysis, as are attested cases of structural mismatch and situationally controlled NUs.\footnote{The rarity of NUs with nonlinguistic antecedents can be understood as a function of how easily a situational context can give rise to a MAX-QUD and thus license ellipsis (see Miller \& Pullum 2014 for this point with regard to VPE).} However, some loose ends still remain. (\ref{hf-cx}) currently has no means of capturing certain connectivity effects: it can't rule preposition drop out under sprouting and it incorrectly rules out case mismatch in languages like Hungarian for speakers that do accept it (see discussion around example (\ref{11}).\footnote{See, however, Kim (2015) for proposals envoking a case hierarchy specific to Korean to explain case mismatch and introducing an additional constraint to block preposition drop under sprouting.}


\section{Analyses of predicate/argument ellipsis}
The first issue in the analysis of VPE is the status of an elided VP. It is assumed to be a {\it pro} element due to its pronominal properties (see Lobeck 1995, Lopez 2000, Kim 2006, Aelbrecht and Harwood 2015, Ginzburg and Miller 2018). For instance, VPE applies only to phrasal categories (\ref{42}--\ref{43}),
can cross utterance boundaries (\ref{44}), can override island constraints (\ref{45}--\ref{46}), is subject to the Backwards Anaphora Constraint (\ref{47}--\ref{48}).

\ea *Mary will meet Bill at Stanford because she didn't  \jbtr John.\label{42}\z
\ea Mary will meet Bill at Stanford because she didn't \jbtr at Harvard.\label{43}\z
\ea A: Tom won't leave Seoul soon.\\
B: I don't think Mary will \jbtr either.\label{44}\z
\ea John didn't hit a home run, but I know a woman who did. (CNPC)\label{45}\z
\ea That Betsy won the batting crown is not surprising, but that
  Peter didn't know she did \jbtr is indeed surprising. (SSC)\label{46}\z
\ea *Sue didn't [e] but John ate meat.\label{47}\z
\ea Because Sue didn't [e], John ate meat.\label{48}\z

Argument ellipsis we find in languages like Polish and Korean can also be taken to be ellipsis of a pronominal expression, as in (\ref{49}).

\ea
\gll Mimi-ka {\it pro} po-ass-ta.\\
Mimi.NOM  pro see.PST.DECL\\
\glt `Mimi saw (him)'\label{49}\z

 In accounting for {\it pro}-drop phenomena, we do not need to posit a phonologically empty pronoun if the level of argument structure is
available (see Bresnan 1982). We simply encode the required pronominal properties in the argument structure. In the framework of HPSG, we represent this as the following Argument Realization Constraint allowing mismatch between argument-structure and syntactic-valence features:\footnote{Expressions have two subtypes: overt and covert
ones, the latter of which has two subtypes, {\it pro} and {\it gap}. See Sag (2012) for details.}

\ea
Argument Realization Constraint (ARC):\\
\begin{avm}
{\it v-word} \;   $\Rightarrow$ \;
\[SYN\|VAL \[SUBJ & \@A\\
                 COMPS & \@B $\ominus$ list({\it pro})\]\\
  ARG-ST \@A $\oplus$ \@B\]
  \end{avm}\label{50}
\z
The argument realization here tells us that a {\it pro} element
in the argument structure need not be realized in the syntax. For
example, as represented in (\ref{51}), the transitive verb {\it
po-ass-ta} `see.PST.DECL' takes a {\it pro} object NP as its argument and
thus the {\it pro} NP is not instantiated as the syntactic
complement of the verb. \index{transitive}

\ea
Lexical entry for {\it po-ass-ta}\\
\begin{avm}
\[{\it v-word}\\
 FORM \q<po-ass-ta\q>\\
 SYN\[HEAD \|VFORM \ \ {\it fin}\\
      VAL \[SUBJ & \q<\@1\q>\\
           COMPS & \q<  \; \q>\]\]\\
ARG-ST  \q<\@1NP, NP[{\it pro}]\q>\]
 \end{avm} \label{51}\z
  
  
Similarly, English VPE is analyzed as a language-particular VP {\it pro} drop phenomenon normally licensed by auxiliary verbs. This idea can be formalized as in (\ref{52}):

\ea\label{52}
Aux-Ellipsis Construction:\\
\begin{forest}
[
{\begin{avm}
 \[{\it aux-ellipsis-wd}\\
   ARG-ST \q<\@1XP, \@2YP[{\it pro}]\q>\]
 \end{avm}}
 [{\begin{avm}
 \[{\it aux-v-lxm}\\
   %\SYN\|\HEAD\|\AUX\ $+$\\
   ARG-ST \q<\@1XP, \@2YP\q>\]\end{avm}}]]
\end{forest}
\z
What this tells us is that an auxiliary verb selecting two arguments
can be projected into an elided auxiliary verb whose second argument
is realized as a small {\it pro}. Note that this argument is not mapped
onto the syntactic grammatical function COMPS. (\ref{52}) will also project a structure like the one
in Figure~\ref{fig-53}.

\begin{figure}
\begin{forest}
[S
  [\begin{avm}\@1NP\end{avm}
      [Sandy]]
  [\begin{avm} \avml \hfil VP\\
      \[%{\it head-only-cxt} \& {\it ellip-cxt}\\
      HEAD \@2\\
      SUBJ \q< \@1\q>\]\avmr \end{avm}
    [{\begin{avm} \avml \hfil  V\\
        \[HEAD \@2\[\AUX\ +\]\\
        SUBJ \q<\@1\q>\\
        COMPS \q<\; \;\q>\\
        ARG-ST \q<\@1NP, VP[{\it pro}]\q>\] \avmr \end{avm}}
      [can]]]]
\end{forest}
\caption{add caption}\label{fig-53} 
\end{figure}
The head daughter's COMPS list (VP[bse]) is empty because the second element in the ARG-ST\ is
a {\it pro}. 





\section{Analyses of nonconstituent coordination and gapping}
In this section, we focus on RNR and gapping, whose analyses we address in separate subsections below.

\subsection{Right Node Raising}
A characteristic property of RNR is that it's the only phenomenon where seemingly incomplete structure has consistently attracted HPSG analyses involving deletion of silent material. All existing analyses of RNR (Abeill\'{e} et al. 2016, Beavers \& Sag 2004, Chaves 2008, 2014, Crysmann 2003, Yatabe 2001, 2012) agree on this point, although some of this work proposes more than one mechanism for accounting for coordination of nonconstituents (Chaves 2014, Yatabe 2001, 2012, Yatabe \& Tam 2018).

The RNR literature engages with the question of what kind of deletion it is that targets shared material, based on the kind of material that may be RNRaised and the range of mismatches permitted between the left and right conjuncts.\footnote{Although we refer to the material on the left and right as conjuncts, it is been known since Hudson (1976, 1984) that RNR extends to other syntactic environments than coordination (see Chaves (2014) for stressing this point).} For instance, Chaves (2014: 839--840) concludes that RNR can't be syntactic deletion because it exhibits various argument-structure mismatches (\ref{54}--\ref{55}) and can target material below the word level (\ref{56}--\ref{57}).


\ea Sue gave me---but I don't think I will ever read---[a book about relativity]. \label{54}\z

\ea Never let me---or insist that I---[pick the seats].\label{55}\z

\ea We ordered the hard- but they got us the soft-[cover edition].\label{56}\z

\ea Your theory under- and my theory over[generates].\label{57}\z
Furthermore, RNR can target strings that are not subject to any known syntactic operations, such as rightward movement (Chaves 2014: 865).

\ea I thought it was going to be a good but it ended up being a very bad [reception].\label{58}\z

\ea Tonight a group of men, tomorrow night he himself, [would go out there somewhere and wait].\label{59}\z

\ea They were also as liberal or more liberal [than any other age group in the 1986 through 1989 surveys].\label{60}\z
RNRaised material can also be discontinuous, as in (\ref{61}--\ref{62}) (Chaves 2014: 868; citing Whitman 2009: 238--240).

\ea Please move from the exit rows if you are unwilling or unable [to perform the necessary actions] without injury.\label{61}\z

\ea The blast upended and nearly sliced [an armored Chevrolet Suburban] in half.\label{62}\z
This evidence leads Chaves (2014) to propose that: (1) only 'true' RNR should be accounted for via the mechanism of surface-based deletion, (2) this deletion is sensitive to morph form identity, and (3) the targets of RNR are any linearized strings, whether constituents or otherwise. Chaves' (2014: 874) constraint licensing RNR is given in \ref{50}. It permits the M(orpho)P(honology) feature of the mother to contain only one instance (represented as $L_{3}$ in (\ref{63})) of the two morphophonologically identical sequences [FORM $F_{1}$],...,[FORM $F_{n}$] present in the daughters; the leftmost of these sequences undergoes deletion. The final list in the mother, $L_{4}$, may be empty or nonempty, depending on whether RNR is discontinuous.


\ea{\label{63}
Backward periphery deletion construction\\
\begin{avm}
\[phrase\\
  MP $L_{1}$:\type{ne-list} $\circ L_{2}$:\type{ne-list} $\circ L_{3} \circ L_{4}$ \] $\Rightarrow$
  \end{avm}


   \begin{avm}
\[ phrase\\
   MP $L_{1} \circ \langle$ \[ FORM $F_{1}$\],\ldots,\[ FORM $F_{n}$\]$\rangle \circ L_{2} \circ L_{3}:\langle$\[FORM $F_{1}$\],\ldots,\[FORM $F_{n}$\]$\rangle \circ L_{4}$
\]
\end{avm}}\z



Chaves' (2014) proposal reflects the idea that nonconstituent coordination is a multi-faceted phenomenon, requiring more than one kind of analysis. Indeed, the HPSG literature includes analyses based on NP-ellipsis or A(cross) T(he) B(oard) extraposition (Chaves 2014) and on phonological vs syntactic deletion (Yatabe 2001, 2012, Yatabe \& Tam 2018). Abeill\'{e} et al. (2016) argue for a finer-grained analysis of French RNR based on phonological deletion. Their empirical evidence reveals a split between functional and lexical categories in French such that the former permit mismatch between the two conjuncts (where determiners or prepositions differ) under RNR, while the latter don't.



\subsection{Gapping}
HPSG analyses of gapping fall into two kinds: one kind draws on Beavers and Sag's (2004) deletion-based analysis of nonconstituent coordination (Chaves 2009) and the other on Ginzburg and Sag's (2000) analysis of NUs (Abeill\'{e} et al. 2014, Park et al. 2018). The latter analyses align gapping with analyses of NUs, as discussed in section 4, more than with analyses of nonconstituent coordination, and therefore gapping could be classified together with other NUs. We use the analysis due to Abeill\'{e} et al. (2014) for illustration below.

Abeill\'{e} et al. (2014) assume an identity condition on gapping requiring that gapping remnants match major constituents in the antecedent clause, which they term source clause. In other words, gapping remnants are constrained by elements of the argument structure of the verbal head present in the antecedent and absent from the rightmost conjunct, as originally proposed by Hankamer (1971). This identity condition correctly predicts that gapping remnants needn't appear in the same order as their counterparts in the antecedent (\ref{64}) (see Sag et al. 1985: 156--158), nor are they required to be the same syntactic category as their counterparts (\ref{65}). However, note that despite the syntactic category mismatch, the NP {\it an incredible bore} belongs to the subcategorization frame of the predicate {\it become}.

\ea A policeman walked in at 11, and at 12, a fireman. \label{64}\z

 \ea Pat has become [crazy]$_{AP}$ and Chris [an incredible bore]$_{NP}$.  \label{65}\z
Abeill\'{e} et al. (2014) offer additional evidence from Romance (e.g., case mismatch between gapping remnants and their counterparts and even more possibilities of ordering remnants than is the case in English) to strengthen their point that syntactic identity is relaxed under gapping.


The key assumption in Abeill\'{e} et al.'s (2014) analysis is that (two or more) gapping remnants form a cluster whose mother has an underspecified syntactic category, that is, is a non-headed phrase. This phrase then serves as the head daughter of a head-fragment phrase, whose syntactic category is also underspecified. This means that there is no unpronounced verbal head in the phrase to which gapping remnants belong. Furthermore, the contextual attribute SAL-UTT introduced for NUs serves to ensure syntactic identity between gapping remnants and their counterparts such that the latter are SAL-UTTs bearing the specification [Major +] as part of their Head feature and being coindexed with the remnants.

With these ingredients of the analysis in place, we reproduce the gapping construction in (\ref{66}). The construction represents asymmetric coordination in the sense that the daughters include both nonelliptical verbal conjuncts and head-fragment phrases with an underspecified syntactic category. The mother only shares its syntactic category with the nonelliptical conjuncts so that its own category is specified to be verbal.


\ea 
Gapping construction\\\type{gapping-ph} $\rightarrow$ \type{coord-ph} \&\\
\begin{avm}
\[HEAD \fbox{H} \type{verbal}\\
CONTXT | BACKGROUND \{ ..., sym-discourse-rel(\fbox{M$_{1}$},..., \fbox{M$_{j}$}, \fbox{M$_{j+1}$},..., \fbox{M$_{n}$} ), ... \}\\

DTRS $\langle$ \[ HEAD \fbox{H} \[ verbal \\ CLUSTER elist \]\\CONTENT \fbox{M$_{1}$} \] ,..., \[
HEAD \fbox{H} \[ verbal \\ CLUSTER elist \]\\CONTENT \fbox{M$_{j}$} \]\]$\rangle \bigoplus$
\end{avm}
\begin{avm}
$\langle$ \[ HEAD \[CLUSTER $\langle \fbox{1},...,\fbox{n}\rangle$\] \\
             SOURCE \fbox{M$_{j}$}\\
             CONTENT \fbox{M$_{j+1}$} \],..., \[ HEAD \[CLUSTER $\langle$ \fbox{1$'$},...,\fbox{n$'$}$\rangle$\]\\
                                                SOURCE \fbox{M$_{j}$}\\
                                                CONTENT \fbox{M$_{n}$}\]$\rangle$ 
\end{avm}
\label{66}
\z



\section{Summary}
This chapter has reviewed three types of ellipsis, nonsentential utterances, predicate ellipsis, and nonconstituent coordination, corresponding to three kinds of analysis within HPSG. The pattern that emerges from this overview is that HPSG favors the `what you see is what get' approach to ellipsis and limits a deletion-based approach, common in the minimalist literature on ellipsis, to a subset of nonconstituent coordination phenomena.

%\citep{Chomsky1957}.
%\citep{Comrie1981}




%\begin{table}
%\caption{Frequencies of word classes}
%\label{tab:1:frequencies}
% \begin{tabular}{lllll}
%  \lsptoprule
            %& nouns & verbs & adjectives & adverbs\\
 % \midrule
  %absolute  &   12 &    34  &    23     & 13\\
%  relative  &   3.1 &   8.9 &    5.7    & 3.2\\
 % \lspbottomrule
 %\end{tabular}
%\end{table}




\section*{Abbreviations}
\begin{description}
\item NUs Nonsentential utterances
\item BAE Bare Argument Ellipsis
\item VPE Verb Phrase Ellipsis
\item NCA Null Complement Anaphora
\item SAL-UTT Salient Utterance
\item MAX-QUD Maximal-Question-under-Discussion
\item DGB Dialog Game Board
\end{description}
\section*{Acknowledgements}
We thank the editors and Yusuke Kubota for helpful comments.

%{\sloppy
%\printbibliography[heading=subbibliography,notkeyword=this]
%}

}% AVM options


\section{Handmade list of references: Fix!}
\begin{description}
\item Abeill\'{e}, Anne, Gabriela B\^{i}lb\^{i}ie, and Fran\c{c}ois Mouret. 2014. A Romance perspective on
gapping constructions. In Hans Boas and F. Gonz\'{a}lvez Garcia, (eds.), Romance Perspectives
on Construction Grammar, 227--267. Amsterdam: Benjamins.
\item Abeill\'{e}, Anne, Berthold Crysmann, and Aoi Shira\"{i}shi. 2016. Syntactic mismatch in French peripheral ellipsis. In Christopher Pi\~{n}n (ed.), Empirical Issues in Syntax and Semantics 11, 1--30. Paris: CSSP.
\item Abeill\'{e}, Anne and Shrita Hassamal. 2017. Sluicing in Mauritian: a fragment analysis. Ms.
\item Abels, Klaus. 2017. On the interaction of P-stranding and sluicing in Bulgarian. In Olav Mueller-Reichau and Marcel Guhl (eds.), Aspects of Slavic linguistics: Formal grammar, lexicon and communication, volume 16 of Language, Context and Cognition. Berlin: De Gruyter Mouton.
\item Aelbrecht, Lobke and William Harwood. 2015. To be or not to be elided: VP ellipsis revisited. Lingua 153:66--97.
\item Arregui, Ana, Charles Clifton, Jr., Lyn Frazier, and Keir Moulton. 2006. Processing elided
verb phrases with flawed antecedents: The recycling hypothesis. Journal of Memory and
Language 55:232--246.
\item Barton, Ellen. 1998. The grammar of telegraphic structures: Sentential and nonsentential derivation. Journal of English Linguistics 26:37--67.
\item Beavers, John and Ivan A. Sag. 2004. Coordinate ellipsis and apparent non-constituent
coordination. In Stefan M\"{u}ller, (ed.), Proceedings of the 11th International Conference on
Head-Driven Phrase Structure Grammar, HPSG Online Proceedings, 48--69. Stanford: CSLI Publications.
\item Beecher, Henry. 2008. Pragmatic inference in the interpretation of sluiced prepositional
phrases. San Diego Linguistics Papers 3. http://escholarship.org/uc/item/2261c0tg.
\item Bresnan, Joan (ed.). 1982. The Mental Representation of Grammatical Relations. Cambridge, Massachusetts: MIT Press.
\item Chaves, Rui P. 2008. Linearization-based word-part ellipsis. Linguistics and Philosophy 3(31):261--307.
\item Chaves, Rui P. 2009. A linearization-based approach to gapping. In James Rogers, (ed.),
Proceedings of FG-MoL 2005, 205--218. Stanford, CA: CSLI Publications.
\item Chaves, Rui P. 2014. On the disunity of right-node raising phenomena. Language 90:834--886.
\item Chung, Sandra, William A. Ladusaw and James McCloskey. 1995. Sluicing and logical form. Natural Language and Linguistic Theory 3:239--282.
\item Crysmann, Berthold. 2008. An asymmetric theory of peripheral sharing in HPSG:
Conjunction reduction and coordination of unlikes. In Gerald Penn, (ed.), Proceedings of
FGVienna: The 8th Conference on Formal Grammar, 45--64. Stanford, CA: CSLI
Publications.
\item Culicover, Peter W. and Ray S. Jackendoff. 2005. Simpler Syntax. Oxford: Oxford University
Press.
\item Fern\'{a}ndez, Raquel. 2006. Non-Sentential Utterances in Dialogue: Classification, Resolution
and Use. Ph.D. thesis, King's College, London.
\item Fern\'{a}ndez, Raquel and Jonathan Ginzburg. 2002. Non-sentential utterances: A corpus study.
Traitement automatique des languages Dialogue 43(2):13--42.
\item Fern\'{a}ndez, Raquel, Jonathan Ginzburg, and Shalom Lappin. 2007. Classifying ellipsis in
dialogue: A machine learning approach. Computational Linguistics 33(3):397--427.
\item Fortin, Catherine. 2007. Indonesian sluicing and verb phrase ellipsis: Description and
explanation in a minimalist framework. Doctoral dissertation, University of Michigan, Ann
Arbor.
\item Ginzburg, Jonathan. 1994. An update semantics for dialogue. In H. Bunt, (ed.), Proceedings of
the 1st International Workshop on Computational Semantics. Tilburg: ITK, Tilburg
University.
\item Ginzburg, Jonathan. 2012. The Interactive Stance: Meaning for Conversation. Oxford:
Oxford University Press.
\item Ginzburg, Jonathan. 2013. Rethinking deep and surface: towards a comprehensive model of anaphoric processes in
dialogue.
\item Ginzburg, Jonathan and Robin Cooper. 2004. Clarification, ellipsis, and the nature of
contextual updates. Linguistics and Philosophy 27(3):297?66.
\item Ginzburg, Jonathan and Robin Cooper. 2014. Quotation via dialogical interaction. Journal of
Logic, Language, and Information 23(3):287?11.
\item Ginzburg, Jonathan and Raquel Fern\'{a}ndez. 2010. Computational models of dialogue. In Alex
Clark, Chris Fox, and Shalom Lappin, (eds.), Handbook of Computational Linguistics and
Natural Language. Oxford: Blackwell.
\item Ginzburg, Jonathan, Raquel Fern\'{a}ndez, and David Schlangen. 2014. Disfluencies as
intra-utterance dialogue moves. Semantics and Pragmatics 7(9):1?4.
\item Ginzburg, Jonathan and Ivan A. Sag. 2000. Interrogative Investigations: the form, meaning
and use of English Interrogatives. Number 123 in CSLI Lecture Notes. Stanford:
California: CSLI Publications.
\item Ginzburg, Jonathan and Philip Miller. 2018. Ellipsis in Head-Driven Phrase Structure Grammar. In Jeroen van Creanen and Tanja Temmerman (eds.), The Oxford handbook of ellipsis. Oxford: OUP.
\item Grant, Margaret, Charles Clifton, and Lyn Frazier. 2012. The role of non-actuality
implicatures in processing elided constituents. Journal of Memory and Language
66:326--343.
\item Griffiths, James, and Anik\'{o} Lipt\'{a}k. 2014. Contrast and island sensitivity in clausal ellipsis. Syntax 17:189--234.
\item Hankamer, Jorge and Ivan A. Sag. 1976. Deep and Surface Anaphora. Linguistic
Inquiry 7(3):391--428.
\item Hardt, Daniel. 1993. Verb Phrase Ellipsis: Form, Meaning, and Processing. Ph.D. thesis,
University of Pennsylvania. Distributed as IRCS Report 93--23.
\item Jacobson, Pauline. 2016. The short answer: Implications for direct compositionality (and vice
versa). Language 92(2):331--375.
\item Kehler, Andrew. 2002. Coherence, Reference, and the Theory of Grammar. Stanford: CSLI
Publications.
\item Kertz, Laura. 2013. Verb phrase ellipsis: The view from information structure. Language
89:390--428.
\item Kim, Christina S., Gregory M. Kobele, Jeffrey T. Runner, and John T. Hale. 2011. The
acceptability cline in VP ellipsis. Syntax 14:318--354.
\item Kim, Christina and Jeffrey Runner. 2017. The division of labor in explanations of verb phrase ellipsis. Linguistics and Philosophy 41:41--85.
\item Kim, Jong-Bok. 2006. Similarities and differences between English VP ellipsis and VP fronting: An HPSG analysis. Studies in Generative Grammar 13.3:429--459.
\item Kim, Jong-Bok. 2015. Syntactic and semantic identity in Korean sluicing: A direct interpretation approach. Lingua 66:260--293.
\item Kim, Jong-Bok. 2016. The syntactic structures of Korean: A construction-based perspective. Cambridge: Cambridge University Press.
\item Larsson, Steffan. 2002. Issue based Dialogue Management. Ph.D. thesis, Gothenburg
University.
\item Lee, Hanjung. 2016. Usage probability and subject--object asymmetries in Korean case ellipsis: Experiments with subject case ellipsis. Journal of Linguistics 52:70--110.
\item Leung, Tommy. 2014. The Preposition Stranding Generalization and conditions on sluicing: Evidence from Emirati Arabic. Linguistic Inquiry 45:332--340.
\item Lobeck, Anne. 1995. Ellipsis. Functional heads, licensing, and identification. Oxford: Oxford University Press.
\item L\'{o}pez, Luis. 2000. Verb Phrase Ellipsis and Discourse-Linking. Lingua 110:183--213.
\item Manning, Christopher and Ivan Sag. 1999. Argument structure, valence, and binding. Nordic Journal of Linguistics 21:107--144.
\item Melnik, Nurit. 2007. Extending partial pro-drop in modern Hebrew: A comprehensive analysis. In Stefan M\"{u}ller, (ed.), Proceedings of the 14th International Conference on Head-Driven Phrase Structure Grammar, HPSG Online Proceedings, 173--193. Stanford: CSLI Publications. URL http://csli-publications.stanford.edu/HPSG/2007.
\item Merchant, Jason. 2001. The Syntax of Silence. Sluicing, Islands, and the Theory of Ellipsis.
Oxford: Oxford University Press.
\item Merchant, Jason. 2004. Fragments and ellipsis. Linguistics and Philosophy 27:661--738.
\item Merchant, Jason. 2005. Revisiting syntactic identity conditions. Workshop on ellipsis, University of California, Berkeley.
\item Merchant, Jason. 2013. Voice and ellipsis. Linguistic Inquiry 44:77--108.
\item Miller, Philip. 2011. The choice between verbal anaphors in discourse. In I. Hendrickx,
S. Lalitha Devi, A. Branco, and R. Mitkov, (eds.), Anaphora Processing and Applications:
8th Discourse Anaphora and Anaphor Resolution Colloquium, DAARC 2011, Volume 7099
of Lecture Notes in Artificial Intelligence, 82--95. Berlin: Springer.
\item Miller, Philip and Barbara Hemforth. 2014. Verb phrase ellipsis with nominal antecedents.
Ms. Universit\'{e} Paris Diderot.
\item Miller, Philip and Geoffrey K. Pullum. 2014. Exophoric VP ellipsis. In Philip Hofmeister and
Elisabeth Norcliffe, (eds.), The Core and the Periphery: Data-driven Perspectives on Syntax Inspired by Ivan A. Sag, 5--32. Stanford, CA: CSLI Publications.
\item Molimpakis, Emilia. 2018. P-stranding under sluicing: Acceptability and processing. Workshop Relating Elliptical Utterances to Information in Context. Stuttgart, Germany.
\item Morgan, Jerry. 1989. Sentence fragments revisited. In Bradley Music, Randolph Graczyk, and Caroline Wiltshire (eds.), CLS 25, 228--241. Chicago Linguistic Society, Chicago.
\item Nykiel, Joanna. 2013. Clefts and preposition omission under sluicing. Lingua 123:74--117.
\item Park, Sang-Hee, Koenig, Jean-Pierre, and Rui P. Chaves. 2018. Scope ambiguities and Lexical Resource Semantics: The case of gapping. Paper presented at the 25th International Conference on Head-Driven Phrase Structure Grammar.
\item Philippova, Tatiana. 2014. P-omission under sluicing, [P clitic] and the nature of P-stranding. In Martin Kohlberger, Kate Bellamy and Eleanor Dutton (eds.), Proceedings of ConSOLE XXII, 133--155.
\item Purver, Matthew. 2006. Clarie: Handling clarification requests in a dialogue system. Research
on Language \& Computation 4(2):259--288.
\item Rodrigues, Cillene, Andrew Nevins and Luis Vicente. 2009. Cleaving the interactions between sluicing and preposition stranding. In D. Torck and L. Wetzels (eds.), Romance Languages and Linguistic Theory 2006. 175--198. Amsterdam: John Benjamins Publishing Company.
\item Ross, John R. 1967. Constraints on variables in syntax. (Ph.D. Dissertation). MIT.
\item Ross, John R. 1969. Guess who. In Proceedings of the 5th Annual Meeting of the Chicago
Linguistic Society, 252--286. CLS.
\item Ruppenhofer, Josef  and Laura A. Michaelis. 2014. Frames and the interpretation of omitted arguments in English.
In Katz Bourns, Stacey and Lindsy L. Myers (eds.), Perspectives on Linguistic Structure and Context: Studies in honor of Knud Lambrecht, 57--86. John Benjamins.
\item Sag, Ivan A. 2012. Sign-Based Construction Grammar: An informal synopsis. In Boas and Sag (eds.), Sign-Based Construction Grammar, 69--202. Stanford: CSLI Publications.
\item Sag, Ivan A. and Joanna Nykiel. 2011. Remarks on sluicing. In Stefan M\"{u}ller, (ed.),
Proceedings of the 18th International Conference on Head-Driven Phrase Structure
Grammar, HPSG Online Proceedings, 188--298. Stanford: CSLI Publications. URL http://
web.stanford.edu/group/cslipublications/cslipublications/HPSG/2011/sag-nykiel.pdf.
\item Sag, Ivan A., Gerold Gazdar, Thomas Wasow, and Steven Weisler. 1985. Coordination and how to distinguish categories.
Natural Language and Linguistic Theory, 3, 117--171.
\item Schachter, Paul. 1977. Does she or doesn't she? Linguistic Inquiry 8:763--767.
\item Schmeh, Katharina, Peter W. Culicover, Jutta Hartmann and Susanne Winkler. 2015. Discourse function ambiguity of fragments: A linguistic puzzle. In S. Winkler (ed.), Ambiguity, 368--400. de Gruyter.
\item Stainton, Robert. 2006. Words and Thoughts. Oxford University Press.
\item Stjepanovi\'{c}, Sandra. 2008. P-stranding under sluicing in a non-P-stranding language? Linguistic Inquiry 37:179--190.
\item Stjepanovi\'{c}, Sandra. 2012. Two cases of violation repair under sluicing. In Jason Merchant and Andrew Simpson (eds.), Sluicing. Cross-linguistic perspectives. 68--82. Oxford: Oxford University Press.
\item Szczegielniak, Adam. 2008. Islands in sluicing in Polish. In Natasha Abner and Jason Bishop (eds.), Proceedings of the 27th West Coast Conference on Formal Linguistics. Cascadilla Proceedings Project, 404--412. Somerville, MA, USA.
\item Vicente, Luis. 2015. Morphological case mismatches under sluicing. Snippets 29.
\item Webber, Bonnie. 1979. A Formal Approach to Discourse Anaphora. Outstanding
Dissertations in Linguistics. New York: Garland.
\item Yatabe, Sh\^{u}ichi. 2001. The Syntax and Semantics of Left-Node Raising in Japanese. In
D. Flickinger and A. Kathol (eds.), Proceedings of the 7th International Conference
on Head-Driven Phrase Structure Grammar, CSLI, Stanford, pp. 325--344.
\item Yatabe, Sh\^{u}ichi. 2012. Comparison of the Ellipsis-Based Theory of Non-Constituent Coordination with Its Alternatives. In S. M\"{u}ller (ed.), Proceedings of the 19th International
Conference on Head-Driven Phrase Structure Grammar, CSLI, Stanford, 454--474.
\item  Yatabe, Sh\^{u}ichi and Wai Lok Tam. 2018. In defense of an HPSG-based theory of
non-constituent coordination: A reply to Kubota and Levine. Ms. University of Tokyo and Rozetta Corp.
\end{description}

 
\printbibliography[heading=subbibliography,notkeyword=this] 
\end{document}
