\documentclass[output=paper,biblatex,babelshorthands,newtxmath,draftmode,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.13644999}

\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
  \usepackage{../nomemoize}
  \input{../localpackages}
  \input{../localcommands}
  \input{../locallangscifixes.tex}

  \togglepaper[18]
}{}

\author{Manfred Sailer\orcid{0000-0002-1693-8859}\affiliation{Goethe-Unversität Frankfurt}}
\title{Idioms}


\abstract{This chapter first sketches basic empirical properties of idioms. The state of the art before the emergence of HPSG is presented, followed by a discussion of four types of HPSG approaches to idioms. A section on future research closes the discussion.}

\begin{document}
\maketitle
\label{chap-idioms}


%%% Index cross references
%\is{MWE|see{multiword expression}}
%%%

\section{Introduction}
\label{Sec-Intro}
%\section{Introduction}

In this chapter, I will use the term \emph{idiom} interchangeably with broader terms such as
\emph{phraseme}\is{idiom|(}, \emph{phraseologism}\is{phraseologism|see{idiom}},
\emph{phraseological unit}\is{phraseological unit|see{idiom}}, or \emph{multiword expression}. 
This means, that I will subsume under this notion expressions such as prototypical idioms
(\bspT{kick the bucket}{die}), support verb constructions (\bsp{take advantage}), formulaic
expressions (\bsp{Good morning!}), and many more.\footnote{I will provide a paraphrase for all
  idioms at their first mention. They are also listed in the appendix, together 
  with their paraphrase and a remark on which aspects of the idiom are discussed in the text.} The
main focus of the discussion will, however, be on prototypical idioms.

%In the rest of this section, 
I will sketch some empirical aspects of idioms in Section~\ref{Sec-EmpiricalDomain}.
%and, then, characterize theoretical issues that arise in the formal modelling of idioms, both in general and with respect to HPSG in particular. 
In Section~\ref{Sec-Predecessors}, I will present the theoretical context within which idiom analyses arose in HPSG.
An overview of the development within HPSG will be given in Section~\ref{Sec-Analyses}. 
%Section~\ref{Sec-RecentOtherFrameworks} contains a brief sketch of the theoretical development outside HPSG. 
Desiderata for future research are mentioned in Section~\ref{Sec-WhereToGo}, before I close with a short conclusion.

%\subsection
\section{Empirical domain}
\label{Sec-EmpiricalDomain}
%Defining the empirical domain of idioms and phraseology. I will aim at a very inclusive definition, i.e., more in lines with ``phraseology'' than with ``idioms'' in the strict sense.

%I will assume the basic characterization of a phraseological unit from \citet{Fleischer97a-u} and \citet{Burger:98} as \emph{complex} units that show at least one of \emph{fixedness}, \emph{(semantic) idiomaticity}, and \emph{lexicalization}.

\is{idiomaticity|(}
In the context of the present handbook, the most useful characterization of idioms might be the
definition of \emph{multiword expression}\is{multiword expression} from  \citet[\page 269]{Baldwin:Kim:10}.
%\footnote{See also \citet{Sailer:18SemComp} for a more detailed summary of }
For them, any combination of words counts as a multiword expression if it is syntactically complex
and shows some degree of \emph{idiomaticity} (i.e., irregularity), be it lexical, syntactic,
semantic, pragmatic, or statistical.% 
\footnote{In the phraseological tradition, the aspect of \emph{lexicalization} is added
  \citep{Fleischer97a-u,Burger:98}. This means that an expression is stored in the lexicon. This
  criterion might have the same coverage as \emph{conventionality} used in
  \citet[492]{NSW94a}. These criteria are addressing the mental representation of idioms as a
  unit and are, thus, rather psycholinguistic in nature.} 
%
I speak of a ``combination of words'' in the sense of a \emph{substantive} or \emph{lexically filled idiom}, which 
contrasts with \emph{formal} or \emph{lexically open idioms} \citep[505]{FKoC88a}. 

\citeauthor{Baldwin:Kim:10}'s criteria can help us structure the data presentation in this section, expanding their criteria where it seems suitable.
My expansions concern the aspect known as \emph{fixedness} in the phraseological tradition as in \citet{Fleischer97a-u}.%
\footnote{\citet{Baldwin:Kim:10} describe idioms in terms of syntactic fixedness, but they seem to consider fixedness a derived notion.}

\begin{sloppypar}
\is{idiomaticity!lexical|(}%
For \citet{Baldwin:Kim:10}, \emph{lexical idiosyncrasy} concerns expressions with words that only
occur in an idiom, so-called \emph{phraseologically bound words}\is{bound word}, or \emph{cranberry
  words}\is{cranberry word|see{bound word}} \citep[\page 15]{Aronoff76a-u}. Examples include \bspT{make headway}{make progress}, 
\bspT{take umbrage}{take offense},
%\bspT{at first blush}{at first sight}, 
\bspT{in a trice}{in a moment/""very quickly}.%
\footnote{
See https://www.english-linguistics.de/codii/, accessed 2019-09-03, for a list of bound words\is{bound word} in \ili{English} and \ili{German} \citep{Trawinski:al:08lrec}.}
For such expressions, the grammar has to make sure that the bound word does not occur outside the idiom, i.e., we need to prevent combinations such as \refer{trice-ko}.%
\footnote{Tom Wasow (p.c.) points out that there are attested uses of many alleged bound words\is{bound word} outside their canonical idiom, as in \pref{not-a-trice}. Such uses are, however, rare and restricted.
\ea
[]{Not a trice later, the sounds of gunplay were to be heard echoing from Bad Man's Rock. (COCA)\label{not-a-trice}}
\zlast
}
\end{sloppypar}

\eal \label{trice}
\ex []{They fixed the problem in a trice.}
\ex [*]{It just took them a trice to fix the problem.\label{trice-ko}}
\zl 

\largerpage[-1]
\noindent
We can expand this type of idiosyncrasy to include  a second important property of idioms. 
Most idioms have a fixed inventory of words. In their summary of this aspect of idioms, \citet[\page 827--828]{Gibbs:Colston:07} include the following examples: \bsp{kick the bucket} means `die', but \bsp{kick the pail}, \bsp{punt the bucket}, or \bsp{punt the pail} do not have this meaning. However, some degree of lexical variation seems to be allowed, as
the idiom \bspT{break the ice}{relieve tension in a strained situation} can be varied into \bsp{shatter the ice}.%
\footnote{\label{fn-semmeln}While \citet{Gibbs:Colston:07}, following \citet{Gibbs:al:89}, present this example as a lexical variation, \citew[\page 85]{Glucksberg:01}, from which it is taken, characterizes it as having a somewhat different aspect of an ``abrupt change in the social climate''. Clear cases of synonymy under lexical substitution are found with \ili{German} \bspTL{wie warme Semmeln/Brötchen/Schrippen weggehen}{like warm rolls vanish}{sell like hotcakes} in which some regional terms for rolls can be used in  the idiom.}
%For example, \citet[\page 85]{Glucksberg:01} mentions the variation \bsp{shatter the ice} of the idiom \bspT{break the ice}{XXX}.
 So, a challenge for idiom theories is to guarantee that the 
right lexical elements are used in the right constellation.
\is{idiomaticity!lexical|)}

%\medskip%
\is{idiomaticity!syntactic|(}
\emph{Syntactic idiomaticity} is used in \citet{Baldwin:Kim:10} to describe expressions that are not
formed according to the productive rules of \ili{English} syntax, following \citet{FKoC88a}, such as
\emph{by and large}`on the whole'/""`everything considered' and \bspT{trip the light fantastic}{dance}.  

In my expanded use of this notion, syntactic idiomaticity also subsumes
irregularities/""restrictions in the syntactic flexibility of an idiom, i.e., whether an idiom can
occur in the same syntactic constructions as an analogous non"=idiomatic combination. In
Transformational Grammar\is{Transformational Grammar}, such as \citet{Weinreich:69} and
\citet{Fraser:70}, lists of different syntactic transformations were compiled and it was observed
that some idioms allow for certain transformations but not for others. This method has been pursued
systematically in the framework of \emph{Lexicon-Grammar}\is{Lexicon-Grammar} \citep{Gross:82}.% 
\footnote{See \citet{Laporte:18} for a recent discussion of applying this method for a classification of idioms.}
%
\citet{SBBCF2002a-u} distinguish three levels of fixedness: \emph{fixed}, \emph{semi-fixed}, and \emph{flexible}. 
Completely fixed idioms include \bsp{of course}, \bsp{ad hoc} and are often called \emph{words with spaces}\is{word with spaces}.
Semi-fixed idioms allow for morphosyntactic variation such as inflection. These include some
prototypical idioms (\bsp{trip the light fantastic}, \bsp{kick the bucket}) and complex proper
names. In \ili{English}, semi-fixed idioms show inflection, but cannot easily be passivized, nor do
they allow for parts of the idiom being topicalized or pronominalized, see \refer{kick-ex}.

\eal \label{kick-ex} 
\ex[]{Alex kicked / might kick the bucket.}
\ex[*]{The bucket was kicked by Alex.}
\ex[*]{The bucket, Alex kicked.}
\ex[*]{Alex kicked the bucket and Kim kicked it, too.}
\zl 

\largerpage[-1]
\noindent
Flexible idioms pattern with free combinations. For them, we do not only find inflection, but also
passivization, topicalization, pronominalization of parts, etc. Free combinations include some
prototypical idioms (\bspT{spill the beans}{reveal a secret}, \bspT{pull strings}{exert
  influence'/""`use one's connections}), but also collocations (\bsp{brush one's teeth}) and light
verbs (\bsp{make a mistake}).

The assumption of two flexibility classes is not uncontroversial: 
\citet{Horn:03} distinguishes two types among what \citet{SBBCF2002a-u} consider flexible idioms. 
\citet{Fraser:70} assumes six flexibility classes, looking at a wide range of syntactic operations.
\citet{Ruwet:91} takes issue with the cross-linguistic applicability of the classification of
syntactic operations. Similarly, \citet{Schenk:95} claims that for languages such as \ili{Dutch} and
\ili{German}, automatic/meaningless syntactic processes other than just inflection are possible for
semi-fixed idioms, such as verb-second movement\is{word order!V2} and some types of fronting.

The analytic challenge of syntactic idiomaticity is to capture the difference in flexibility in a
non"=ad hoc way. It is this aspect of idioms that has received particular attention in Mainstream
Generative\is{Generative Grammar} Grammar (MGG),\footnote{%
  I follow \citet[\page 3]{CJ2005a} in using the term \emph{Mainstream Generative Grammar} to
  refer to work in Minimalism and the earlier Government \& Binding framework.}$^,$\footnote{See the
  references in \citet{Corver:al:19} for a brief up-to-date overview of MGG work.}
but also in the HPSG approaches sketched in Section~\ref{Sec-Analyses}.
\is{idiomaticity!syntactic|)}

%\medskip%
\is{idiomaticity!semantic|(}
\emph{Semantic idiomaticity} may sound pleonastic, as, traditionally, an expression is called idiomatic if it has a conventional meaning that is different from its literal meaning. 
Since I use the terms idiom and idiomaticity in their broad senses of phraseological unit and irregularity, respectively, the qualification \emph{semantic} idiom(aticity) is needed. 

One challenge of the modeling of idioms is capturing the relation between the literal and the idiomatic meaning of an expression.
\citet{Gibbs:Colston:07} give an overview of 
psycholinguistic research on idioms. Whereas it was first assumed that speakers would compute the literal meaning of an expression and then derive the idiomatic meaning, evidence has been accumulated that the idiomatic meaning is accessed directly.

\citet*{WSN84a-u} and \citet*{NSW94a} explore various semantic relations for idioms, in particular \emph{decomposability}\is{decomposability} and \emph{transparency}\is{transparency}.
An idiom is \emph{decomposable} if its idiomatic meaning can be distributed over its component parts in such a way that we would arrive at the idiomatic meaning of the overall expression if we interpreted the syntactic structure on the basis of such a meaning assignment. 
The idiomatic meaning of the expression \bsp{pull strings} can be decomposed by interpreting \bsp{pull} as \bsp{exploit/use} and \bsp{strings} as \bsp{connections}. 
%To make this criterion non"=arbitrary, it is now common to require that there be empirical support for such a decomposition. In \ili{English}, we can often insert an adjective 
The expressions \bsp{kick the bucket} and \bspT{saw logs}{snore} are not decomposable.

\largerpage
An idiom is \emph{transparent} if there is a synchronically accessible relation between the literal
and the idiomatic meaning of an idiom.  For some speakers, \bsp{saw logs} is transparent in this
sense, as the noise produced by this activity is similar to a snoring noise.  For \bsp{pull
  strings}, there is an analogy to a puppeteer controlling the puppets' behavior by pulling
strings. A non"=transparent idiom is called \emph{opaque}.

Some idioms do not show semantic idiomaticity at all, such as collocations\is{collocation} (\bsp{brush one's teeth}) or support verb constructions (\bsp{take a shower}). 
Many body-part expressions such as \bspT{shake hands}{greet} or \bspT{shake one's head}{decline/""negate} constitute a more complex case: they describe a conventionalized activity and denote the social meaning of this activity.% 
\footnote{The basic reference for the phraseological properties of body-part expressions is \citew{Burger:76}.}

In addition, we might need to assume a \emph{figurative} interpretation. For some expressions, in particular proverbs\is{proverb} or cases like 
\bspT{take the bull by the horns}{approach a problem directly} we might get a figurative reading rather than an idiomatic reading. 
%
\citet{Glucksberg:01} explicitly distinguishes between idiomatic and figurative interpretations. In his view, the above-mentioned case of \bsp{shatter the ice} would be a figurative use of the idiom \bsp{break the ice}. 
While there has been a considerable amount of work on figurativity in psycholinguistics, the integration of its results into formal linguistics is still a desideratum.%
\is{idiomaticity!semantic|)}



%\medskip%
\is{idiomaticity!pragmatic|(}
\emph{Pragmatic idiomaticity} covers expressions that have a \emph{pragmatic point} in the terminology of \citet{FKoC88a}. These include complex formulaic expressions (\bsp{Good morning!}). There has been little work on this aspect of idiomaticity in formal phraseology.
\is{idiomaticity!pragmatic|)}

%\medskip%
\is{idiomaticity!statistical|(}%
The final type of idiomaticity is \emph{statistical idiomaticity}.
Contrary to the other idiomaticity criteria, this is a usage-based aspect. If we find a high degree
of co-occurrence of a particular combination of words that is idiosyncratic for this combination, we
can speak of a statistical idiomaticity. This category includes
\emph{collocations}\is{collocation}. \citet{Baldwin:Kim:10} mention \bsp{immaculate performance} as
an example. Collocations are important in computational linguistics and in foreign-language
learning, but their status for theoretical linguistics and for a competence-oriented\is{competence}
framework such as HPSG is unclear.%
\is{idiomaticity!statistical|)}

%\bigskip%
\largerpage
This discussion of the various types of idiomaticity shows that idioms do not form a homogeneous empirical domain but rather are defined negatively. 
This leads to the basic analytical challenge of idioms: while the empirical domain is defined by  absence of regularity in at least one aspect, idioms largely obey the principles of grammar. 
In other words, there is a lot of regularity in the domain of idioms, while any approach still needs to be able to model the irregular properties.%
%I have tried to introduce the notions that are most commonly used in HPSG research and to identify the analytical problems related to them. 
\is{idiomaticity|)}

%\subsection{Language-theoretical interest in idioms}
%\label{Sec-TheoreticalInterest}

%\begin{itemize}
%\item between rule-based and idiosyncratic
%\item compositional challenge and collocational challenge
%\end{itemize}


\section{Predecessors to HPSG analyses of idioms}
\label{Sec-Predecessors}


In this section, I will sketch the theoretical environment within which HPSG and HPSG analyses of idioms have emerged.

The general assumption about idioms in MGG\is{Generative Grammar} is that they
must be represented as a complex phrasal form-meaning unit. 
Such units are inserted \emph{en bloc}\is{en bloc insertion} into the structure rather than built by syntactic operations.
This view goes back to \citet[\page 190]{Chomsky:65}. 
With this unquestioned assumption, arguments for or against particular analyses can be constructed. 
To give just one classical example, \citet[\page 85]{Chomsky81a} uses the passivizability of some idioms as an argument for the existence of Deep Structure, i.e., a structure on which the idiom is inserted holistically. 
%
\citet{Ruwet:91} and \citet{NSW94a} go through a number of such lines of argumentation showing their basic problems. 

The holistic view on idioms is most plausible for idioms that show many types of idiomaticity at the same time, though it becomes more and more problematic if only one or a few types of idiomaticity are attested.
HPSG is less driven by analytical pre-decisions than other frameworks; see
\crossrefchaptert[Section~\ref{minimalism:sec-formalization-exhaustivity}]{minimalism}. Nonetheless, idioms have been used to motivate assumptions about the architecture of linguistic signs in HPSG as well.

\citet{WSN84a-u} and \citet{NSW94a} are probably the two most influential papers in formal phraseology in the last decades. 
%These papers have also shaped the analysis of idioms in \emph{Generalized Phrase Structure Grammar} \citep{GKPS85a}\is{Generalized Phrase Structure Grammar} and, consequently in HPSG. 
While there are many aspects of \citet{NSW94a} that have not been integrated into the formal modeling of idioms, 
there are at least two insights that have been widely adapted in HPSG.
First, not all idioms should be represented holistically. 
Second, the syntactic flexibility of an idiom is related to its semantic decomposability. 
In fact, \citet{NSW94a} state this last insight even more generally:%
\footnote{Aspects of this approach are already present in \citet{Higgins:74} and \citet{Newmeyer:74}.}

%\ea \label{NSW-quote} \citet[\page 531]{NSW94a}:
\largerpage
\begin{quote}
We predict that the syntactic flexibility of a particular idiom will ultimately be explained in
terms of the compatibility of its semantics with the semantics and pragmatics of various
constructions. 
\citep[\page 531]{NSW94a}\label{NSW-quote}
\end{quote}
%\z 
%One of the theoretical predecessors of HPSG is \emph{Generalized Phrase Structure Grammar} \citep{GKPS85a}\is{Generalized Phrase Structure Grammar}. 

\noindent
\citet{WSN84a-u} and \citet{NSW94a} propose a simplified first approach to a theory that would be in
line with this quote. They argue that, for \ili{English}, there is a correlation between syntactic
flexibility and semantic decomposability in that non"=decomposable idioms are only semi-fixed,
whereas decomposable idioms are flexible, to use our terminology from
Section~\ref{Sec-EmpiricalDomain}.  
\indexgpsgstart
This idea has been directly encoded formally in the idiom theory of
\citet*[Chapter~7]{GKPS85a}, who define the framework of 
\emph{Generalized Phrase Structure Grammar}\indexgpsg (GPSG).

\citet{GKPS85a} assume that non"=decomposable idioms are inserted into sentences \emph{en
  bloc}\is{en bloc insertion}, i.e., as fully specified syntactic trees which are assigned the
idiomatic meaning holistically. This means that the otherwise strictly context-free grammar of GPSG
needs to be expanded by adding a (small) set of larger trees.  Since non"=decomposable idioms are
inserted as units, their parts cannot be accessed for syntactic operations such as passivization or
movement. Consequently, the generalization about semantic non"=decomposability and syntactic
fixedness of \ili{English} idioms from \citet{WSN84a-u} is implemented directly.

Decomposable idioms are analyzed as free combinations in syntax. The idiomaticity of such
expressions is achieved by two assumptions: First, there is lexical ambiguity, i.e., for an idiom
like \bsp{pull strings}, the verb \bsp{pull} has both a literal meaning and an idiomatic
meaning. Similarly for \bsp{strings}.  Second, \citet{GKPS85a} assume that lexical items are not
necessarily translated into total functions but can be partial functions. Whereas the literal
meaning of \bsp{pull} might be a total function, the idiomatic meaning of the word would be a
partial function that is only defined on elements that are in the denotation of the idiomatic
meaning of \bsp{strings}. This analysis predicts syntactic flexibility for decomposable idioms, just
as proposed in \citet{WSN84a-u}.  \indexgpsgend



%Related to this general dilemma are two more concrete analytical challenges, which \citet[\page 12]{Bargmann:Sailer:18} call the \emph{compositional challenge} and the \emph{collocational challenge}. The compositional challenge consists in associating a sequence of words with a non"=literal, idiomatic, meaning. The collocational challenge consists in making sure that the components of an idiom all occur together in the right constellation.


%\section{Predecessors to HPSG analyses of idioms}
%\label{Sec-Predecessors}

%\begin{itemize}
%\item Generalized Phrase Structure Grammar \citep{GKPS85a}
%\item Semi-formal, influential papers \citep{WSN84a-u,NSW94a}
%\item Construction Grammar \citep{FKoC88a}
%\end{itemize}


\citet[511--514]{NSW94a} show that the connection between semantic decomposability and syntactic flexibility is not as straightforward as suggested. They say that, in \ili{German} and \ili{Dutch}, ``noncompositional\is{idiom!non-decomposable} idioms  are syntactically versatile'' \citep[514]{NSW94a}. Similar observations have been brought forward for \ili{French} in \citet{Ruwet:91}. 
\citet{Bargmann:Sailer:18}, \citet{Fellbaum:19}, and \citet[Chapter~3]{Bargmann:23} argue that even for \ili{English}, passive examples are attested for non"=decomposable idioms such as \pref{ex-kick-fellbaum}.

\ea
\label{ex-kick-fellbaum}
Live life to the fullest, you never know when the bucket will be kicked.\footnote{
\citew[756]{Fellbaum:19}
}
\z 

\noindent
The current state of our knowledge of the relation between syntactic and semantic idiosyncrasy is that the semantic idiomaticity of an idiom does have an effect on its syntactic flexibility, though the relation is less direct than assumed in the literature based on \citew{WSN84a-u} and \citew{NSW94a}.


\section{HPSG analyses of idioms}
\label{Sec-Analyses}

%\subsection{HPSG-specific research questions}

%\largerpage
HPSG does not make a core-periphery\is{core grammar} distinction; see \citet{MuellerKernigkeit}. Consequently, idioms belong to the empirical domain to be covered by an HPSG grammar.
Nonetheless, idioms are not discussed in \citet{ps2} and their architecture of grammar does not have a direct place for an analysis of idioms.%
\footnote{This section follows the basic structure and argument of \citet{Sailer:12} and \citet{Richter:Sailer:14}.} 
They situate all idiosyncrasy in the lexicon, which consists of lexical entries for basic words. 
Every word has to satisfy a lexical entry and all principles of grammar; see \crossrefchaptert{lexicon}.%
\footnote{I refer to the lexicon\is{lexicon} in the technical sense as the collection of lexical entries, i.e., as \emph{descriptions}, rather than as a collection of lexical items, i.e., linguistic signs. 
Since \citet{ps2} do not discuss morphological processes, their lexical entries describe full forms. 
If there is a finite number of such lexical entries, the
lexicon can be expressed by a \emph{Word Principle}\is{principle!Word}, a constraint on words that contains a disjunction of all such lexical entries. 
Once we include morphology, lexical rules, and 
idiosyncratic, lexicalized phrases
in the picture, we need to refine this simplified view.
%
}
%
All properties of a phrase can be inferred from the properties of the lexical items occurring in the phrase and the constraints of grammar. 

In their grammar, \citet{ps2} adhere to the \emph{Strong Locality Hypothesis}
(SLH),\is{locality!Strong Locality Hypothesis}\is{locality|(} i.e., all lexical entries describe
leaf nodes in a syntactic structure and all phrases are constrained by principles that only refer to
local (i.e., \type{synsem}) properties of the phrase and to local properties of its immediate
daughters. This hypothesis is summarized in \refer{slh}. 

%\vbox{
\eanoraggedright
Strong Locality Hyphothesis\label{slh} (SLH)\\
The rules and principles of grammar are statements on a single node of a linguistic structure or on nodes that are immediately dominated by that node.\is{SLH|see{locality!Strong Locality Hypothesis}}
\z 
%}

\noindent
This precludes any purely phrasal approaches to idioms.  Following the heritage of GPSG\indexgpsg,
we would assume that all regular aspects of linguistic expressions can be handled by mechanisms that
follow the SLH, whereas idiomaticity would be a range of phenomena that may violate it.  It is,
therefore, remarkable that a grammar framework that denies a core-periphery distinction would start
with a strong assumption of locality, and, consequently, of regularity.

%\largerpage[1.3]
This is in sharp contrast to the basic motivation of Construction Grammar\indexcxg, which assumes
that constructions can be of arbitrary depth and of an arbitrary degree of idiosyncrasy.
\citet{FKoC88a} use idiom data and the various types of idiosyncrasy discussed in
Section~\ref{Sec-EmpiricalDomain} as an important motivation for this assumption.  To contrast this
position clearly with the one taken in \citet{ps2}, I will state the \emph{Strong Non"=locality
  Hypothesis}\is{locality!Strong Non"=locality Hypothesis} (SNH) in \refer{snh}.

%\vbox{
\eanoraggedright
Strong Non"=locality Hypothesis\is{locality!Strong Non"=locality Hypothesis} (SNH)\is{SNH|see{locality!Strong Non-locality Hypothesis}}\label{snh}\\
The internal structure of a construction can be arbitrarily deep and show an arbitrary degree of irregularity at any substructure.
\z
%}

\noindent
The actual formalism used in \citet{ps2} and \citet{King89} -- see
\crossrefchaptert{formal-background} -- does not require the strong versions of the locality and the
non"=locality hypotheses, but is compatible with weaker versions. I will call these the \emph{Weak
  Locality Hypothesis} (WLH),\is{locality!Weak Locality Hypothesis} and the \emph{Weak Non"=locality
  Hypothesis} (WNH); see \refer{wlh} and \refer{wnh} respectively.

%\vbox{
\eanoraggedright
Weak Locality Hypothesis (WLH)\label{wlh}\\
At most the highest node in a structure is licensed by a rule of grammar or a lexical entry.\is{WLH|see{locality!Weak Locality Hypothesis}}
%The rules and principles of grammar can constrain the internal structure of a linguistic sign at arbitrary depth, but each sign needs to be licensed independently.
\z 
%}

\noindent
According to the WLH, just as in the SLH, each sign needs to be licensed by the lexicon and/""or the grammar. 
This precludes any \emph{en bloc}-insertion analyses, which would be compatible with the SNH.
%
According to the WNH, in line with the SLH, a sign can, however, impose further constraints on its
component parts, that may go beyond local (i.e., \type{synsem}) properties of its immediate
daughters. 

\eanoraggedright
Weak Non"=locality Hypothesis\is{locality!Weak Non-locality Hypothesis} (WNH)\is{WNH|see{locality!Weak Non-locality Hypothesis}}\label{wnh}\\
The rules and principles of grammar can constrain -- though not license -- the internal structure of a linguistic sign at arbitrary depth.
\z 

\noindent
This means that all substructures of a syntactic node need to be licensed by the grammar, but the
node may impose idiosyncratic constraints on which particular well-formed substructures it may
contain. 


%\bigskip%
In this section, I will review four types of analyses developed within HPSG in a mildly
chronological order: First, I will discuss a conservative extension of \citet{ps2} for idioms
\citep{KE94a} that sticks to the SLH.  Then, I will look at attempts to incorporate constructional
ideas more directly, i.e., ways to include a version of the SNH. The third type of approach will
exploit the WLH. Finally, I will summarize recent approaches, which are, again, emphasizing the
locality of idioms.
\is{locality|)}

\subsection{Early lexical approaches}
\label{Sec-EarlyLexical}

%\largerpage[1.3]
\citet{KE94a}, based on \citet{Erbach92a}, present the first comprehensive HPSG account of idioms.
They look at a wide variety of different types of \ili{German} idioms, including support verb
constructions.  They only modify the architecture of \citet{ps2} marginally and stick to the Strong
Locality Hypothesis.  They base their analysis on the apparent correlation between syntactic
flexibility and semantic decomposability from \citet{WSN84a-u} and \citet{NSW94a}.  Their analysis
is a representational variant of the analysis in \citet{GKPS85a}.

To maintain the SLH, \citet{KE94a} assume that the information available in syntactic selection is
slightly richer than what has been assumed in \citet{ps2}: first, they use a lexeme-identification
feature, \attrib{lexeme}\isfeat{lexeme}, which is located inside the \attrib{index} value and whose value is the
semantic constant associated with a lexeme.  Second, they include a feature \feat{theta-role}, whose
value indicates which thematic role a sign is assigned in a structure. In addition to standard
thematic roles, they include a dummy value \type{nil}.  Third, as the paper was written in the
transition phase between \citet{ps} and \citet{ps2}, they assume that the selectional attributes
contain complete \type{sign} objects rather than just \type{synsem} objects.  Consequently,
selection for phonological properties and internal constituent structure is possible, which we could
consider a violation of the SLH.

The effect of these changes for the treatment of idioms can be seen in \refer{ke-spill} and
\refer{ke-kick}, which are adaptations of the analyis in \citew[Section~11.4]{KE94a}. In \refer{ke-spill}, I sketch the analysis of the syntactically flexible, decomposable idiom \emph{spill the beans}.  There are individual lexical items for the idiomatic words.%
\footnote{We do not need to specify the \attrib{reln}\isfeat{reln} value for the noun \emph{beans}, as the \attrib{lexeme} and the \attrib{reln} value are usually identical.}

\eal % Analysis of \emph{spill the beans} in the spirit of \citet{KE94a}
\label{ke-spill}
\ex 
\label{le-idiomatic-spill}
\avm{
[phon < \phonfont{spill} > \\
synsem|loc [cat & [subj  & < NP[index & \1] >\\
                   comps & < NP[index & \2 [lexeme & beans\_i]] >] \\
            cont & [reln    & spill\_i\\
                    spiller & \1\\
                    spilled & \2 ] ] ]
}
\ex 
\avm{
[phon < \phonfont{beans} > \\
synsem|loc|content|index|lexeme  \type{beans\_i} ]
}
\zl 

\noindent
The \attrib{lexeme} values of these words can be used to distinguish them from their ordinary,
non"=idiomatic homonyms.  Each idiomatic word comes with its idiomatic meaning, which models the
decomposability of the expression.  For example, the lexical items satisfying the entry in
\refer{le-idiomatic-spill} can undergo lexical rules such as passivization.

%\largerpage[1]
The idiomatic verb \emph{spill} selects an NP complement with the \attrib{lexeme} value
\type{beans\_i}.  The lexicon is built in such a way that no other word selects for this
\attrib{lexeme} value.  This models the lexical fixedness of the idiom.

The choice of putting the lexical identifier into the \attrib{index} guarantees that it is shared
between a lexical head and its phrase, which allows for syntactic flexibility inside the NP.
Similarly, the information shared between a trace and its antecedent contains the \attrib{index}
value. Consequently, participation in unbounded dependency constructions is equally accounted for.
Finally, since a pronoun has the same \attrib{index} value as its antecedent, pronominalization as in
(\ref{ex-pron-spill}) is also possible.

\ea
 \label{ex-pron-spill}
Eventually, she spilled all the beans$_i$. But it took her a few days to spill them$_i$
all.\footnote{
\citew[207]{Riehemann2001a}
}
\z

\noindent
I sketch the analysis of a non"=decomposable, fixed idiom, \emph{kick the bucket}, in
\refer{ke-kick}.  In this case, there is only a lexical entry of the syntactic head of the idiom,
the verb \emph{kick}.  It selects the full \isi{phonology} of its complement. This blocks any syntactic
processes inside this NP. It also follows that the complement cannot be realized as a trace, which
blocks extraction.\footnote{See \crossrefchaptert{udc} for
  details on the treatment of extraction in HPSG.}  
%
The special \feat{theta-role} value \type{nil} will be used to restrict the lexical rules that can
be applied.  The passive lexical rule, for example, would be specified in such a way that it cannot
apply if the NP complement in its input has this theta-role.


\ea % Analysis of \emph{spill the beans} in the spirit of \citet{KE94a} 
\label{ke-kick}
\avm{
[phon < \phonfont{kick} > \\
 synsem|loc [cat & [subj  & < NP[index & \1] > \\
                    comps & < NP[\punk{phon}{\phonliste{ the, bucket }} \\
                                 theta-role & nil] > ] \\
             cont & [reln  & \type{die}\\
                     dying & \1]] ]
}
\z 

\noindent
With such an analysis, \citet{KE94a} capture both the idiosyncratic aspects and the regularity of idioms. 
They show how it generalizes to a wide range of idiom types. 
I will briefly mention some problems of the approach, though.
%There are, however, a number of problems. I will just mention few of them here.

There are two problems for the analysis of non"=decomposable idioms.  First, the approach is too
restrictive with respect to the syntactic flexibility of \emph{kick the bucket}, as it excludes
cases such as \emph{kick the social/figurative bucket}, which are discussed in \citet{Ernst:81}.
Second, it is built on equating the class of non"=decomposable idioms with that of semi-fixed
idioms. As shown in my discussion around example \pref{ex-kick-fellbaum}, this cannot be maintained.

%\largerpage
There are also some undesired properties of the \textsc{lexeme} value selection. The index identity
between a pronoun and its antecedent would require that the subject of the relative
clause\is{relative clause} in \refer{strings-relcl} has the same \attrib{index} value as the head noun
\emph{strings}. However, the account of the lexical fixedness of idioms is built on the assumption
that no verb except for the idiomatic \emph{pull} selects for an argument with \attrib{lexeme} value
\type{strings\_i}.\footnote{\citet{Pulman:93} discusses an analogous problem for the denotational
  theory of \citet{GKPS85a}.}

\ea \label{strings-relcl}
Parky pulled the strings that got me the job.
\citep[137]{McCawley:81}
\z 

\noindent
Notwithstanding these problems, the analytic ingredients of \citet{KE94a} constitute the basis of
later HPSG analyses. In particular, a mechanism for lexeme-specific selection has been widely
assumed in most approaches. The attribute \feat{theta-role} can be seen as a simple form of an
\emph{inside-out} mechanism\is{inside-out constraint}, i.e., as a mechanism of encoding information
about the larger structure within which a sign appears.
%
%\citet{SS2003a}: inside out, 
%\citet{Sag2012a}

\subsection{Phrasal approach}
\label{Sec-Phrasal}

%\largerpage[1]
\is{Constructional HPSG|(}%
With the advent of constructional analyses within HPSG, starting with
\citet{Sag97a}, it is natural to expect phrasal accounts of idioms to emerge as well, as
idiomaticity is a central empirical domain for Construction Grammar\indexcxg; see
\crossrefchaptert{cxg}.  In this version of HPSG, there is an elaborate type hierarchy below
\type{phrase}.  \citet{Sag97a} also introduces \emph{defaults}\is{default} into HPSG, which play an
important role in the treatment of idioms in Constructional HPSG.  The clearest phrasal approach to
idioms can be found in \citet{Riehemann2001a}, which incorporates insights from earlier publications
such as \citet{Riehemann97a} and \citet{RB99a}.
%
The overall framework of \citet{Riehemann2001a} is Constructional HPSG with \emph{Minimal Recursion Semantics}\indexmrs  \citep{CFMRS95a-u,CFPS2005a}; see also \crossrefchaptert[Section~\ref{semantics:sec-mrs}]{semantics}.

\begin{sloppypar}
For Riehemann, idioms are phrasal units.  Consequently, she assumes a subtype of \type{phrase} for each idiom, such as \type{spill-beans-idiomatic-phrase} or \type{kick-bucket-idiomatic-phrase}.  The proposal in \citet{Riehemann2001a} simultaneously is phrasal and obeys the SLH. To achieve this, \citet{Riehemann2001a} assumes an attribute \attrib{words}, whose value contains all words dominated by a phrase. This makes it possible to say that a phrase of type \type{spill-beans-idiomatic-phrase} dominates the words \emph{spill} and \emph{beans}. This is shown in the relevant type constraint for the idiom \emph{spill the beans} in \refer{sr-spillbeans}.%
\footnote{The percolation mechanism for the feature \attrib{words} is rather complex. 
%In fact, in \citew[Section 5.2.1]{Riehemann2001a} the idiom-specific words appear within a \feat{c-words} value and the other words are dominated by the idiomatic phrase in the value of an attribute \feat{other-words}, both of which together form the value of \attrib{words}. While all the values of these features are subject to local percolation principles, 
The fact that entire words are percolated undermines the locality intuition behind the SLH.
}
\end{sloppypar}


\vbox{
\ea Constraint on the type \type{spill-beans-idiomatic-phrase} from \citet[185]{Riehemann2001a}:\label{sr-spillbeans}\\
\avm{
[\type*{spill-beans-ip}
words & \{	[\type*{i\_spill}
			\ldots{} liszt & <[\type*{i\_spill\_rel}
							undergoer & \1]> ] \srdefault
									[\ldots{} liszt & <\type{\_spill\_rel}>],\\
			[\type*{i\_beans}
			\ldots{} liszt &	<[\type*{i\_beans\_rel}
							inst & \1]> ] \srdefault 
							[\ldots{} liszt & <\type{\_beans\_rel}>], \ldots{} 

		\}
]
}
\z 
}

\noindent
The \attrib{words} value of the idiomatic phrase contains at least two elements, the idiomatic words
of type \type{i\_spill} and \type{i\_beans}.  The special symbol {\scriptsize \srdefault} used in
this constraint expresses a default\is{default}. It says that the idiomatic version of the word
\emph{spill} is just like its non"=idiomatic homonym, except for the parts specified in the
left"=hand side of the default.  In this case, the type of the words and the type of the semantic
predicate contributed by the words are changed.  \citet{Riehemann2001a} only has to introduce the
types for the idiomatic words in the type hierarchy but need not specify type constraints on the
individual idiomatic words, as these are constrained by the default statement within the constraints
on the idioms containing them.

%\largerpage
As in the account of \citet{KE94a}, the syntactic flexibility of the idiom follows from its free
syntactic combination and the fact that all parts of the idiom are assigned an independent semantic
contribution. The lexical fixedness is a consequence of the requirement that particular words are
dominated by the phrase, namely the idiomatic versions of \emph{spill} and \emph{beans}.

The appeal of the account is particularly clear in its application to non"=de\-com\-posable,
semi-fixed idioms such as \emph{kick the bucket} \citep[\page 212]{Riehemann2001a}.  For such
expressions, the idiomatic words that constitute them are assumed to have an empty semantics and the
meaning of the idiom is contributed as a constructional semantic contribution only by the idiomatic
phrase.  Since the \attrib{words} list contains entire words, it is also possible to require that the
idiomatic word \emph{kick} be in active voice and/""or that it take a complement compatible with the
description of the idiomatic word \emph{bucket}.  This analysis captures the syntactically regular
internal structure of this type of idioms and is compatible with the occurrence of modifiers such as
\emph{proverbial}. At the same time, it prevents passivization and excludes extraction of the
complement as the \synsemv of the idiomatic word \emph{bucket} must be on the \compsl of the
idiomatic word \emph{kick}.\footnote{%
This assumes that extracted elements are not members of the valence lists. See
\crossrefchapterw[\pageref{udc:ex-slashed-verb-traceless}]{udc} for details.
}

Riehemann's approach clearly captures the intuition of idioms as phrasal units much better than any
other approach in HPSG.  However, it faces a number of problems.  First, the integration of the
approach with Constructional HPSG is done in such a way that the phrasal types for idioms are
cross-classified in complex type hierarchies with the various syntactic constructions in which the
idiom can appear.  This allows Riehemann to account for idiosyncratic differences in the syntactic
flexibility of idioms, but the question is whether such an explicit encoding misses generalizations
that should follow from independent properties of the components of an idiom and/""or of the
syntactic construction -- in line with the quote from \citet{NSW94a} on page \pageref{NSW-quote}.

Second, the mechanism of percolating dominated words to each phrase is not compatible with the
intuitions of most HPSG researchers.  Since no empirical motivation for such a mechanism aside from
idioms is provided in \citet{Riehemann2001a}, this idea has not been pursued in other papers.

Third, the question of how to block the free occurrence of idiomatic words, i.e., the occurrence of
an idiomatic word without the rest of the idiom, is not solved in \citet{Riehemann2001a}. While the
idiom requires the presence of particular idiomatic words, the occurrence of these words is not
restricted.\footnote{Since the problem of free occurrences of idiomatic words is not an issue for
  parsing, versions of Riehemann's approach have been integrated into practical parsing systems
  \citep{Villavicencio:Copestake:02}; see \crossrefchaptert{cl}.  Similarly, the approach to idioms
  sketched in \citet{Flickinger:15Slides2} is part of a system for parsing and machine
  translation\is{machine translation}. Idioms in the source language are identified by bits of
  semantic representation -- analogous to the elements in the \attrib{words} set. This approach,
  however, does not constitute a theoretical modeling of idioms; it does not exclude ill-formed uses
  of idioms but identifies potential occurrences of an idiom in the output of a parser.}  Note that
idiomatic words may sometimes be found without the other elements of the idiom -- evidenced by
expressions such as in \emph{bucket list} `list of things to do before one dies'. Such data may be
considered as support of Riehemann's approach; however, the extent to which we find such free
occurrences of idiomatic words is extremely small.\footnote{See the discussion around \pref{trice}
  for a parallel situation with bound words.}

Before closing this subsection, I would like to point out that \citet{Riehemann2001a} and
\citet{RB99a} are the only HPSG papers on idioms that address the question of statistical
idiomaticity\is{idiomaticity!statistical}, based on the variationist study in \citet{Bender2000a}.
In particular, \citet[\page 297--301]{Riehemann2001a} proposes phrasal constructions for
collocations even if these do not show any lexical, syntactic, semantic, or pragmatic idiosyncrasy
but just a statistical co-occurrence preference.  She extends this into a larger plea for an
\emph{experience-based HPSG}\is{experience-based HPSG}.
%
\citet{Bender2000a} discusses the same idea under the notions of \emph{minimal} versus
\emph{maximal} grammars, i.e., grammars that are as free of redundancy as possible to capture the
grammatical sentences of a language with their correct meaning versus grammars that might be open to
a connection with usage-based approaches\is{usage-based grammar} to language modeling.  \citet[\page
292]{Bender2000a} sketches a version of HPSG with frequencies/probabilities attached to lexical and
phrasal types.\footnote{An as-yet unexplored solution to the problem of free occurrence of idiomatic
  words within an experience-based version of HPSG could be to assign the type \type{idiomatic-word}
  an extremely low probability of occurring. This might have the effect that such a word can only be
  used if it is explicitly required in a construction. However, note that neither
  defaults\is{default} nor probabilities are well-defined part of the formal foundations of
  theoretical work on HPSG; see \crossrefchaptert{formal-background}.}\is{Constructional HPSG|)}

\subsection{Mixed lexical and phrasal approaches}
\label{Sec-Mixed}

While \citet{Riehemann2001a} proposes a parallel treatment of decomposable and non"=decomposable
idioms -- and of flexible and semi-fixed idioms -- the division between fixed and non"=fixed
expressions is at the core of another approach, the \emph{two-dimensional theory of
  idioms}\is{two-dimensional theory of idioms}. This approach was first outlined in
\citet{Sailer2000a} and referred to under this label in
\citet{Richter:Sailer:09,Richter:Sailer:14}. It is intended to combine constructional and
collocational approaches to grammar.

The basic intuition behind this approach is that signs have internal and external properties.  All
properties that are part of the feature structure of a sign are called \emph{internal}.  Properties
that relate to larger feature structures containing this sign are called its \emph{external}
properties.  The approach assumes that there is a notion of \emph{regularity} and that anything
diverging from it is \emph{idiosyncratic} -- or idiomatic, in the terminology of this chapter.

This approach is another attempt to reify the GPSG\indexgpsg analysis
within HPSG.  \citet{Sailer2000a} follows the distinction of \citet{NSW94a} into non"=decomposable
and non"=flexible idioms on the one hand and decomposable and flexible idioms on the other. The
first group is considered internally irregular and receives a constructional analysis in terms of a
\emph{phrasal lexical entry}\is{phrasal lexical entry}. The second group is considered to consist of
independent, smaller lexical units that show an external irregularity in being constrained to
co-occur within a larger structure.  Idioms of the second group receive a collocational
analysis. The two types of irregularity are connected by the \emph{Predictability Hypothesis}, given
in \refer{PredHypo}.

\vbox{
\eanoraggedright 
\label{PredHypo}
Predictability Hypothesis\is{Predictability Hypothesis} \citep[\page 366]{Sailer2000a}:\\
For every sign whose internal properties are fully predictable, the distributional
behavior of this sign is fully predictable as well.
\z 
}

\noindent
In the most recent version of this approach, \citet{Richter:Sailer:09,Richter:Sailer:14}, there is a
feature \attrib{coll} defined on all signs. The value of this feature specifies the type of internal
irregularity.  The authors assume a cross-classification of regularity and irregularity with respect
to syntax, semantics, and phonology -- ignoring pragmatic and statistical (ir)regularity in their
paper.  Every basic lexical entry is defined as completely irregular, as its properties are not
predictable.  Fully regular phrases such as \emph{read a book} have a trivial value of \attrib{coll}.
A syntactically internally regular but fixed idiom such as \emph{kick the bucket} is classified as
having only semantic irregularity, whereas a syntactically irregular expression such as \emph{trip
  the light fantastic} is of an irregularity type that is a subsort of syntactic and semantic
irregularity, but not of phonological irregularity.  Following the terminology of \citet{FKoC88a},
this type is called \type{extra-grammatical-idiom}.
%
The phrasal lexical entry for \emph{trip the light fantastic} is sketched in \refer{rs-trip},
adjusted to the feature geometry of \citet{Sag97a}.
The feature \attrib{listeme} is used to identify individual lexemes.

\vbox{
\ea Phrasal lexical entry for the idiom \emph{trip the light fantastic}:\label{rs-trip}\\
% todo avm
\avm{
[\type*{headed-phrase}
phon \1 \+ <\type{the, light, fantastic}> \\
synsem|loc [ cat & [ head & \2 \\
	             listeme & trip-the-light-fantastic \\
                     subj & < \3 [\ldots{}{} index & \4] >  \\
                     comps & < > ] \\
             cont & [\type*{trip-light-fant}
                     dancer & \4] ] \\
head-dtr  [phon \1 \\
           synsem|loc [ cat & [ head & \2 verb \\
                                 listeme & trip \\
                                 subj  & < \3 > \\
                                 comps & < > ] ] ] \\
coll  \type{extra-grammatical-idiom}~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ ]
}
\z 

}

In \refer{rs-trip}, the constituent structure of the phrase is not specified, but the phonology is
fixed, with the exception of the head daughter's phonological contribution. This accounts for the
syntactic irregularity of the idiom. The semantics of the idiom is not related to the semantic
contributions of its components, which accounts for the semantic idiomaticity.

\citet{Soehn2006a} applies this theory to \ili{German}\il{German}. He solves the problem of the
relatively large degree of flexibility of non"=decomposable idioms in \ili{German} by using
underspecified descriptions of the constituent structure dominated by the idiomatic phrase.

%\largerpage
For decomposable idioms, the two-dimensional theory assumes a collocational component. This
component is integrated into the value of an attribute \attrib{req}, which is only defined on
\type{coll} objects of one of the irregularity types. This encodes the Predictability Hypothesis.
%
The most comprehensive version of this collocational theory is given in \citet{Soehn:09},
summarizing and extending ideas from \citet{Soehn2006a} and \citet{richter-soehn:2006}.  Soehn
assumes that collocational requirements can be of various types: a lexical item can be constrained
to co-occur with particular \emph{licensers} (or collocates). These can be other lexemes, semantic
operators, or phonological units. In addition, the domain within which this licensing has to be
satisfied is specified in terms of syntactic barriers, i.e., syntactic nodes dominating the
externally irregular item.

To give an example, the idiom \emph{spill the beans} would be analyzed as consisting of two
idiomatic words \emph{spill} and \emph{beans} with special \attrib{listeme} values \type{spill-i} and
\mbox{\type{beans-i}}. The idiomatic verb \emph{spill} imposes a lexeme selection on its
complement. The idiomatic noun \emph{beans} has a non"=empty \attrib{req} value, which specifies that
it must be selected by a word with \attrib{listeme} value \type{spill-i} within the smallest complete
clause dominating it.

%\citet{Richter:Sailer:09,Richter:Sailer:14} look at idioms with 

%\bigskip
The two-dimensional approach suffers from a number of weaknesses.  First, it presupposes a notion of
regularity. This assumption is not shared by all linguists.  Second, the criteria for whether an
expression should be treated constructionally or collocationally are not always clear.  Idioms with
irregular syntactic structure need to be analyzed constructionally, but this is less clear for
non"=decomposable idioms with regular syntactic structure such as \emph{kick the bucket}.


%\begin{itemize}
%\item \citet{Sailer2000a}, \citet{Soehn2006a}, \citet{Richter:Sailer:09}
%\end{itemize}

\subsection{Recent lexical approaches}
\label{Sec-RecentLexical}\label{idioms:recent-lexical-approaches}

%\largerpage
%\enlargethispage{3pt}
\citet{KSF2015a} marks an important re-orientation in the analysis of idioms: the lexical analysis
is extended to all syntactically regular idioms, i.e., to both decomposable (\emph{spill the beans})
and non"=decomposable idioms (\emph{kick the bucket}).\footnote{This idea has been previously
  expressed within a Minimalist perspective by \citew{Everaert:10} and G.\ \citew[\page
  213--214]{GMueller2011a}.  }
%\citet{KSF2015a} use Sign-based Construction Grammar\is{Sign-based Construction Grammar}
%
\citet{KSF2015a} achieve a lexical analysis of non"=decomposable idioms by two means: (i), an
extension of the HPSG selection mechanism, and (ii), the assumption of semantically empty idiomatic
words.  

As in previous accounts, the relation among idiom parts is established through lexeme-specific
selection, using a feature \attrib{lid} (for \emph{lexical identifier}).  The authors assume that
there is a difference between idiomatic and non"=idiomatic \attrib{lid} values.  Only heads that are
part of idioms themselves can select for idiomatic words.%
\footnote{I mentioned three attributes which are used to encode lexeme-specific selection: \attrib{lexeme}\isfeat{lexeme}  (defined on \type{index}), \attrib{listeme}\isfeat{listeme} (defined on \type{category}), and \attrib{lid}\isfeat{lid} (defined on \type{head}). Each of them comes with a different motivation and different limitations.}
%Quote: Ordinary, non"=idiom predicators are lexically specified as requiring all members of their VAL list to be nonidiomatic.

For the idiom \emph{kick the bucket}, \citet{KSF2015a} assume that all meaning is carried by the
lexical head, an idiomatic version of \emph{kick}, whereas the other two words, \emph{the} and
\emph{bucket} are meaningless.  This meaninglessness allows \citeauthor{KSF2015a} to block the idiom from
occurring in constructions which require meaningful constituents, such as questions,
\emph{it}-clefts, middle voice, and others.  To exclude passivization, the authors assume that the
\ili{English} passive cannot apply to verbs selecting a semantically empty direct object.

The approach in \citet{KSF2015a} is a recent attempt to maintain the SLH as much as possible.  Since
the SLH has been a major conceptual motivation for Sign-Based Construction Grammar\indexsbcg, \citeauthor{KSF2015a}'s paper is an important contribution showing the
empirical robustness of this assumption.

%\medskip%
\citet{Bargmann:Sailer:18} propose a similar lexical approach to non"=de\-com\-pos\-able idioms.
They take as their starting point the syntactic flexibility of semantically non"=decomposable idioms
in \ili{English} and, in particular, in \ili{German}.  There are two main differences between
\citeauthor{KSF2015a}'s paper and \citeauthor{Bargmann:Sailer:18}'s: (i),
\citeauthor{Bargmann:Sailer:18} assume a collocational rather than a purely selectional mechanism to
capture lexeme restrictions of idioms, and (ii), they propose a redundant semantics rather than an
empty semantics for idiom parts in non"=decomposable idioms. In other words,
\citet{Bargmann:Sailer:18} propose that both \emph{kick} and \emph{bucket} contribute the semantics
of the idiom \emph{kick the bucket}.  \citeauthor{Bargmann:Sailer:18} argue that the semantic
contributions of parts of non"=decomposable, syntactically regular idioms are the same across
languages, whereas the differences in syntactic flexibility are related to the different syntactic,
semantic, and pragmatic constraints imposed on various constructions.  To give just one example,
while there are barely any restrictions on passive subjects in \ili{German}, there are strong
discourse-structural constraints on passive subjects in \ili{English}.

%\largerpage[1.3]
Both \citet{KSF2015a} and \citet{Bargmann:Sailer:18} attempt to derive the (partial) syntactic
inflexibility of non"=decomposable idioms from independent properties of the relevant constructions.
As such, they subscribe to the programmatic statement of \citet{NSW94a} quoted on page
\pageref{NSW-quote}.  In this respect, the extension of the lexical approach from decomposable
idioms to all syntactically regular expressions has been a clear step forward.

%\bigskip%
\citet{Findlay:17} provides a recent discussion and criticism of lexical approaches to idioms in
general, which applies in particular to non"=decomposable expressions.  His reservations comprise
the following points.  First, there is a massive proliferation of lexical entries for otherwise
homophonous words.
%Is is unclear, for example, if a separate definite article is needed for each idiom which contains one, i.e., it might turn out that we need different lexical entries for the word \emph{the} in \emph{kick the bucket}, \emph{shoot the breeze}, and \emph{shit hits the fan}.%
%
Second, the lexical analysis does not represent idioms as units, which might make it difficult to
connect their theoretical treatment with processing evidence.  Findlay refers to psycholinguistic
studies, such as \citet{Swinney:Cutler:79}, that point to a faster processing of idioms than of free
combinations.  While the relevance of processing arguments for an HPSG analysis are not clear, I
share the basic intuition that idioms, decomposable or not, are a unit and that this should be part
of their linguistic representation.


\section{Where to go from here?}
\label{Sec-WhereToGo}

The final section of this article contains short overviews of research that has been done in areas
of phraseology that are outside the main thread of this chapter. I will also identify desiderata.


%In this final section, I would like to point to directions for future research. 
\subsection{Neglected phenomena}
\label{Sec-Neglected}

Not all types of idioms or idiomaticity mentioned in Section~\ref{Sec-EmpiricalDomain} have received
an adequate treatment in the (HPSG) literature.  I will briefly look at three empirical areas that
deserve more attention: neglected types of idiom variation, phraseological patterns, and the literal
and non"=literal meaning components of idioms.

%\bigskip%Variation
Most studies on idiom variation have looked at verb- and sentence-related syntactic constructions,
such as passive and topicalization.  However, not much attention has been paid to lexical variation
in idioms. This variation is illustrated by the following examples from \citet[\page 184,
191]{Richards:01}.

\eal  \label{creeps}
\ex The Count gives everyone the creeps.
\ex You get the creeps (just looking at him).
\ex I have the creeps.
\zl 

\noindent
In~\refer{creeps}, the alternation of the verb seems to be very systematic -- and has been used by
\citet{Richards:01} to motivate a lexical decomposition of the involved verbs.  A similar argument has been
made in \citet{Mateu:Espinal:07} for similar idioms in \ili{Catalan}.  We are lacking systematic,
larger empirical studies of this type of substitution, and it would be important to see how it can
be modeled in HPSG.  One option would be to capture the \emph{give}--\emph{get}--\emph{have}
alternation(s) with lexical rules. Such lexical rules would be different from the standard cases,
however, as they would change the lexeme itself rather than just alternating its morpho-syntactic
properties or its semantic contribution.

In the case mentioned in footnote~\ref{fn-semmeln}, the alternation consists of substituting a word
with a (near) synonym and keeping the meaning of the idiom intact. Again, HPSG seems to have all the
required tools to model this phenomenon -- for example, by means of hierarchies of
\textsc{lid} values.  However, the extent of this phenomenon across the set of idioms is not
known empirically.

%\medskip%
Concerning syntactic variation, the nominal domain has not yet received the attention it might
deserve.  There is a well-known variation with respect to the marking of possession within idioms.
This has been documented for \ili{English}\il{English} in \citet{Ho:15}, for Modern
\ili{Hebrew}\il{Hebrew} in \citet{Almog:12}, and for Modern \ili{Greek} and \ili{German}\il{German}
in \citet{Markantonatou:Sailer:16}.  In \ili{German}, we find a relatively free alternation between
a plain definite and a possessive; see \refer{ex-verstand}. This is, however, not possible with all
idioms; see \refer{ex-frieden}.

\eal \label{ex-verstand-herz}
\ex 
\gll Alex hat den / seinen Verstand verloren.\\
Alex has the {} his mind lost\\\hfill(\ili{German})
\glt `Alex lost his mind.'\label{ex-verstand}
\ex 
\gll Alex hat *den / ihren Frieden mit der Situation gemacht.\\
     Alex has \hphantom{*}the {} her peace with the situation made\\
\glt `Alex made her peace with the situation.'\label{ex-frieden}
\zl 

\noindent
We can also find a free dative in some cases, expressing the possessor.  In \refer{ex-herz}, a
dative possessor may co-occur with a plain definite or a coreferential possessive determiner; in
\refer{ex-augen}, only the definite article but not the possessive determiner is possible.


\eal
\label{ex-herz-augen}
\ex 
\gll Alex hat mir das / mein Herz gebrochen.\\
Alex has me.\textsc{dat} the {} my heart broken\\\hfill(\ili{German})
\glt `Alex broke my heart.'\label{ex-herz}
\ex 
\gll Alex sollte mir lieber aus den / *meinen Augen gehen.\\
Alex should me.\textsc{dat} rather {out of} the {} \hphantom{*}my eyes go\\
\glt `Alex should rather disappear from my sight.'\label{ex-augen}
\zl 

\noindent
While they do not offer a formal encoding, \citet{Markantonatou:Sailer:16} observe that a particular
encoding of possession in idioms is only possible if it would also be possible in a free
combination. However, an idiom may be idiosyncratically restricted to a subset of the realizations
that would be possible in a corresponding free combination. A formalization in HPSG might consist of
a treatment of possessively used definite determiners, combined with an analysis of free datives as
an extension of a verb's argument structure.%
\footnote{See \citew{Koenig1999b} for an analysis of possessively interpreted definites and
  \citew[68]{MuellerLFGphrasal} for an extension of the argument structure as suggested in the main
  text.
}

Related to the question of lexical variation are \emph{phraseological
  patterns}\is{phraseological pattern}, i.e., very schematic idioms in which the lexical material is largely free. Some
examples of phraseological patterns are the \emph{Incredulity Response Construction} as in
\emph{What, me worry?} \citep{Akmajian:84,Lambrecht:90}, or the \emph{What's X doing Y?}
construction \citep{KF99a}.  Such patterns are of theoretical importance as they typically involve a
non"=canonical syntactic pattern.  The different locality and non"=locality hypotheses introduced
above make different predictions.  \citet{FKoC88a} have presented such constructions as a motivation
for the non"=locality of constructions, i.e., as support of a SNH. However, \citet{KF99a} show that
a lexical analysis might be possible for some cases at least, which they illustrate with the
\emph{What's X doing Y?} construction. The N-\textit{after}-N construction seemed to be an example
of a phraseological pattern that can very naturally modelled in accordance with the SLH, see
\citet{Poss:10} and Chapter~6 of \citet{Bargmann:23} for an SBCG and an HPSG account
respectively. 

\citet{Borsley2004a} looks at another phraseological pattern, the \emph{the X-er the Y-er}
construction, or \emph{Comparative Correlative Construction}\is{comparative correlative} -- see
\crossrefchaptert[Section~\ref{coord:sec-comparative-correlatives}]{coordination} and
\crossrefchapterw[\pageref{udc:page-correlatives-start}--\pageref{udc:page-correlatives-end}]{udc}.
Borsley analyzes this construction by means of two special (local) phrase structure types: one for
the comparative \emph{the}-clauses, and one for the overall construction. He shows that (i), the
idiosyncrasy of the construction concerns two levels of embedding and is, therefore, non"=local;
however, (ii), a local analysis is still possible. This approach raises the question of whether the
WNH is empirically vacuous since we can always encode a non"=local construction in terms of a series
of idiosyncratic local constructions.  Clearly, work on more phraseological patterns is needed to
assess the various analytical options and their consequences for the architecture of grammar.

A major charge for the conceptual and semantic analysis of idioms is the interaction between the
literal and the idiomatic meaning.  I presented the basic empirical facts in
Section~\ref{Sec-EmpiricalDomain}.  All HPSG approaches to idioms so far basically ignore the
literal meaning.  This position might be justified, as an HPSG grammar should just model the
structure and meaning of an utterance and need not worry about the meta-linguistic relations among
different lexical items or among different readings of the same (or a homophonous) expression.
Nonetheless, this issue touches on an important conceptual point.  Addressing it might immediately
provide possibilities to connect HPSG research to other disciplines and/""or frameworks like
Cognitive Linguistics, such as in \citet{Dobrovolskij:Piirainen:05}, and
psycholinguistics.\footnote{%
\citet{Bargmann:al:21} provide an in-depth discussion of examples that show the complexity of the
interaction of literal and idiomatic meaning.
}



\subsection{Challenges from other languages}
\label{Sec-OtherLanguages}

The majority of work on idioms in HPSG has been done on \ili{English} and \ili{German}. 
%This led to a limitation of the possible phenomena that can be studied on idioms. 
As discussed in Section~\ref{Sec-RecentLexical}, the recent trend in HPSG idiom research necessitates a detailed study of individual syntactic structures. 
Consequently, the restriction to two closely related languages limits the possible phenomena that can be studied concerning idioms. 
It would be essential to expand the empirical coverage of idiom analyses in HPSG to as many different languages as possible. 
The larger degree of syntactic flexibility of \ili{French}, \ili{German}, and \ili{Dutch} idioms \citep{Ruwet:91,NSW94a,Schenk:95} has led to important refinements of the analysis in \citet{NSW94a} and, ultimately, to the lexical analyses of all syntactically regular idioms. 

Similarly, the above-mentioned data on possessive alternations only become prominent when languages beyond \ili{English} are taken into account. Modern \ili{Greek}\il{Greek}, \ili{German}\il{German}, and many others
%mentioned above \refer{ex-herz-augen} 
show the type of external possessor classified as a European areal phenomenon in \citet{Haspelmath:99}. 
It would be important to look at idioms in languages with other types of external possessors.


In a recent paper, \citet{Sheinfux:al:19} provide data from Modern \ili{Hebrew}\il{Hebrew} that show that opacity and figurativity of an idiom are decisive for its syntactic flexibility, rather than decomposability.
This result stresses the importance of the literal reading for an adequate account of the syntactic behavior of idioms. 
%
It shows that the inclusion of other languages can cause a shift of focus to other types of idioms or other types of idiomaticity. 

To add just one more example, HPSG(-related) work on \ili{Persian} such as \citet{MuellerPersian}
and \citet{Samvelian:Faghiri:16} establishes a clear connection between complex predicates and
idioms. Their insights might also lead to a reconsideration of the similarities between light verbs
and idioms, as already set out in \citet{KE94a}. 


As far as I can see, the following empirical phenomena have not been addressed in HPSG approaches to
idioms, as they do not occur in the main object languages for which we have idiom analyses, i.e.,
\ili{English} and \ili{German}. They are, however, common in other languages: the occurrence of
clitics\is{clitic} in idioms (found in \ili{Romance}\il{Romance} and \ili{Greek}\il{Greek});
aspectual\is{aspect} alternations in verbs (\ili{Slavic}\il{Slavic} and \ili{Greek}); argument
alternations other than passive and dative alternation, such as anti-passive\is{anti-passive},
causative\is{causative}, inchoative\is{inchoative}, etc.\ (in part found in \ili{Hebrew} and
addressed in \citealt{Sheinfux:al:19}); and displacement of idiom parts into special syntactic
positions (focus position in \ili{Hungarian}).

Finally, so far, idioms have usually been considered as either offering irregular structures or as
being more restricted in their structures than free combinations. In some languages, however, we
find archaic syntactic structures and function words in idioms that do not easily fit these two
analytic options. To name just a few, \citet{Lodrup:09} argues that \ili{Norwegian}\il{Norwegian}
used to have an external possessor construction similar to that of other European languages, which
is only conserved in some idioms. Similarly, \ili{Dutch}\il{Dutch} has a number of archaic case
inflections in multiword expressions \citep[\page 129]{Kuiper:18}, and there are archaic forms in
Modern \ili{Greek}\il{Greek} multiword expressions. It is far from clear what the best way would be
to integrate such cases into an HPSG grammar.


\section{Conclusion}
\label{Sec-Summary}

Idioms are among the topics in linguistics for which HPSG-related publications have had a clear
impact on the field and have been widely quoted across frameworks.  This handbook article aimed at
providing an overview over the development of idiom analyses in HPSG.  There seems to be a
development towards ever more lexical analyses, starting from the holistic approach for all idioms
in Chomsky's work, to a lexical account for all syntactically regular expressions.  Notwithstanding
the advantages of the lexical analyses, I consider it a basic problem of such approaches that the
unit status of idioms is lost. Consequently, I think that the right balance between phrasal and
lexical aspects in the analysis of idioms has not yet been fully achieved.\footnote{%
A recent, the  attempt of combining a lexical and a phrasal approach to idioms is
\citet{Sailer:Bargmann:21}.
}

The sign-based character of HPSG seems to be particularly suited for a theory of idioms as it allows
one to take into consideration syntactic, semantic, and pragmatic aspects and to use them to
constrain the occurrence of idioms appropriately.

% moved abbrevs and ackno before appendix to fill the page
\section*{Abbreviations}

\begin{tabular}{@{}ll}
GPSG & Generalized Phrase Structure Grammar \citep{GKPS85a}\\
%MRS & Minimal Recursion Semantics \citep{CFPS2005a}\\
%MWE & multiword expression\\
MGG & Mainstream Generative Grammar\\
SLH & Strong Locality Hypothesis\\%, see page \pageref{slh}\\
SNH & Strong Non"=locality Hypothesis\\%, see page \pageref{snh}\\
WLH & Weak Locality Hypothesis\\%, see page \pageref{wlh}\\
WNH & Weak Non"=locality Hypothesis\\%, see page \pageref{wnh}\\
\end{tabular}

\section*{\acknowledgmentsUS}

I have perceived Ivan A.\@ Sag\ia{Sag, Ivan A.} and his work with various colleagues as a major inspiration for my own work on idioms and multiword expressions. 
This is clearly reflected in the structure of this paper, too. 
I apologize for this bias, but I think it is legitimate within an HPSG handbook.
%
I am grateful to Jean-Pierre Koenig, Stefan Müller and Tom Wasow for comments on the outline and the first version of this chapter. 
%
I would not have been able to format this chapter without the support of the Language Science Press team, in particular Sebastian Nordhoff.
%
I would like to thank Elizabeth Pankratz for comments and proofreading.


%\appendix
\section*{Appendix: List of used idioms}

\subsection*{English}

\begin{tabular}{@{}lL{3.8cm}L{3.8cm}@{}}
idiom & paraphrase & comment\\\hline
break the ice      & relieve tension in a strained situation  & figurative\\
brush one's teeth  & clean one's teeth with a tooth brush     & collocation, no idiomaticity\\
give s.o.\ the creeps & make s.o.\ feel uncomfortable               & systematic lexical variation\\
Good morning!      & (morning greeting)                       & formulaic expression\\
immaculate performance & perfect performance & statistical idiomaticity\\
in a trice & in a moment & bound word: \emph{trice}\\
kick the bucket & die & non"=decomposable\\
make headway & make progress & bound word: \emph{headway}\\
pull strings & exert influence/""use one's connections & flexible\\
saw logs & snore & {transparent, non"=decomposable, semi"=flexible}\\
shake hands & greet & body-part expression\\
shake one's head & decline/""negate & {body-part expression, possessive idiom}\\
shit hit the fan & there is trouble & {subject as idiom component, transparent/""figurative, non"=decomposable}\\
shoot the breeze & chat & non"=decomposable\\
spill the beans & reveal a secret & flexible\\
take a shower & clean oneself using a shower & collocation, light verb construction\\
take the bull by the horns & {approach a problem directly} & figurative expression\\
take umbrage & take offense & bound word: \emph{umbrage}\\
trip the light fantastic & dance & syntactically irregular\\
\end{tabular}


\subsection*{German}

\begin{sideways}
\begin{tabular}{@{}L{4cm}lL{4.0cm}L{4.0cm}}
idiom & gloss & translation & comment\\\hline
%
 {den/seinen Verstand verlieren}
 & {the/""one's mind lose}
 & {lose one's mind}
 & {alternation of possessor marking}
 \\
 %
 {jdm.\ das Herz brechen} & s.o.\ the heart break & break s.o.'s heart
 & {dative possessor and possessor alternation}\\
 %
 {jdm.\ aus den Augen gehen} & s.o.\ out of the eyes go
 & {disappear from s.o.'s sight} &
 {dative possessor, restricted possessor alternation}\\
 %
 {seinen Frieden machen mit}
 & {one's peace make with}
 & {make one's peace with}
 & {no possessor alternation possible}\\
 %
 {wie warme Semmeln/""Brötchen/""Schrippen weggehen}
 & {like warm rolls vanish} & sell like hotcakes & 
 {parts can be exchanged by synonyms}\\
\end{tabular}
\end{sideways}
\is{idiom|)}


{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this] }
\end{document}


%      <!-- Local IspellDict: en_US-w_accents -->
