\documentclass[output=paper,biblatex,babelshorthands,newtxmath,draftmode,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.13644997}

\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
  \usepackage{../nomemoize}
  \input{../localpackages}
  \input{../localcommands}
  \input{../locallangscifixes.tex}

  \togglepaper[16]
}{}

\author{Rui P. Chaves\orcid{0000-0003-0820-6145}\affiliation{University at Buffalo, SUNY}}
\title{Island phenomena and related matters}


\abstract{Extraction constraints on long-distance dependencies -- so-called \emph{islands} -- have been the subject of intense linguistic and psycholinguistic research for the last half century. Despite of their importance in syntactic theory, the heterogeneity of island constraints has posed many difficult challenges to linguistic theory, across all frameworks. The HPSG perspective of island phenomena is that they are unlikely to be due to a unitary syntactic constraint given the fact that virtually all such island constraints have known exceptions.  Rather, it is more plausible that island constraints result from a combination of independently motivated syntactic, semantic, pragmatic and processing phenomena.  The present chapter is somewhat different from others in this volume in that its focus is not on HPSG analyses of some phenomena, but rather on the nature of the phenomena itself.  This is because there is evidence that most of the phenomena are not purely grammatical, and to that extent independent from HPSG or indeed any theory of grammar. One may call this view of island phenomena ``minimalist'' in the sense that much of it does not involve formal grammar.}


\begin{document}
\maketitle
\label{chap-islands}



\section{Introduction} 

This chapter provides an overview of various island effects that have received attention from
members of the HPSG community. I begin with the extraction constraints peculiar to coordinate
structures, because they not only have a special status in the history of HPSG, but also because
they illustrate well the non-unitary nature of island constraints. I then argue that, at a deeper
level, some of these constraints are in fact present in many other island types, though not
necessarily all.  For example, I take it as relatively clear that \emph{factive islands} are purely
pragmatic in nature \citep{Oshima:2006:FIP:1761528.1761544}, as are \emph{negative islands}
\citep{Kroch98a,sza,abrusan,foxneg,abrusanspec}, although one can quibble about the particular
technical details of how such accounts are best articulated.  Similarly, the \emph{NP Constraint} in
the sense of \citet{horn72} is likely to be semantic-pragmatic in nature
\citep{kuno87,godard92f,dubinsky2009,SS2013b-u,chavesking2020}.  Conversely, I take it as relatively
uncontroversial that the \emph{Clause Non-Final Incomplete Constituent Constraint} is due to
processing difficulty \citep{lev91,fodor92}. See also \citet{kothari}, \citet{ambridge:2008}, and
\citet{richterchaves20} for evidence that ``bridge'' effects in filler-gap dependencies at least in
part due to pragmatics.

In the present chapter I focus on islands that have garnered more attention from members of the HPSG
community, and that have caused more controversy cross-theoretically. My goal is to provide an
overview of the range of explanations that have been proposed to account for the complex array of
facts surrounding islands, and to show that no single unified account is likely.
For a more comprehensive overview of islands and related phenomena see \citet[Chapter~3]{CP2020a-u}.



\section{Background}

As already detailed in \crossrefchaptert{udc}, HPSG encodes filler-gap dependencies in terms of a
set-valued feature \textsc{slash}. Because the theory consists of a feature-based declarative system of
constraints, virtually all that goes on in the grammar involves constraints stating which value a
given feature takes. By allowing \textsc{slash} sets to be identified (or unioned), it
follows that constructions in which multiple gaps are linked to the same filler are trivially
obtained, as in (\ref{cc5}).

\eal
\label{cc5}
\ex Which celebrity did  [the article insult \spc more than it praised \spc]?
\ex Which celebrity  did you expect [[the pictures of \spc ] to bother \spc the most]?
\ex Which celebrity did you [inform \spcs [that the police was coming to arrest \spc]]?
%\item What do you think the best job [I ever got fired from \spc] was \spc?
\ex Which celebrity did you [compare [the memoir of \spc] [with a movie about \spc]?
\ex Which celebrity  did you [hire \spcs [without auditioning \spcs first]]?
\ex Which celebrity  did you [[meet \spcs at a party] and [date \spcs for a few months]]?
\zl 

\noindent
But another advantage of encoding the presence of filler-gap dependencies as a feature is that
certain lexical items and constructions can easily impose idiosyncratic constraints on \textsc{slash}
values. For example, to account for  languages that do not allow preposition stranding,
it suffices to state that prepositions are necessarily specified as [\textsc{slash} \{ \}]. Thus,
their complements cannot appear in \textsc{slash} instead of \textsc{comps}. 
The converse also occurs. Certain uses of the verb \emph{assure}, for example, 
are lexically required to have one complement in \textsc{slash} rather than in \textsc{comps}. Thus,
extraction is obligatory as (\ref{suggest}) shows, based on \citew[\page 4]{Kayne84a-u}.
   
\eal \label{suggest}   
\ex[*]{I can assure you him to be the most competent.}
\ex[]{Who$_i$ can you assure me \trace$_i$ to be the most competent?
}


% @book{kayne84,
%  author={Richard Kayne},
% title={Connectedness and Binary Branching},
%    year={1984},
%  publisher={Dordrecht: Foris}}

% https://as.nyu.edu/content/dam/nyu-as/linguistics/documents/richie/Kayne%200217%20Connectedness%20and%20Binary%20Branching.pdf

\zl

\noindent
As we shall see, it would be rather trivial to impose the classic island constraints in the standard syntactic
environments in which they arise.\footnote{In such a view, island effects could perhaps result from
grammaticized constraints, induced by parsing and performance considerations  
\citep{prichett,fodor78,Fodor83}.}
 The problem is that island effects are riddled with exceptions which
defy purely syntactic accounts of the phenomena.  Hence, HPSG has generally refrained from
assuming that islands are syntactic, in contrast to Mainstream Generative Grammar.


\section{The Coordinate Structure Constraint}

\citet{Ross67} first observed that coordinate structures impose various constraints on long-distance
dependencies, shown in (\ref{bors}), collectively dubbed the \emph{Coordinate Structure
  Constraint}. For perspicuity, I follow \citet{grosu73} in referring to (i) as the \emph{Conjunct
  Constraint} and to (ii) as the \emph{Element Constraint}.

\eanoraggedright \label{bors}
Coordinate Structure Constraint (CSC):

In a coordinate structure, (i) no conjunct may be moved, (ii) nor may any element contained in a
conjunct be moved out of that conjunct \ldots \,unless each conjunct properly contains a gap paired
with the same filler.  
\z

\largerpage
\noindent
The \emph{Conjunct Constraint} (CC) is illustrated by the unacceptability of the extractions in
(\ref{islands-cc}). No such constraint is active in other constructions like those in (\ref{com})
and (\ref{compar0}), for example.
 

\eal
\label{islands-cc}  
\ex[*]{Which celebrity did you see [Priscilla and \spc]?\\
(cf.\ with `Did you see Priscilla and Elvis?')}
\ex[*]{Which celebrity did you see  [ \spc and Priscilla]?\\
(cf.\ with `Did you see Elvis and Priscilla?')}
\ex[*]{Which celebrity did you see  [ \spc or/and a picture of \spc]?\\
(cf.\ with `Did you see Elvis or/and a picture of Elvis?')}
%\item \bad{*}Who did you compare  [[\spc] and [\spc]]?
\zl 

\eal
\label{com}
\ex[]{Which celebrity did you see Priscilla with \spc?\\
(cf.\ with `Did you see Priscilla with Elvis?')}
\ex[]{Which celebrity did you see \spc with Elvis?\\
(cf.\ with `Did you see Priscilla with Elvis?')}
%\item Which celebrity did you prefer \spc to a picture of \spc?
%\item Who did you compare  [[\spc] with [\spc]]?
\zl
% the memoir of with a movie about ?


\eal
\label{compar0}
\ex[]{Which celebrity  is Kim as tall as \spc?}
\ex[]{Which celebrity did you say Robin arrived earlier than \spc?}
\zl


\noindent
In HPSG accounts of extraction that assume the existence of traces \citep{pollardsag,levhubook} the
CC must be stipulated at the level of the coordination construction, by stating that conjuncts
cannot be empty elements.\footnote{See however \citet[317--318]{levine17} for the claim that each
  conjunct must contain at least one stressed syllable. Given that traces are phonologically silent,
  nothing is there to bear stress and the CC is obtained. This raises the question of why no such
  stress constraint exists in P-stranding, for example, or indeed in any kind of extraction.}  On
the other hand, the CC follows immediately in a traceless account of filler-gap dependencies
\citep{fodorsagt,bouma,ginzsag,fgsag08} since there is simply nothing to conjoin in
(\ref{islands-cc}), and thus nothing else needs to be said about conjunct extraction; see
\citet{sagonline} for more criticism of traces.

\largerpage
HPSG's\label{page-hpsg-traceless-account-arg-st-extraction-conjuncts} traceless account of the CC is
semantic in nature, in a sense.  Coordinators like \emph{and}, \emph{or}, \emph{but} and so on are
not regarded as heads that select arguments, and therefore have empty \textsc{arg-st} and valence
specifications. And given that HPSG assumes that the signs that can appear in a given lexical head
\textsc{slash} values are valents, then it follows that the signs that coordinators combine with cannot
instead be registered in the coordinator's \textsc{slash} feature. Hence, words like \emph{and} have no
valents, no arguments and therefore no conjunct extraction. Incidentally, adnominal adjectives
cannot be extracted either, for exactly the same reason, as they are not selected by any head, and
therefore are not listed in any \textsc{arg-st} list.

In order to allow certain adverbials to be extractable, \citet{ginzsag} assume that those adverbials
are members of \textsc{arg-st}.  See \citet{levhubook} for more on adverbial extraction, and see
\crossrefchaptert{udc} for further discussion.\footnote{The empirical facts are less clear when it
  comes to adnominal PPs, however. Even PPs that are usually regarded as modifiers can sometimes be
  extracted, as in \emph{From which shelf am I not supposed to read any books?} In many such
  extractions the PP can alternatively be parsed as VP modifier, which complicates
  judgements. See also \citew[176]{deKuthy2002a} and \crossrefchapterw[\pageref{inf-struc:page-np-pp-split}]{information-structure}
  on NP-PP split.}

 
Let us now turn to the \emph{Element Constraint}, illustrated in (\ref{cc2}). As before, the
constraint appears to be restricted to coordination structures, as no oddness arises in the
comitative counterparts like (\ref{cc3}), or in comparatives like (\ref{compar}).\footnote{Although
  \citet[83]{winter01} and others claim that coordination imposes semantic scope islands,
  \citet[\S3.6]{chavesthesis} shows that this is not the case, as illustrated in examples like those
  below.

\eal
\ex The White House is very careful about this. An official representative [[will
              personally read each document] and [reply to every letter]].\\
              ($\forall$ doc-letter $>$ $\exists$ representative / $\exists$ representative $>$ $\forall$ doc-letter)

\ex We had to do this ourselves. By the end of the year,
some student [[had proof-read every document] and [corrected each
theorem]].\\
($\forall$  doc-theorem $>$ $\exists$ student / $\exists$ student $>$ $\forall$ doc-theorem)

\ex Your task is  to document the social interaction
between [[each female] and [an adult male]].\\
($\forall$ female $>$ $\exists$ adult male / $\exists$ adult male $>$ $\forall$ female)
\zllast
}


 

\eal
\label{cc2}
\ex[*]{Which celebrity did you see [Priscilla and a picture of \spc]?\\
(cf.\ with `Did you see Priscilla and a picture of Elvis?')}
\ex[*]{Which celebrity did you see [a picture of \spc and Priscilla]?\\
(cf.\ with `Did you see a picture of Elvis and Priscilla?')}
\zl

\eal 
\label{cc3} 
\ex Which celebrity did you see [the brother of \spc with Priscilla]?
\ex Which celebrity did you see [Priscilla with the brother of \spc]?
\zl

\eal
\label{compar} 
\ex Which celebrity did  [[you enjoy the memoir of \spcs more] than
                                 [any other non-fiction book]]?

\ex Which celebrity did you say that [[the sooner we take a picture of \spc ],
[the quicker we can go home]]?
\zl

\largerpage[2]
%\enlargethispage{3pt}
\noindent
The Across-The-Board (ATB) exception to the CSC is illustrated by the acceptability of (\ref{cc4}),
where each conjunct hosts a gap, linked to the same filler. As already noted above in (\ref{cc5}),
the fact that multiple gaps can be linked to the same filler is not unique to coordination.

\eal\label{cc4}
\ex Which celebrity did you buy [[a picture of  \spcs and a book about \spcs]]?
\ex Which celebrity  did you [[meet \spcs at a party] and [date \spcs for a few months]]?
\zl


\noindent
\citet{gazdar} and \citet{gpsg} assumed that the coordination rule requires \textsc{slash} values to be
structure-shared across conjuncts and the mother node, thus predicting both the Element Constraint
and the ATB exceptions. The failure of movement-based grammar to predict multiple gap extraction
facts was also seen as a major empirical advantage of GPSG/HPSG. A similar constraint is assumed in
\citet[202]{pollardsag} and \citet[60]{Beavers}, among others, illustrated in (\ref{cschpsgr}). See
\crossrefchaptert{coordination} for more discussion about coordination.

\ea
\label{cschpsgr}
Coordination Schema\is{schema!Coordination} (abbreviated):\\
\avm{
\emph{coordinate-phrase} \impl\\
[\punk{synsem|nonlocal|slash}{\1} \\
 dtrs & <[synsem|nonlocal|slash & \1 ], [synsem|nonlocal|slash & \1 ] > ]
}
\z

\noindent
Because the \textsc{slash} value \ibox{1} is structure-shared between the mother and the daughters in
(\ref{cschpsgr}), all three nodes must bear the same \textsc{slash} value.  This predicts the CSC and
the ATB exceptions straightforwardly. The failure of Mainstream Generative Grammar to predict these
and related multiple gap extraction facts in a precise way is regarded as one of the major empirical
advantages of HPSG over movement-based accounts.

But the facts about extraction in coordination structures are more complex than originally assumed,
and than (\ref{cschpsgr}) allows for. A crucial  difference between the Conjunct Constraint and the
Element Constraint is  that the latter is only in effect if the coordination has a symmetric
interpretation  \citep{Ross67,goldsmith,lakoff86,levinprince86}, as in  (\ref{asym}).\footnote{In
  asymmetric coordination, the order of the conjuncts has a major effect on the interpretation. 
  Thus, \emph{Robin jumped on a horse and rode into the sunset} does not 
mean the same as  \emph{Robin rode into the sunset and jumped on a horse}. Conversely,
in symmetric coordination the order of the conjuncts leads to no interpretational  differences,
as illustrated by the paraphrases \emph{Robin drank a beer and Sue ate a burger}
and \emph{Sue ate a burger and Robin drank a beer}.}


\eal  \label{asym}
\ex
 Here's the whiskey  which I [[went to the store] and [bought \spc]].
\ex Who did Lizzie Borden [[take an ax] and [whack \spcs to death]]?
\ex How much can you [[drink \spc] and [still stay sober]]?
\zl

% OR

\noindent
The coordinate status of (\ref{asym}) has been questioned since \citew{Ross67}. After all, if these
are subordinate structures rather than coordinate structures, then the possibility for non-ATB
long-distance dependencies ceases to be exceptional. But as \citet{schmerling72}, \citet{lakoff86},
\citet{levinepostal} and \citet[Chapter~5]{kehler} point out, there is no empirical reason to assume that the
examples in (\ref{asym}) are anything other than coordination structures.

Another reason to reject the idea that  the \textsc{slash} values of the daughters and the mother node
are simply equated in ATB extraction is the fact that sometimes multiple gaps are ``cumulatively''
combined into a ``pluralic gap''.\footnote{See for example  \citet{munn98atb,munn99atb},  \citet[136,
  160]{postal98}, \citet[125]{kehler}, \citet{gawronkehlersalt}, 
 \citet{zhang}, \citet{chavessubjexp}, and \citet{Vicente2016-NELS46}.}
As an example, consider the extractions in (\ref{cumulative}). There are two possible interpretations
for such extractions: one in which the \emph{ex situ} signs (i.e.\ the gap signs) and the filler
phrase are co-indexed, and therefore co-referential, and a second reading in which the two \emph{ex
  situ} phrases are not co-indexed even though they are linked to the same filler phrase.  Rather,
the filler phrase refers to a plural referent composed of the referents of the \emph{ex situ} signs,
as indicated by the subscripts in (\ref{cumulative}). For different speakers, the preferred reading
is the former, and in other cases, the latter, often depending on the example.

\eal \label{cumulative}
\ex
{}[What]$_{\lbrace i,j \rbrace}$ did Kim eat \trace$_i$ and drink \trace$_j$ at the party?\\
(answer: `Kim ate pizza and drank beer.')
\ex {}[Which city]$_{\lbrace i,j \rbrace}$ did Jack travel to \trace$_i$ and Sally decide to live in \trace$_j$?\\
(answer: `Jack traveled to London and Sally decided to live in Rome.')
\ex {}[Who]$_{\lbrace i,j \rbrace}$ did the pictures of \trace$_i$ impress \trace$_j$ the most?\\
(answer: `Robin's pictures impressed Sam the most.')
\ex {}[Who]$_{\lbrace i,j \rbrace}$ did the rivals of \trace$_i$ shoot \trace$_j$?\\
(answer: `Robin's rivals shot Sam.')
\ex {}[Who]$_{\lbrace i,j \rbrace}$ did you send nude photos of \trace$_i$ to \trace$_j$?\\
(answer: `I sent photos of Sam to Robin.')
%\item Where$_{\lbrace i,j \rbrace}$ did the train line run to \trace$_i$ from \trace$_j$?\\
%(possible answer: `The train line ran from NYC to Montreal.')
\zl


\noindent
In conclusion, the non-ATB exceptions in (\ref{asym}) suggest that the coordination rule should not
constrain \textsc{slash} at all, as argued for in \citet{chaves}.  Rather, the Element Constraint, its
ATB exceptions in (\ref{cc4}a,b) and the asymmetric non-ATB exceptions in (\ref{asym}) are more
likely to be the consequence of an independent semantic-pragmatic constraint that requires the
filler phrase to be ``topical'' relative to the clause
\citep{lakoff86,kuno87,kehler,kubotalee}. Thus, if the coordination is symmetric, then the
topicality requirement distributes over each conjunct, to require that the filler phrase be topical
in each conjunct. Consequently, extraction must be ATB in symmetric coordination. No distribution
needs to take place in asymmetric coordination, and thus both ATB and non-ATB extraction is licit in
asymmetric coordination. For an attempt to transfer some of Kuno's and Kehlers' insights into HPSG
see \citet{chaves}. In the latter proposal, the coordination rule is like most other rules in the
grammar in that it says nothing about the \textsc{slash} values of the mother and the daughters, along
the lines of \citet[354]{levhubook}. In other words, the constraints on \textsc{slash} in
(\ref{cschpsgr}) are unnecessary.  Rather, pragmatics is the driving force behind how long-distance
dependencies propagate one or more conjuncts, depending on the coordination being interpreted
symmetrically or not.  
%See \crossrefchaptert{coordination}\addpages for more discussion.
% The coordination chapter says nothing about pragmatics.



Let us take stock. The CSC does not receive a unitary account in modern HPSG, given that the
Conjunct Constraint and the Element Constraint are of a very different nature. Whereas the former
does not admit ATB extraction, and is predicted by a traceless analysis, the latter allows ATB
extraction as seen by the contrast between (\ref{islands-cc}c) and (\ref{cc4}).  Upon closer
inspection, the Element Constraint and the ATB exceptions are semantic-pragmatic in nature.  As we
shall see, a similar conclusion is plausible for various other island phenomena.


\section{Complex NP Constraint}

The Complex NP Constraint concerns the difficulty in extracting out of complex NPs formed with
either relative clauses (\ref{cpnpb}) or complement phrases (\ref{cpnpb2}).

\eal \label{cpnpb}
\ex[*]{ [What]$_i$ does Robin know $[$someone who has \trace$_i]$?\\
(cf.\ with `Does Robin know someone who has a drum kit?')}
\ex[*]{[Which language]$_i$ did they hire [someone [who speaks \trace$_i$]]?\\
       (cf.\ with `Did they hire someone who speaks Arabic?')}
\zl

\eal  \label{cpnpb2}
\ex[*]{[Which book]$_i$ do you believe the claim [that Robin plagiarized  \trace$_i$]?\\
      (cf.\ with `Do you believe the claim that Robin plagiarized  \emph{this book}?')}
\ex[*]{What$_i$  did you believe [the rumor [that Ed disclosed \trace$_i$]]?\\
      (cf.\ with `Did you believe the rumor that Ed disclosed \emph{that}?')}
\zl 

\noindent
It is tempting to prevent extractions out of adnominal clauses by simply stipulating that the \textsc{slash} value of the modifier must be empty, as (\ref{relcxt}) illustrates.  Perhaps, along the
lines of \citet{fodor78,Fodor83}, \citet{berwickwein}, and \citet{hawkins,hawbook}, processing
difficulties lead to the grammaticization of such a constraint, effectively blocking any modified
head from hosting any gaps.

\vbox{
\ea \label{relcxt}
Head-Modifier Schema\is{schema!Head-Adjunct} (abbreviated):\\
% todo avm
\avm{
[ head-dtr       [ synsem \1 ] \\
  non-head-dtrs  < [ synsem & [loc|cat|head|mod \1\\
                               nonloc|slash ~!\{ \}!~~~~~~]] > ]
}
\z 
}

\noindent
However, the robustness of the CNPC has been challenged by various counterexamples over the years
\parencites[139]{Ross67}[\page 206--207]{pollardsag}{kluender}[\page 9]{postal98}{saghof}.
% saghof is the complete paper
The sample in (\ref{csubok}) involves acceptable extractions from NP-embedded complement CPs (some of which are definite), and
(\ref{csubok2}) involves acceptable extractions from NP-embedded relative
clauses.\footnote{Counterexamples to the CNPC can be found in a number of languages, including
  \ili{Japanese} and \ili{Korean} \citep{kunojap,Nishigauchi99}, Ahan \citep{saah}, \ili{Danish}
  \citep[Chapter 2]{shir}, \ili{Swedish} \citep{allwood,engdahl82}, \ili{Norwegian}
  \citep{taraldsen82} and \ili{Romance} languages \citep{Cinque10}.  In some languages that have
  support verbs (or light verbs like ‘make the claim’), the CNPC is apparently not active, which is consistent with a complex
  predicate analysis for such constructions \citep{vives}.}


%Cases like ROSS have been taken to be lexically conditioned and  treated in terms of complex-predicate formation /N-V reanalysis/N-incorporation (Chomsky 1980, Kayne 1981, Cinque 1990, Davies and Dubinsky 2003).

\eal
 \label{csubok}
\ex The money [which]$_i$ I am making [the claim [that the company squandered \trace$_i$]] amounts to \$400,000.\footnote{
 \citew[139]{Ross67a}
}

\ex {}[Which rebel]$_i$ leader would you favor [a proposal [that the CIA assassinate \trace$_i$]]?\footnote{
\citew[206]{pollardsag}
}
%\itdopt{The sentence you provided was a mix of the two sentences by Pollard \& Sag. I fixed this.}

\ex {}[Which company]$_i$ did Simon spread [the rumor [that he had started \trace$_i$]]?
%\itdopt{Where are these examples from?\\
%R: The above is a modified version of an example in Kluender'96:246 `deriving island constraints from principles of predication', and it turns his datum is
%a modified version of Ross's. I say remove the citation to Kluender.
%}
\ex {}[What]$_i$ did you get [the impression [that the problem really was \trace$_i$]]?
%\footnote{
%\citew[\iaddpages]{kluender}
%}
%\itdobl{The example is not in \citew{kluender}!\\
%Rui: Remove citation to Kluender.}
\zl

\eal \label{csubok2}
\ex This is the kind of weather$_i$ that there are [many people [who like \trace$_i$]].\footnote{
\citew[\page 58]{shirlappin}
}
\ex Violence is something$_i$ that there are [many Americans [who condone \trace$_i$]].\footnote{
\citew[108]{mccawley81}
}
\ex There were several old rock songs$_i$ that she and I were [the only two [who knew \trace$_i$]].\footnote{
%\citew[\iaddpages]{chungmc}
\citew[\page 454]{Sag97a}
}
\ex This is the chapter$_i$ that we really need to find [someone [who understands \trace$_i$]].\footnote{
\citew[238]{kluender92}
}

\ex Which diamond ring did you say there was [nobody in the world [who could
buy \trace$_i$]]?\footnote{
\citew[206]{pollardsag}
}
\ex John is the sort of guy that I don't know [a lot of people [who think well of
\trace$_i$]].\footnote{
\citew[230]{culicover99}
}
\zl

\noindent
In the above counterexamples, the relative clauses contribute to the main assertion of the
utterance, rather than expressing background information. For example, (\ref{csubok2}a) asserts
`There are many people who like this kind of weather', and so on.  Some authors have argued that it
is precisely because such relatives express new information that the extraction can escape the
embedded clause \citep{shirlappin,kuno87,Dean,goldberg13}.  If this is correct, then the proper
account of CNPC effects is not unlike that of the CSC. In both cases, the information structural
status of the clause that contains the gap is crucial to the acceptability of the overall
long-distance dependencies.\footnote{Although it is sometimes claimed that such island effects are
  also active in logical form and semantic scope \citep{may85,ruys,fox,sab,katzira}, there is reason
  to be skeptical.  For example, the universally quantified noun phrases in (i) and (ii) is embedded
  in a relative clause but can have wide scope over the indefinite \emph{someone}, constituting a
  semantic CNPC violation. Note that these relatives are not presentational, and therefore are not
  specially permeable to extraction.

\eal
\ex  We were able to find someone who was an expert on each of the
     castles we planned to visit.  \citep[304]{MRS}
\ex John was able to find someone  who is willing to learn every Germanic language
    that we intend to study. \citep[\page 853]{chavesrnr}
\zllast}
  
In addition to pragmatic constraints, \citet{kluender92,kluender} proposed that processing factors
also influence the acceptability of CNPC violations.  Consider for example the acceptability
hierarchy in (\ref{k}); more specific filler phrases increase acceptability, whereas the presence of
more specific phrases between the filler and the gap seem to cause increased processing difficulty,
and therefore lower the acceptability of the sentence.  The symbol `$<$' reads as ``is less
acceptable than''.

\eal \label{k}
\ex What do you need to find the expert who can translate \spc?  $<$
\ex What do you need to find an expert who can translate \spc?   $<$
\ex What do you need to find someone who can translate \spc? $<$
\ex
\label{ex-which-document-do-you-need}
Which document do you need to find an expert who can translate \spc?
\zl

\largerpage
\noindent
There is on-line sentence processing evidence that CNPC violations with more informative fillers are
more acceptable and are processed faster at the gap site than violations with less informative
fillers \citep{hofsaglang}, as in (\ref{dd4}).
 
\eal
\label{dd4}
\judgewidth{?}
\ex[?]{Who  did you say that nobody in the world could ever depose \spc?}
\ex[]{
\label{ex-which-military-dictator}
Which  military dictator did you say that nobody in the world could ever depose \spc?}
\zl



\noindent
The same difference in reading times is found in sentences without CPNP violations, in fact. For
example, (\ref{hs}b) was found to be read faster at \emph{encouraged} than (\ref{hs}a). Crucially,
that critical region of the sentence is not in the path of any filler-gap dependency.
 
\ealnoraggedright \label{hs}
\ex The diplomat contacted the dictator who the activist looking for more contributions encouraged to preserve natural habitats and resources.
 
\ex The diplomat contacted the ruthless military dictator who the activist looking for more contributions encouraged to preserve natural habitats and resources.
\zl


\noindent
Given that finite tensed verbs can be regarded as definite, and infinitival verbs as indefinite
\citep{partee84}, and given that finiteness can create processing difficulty
\citep{kluender92,gibson0000}, then acceptability clines like (\ref{fininf}) are to be expected. See
\citet[Chapter 5]{levhubook} and \citet[308]{levine17} for more discussion.

\eal \label{fininf}
\ex Who did you wonder what Mary said to \spc?  $<$
\ex Who did you wonder what to say to \spc?  $<$
\ex\label{ex-which-of-the-people-at-the-party} 
Which of the people at the party did you wonder what to say to \spc? 
\zl

\subsection{On D-Linking}

The amelioration caused by more specific (definite) \emph{wh}-phrases as in (\ref{k}d), (\ref{dd4}b)
and (\ref{fininf}c) has been called a ``D-Linking'' effect \citep{pesetskydlink,pesetskybook}. It
purportedly arises if the set of possible answers is pre-established or otherwise salient. But there
are several problems with the D-Linking story.  First, there is currently no non-circular definition
of D-Linking; see \citet[16]{pesetskybook}, \citet[247--250]{ginzsag}, \citet[33, 39]{chung94} and
\citet[242, 268--271]{levhubook}. Second, the counterexamples above in (\ref{ex-which-document-do-you-need}), (\ref{ex-which-military-dictator}) and (\ref{ex-which-of-the-people-at-the-party}) are given out of the blue, and
therefore cannot evoke any preexisting set of referents, as D-Linking requires.  Furthermore,
nothing should prevent D-Linking with a bare \emph{wh}-item, as Pesetsky himself acknowledges, but
on the other hand there is no experimental evidence that context can lead to D-linking of a bare
\emph{wh}-phrase \citep{sprousediss07,villata}.\footnote{For more detailed criticism of D-Linking
  see \citet{hof2007}.}

\citet{kluenderkustas}, \citet{saghof}, \citet{philipt07,philcls} and \citet{hofsaglang}
argue that more definite \emph{wh}-phrases improve the acceptability of extractions because they
resist memory decay better than indefinites, and are compatible with fewer potential gap sites. In
addition, \citet{Kroch98a} and \citet[270]{levhubook} point out that D-Linking amelioration effects
may simply result from the plausibility of background assumptions associated with the proposition.

\subsection{On memory limitations}

\enlargethispage{3pt} 
\citet{sprouse12} use $n$-back and serial recall tasks to argue that there is no evidence that
working memory limitations correlate with island acceptability, and therefore that the
``processing-based'' account of islands put forth by \citet{kluender92,kluender},
\citet{kluenderkustas}, \citet{hofsaglang} and others is unfounded. To be sure, it cannot be
stressed enough that the accounts in \citet{kluender92} and \citet{hofsaglang} are not strictly
based on performance, and involve other factors as well, most notably plausibility and pragmatic
factors. See in particular \citet[49]{hoflangreply}, where it is argued that at least some
extraction constraints may be due to a combination of syntactic, semantic, pragmatic, and
performance factors.  Basically, if the correct location of a gap is syntactically, semantically, or
pragmatically highly unlikely in that particular utterance, then it is less likely for the sentence
to be acceptable.  Indeed, there is independent experimental evidence that speakers attend to
probabilistic information about the syntactic distribution of filler-gap dependencies
\citep{culcogsci}, and that gap predictability is crucial for on-line processing of islands
\citep{michelt}.\footnote{More broadly, there is good evidence that speakers deploy probabilistic
  information when processing a variety of linguistic input, including words
  \citep{altman99,arai,creel,delong,kutas84}, lexical categories \citep{gibson07,levy13,tabor97},
  syntactic structures \citep{levyted,lau06,levy08,staub}, semantics
  \citep{altman99,federmeier,kamide03}, and pragmatics \citep{shankweiler,mak,roland12}.} But as
\citet{reply2} point out, there is no reason to believe that $n$-back and serial recall tasks are
strongly correlated to working memory capacity to begin with. Second, one of the main points of
\citet{hofsaglang} is that the literature on experimental island research has not systematically
controlled for multiple factors that can impact the processing and comprehension of complex
sentences. If the experimental items are excessively complex, then readers are more likely to give
up understanding the utterances and subtler effects will not be measurable.  \citet{phil13},
however, regard such concerns as irrelevant.  Although it is unclear to what extent expectations and
processing constraints contribute to island effects, it is likely that they play some role in CNPC
effects, as well as other island types discussed below.

\section{Subjacency}

\is{subjacency|(}
One general constraint called Subjacency (Chomsky \citeyear[\page 271]{Chomsky73a}; \citeyear[\page 40]{Chomsky86b};
\citealp{Baltin81a,Baltin2006a}) was introduced to try to capture many of the constraints discussed here.
The claim was that movement cannot cross so-called bounding nodes, and what exactly counts as bounding
node was assumed to be a language-specific parameter in a universal principle. Theoretically, this
seems questionable, since it requires innate knowledge involving part of speech information
\parencites[Section~13.1.5.1]{MuellerGT-Eng4}[\page 539--540]{Newmeyer2004a}, and specific claims
concerning German and English are empirically wrong as well, as \citet{Mueller2004d},
\citet{Mueller2007c}, \citet{MM2009a} and \citet{SS2013b-u} showed with corpus examples.\footnote{
  The observation that arbitrarily many NP nodes can be crossed by \isi{extraposition} goes back
  to at least \citet[\page 52]{Koster78b-u}, who discussed Dutch data.
}

The original claim by \citet{Baltin81a} and \citet[\page 40]{Chomsky86b} was that the extraposed
relative clauses in (\mex{1}) can only be interpreted as referring to the embedding NP, that is, an
assumed \isi{extraposition} starts in t$'$ rather than t.
\eal
\label{ex-chomsky-sub}
\ex {}[\sub{NP} Many books [\sub{PP} with [stories t]] t$'$]  were sold\\
    {}[that I wanted to read].
\ex {}[\sub{NP} Many proofs [\sub{PP} of [the theorem t]] t$'$] appeared\\
    {}[that I wanted to think about].
\zl
The authors assume that NP, PP, VP and AP are bounding nodes for rightward movement in English and
that the unavailable interpretation is ruled out by the Subjacency Principle \citep[\page
262]{Baltin81a}. However, the attested examples in (\ref{nosubza}) show that subjacency does not
hold for \isi{extraposition} out of NPs or PPs. The examples in (\ref{nosubza}a--c) are adapted from
\citet[\page 106, 109, 111]{SS2013b-u}, and those in (\ref{nosubza}d--f) are from \citet[863]{chavesrnr}.
\eal \label{nosubza}
\ex {[In [what noble capacity \spc]] can I serve him
[that would glorify him and magnify his name]?}
% Strunk2013a p 106

\ex  {We drafted [a list of basic demands \spc] last night [that have
to be unconditionally met or we will  go on strike].}
% Strunk2013a p 109

\ex { For example, we understand that Ariva buses have won [a number
of contracts for routes in London \spc] recently,
[which will not be run by low floor accessible buses].}
% Strunk2013a p. 111 

\ex {Robin bought [a copy of a book \spc] yesterday [about ancient \ili{Egyptian} culture].}

\ex {I'm reading [a book written by  a famous physicist
\spc]
right now, [who was involved in the Manhattan Project].}

\ex {I saw [your ad in a magazine \spc] yesterday [on the table at the dentist office].}
\zl
\is{subjacency|)}


\section{Right Roof Constraint}

\is{Right Roof Constraint|(}\is{extraposition|(}
Rightward movement is traditionally regarded as being clause bounded. Such \emph{Right Roof
    Constraint} \citep[Section~5.1.2]{Ross67} effects are illustrated in (\ref{rrc}), in which a
phrase appears \emph{ex situ} in a position to the right of its \emph{in situ} counterpart; see
\citet{akma75}, \citet{baltin78}, and \citet{stowelldiss}, among others.

\eal \label{rrc}
\ex[*]{I [met a man [who knows \trace$_i$] yesterday] [all of your songs]$_i$.\footnote{
    \citew[861]{Chaves2014a-u}
}
}
\ex[*]{[[That a review \trace$_i$ came out yesterday] is catastrophic] [of this book]$_i$.\footnote{
\citew[\page 375]{rochemont}}
}
\ex[*]{It was believed \trace$_i$ that [there walked into the room \trace$_j$ ]  [by everyone]$_i$ [a man with  long blond hair]$_j$].\footnote{
\citew[\page 386]{rochemont}}}
\zl



\noindent
When treated as a form of extraction, rightward movement has been predominantly accounted for via a
feature \textsc{extra(posed)} \parencites{kellerverb,Keller95b,Bouma96,eynde96}[Section~13.2]{Mueller99a}{KimSag2005},
rather than by \textsc{slash}.  Thus, Right-Roof Constraint (RRC) island effects can be easily modeled
by stipulating that the \textsc{extra} value of an S node must be empty. One way to do so is to state
that any S dependent (valent or adjunct) must be [\textsc{extra} \eliste].  Thus, no extraposed element
may escape its clause. However, the oddness of (\ref{rrc}) may not be due to any such syntactic
stipulation, given the acceptability of counterexamples like (\ref{longb}). Note that the adverbial
interveners in such examples do not require parenthetical prosody. Conversely, even strong
parenthetical prosody on the adverbs in (\ref{rrc}) fails to improve those sentences.


\eal \label{longb}
\ex  I've [been requesting [that you pay back \spc] [ever since May]] [the money
I lent to you a year ago].\footnote{
  Adapted from \citew[167]{Kayne98a-u}.
}

\ex I've [been wanting to [meet someone who \textsc{knows} \spc] [ever since I was little]] [exactly what happened to Amelia
Earhart].\footnote{
  \citew[861]{chavesrnr}, adapted from \citew[177]{Gazdar81a}
}

\ex  I've been wondering  if it is possible   \spc]
[for many years now] [for anyone to memorize the Bible word for word].\footnote{
\citew[861]{chavesrnr}
}
\zl


\noindent
The durative semantics of \emph{I've been wanting/requesting/wondering} raises an expectation about
the realization of a durative adverbial expression like \emph{ever since} or \emph{for many years}
that provides information about the durative semantics of the main predicate.  Hence, the adverb is
cued by the main predication, in some sense, and coheres much better in a high attachment than with
a lower one.

The fact that the RRC is prone to exceptions has been noted by multiple authors as the sample in
(\ref{unb}) illustrates. In all such cases, a phrase is right-extracted from an embedded clause,
which should be flat out impossible if \isi{extraposition} is clause-bounded. Again, the adverbial
interveners in (\ref{unb}) do not require any special prosody, which means that these data cannot be
easily discarded as parenthetical insertions.
  
\eal \label{unb}
\ex I have  [wanted [to know \spc] for many years] [exactly what happened to Rosa Luxemburg].\footnote{
Attributed to \citew{witten} in \citew[92n]{postal74}.}

\ex I have  [wanted [to meet \spc] for many years] [the man who spent so much money planning the assassination of Kennedy].\footnote{
Attributed to Janet Fodor (p.c.) in \citew[177]{gazdar}.}

\ex Sue [kept [regretting \spc] for years] [that she had not turned
him down].\footnote{
\citew[\iaddpages]{eynde96}
}

\ex She has been [requesting that he [return \spc] [ever since last Tuesday]] [the book
that John borrowed from her last year].\footnote{
Adapted from \citew[167]{Kayne98a-u}.
}

\ex  Mary [wanted [to go \spc] until yesterday]  [to the public lecture].\footnote{\citew{lasnik07}}
\zl

\noindent
\citet{grosurrc}, \citet{gazdar} and \citet{stucky} argued that the RRC is the result of performance
factors such as syntactic and semantic parsing expectations and memory resource limitations, not
grammar proper.  Indeed, we now know that there is a general well-known tendency for the language
processor to prefer attaching new material to the more recent constituents
\citep{frazcl,gibetal,trax98,fodor02b,fernandez03}.  Indeed, eye-tracking studies like \citet{hnps}
indicate that the parser is reluctant to adopt extraposition parses.  This explains why
extraposition in written texts is less common in proportion to the length of the intervening
material \citep{UBDKKOS98b}: the longer the structure, the bigger the processing burden.  Crucially,
however, the preference for the closest attachment can be weakened by many factors
\citep{fernandez03,desmet,devic,carreras}.  For example, \citet{levyted} show that relative clause
extraposition creates significant processing difficulty when compared with non-extraposed
counterparts of the same sentences, but that a preceding context that sets up a strong expectation
for a relative clause modifying a given noun can facilitate comprehension of an extraposed relative
clause modifying that noun.  In other words, in spite of a larger processing burden, some
extrapositions can be made easier to process by parsing expectations.

A detailed account of extraposition island phenomena does not exist in any framework, as far as I am
aware. But the line of inquiry first proposed by \citet{grosurrc}, \citet{gazdar} and
\citet{stucky}, and later experimentally supported by \citet{levyted}, \citet{SS2013b-u} seems to be
on the right track. If so, then there is no syntactic constraint on \textsc{extra}. Rather, RCC effects
are to a large extent the result of difficulty in integrating the extraposed phrase \emph{in situ}.
\is{Right Roof Constraint|)}
\is{extraposition|)}

\section{Freezing}
\label{sec-freezing}

\is{Freezing|(}
A related island phenomenon also involving rightward displacement, first noted in
\citet[305]{Ross67}, is \emph{Freezing}: leftward extraction (\ref{z1}a) and extraposition
(\ref{z1}b) cause low acceptability when they interact, as seen in (\ref{f}). In (\ref{f}a) there is
extraction from an extraposed PP, in (\ref{f}b) there is extraction from an extraposed NP, and in
(\ref{f}c) an extraction from a PP crossed with direct object extraposition.
 
 
\eal \label{z1}
\ex Who$_j$ did you [give [a picture of \trace$_j$] [to Robin]]?
\ex Did you [give \trace$_i$ [to Robin] [a picture of my brother]$_i$]?
\zl

\eal \label{f}
\ex[*]{Who$_j$ did you [give a picture  \trace$_i$]  [to Robin] [of \trace$_j$]$_i$?}    
\ex[*]{Who$_j$ did you [give \trace$_i$ [to Robin] [a picture of \trace$_j$]$_i$]?} 
\ex[*]{Who$_j$ did you [give \trace$_i$ [to \trace$_j$] [a picture of my brother]$_i$]?}
\zl


\noindent
\citet[457]{fodor78} notes that (\ref{f}c) has a syntactically highly probable temporary alternative
parse in which \emph{to} combines with the NP \emph{a picture of my brother}. The existence of this
local ambiguity likely disrupts parsing, especially as it occurs in a portion of the sentence that
contains two gaps in close succession. Indeed, constructions with two independent gaps in close
proximity are licit, but not trivial to process, as seen in (\ref{ddg}), specially if the extraction
paths cross \citep{fodor78}, as in (\ref{ddg}b).

{\judgewidth{?}
\eal \label{ddg}
\ex[?]{This is a problem which$_i$ John$_j$ is difficult to talk to  \trace$_j$ about \trace$_i$.}
\ex[?]{Who$_j$  can't  you remember which papers$_i$ you sent copies of \trace$_i$ to \trace$_j$?}
\zl
}

\noindent
A similar analysis is offered by \citet[477]{freezing}, who note that constructions like (\ref{f}c)
must cause increased processing effort since the point of retrieval and integration coincides with
the point of reanalysis. The existence of a preferential alternative parse that is locally licit but
globally illicit can in turn lead to a ``digging-in'' effect
\citep{ferreirahend,ferreirahend2,tabor3}, in which the more committed the parser becomes to a
syntactic parse, the harder it is to backtrack and reanalyze the input. The net effect of these
factors is that the correct parse of (\ref{f}c) is less probable and therefore harder to identify
than that of (\ref{f}b), which suffers from none of these problems, and is regarded to be more
acceptable than (\ref{f}c) by \citet[453]{fodor78} and others. See \citet{chavesf} for experimental
evidence that speakers can adapt and to some extent overcome some of these parsing biases.
 
Finally, prosodic and pragmatic factors are likely also at play in (\ref{f}), as in the
RRC. \citet{huck} show that when an unstressed stranded preposition is separated from its selecting
head by another phrase, oddness ensues for prosodic reasons.  Finally, \citet{huck} and
\citet{boling92} also argue that freezing effects are in part due to a pragmatic conflict created by
extraposition and extraction: \emph{wh}-movement has extracted a phrase leftward, focusing interest
on that expression, while at the same time extraposition has moved a constituent rightward, focusing
interest on that constituent as well.  Objects tend to be extraposed when they are discourse new,
and even more so when they are heavy \citep[71]{wasowbook}. Therefore, the theme phrase \emph{a
  picture of John} in (\ref{f}c) is strongly biased to be discourse new, but this clashes with the
fact that an entirely different entity, the recipient, is leftward extracted, and therefore is the
\emph{de facto} new information that the open proposition is about.  No such mismatch exists in
(\ref{f}a) or (\ref{f}b), in contrast, where the extraposed theme is more directly linked to the
entity targeted by leftward extraction.
\is{Freezing|)}

% \itdopt{I have several extractions out of extrapositions in German. If you extrapose an infinitval
%   clause and extract out of it, this is pretty normal. If you want, I can provide examples. They are
%   in \citew{Mueller99a}.}


\section{Subject islands}

Extraction out of subject phrases like (\ref{all}) is broadly regarded to be impossible in several
languages, including \ili{English} \citep[106]{chomsky73}, an effect referred to as a
\emph{Subject Island} (SI).  This constraint is much less severe in languages like \ili{Japanese},
\ili{German}, and \ili{Spanish}, among others
\citep{stepanov,jurkaetal,goodall11,greco,fukuda,polinsky13}.


\eal \label{all}
\ex[*]{Who did stories about terrify John?\footnote{
 \citew[106]{chomsky77b}}
}
\ex[*]{ Who was a picture of laying there?\footnote{
\citew[114]{kayne81}}
}
\ex[*]{Who do you think pictures of would please John?\footnote{
\citew[497]{huang82}}
}
\ex[*]{Who does the claim that Mary likes upset Bill?\footnote{
\citew[42]{lasniksaito}}
}
\ex[*]{Which candidate were there posters of  all over town?\footnote{
\citew[651]{lasnikpark}
}
}
\zl

% https://muse.jhu.edu/article/48037

\noindent
However, \ili{English} exceptions were noticed early on, and have since accumulated in the
literature.  In fact, for \citet{Ross67}, \ili{English} extractions like (\ref{sie}a) are not
illicit, and more recently \citet[147]{chomsky08} has added more such counterexamples.  Other
authors noted that certain extractions from subject phrases are naturally attested, as in
(\ref{sie}b,c). Indeed, \citet{AHWG2020a-u} shows that extractions like those in (\ref{sie}c) are in
fact acceptable to native speakers, and that no such island effect exists in \ili{French}
either.\footnote{For completeness, other authors argue that PP extractions from NP subjects are
  illicit, such as \citet[653]{lasnikpark}, among many others.}


\eal \label{sie}
\ex[]{[Of which cars]$_i$ were [the hoods \trace$_i$] damaged by the explosion?\footnote{
\citew[\page 242]{Ross67}
% ex 4.252
}}

\ex[]{They have eight children [of whom]$_i$ [five \trace$_i$] are still living at home.\footnote{
\citew[1093]{hud12} }
}

\ex[]{ Already Agassiz had become interested in the rich stores of the extinct fishes of Europe, especially those of Glarus in Switzerland and of Monte Bolca near Verona, [of which]$_i$, at that time, [only a few \trace$_i$] had been critically studied.\footnote{
Encyclopaedia Britannica Online, Agassiz, (Jean) Louis (Rodolphe). Quoted from \citew{santorini}}
}
\zl

\noindent
\ili{English} exceptions to the SI constraint are not restricted to PP extractions,
however. Although \citet[265]{Ross67} claimed NP extractions from NP subjects like (\ref{rosssi}) are
illicit, it was arguably premature to generalize from such a small sample.

\eal  \label{rosssi}
\ex[]{
The hoods of these cars were damaged by the explosion. 
}
\ex[*]{Which cars were the hoods of damaged by the explosion?
%\citep[ft.\,31,254]{Ross67}}
%\citep{Ross67}
}
\zl


\noindent
Indeed, a number of authors have noted that some NP extractions from subject NPs are either passable
or fairly acceptable, as illustrated in (\ref{etal}). See also \citet[195, ft.\,32]{pollardsag},
\citet{postal98}, \citet[304]{sauerelb}, \citet[230]{culicover99}, \citet[265]{levhubook},
\citet[470, 471]{chavesextr}, and \citet[97]{chavesjeruen}.

  
\ealnoraggedright  
\label{etal}
%?[Who]$_i$ did [John's talking to \trace$_i$] bother you most?\\
 %\citep{engdahl}

\ex[]{[What]$_i$ were [pictures of \trace$_i$] seen around the globe?\footnote{
  \citew[268]{kluender} }
}

\ex[]{ It's [the kind of policy statement]$_i$ that [jokes about \trace$_i$] are a dime a dozen.\footnote{
\citew[204]{levineetal}}
}

\ex[]{ There are [certain topics]$_i$ that [jokes about \trace$_i$] are completely unacceptable.\footnote{
\citew[252, ft.\,6]{Levine:Sag:03}}
}

\ex[]{  [Which car]$_i$ did [some pictures of \trace$_i$] cause a scandal?\footnote{
 \citew[111]{fernandez}}
}
 
 \ex[]{ [What]$_i$ did [the attempt to find \trace$_i$] end in failure?\footnote{
 \citew[370]{hofsaglang}}
}
 
 \ex[]{  [Which president]$_i$ would [the  impeachment of \trace$_i$] cause more  outrage?\footnote{
 \citew[\page 467]{chavesextr}}
}
 
\ex[]{ I have a question$_i$ that [the probability of you knowing the answer to \trace$_i$] is zero.\footnote{
\citew[\page 305]{chavessubj}}
}

\zl



\noindent
Whereas SI violations involving subject CPs are not attested, those involving infinitival VP
subjects like (\ref{vpsia}) are.  See \citet[471]{chavesextr} for more natural occurrences.


\ealnoraggedright
\label{vpsia}
\ex []{The eight dancers and their caller, Laurie Schmidt, make up the Farmall Promenade of nearby Nemaha, a town$_i$
that [[to describe \trace$_i$ as tiny] would be to overstate its size].\footnote{
 \citew[1094, ft.\,27]{hud12} }
}

\ex[]{ In his bedroom, [which]$_i$ [to describe \trace$_i$ as small] would be a gross
  understatement, he has an audio studio setup.\footnote{
\url{http://pipl.com/directory/name/Frohwein/Kym}, accessed 2021-04-03, quoted from
  \citew[\page 303]{Chaves2013a-u}
}
}
\ex[]{They amounted to near twenty thousand pounds, [which]$_i$ [to pay \trace$_i$] would have ruined me.\footnote{ 
Benjamin Franklin, William Temple Franklin and  William Duane. 1834. Memoirs of Benjamin Franklin,
vol 1.\ p.\,58}
}
\zl 

\noindent
Incidentally, subject phrases are not extraposition islands either, as shown in (\ref{subxxx}) from
\citew[\page 864]{Chaves2014a-u}. See also \citet{Gueron:May:84}. The oddness of examples
like *\mbox{[\emph{Pictures} \trace]} \emph{frighten people} [\emph{of John}] from
\citet[46]{drummond} is more likely due to a digging-in effect, caused by speakers assuming that the
subject is syntactically and semantically complete by the end of the verb phrase. 


\eal \label{subxxx}
\ex {}[The circulation of a rumor \trace$_i$] has started [that Obama will not seek re-election]$_i$.
\ex {}[A  copy of a new book \trace$_i$] arrived yesterday [about ancient \ili{Egyptian} culture]$_i$.
%\ex {}[Concerns about the deaths  \trace$_i$] were
%raised [of two diplomatic envoys recently abducted in Somalia]$_i$.
\zl
%\itdopt{Also Keller \& Müller.\\
%Rui: If you say so. What's the bitext reference?}

For an extremely in-depth and detailed study on French Subject Islands, using corpora and controlled experiments, as well as a HPSG formalization of a pragmatic account
of the phenomena, see \citew{Winckel2024a}.

\subsection{Clausal Subject Constraint}

Let us now consider SI effects involving more complex subjects.  Infinitival subject clauses seem to
impose no SI constraint, an observation going back to \citet{kunotakamib}, but noted elsewhere a few
times:

\eal \label{vpsi}
\ex[]{ This is something [which]$_i$  -- for you to try to understand \trace$_i$  -- would be futile.\footnote{
\citew[49]{kunotakamib}}
}

\ex[]{ I just met Terry's eager-beaver research assistant [who]$_i$  -- for us
to talk to \trace$_i$ about  any subject other than linguistics -- would be absolutely pointless.\footnote{
\citew[265]{levhubook}}
}

\ex[]{ There are [people in this world]$_i$ that -- for me to describe \trace$_i$ as despicable -- would be an understatement.\footnote{
\citew[468]{chavesextr}}
}
%\itdopt{Should the -- be brackets?\\
%No, it is intended to indicate parenthetical-like prosody.}
\zl

\noindent
Infinitival subjects contrast dramatically with finite subjects. The latter are renowned for being
particularly hard to extract from, as shown in (\ref{ssc}).  \citet[\page 243]{Ross67} dubbed this extreme kind
of SI the \term{Sentential Subject Constraint} (SSC). See also \citet{chomsky73}, \citet{huang82},
\citet{Chomsky86b}, and \citet{freidin92}.\footnote{That said, \citet{chavessubj} reports that some
  native speakers find SSC violations like (i) to be fairly acceptable, again raising some doubt
  about the robustness of \ili{English} SI effects:

\ea
{}[Which actress]$_i$ does [whether Tom Cruise marries \trace$_i$] make any difference to you?
\zlast
}


\eal \label{ssc}
\ex[*]{[Who]$_i$ did [that Maria Sharapova beat \trace$_i$] surprise everyone?\\
(cf.\ with `That Maria Sharapova beat Serena Williams surprised everyone.')}

\ex[*]{[Who]$_i$ did [that Robin married \trace$_i$] surprise you?\\
(cf.\ with `Did that Robin married Sam surprise you?')}
\zl

%\noindent
%
%\ea
%?The hat [which]$_i$ -- [[{\slshape that I brought} \trace$_i$] seemed strange to the nurse] -- was a fedora.\\
%\citep{Ross67}
%\z

\noindent
There are some functional reasons for why clausal SI violations may be so strong. First, subject
clauses are notorious for being particularly difficult to process, independent of
extraction. Clausal subjects are often stylistically marked and difficult to process, as
(\ref{that}a) illustrates. Thus, it is extremely hard to embed a clausal subject within another
clausal subject, even though such constructions ought to be perfectly grammatical, like
(\ref{that}b, c). In addition, it is known that tense can induce greater processing costs
\citep{kluender92,gibson0000}.

\eal  \label{that}
\ex[]{ That the food that John ordered tasted good pleased him.\footnote{
%\textcites[\iaddpages]{cowper76}
% The example was not in Cowper but another example that Gibson cited.
\citew[\page 57]{gibson91}}
}

\ex[*]{That that Joe left bothered Susan surprised Max.\footnote{
\citew[\page 33]{kimball}}
}
 
\ex[*]{That that the world is round is obvious is dubious.\footnote{
\citew[\page 119]{kuno74}} 
}
\zl


\noindent
Interestingly, clausal subjects become more acceptable if extraposed as shown in (\ref{exps}).  The
explanation offered by \citet[356--357]{fod74} is that speakers tend to take the initial clause in
the sentence to be the main clause. Thus, \emph{that} is taken to be the subject, but the remainder
of the structure does not fit this pattern. Thus, a sentence like (\ref{exps}a) causes increased
processing load because it has a different structure than the parser expects.  This processing
problem does not arise in the counterpart in (\ref{exps}b).\footnote{See \citet{gibson07} for online
  evidence that the word \emph{that} is preferentially interpreted as a determiner even in syntactic
  contexts where it cannot be a determiner. The use of ``determiner'' corresponds to the traditional
  term, referring to a certain category of prenominal constituent rather than to the whole nominal
  phrase including the noun and all its dependents.  Gibson's evidence suggests that both top-down
  (syntactic) expectations are independent from bottom-up (lexical) frequency-based expectations in
  sentence processing. Thus, a clausal subject phrase starting with the complementizer \emph{that}
  is likely to be misparsed as a matrix clause with sentence-initial pronominal or determiner
  \emph{that}.}



\eal  \label{exps}
\judgewidth{?}
\ex[?]{That [it is obvious that [the world is round]] is dubious.}
\ex[]{It is dubious that [it is obvious that [the world is round]].\footnote{
\citew[\page 130]{kuno74}}
}
\zl


\noindent
Indeed, \citet{fodor67}, \citet{bever}, and \citet{frazier88} also show that extraposed clausal
subject sentences like (\ref{ext}a) are easier to process than their \textit{in-situ} counterparts
like (\ref{ext}b).  Not surprisingly, the former are much more frequent than the latter, which
explains why the parser would expect the former more than the latter.


\eal  \label{ext}
\ex It surprised Max that Mary was happy. 
\ex That Mary was happy surprised Max.
\zl

% Gibson "that"
%* Who did [the fact that the candidate supported __] upset voters?
% http://www.bu.edu/linguistics/UG/course/lx500-s06/handouts/lx500s06q-13b-proc.pdf
% That the food that John ordered tasted good pleased him
% cowper76 gibson 91
  
\noindent
If we add a filler-gap dependency to a sentence that already is complex by virtue of having a
clausal subject, the resulting structure may be too difficult to parse.  This point is illustrated
by the contrast in (\ref{that2}).



\eal \label{that2}
\judgewidth{?*}
\ex[?*]{
What does that he will come prove?
}
\ex[]{
What does his coming  prove?\footnote{
\citew[146]{lewis93}
}
}
\zl


\noindent
As argued by \citet{dubinsky2009}, the low acceptability of extraction in subject-auxiliary
inversion sentences with clausal subjects is more likely to be the result of extragrammatical
factors than of grammatical conditions.  For example, not all extractions like (\ref{v}b) are
unacceptable, as \citet[382--387]{delahunty} and \citet[115]{dubinsky2009} point out.

\eal \label{v}
\ex[]{That the food that John ordered tasted good pleased me.}
\ex[*]{Who did that the food that John ordered tasted good please \spcs?}
\zl

\noindent
The evidence discussed so far suggests that sentences involving extraction and clausal subjects are
odd at least in part due to the likely cumulative effect of various sources of processing
complexity.  Sentences with sentential subjects are unusual structures, which can mislead the parser
into the wrong analysis. A breakdown in comprehension can occur because the parser must hold complex
incomplete phrases in memory while processing the remainder of the sentence. The presence of a
filler-gap dependency will likely only make the sentence harder to process.  It is independently
known that the more committed the parser becomes to a syntactic parse, the harder it is to reanalyze
the string \citep{ferreirahend,ferreirahend2,tabor3}.  For example, unless prosodic or contextual
cues are employed to boost the activation of the correct parse, (\ref{compzeat}) will be
preferentially misanalyzed as having the structure [NP [V [NP]]].

\ea \label{compzeat} Fat people eat accumulates.
\z 

\noindent
The garden-path effect that the digging-in causes in example (\ref{compzeat}) serves as an analogy
for what may be happening in particularly difficult subject island violations.  In both cases, the
sentences have exactly one grammatical analysis, but that parse is preempted by a highly
preferential alternative which ultimately cannot yield a complete analysis of the sentence.  Thus,
without prosodic cues indicating the extraction site, sentences like (\ref{temp}) induce a
significant digging-in effect as well.


\eal \label{temp}
\ex[?]{
Which problem will a solution to be found by you?
}
\ex[?]{
Which  disease will  a cure for be found by you?
} 
\zl

\noindent
This also explains why SI violations like (\ref{excz}) are relatively acceptable: a subject NP with
a subordinate CP is more expectable and easier to process than a CP subject, even though the former
is more complex than the latter.\footnote{For claims that NP-embedded clausal SI violations are
  illicit see \citet[42]{lasniksaito}, \citet[796]{colinphillips}, and \citet[67]{colin_horn}.}
Clausal subjects are unusual structures, inconsistent with the parser expectations \citep{fod74},
and the presence of filler-gap dependency in an NP-embedded clausal subject is less likely to cause
difficulty for the parse to go awry than a filler-gap dependency in a clausal
subject.\footnote{\citet{clausen,clausencuny} provide experimental evidence that complex subjects
  cause a measurable increase in processing load, with and without extraction. Moreover, it is known
  that elderly adults have far more difficulty repeating sentences with complex subjects than
  sentences with complex objects \citep{kemper86}. Similar difficulty is found in timed reading
  comprehension tasks \citep{kynette}, and in disfluencies in non-elderly adults \citep{clarkwasow}.
  Speech initiation times for sentences with complex subjects are also known to be longer than for
  sentences with simple subjects \citep{ferreirasubj,tsiam}.  Finally, \citet{garnsey},
  \citet{kutasetal}, and \citet{vanpetten} show that the processing of open-class words,
  particularly at the beginning of sentences, require greater processing effort than closed-class
  words.}


\eal \label{excz}
\ex {}[Which puzzle]$_i$ did the fact that nobody could solve \trace$_i$ astonish you the most?
\ex {}[Which crime]$_i$ did the fact that nobody was accused of \trace$_i$ astonish you the most?
\ex {}[Which question]$_i$ did the fact that none of us could answer \trace$_i$  surprise you the most?
\ex {}[Which joke]$_i$ did the fact that nobody laughed at \trace$_i$ surprise you the most?
%\item Which ancient script did the fact that nobody has deciphered surprise you the most?
%\item Which poison did the fact that there is no antidote to surprise you?
%\item Which disease did the claim that nobody can cure $|$ astonish you the most? ...
%\item  Which employees did the fact that the company refused to fire $|$ astonish you? ...
%\item Which bill did the fact that some senators tried to amend $|$ surprise you? ...
\zl

\subsection{Accounts of SI effects}

This complex array of effects suggests that the SI constraint is not due to a single factor
\citep{chomsky08,chavessubj,fernandez}, be it grammatical or otherwise.  One possibility is that SIs
are partly due to pragmatic and processing constraints, perhaps not too different from those that
appear to be active in the island effects discussed so far. As \citet[495]{kluender06} notes:
``Subject Island effects seem to be weaker when the \emph{wh}-phrase maintains a pragmatic
association not only with the gap, but also with the main clause predicate, such that the filler-gap
dependency into the subject position is construed as of some relevance to the main assertion of the
sentence''. Indeed, many authors \citep{shir-jrn,valin86,kuno87,ken,Dean,goldberg13} have argued
that extraction is in general restricted to the informational focus of the proposition, and that SIs
(among others) are predicted as a consequence. In a nutshell, since subjects are typically reserved
for topic continuity, subject-embedded referents are unlikely to be the informational focus of the
utterance.  Although it is not easy to construct sentences where a dependent of the subject can be
easily deemed as the informational focus, it is by no means impossible.  For instance, (\ref{kh}a)
is particularly acceptable because whether or not an impeachment causes outrage crucially depends on
who is impeached (cf.\ with \emph{Would the impeachment of Donald Trump cause outrage?}). Similarly,
in (\ref{kh}b) whether or not an attempt failed or succeeded crucially depends on what was attempted
(cf.\ with \emph{The attempt to find the culprit ended in failure}).

%\enlargethispage{2pt}
\eal  \label{kh}
\ex[]{ Which President would [the  impeachment of \spcs] cause more outrage?\footnote{
\citew[467]{chavesextr}}
}

\ex[]{ What did [the attempt to find \spcs]  end in failure?\footnote{
 \citew[370]{hofsaglang}}
}
\zl
 
\noindent
Although experimental research has confirmed that sentences with SI violations tend to be less
acceptable than grammatical controls \citep{sprsat2,goodall11,crawfordwccfl,clausencuny,greco}, and
that their acceptability remains consistently low during repeated exposition
\citep{sprsat2,crawfordwccfl}, other research has found that the acceptability of SI violations is
not consistently low, and can be made to increase significantly
\citep{hiramatsu00,clausencuny,chavessubjexp,chavesjeruen}. This mixed evidence is consistent with
the idea that SI effects are very sensitive to the particular syntax, semantics, and pragmatics of
the utterance in which they occur. If the items are too complex, or stylistically awkward, or
presuppose unusual contexts, then SI effects are strong.  For example, if the extraction is
difficult to process because the sentence gives rise to local garden-path and digging-in effects,
and is pragmatically infelicitous in the sense that the extracted element is not particularly
relevant for the proposition (i.e.\ unlikely to be what the proposition is about) or comes from the
presupposition rather than the assertion, then we obtain a very strong SI effect. Otherwise, the SI
effect is weaker, and in some cases nearly non-existent like (\ref{kh}), (\ref{vpsia}), or the
pied-piping examples studied by \citet{AHWG2020a-u}. The latter involve relative clauses, in which
subjects are not strongly required to be topics, in contrast to the subjects of main clauses.

This approach also explains why subject-embedded gaps often become more acceptable in the presence
of a second non-island gap: since the two gaps are co-indexed, then the fronted referent is
trivially relevant for the main assertion, as it is a semantic argument of the main verb. For
example, the low acceptability of (\ref{pgbq}a) is arguably caused by the lack of plausibility of
the described proposition: without further contextual information, it is unclear how the attempt to
repair an unspecified thing $x$ is connected to that attempt damaging a car.\footnote{
The examples in (\mex{1}) are due to \citet[\page 796]{colinphillips}.}

\eal
\label{pgbq}
\ex[*]{What did [the attempt to repair \spcs] ultimately damage the car?}
\ex[]{What did [the attempt to repair \spcs] ultimately damage \spc?}
\zl


\noindent
The example in (\ref{pgbq}a) becomes more acceptable if it is contextually established that $x$ is a
component of the car. In contrast, (\ref{pgbq}b) is felicitous even out-of-the-blue because it
conveys a proposition that is readily recognized as being plausible according to world knowledge:
attempting to fix $x$ can cause damage to $x$.  If Subject Island effects are indeed contingent on
how relevant the extracted subject-embedded referent is for the assertion expressed by the
proposition, then a wide range of acceptable patterns is to be expected, parasitic or
otherwise. This includes cases like (\ref{symbd}), where both gaps are in SI environments. As
\citet{Levine:Sag:03}, \citet[256]{levhubook} and \citet[161]{Culicover13} note, cases like
(\ref{symbd}) should be completely unacceptable, contrary to fact.


\eanoraggedright 
\label{symbd}
This is a man who $[$friends of \spcs $]$ think that $[$enemies of \spcs $]$ are everywhere.
\z 

\noindent
The conclusion that SI effects are contingent on the particular proposition expressed by the
utterance and its pragmatics thus seems unavoidable \citep{chavesresp}. In order to test this
hypothesis, \citet{chavesresp} examine the acceptability of sentences like (\ref{pairs}), which
crucially express nearly-identical truth conditions and have equally acceptable declarative
counterparts. This way, any source of acceptability contrast must come from the extraction itself,
not from the felicity of the proposition.

\eal  \label{pairs}
\ex Which country does the King of  Spain resemble [the President of \spcs]?
\ex Which country does [the President of \spcs] resemble  the King of Spain?
\zl

\noindent
The results indicate that although the acceptability of the SI counterpart in (\ref{pairs}b) is
initially significantly lower than (\ref{pairs}a), it gradually improves.  After eight exposures,
the acceptability of near-truth-conditionally-equivalent sentences like (\ref{pairs}) becomes
non-statistically different.  What this suggests is that SI effects are at least in part
probabilistic: the semantic, syntactic and pragmatic likelihood of a subject-embedded gap likely
matters for how acceptable such extractions are. This is most consistent with the claim that -- in
general -- extracted phrases must correspond to the informational focus of the utterance
\citep{shir-jrn,valin86,kuno87,ken,Dean,goldberg13}, and in particular with the intuition that SI
violations are weaker when the extracted referent is relevant for the main predication
\citep[495]{kluender06}.






\section{Adjunct islands}

\largerpage
\citet{cattell} and \citet{huang82} noted that adjunct phrases often resist extraction, as
illustrated in (\ref{tad}). The constraint blocking respective extractions is usually referred to as
\emph{The Adjunct Island Constraint} (AIC).

\eal  \label{tad}
\ex[*]{What$_i$ does John dance [whistling \trace$_i$]?\footnote{
\citew[\page 1357]{Truswell2007a}
}
}
\ex[*]{What$_i$ did John arrive while [whistling \trace$_i$]?\footnote{
\citew[\page 1359]{Truswell2007a}
}
}
\ex[*]{Which club$_i$ did John meet a lot of girls [without going to \trace$_i$]?\footnote{
\citew[38]{cattell}}
}
\ex[*]{Who$_i$ did Robin laugh [after Pat called on the phone \trace$_i$]?\footnote{
based on \citew[503]{huang82}
}
}
\zl

\noindent
Although a constraint on \textsc{slash} could effectively ban all extraction from adjuncts, the problem
is that the AIC has a long history of exceptions, noted as early as \citet[38]{cattell}, and by many
others since, including \citet[72]{chomsky82}, \citet{engdahl}, \citet[103]{hegarty90},
\citet[139]{cinque}, \citet[191]{pollardsag}, \citet[253]{culicover87}, and \citet[\page 200]{borg}. A sample of
representative counterexamples is provided in (\ref{catt}).

% Kayne 1983,
% Jones 1991, 
%Manzini 1992, 
%Levine & Sag 2003, 
%Haider 2004,
% Truswell 2007, 2009)
% Chaves

\ealnoraggedright \label{catt}
\ex Who did he buy a book [for \spc]?
\ex Who would you rather sing [with \spc]?
\ex What temperature should I wash my jeans [at \spc]?
\ex That's the symphony that Schubert died [without finishing \spc].
\ex Which report did Kim go to lunch [without reading \spc]?
\ex A problem this important, I could never go home [without solving \spcs first].
\ex What did he fall asleep [complaining about \spc]?
\ex What did John drive Mary crazy [trying to fix \spc]?
\ex Who did you go to Girona [in order to meet \spc]?
\ex Who did you go to Harvard [in order to work with \spc]?
\zl

\largerpage
\noindent
Exceptions to the AIC include tensed adjuncts, as noted by \citet[88]{grosu81}, \citet[29]{deane}, \citet[287]{levhubook}, \citet[144]{gold06}, \citet[471]{chavesextr}, \citet[175,
ft.\,1]{truswellbook} and others. A sample is provided in (\ref{gr}).\footnote{\citet{truswellbook}
  argues that the AIC and its exceptions are best characterized in terms of event-semantic
  constraints, such that the adjunct must occupy an event position in the argument structure of the
  main clause verb. However, recent experimental research has been unable to validate Truswell's
  acceptability predictions \citep{kohrt}, and moreover, such an account incorrectly predicts that
  extractions from tensed adjuncts is impossible \citep[175, ft.\,1]{truswellbook}.}
 
% Kohrt, Annika, Trey Sorensen, Dustin A. Chacn. 2018, to be submitted. The real-time status of semantic exceptions to the adjunct island constraint.
 

\ealnoraggedright
\label{gr}
\ex These are the pills that Mary died [before she could take \spc].
\ex This is the house that Mary died [before she could sell \spc].
\ex The person who I would kill myself [if I couldn't marry \spc] is Jane.
\ex Which book  will Kim understand linguistics better [if she reads \spc]?
\ex This is the watch that I got upset [when I lost \spc].
\ex Robin, Pat and Terry were the people who I lounged around at home all day [without realizing were \spcs coming for dinner].
\ex Which email account would you be in trouble [if someone broke into \spc]?
\ex Which celebrity did you say that [[the sooner we take a picture of \spc ], [the quicker we can go home]]?
\zl

\noindent 
To be sure, some of these sentences are complex and difficult to process, which in turn can lead
speakers to prefer the insertion of an ``intrusive'' resumptive pronoun at the gap site, but they
are certainly more acceptable than the classic tensed AIC violations examples like Huang's
(\ref{tad}d).  Acceptable tensed AIC violations are more frequent in languages like \ili{Japanese},
\ili{Korean}, and \ili{Malayalam}.

Like Subject Islands, AIC violations sometimes improve ``parasitically'' in the presence of a second
gap as in (\ref{parasaic}). First of all, note that these sentences express radically different
propositions, and so there is no reason to assume that all of these are equally felicitous.  Second,
note that (\ref{parasaic}a, c) describe plausible states of affairs in which it is clear what the
extracted referent has to do with the main predication and assertion, simply because of the fact
that \emph{document} is a semantic argument of \emph{read}.  In contrast, (\ref{parasaic}b)
describes an unusual state of affairs in that it is unclear what the extracted referent has to do
with the main predication \emph{read the email}, out of the blue. Basically, what does reading
emails have to do with filing documents?

\eal \label{parasaic}
\ex[]{Which document did John read \spcs before filing \spc? }
\ex[*]{Which document did John read the email before filing \spc?}
\ex[]{Which document did John read \spcs before filing a complaint?}
\zl

\noindent
If AIC violations were truly only salvageable parasitically, then counterexamples like
(\ref{books}a) should not exist.  As \citet{Levine:Sag:03} and \citet[256]{levhubook} note, both
gaps reside in island environments and should be completely out and less acceptable than
(\ref{books}b, c), contrary to fact.

\eal  \label{books}
\ex[]{What kinds of books do [the authors of \spc] argue about royalties [after writing \spc]?}
\ex[*]{What kinds of books do [authors of \spc] argue about royalties after writing malicious pamphlets?}
\ex[*]{What kinds of books do authors of malicious pamphlets argue about royalties
      {}[after writing \spc]?
}
\zl

\noindent
In (\ref{books}a), there is no sense in which the gap inside the subject is parasitic on the gap
inside the adjunct, or vice-versa -- under the assumption that neither gaps is supposed to be licit
without the presence of a gap outside an island environment. In conclusion, the notion of parasitic
gap is rather dubious. See \citet[256--273]{levhubook} for a more in-depth discussion of parasitism and
empirical criticism of null resumptive pronoun accounts.

As in the case of other island phenomena discussed so far, it is doubtful that any purely syntactic
account can describe all the empirical facts.  Rather, extractions out of adjuncts are licit to the
degree that the extracted referent can be interpreted as being relevant for the assertion.


%\subsection{constraints on preposition stranding.
%A (brief) presentation of the way they have been/are treated in movement based approaches would be useful. 

% \subsection{Resumption}
%\section{Factive Islands}

\section{Superiority effects}

Contrasts like those below have traditionally been taken to be due to a constraint that prevents a
given phrase from being extracted if another phrase in a higher position can be extracted instead
\citep{chomsky73,chomsky80}. Thus, the highest \emph{wh}-phrase is extractable, but the lowest is
not.

\eal  \label{badsc1}
\ex[]{Who \spcs saw what?}
\ex[*]{What did who see \spc?}
\zl

\eal  \label{badsc2}
\ex[]{Who did you persuade \spcs to buy what?}
\ex[*]{What did you persuade who to buy \spc?}
\zl

Several different kinds of exceptions to this \emph{Superiority Constraint} (SC) have been 
noted in the literature. First, it is generally recognized that \emph{which}-phrases are
immune to the SC:

\eal \label{book}
\ex I wonder which book which of our students read \spc over the summer?
\ex Which book did which professor buy \spc?
\zl

\noindent
\citet{pesetskydlink} proposed to explain the lack of SC effects in (\ref{book}) by stipulating that
\emph{which}-phrases are interpreted as indefinites which do not undergo LF movement. Rather, they
require ``D-linking'' and obtain wide scope via an entirely different semantic mechanism called
unselective binding. In order for a phrase to be D-linked, it must be associated with a salient set
of referents. But as \citet[248]{ginzsag} note, there is no independent evidence for saliency
interpretational differences between \emph{which} and other \emph{wh}-phrases like \emph{what} and
\emph{who}. For example, it is implausible that speakers have a specific referent in mind for the
\emph{which}-phrases in examples like (\ref{which}).\footnote{
The examples in (\mex{1}) are from \citew[248]{ginzsag}.}

\eal  \label{which}
\ex[]{ I don't know anything about cars. Do you have any suggestions about which car -- if any -- I should buy when I get a raise?}
\ex[]{ I don't know anything about cars. Do you have any suggestions about what  -- if anything -- I should buy when I get a raise?
}
\zl

\noindent
Furthermore, there are acceptable SC violations involving multiple \emph{wh}"=questions such as
those in (\ref{multiple}). See \citet{Bolinger78}, \citet{kayne83} and \citet[109]{pesetskydlink}
for more such examples and discussion.\footnote{\citet{gibson10} and others have found no evidence
  that the presence of a third \emph{wh}-phrase improves the acceptability of a multiple
  interrogative, even with supporting contexts. However, the examples in (\ref{multiple}) require
  peculiar intonation, which may be difficult to elicit with written stimuli.}

\eal \label{multiple}
\ex Who wondered what \textsc{who} was doing \spc?
\ex What did \textsc{who} take \spcs \textsc{where}?
\ex Where did \textsc{who} take \textsc{what} \spc?
\zl

\noindent 
Finally, there are also SC violations that involve echo questions like (\ref{break1}) and reference
questions like (\ref{break2}). See \citet[Chapter 7]{ginzsag} for a detailed argumentation that echo
questions are not fundamentally different, syntactically or semantically, from other uses of
interrogatives.

\eal \label{break1}
\ex What did Agamemnon break?
\ex What did \textsc{who} break \spc?
\zl

\eal  \label{break2}
\ex What did he break?
\ex What did \textsc{who} break \spc?
\zl

\noindent
There are two different, yet mutually consistent, possible explanations for SC effects in HPSG
circles.  One potential factor concerns processing difficulty \citep{arnon07}. Basically,
long-distance dependencies where a \emph{which}-phrase is fronted are generally more acceptable and
faster to process than those where \emph{what} or \emph{who} if fronted, presumably because the
latter are semantically less informative, and thus decay from memory faster, and are compatible with
more potential gap sites before the actual gap.  The second potential factor is prosodic in
nature. Drawing from insights by \citet[170--172]{Ladd96} about the \ili{English} interrogative
intonation, \citet[251]{ginzsag} propose that in a multiple \emph{wh}-interrogative construction,
all \emph{wh}-phrases must be in focus except the first. Crucially, focus is typically -- but not
always -- associated with clearly discernible pitch accent. Thus, (\ref{badsc1}) and (\ref{badsc2})
are odd because the second \emph{wh}-word is unaccented. In this account, a word like \emph{who}
has two possible lexical descriptions, shown in (\ref{whoavm}).


\eal \label{whoavm}
\ex[]{\emph{Ex situ} interrogative \emph{who}:\\*
\avm{
	[\phon < \upshape who/WHO > \\
	synsem &	[loc &	[cat &	[head & noun \\
								spr & < > \\
								comps & < >\\
                                                                arg-st & < >] \\
						cont &	[ind & i \\ 
								restr & \{ \}] \\
						store & \{ \1	[ind & i \\ 
										restr & \{\type{person}($i$)\} ] \} ] \\
				nonloc &	[wh & \{ \1 \} \\
							rel & \{ \} \\
							slash & \{ \} ] ] ]
}
}



\ex[]{ Optionally \emph{ex situ} interrogative \emph{who}:\\*
\avm{
	[\phon < \upshape WHO > \\
	synsem &	[loc &	[cat &	[head & noun \\
								spr & < > \\
								comps & < >\\
                                                                arg-st & < >] \\
						cont &	[ind & i \\ 
								restr & \{ \} ] \\
						store & \{[ind & i \\ 
								restr & \{\type{person}($i$)\} ] \} ] \\
				nonloc &	[wh & \{ \} \\
							rel & \{ \} \\
							slash & \{ \} ] ] ]
}
}
\zl

\noindent
Since only the (optionally accented) lexical entry in (\ref{whoavm}a) is specified with a non-empty
\textsc{wh} value, the theory of extraction proposed in \citet[Chapter~5]{ginzsag} predicts that (\ref{whoavm}a)
must appear \emph{ex situ}.  In contrast, the accented lexical entry in (\ref{whoavm}b) can
appear\emph{in situ}. For more discussion see \citet[261]{levhubook}.

A related range of island phenomena concerns extraction from \emph{whether}"=clauses, which is
traditionally assumed to be forbidden, as (\ref{whex}) illustrates.

\eal \label{whex}
\ex[*]{Which movie did John wonder whether Bill liked \spc?}
\ex[*]{Which movie did John ask why Mary liked \spc?}
\zl 

\noindent
But again, the oddness of (\ref{whex}) is unlikely to be due to syntactic constraints, given the
existence of passable counterexamples like (\ref{whexc}).

\eal \label{whexc}
\ex[]{He told me about a book which I can't figure out whether to buy \spcs or not.\footnote{
\citew[27]{Ross67}}
}
\ex[]{ Which glass of wine do you wonder whether I poisoned \spc?\footnote{
\citew[81]{cresti1995extraction}}
}
\ex[]{ Who is John wondering whether or not he should fire \spc?\footnote{
\citew[477]{chavesextr}}
}
\zl

\noindent
As noted by \citet[28]{Kroch98a}, the reduced acceptability of an example like (\ref{kro}a) is better
explained simply by noting the difficulty of accommodating its presupposition in (\ref{kro}b).


\eal \label{kro}
\ex How much money was John wondering whether to pay?
\ex There was a sum of money about which John was wondering whether
to pay it.
\zl


\section{The Left Branch Condition}

\citet[207]{Ross67} discovered that the leftmost constituent of an NP cannot be extracted, as in
(\ref{lbcv}), a constraint he dubbed the \emph{Left Branch Condition} (LBC).\footnote{As in other
  island environments discussed above, the LBC is not operative in constraining semantic scope, as
  illustrated in the following example from \citew[303]{MRS}:

\ea
Someone took a picture of each student's bicycle.
\zlast}
 
 
\eal \label{lbcv}
\ex[*]{Whose$_i$ did you meet [ \trace$_i$ friend]?\\
 (cf. with `\emph{You met whose friend?}')}
 
\ex[*]{Which$_i$ did you buy [ \trace$_i$ book]?\\
(cf. with `\emph{You bought which book?}')}

\ex[*]{How much$_i$ did you find [ \trace$_i$ money]?\\
(cf. with `\emph{You found how much money?}')}
\zl


\noindent
These facts are accounted for if Determiner Phrases (DPs) are not valents of the nominal head. See also \crossrefchapterw[Section~\ref{np:sec-functor-basics}]{np}. If the DP is not listed in the argument
  structure of the nominal head, then there is no way for the DP to appear in \SLASH.  See
  \citet{runner06} for psycholinguistic evidence that reflexive binding to possessors involves
  binding-theory-exempt logophors, since reflexives in the PPs of NPs containing possessors are not
  in complementary distribution with pronouns.  Rather, the DP selects the nominal head as shown in
(\ref{the68}).

\ea \label{the68}
\avm{
[\phon < the > \\
 synsem & [loc & [cat &	[head &	[\type*{determiner} \\
		                 select N$'$[ind & i \\
					     restr & \1 ] ] \\
			 spr    & < > \\
			 comps  & < >\\
                         arg-st & < >] \\
                  cont & [ind & i \\
			  restr & < > ] \\
			  store & \{[\type*{the-rel}
			             var & i \\
				     arg & \1 ] \} ] \\
	   nonloc & [wh & \{ \} \\
                      rel & \{ \} \\
                      slash & \{ \} ] ] ]
}
\z

\noindent
Based on \citet[133]{cxsag07}, \citet[197, 198]{CP2020a-u} assume that genitive DPs combine
with nominal heads and bind their \textsc{x-arg} index via a dedicated construction, not as valents.
For example, in nominalizations like \emph{Kim's description of the problem} the DP \emph{Kim's} is
not a valent of \emph{description}, and therefore the genitive DP cannot appear in \textsc{slash}.
Rather, genitive DPs are instead constructionally co-indexed with the agent role of the noun
\emph{description} via \textsc{x-arg}.  Moreover, the clitic \emph{s} in \emph{Kim's} must lean
phonologically on the NP it selects, and therefore cannot be stranded for independently motivated
phonological reasons, predicting the oddness of *\emph{It was Kim who I read 's description of the
  problem}.

There are various languages which do not permit extraction of left branches from noun phrases, but
have a particular PP construction that appears to allow LBC violations.  This is illustrated below
in (\ref{langslbc}), with \ili{French} data.

\eal \label{langslbc}
\ex[]{
\gll Combien$_i$  a-t-il  vendu  [~\trace$_i$  de  livres]?\\
     how.many     has-he  sold {} of  books\\\hfill(\ili{French})
\glt `How many books did he sell?'
}
\ex[*]{ 
\gll Quels$_i$  avez-vous  achet\'{e}  [~\trace$_i$  livres]?\\
     how.many   have-you   bought {}  books\\
\glt `How many books have you bought?' 
}

%\ex \gll Quantos  livros$_i$  \'{e}  que  tens 
%[\trace$_i$  de  Matem\'{a}tica]?\\ 
%how-many books  is  that  you-have {} of  mathematics\\
%\glt `How many math books do you have?' \hfill (\ili{Portuguese})
 \zl 


\noindent
But the LBC violation in (\ref{langslbc}a) is only apparent. The \emph{de livres} is in fact a
post-verbal \textit{de}-N$'$ nominal, and thus no LBC violation occurs in (\ref{langslbc}a). See
\citet{Abeille:Bonami:ea:04} for details.  Finally, \citet[236--237]{Ross67} also noted that some languages do
not obey the LBC at all. A small sample is given in (\ref{lbcx}). However, the languages in question
lack determiners, and therefore it is possible that the extracted phrase has a similar
independent status to the \ili{French} \textit{de-N}$'$ phrase in (\ref{langslbc}a).
 
\eal
\label{lbcx}
\ex
\gll Jak\k{a}$_i$  kupi\l{}e\'{s}  [~\trace$_i$ ksi\k{a}\.{z}k\k{e}]\\
     what          you.bought  {} book\\ \hfill (\ili{Polish})
    \glt `Which book did you buy?'
       
\ex 
\gll Cju$_i$  citajes [~\trace$_i$  knigu]?\\
     whose  you.are.reading {}  book\\ \hfill (\ili{Russian})
\glt `Whose book are you reading?'
 
\ex 
\gll Ki-nek$_i$  akarod,  hogy  halljam  [~\trace$_i$  a  hang-j\'{a}-t]?\\
     who-\textsc{dat}  you.want  that  I.hear {} the  voice-\textsc{poss.3sg-acc}\\\hspace{-1.4em}(\ili{Hungarian})
\glt `Whose voice do you want me to hear?' 
\zl
 

 
 
 
\section{The Complementizer Constraint}

%\largerpage
\citet{Perlmutter68} noted that subject phrases have different extraction properties than that of
object phrases, as illustrated in (\ref{subobjasymm2}). The presence of the complementizer hampers
extraction of the subject, but not of the complement.\footnote{There is no evidence that the
  Complementizer Constraint applies at the semantic level, however. The subject phrase of the
  embedded clause can outscope the subject phrase of the matrix:

\eal
\ex Some teacher claimed that each student had cheated.
\ex Every teacher claimed that a student had cheated.
\zllast
} 

\eal   \label{subobjasymm2}
\ex[*]{[Who]$_i$ did Tom say (?that) \trace$_i$ had bought the tickets?}
%\item \,[Who] did he say \spcs bought the rutabaga?\\
\ex[*]{[Who]$_i$ do you believe (?that) \trace$_i$ got you fired? }
\ex[]{[The things]$_i$ that they believe (?that) \trace$_i$ will happen are disturbing to contemplate. }
\ex[*]{[Who]$_i$ did you ask if \trace$_i$ bought the tickets?}
%\item \,[Who] did he say \spcs bought the rutabaga?\\
\ex[*]{[Who]$_i$ do you expect for  \trace$_i$ to fire  you? }
\zl



%These examples are also found in L&H 2006, p. 68
\noindent
\citet[\page 194]{Bresnan:77}, \citet{Culicover1993a-u} and others also noted that  Complementizer Constraint effects can be
reduced in the presence of an adverbial intervening between the complementizer and the gap:

\eal \label{advcirc}
%\ex{[Who]$_i$ did Tom say that -- as far he could remember -- \trace$_i$ had bought the tickets?}
\ex{ [Who]$_i$ do you believe that -- for all intents and purposes --  \trace$_i$ got you fired? }
\ex{ [Who]$_i$ do you think that after years and years of cheating death \trace$_i$ finally died?}
\zl

\noindent
In \citet{bouma} and \citet[181]{ginzsag}, extracted arguments are typed as \type{gap-synsem} rather
than \type{canon-synsem}. Only the latter are allowed to correspond to \emph{in situ} signs and to
reside in valence lists. However, subject extraction is different. If a subject phrase is extracted,
then the \textsc{subj} list contains the respective \type{gap-synsem} sign. If one assumes that the
lexical entry for the complementizer \emph{that} requires S complements specified as
$[$\textsc{subj} $\langle \, \rangle]$ then the oddness of (\ref{subobjasymm2}) follows. For
\citet[181]{ginzsag}, the adverbial circumvention effect in (\ref{advcirc}) is the result of
assuming that the complementizer selects an adverb and a clause as arguments (the second of which is
required to have a subject gap. This analysis seems \emph{ad-hoc} because the adverb would be
expected to adjoin to the clause rather than being a complement of the complementizer.  On the other
hand, the analysis is consistent with what happens in \ili{French}: when the subject of the
complement CP is extracted, the complementizer must be \textit{qui} instead of \textit{que}, which could
easily be captured by such an account.

%\largerpage
A simpler account of the Complementizer Constraint has emerged recently, however, in principle
compatible with any theory of grammar. For \citet{Kandy06,Kandy09} and others, the Complementizer
Constraint is prosodic in nature.  Complementizers must cliticize to the following phonological
unit, but if a pause is made at the gap site then the complementizer cannot do so.  Accordingly, if
the pronunciation of \emph{that} is produced with a reduced vowel [\textipa{D@t}] rather than
[\textipa{D\ae t}] then the Complementizer Constraint violations in (\ref{subobjasymm2}) improve in
acceptability. Though promising, \citet{Richart} found no experimental evidence for amelioration of
the Complementizer Constraint effects either with phonological reduction of the complementizer or
with contrastive focus. Further research is needed to determine the true nature of Complementizer
Constraint effects.



\section{Island circumvention via ellipsis}

Ellipsis somehow renders island constraints inactive, as in (\ref{slc}). A deletion-based analysis
of such phenomena such as \citet{merchantbook} relies on moving the \emph{wh}-phrase before deletion
takes place, but since movement is assumed to be sensitive to syntactic island constraints, the
prediction is that (\ref{slc}) should be illicit, contrary to fact.

\eal  \label{slc}
\ex[]{ Terry wrote an article about Lee and a book about someone else from
East Texas, but we don't know who$_i$ (*Terry wrote an article about \trace$_i$ Lee
and a book about \trace$_i$).\\
 \,[CSC violation]}

\ex[]{ Bo talked to the person who discovered something, but I still don't know
what$_i$ (*Bo talked to the person who discovered \trace$_i$). \\
\,[CNPC violation]}

\ex[]{ That he'll hire someone is possible, but I won't divulge who$_i$
(*that he'll  hire \trace$_i$ is possible).\\
\, [SSC violation]}

\ex[]{  She bought a rather expensive car, but I can't remember how expensive
(*she bought a \spcs car). \\
\,[LBC violation]}
\zl

\noindent
The account adopted in HPSG is one in which remnants are assigned an interpretation based on the
surrounding discourse context \citep{ginzsag,Culicover:Jackendoff:05,jacobson08,sagn}.  See
\crossrefchaptert{ellipsis} for more detailed discussion. In a nutshell, the \emph{wh}-phrases in
(\ref{slc}) are ``coerced'' into a proposition-denoting clause via a unary branching construction
that taps into contextual information.  This straightforwardly explains not only why the antecedent
for the elided phrase need not correspond to overt discourse -- e.g.\ sluices like \emph{What
  floor?} or \emph{What else?} -- but also why the examples in (\ref{slc}) are immune to island
constraints: there simply is no island environment to begin with, and thus, no extraction to violate
it. For more on ellipsis and island effects see \citet[108--109]{CP2020a-u}.




\section{Conclusion}

HPSG remains relatively agnostic about many island types, given the existence of robust exceptions.
It is however clear that many island effects are not purely due to syntactic constraints, and are
more likely the result of multiple factors, including pragmatics, semantics and processing
difficulty.  To be sure, it is yet unclear how these factors can be brought together and articulate
an explicit and testable account of island effects. In particular, it is unclear how to combine
probabilistic information with syntactic, semantic and pragmatic representations, although one
fruitful avenue to approach this problem may be via \emph{Data-Oriented Parsing}
\citep{NF2002a-u,NF99a,Arnold:Linardaki:07,BSS2003a-ed,Bod2009a}.


From its inception, HPSG has been meant to be compatible with models of language comprehension and
production \citep{sagser,Sag:Wasow:ta,Sag:Wasow:ta2}, but not much work has been dedicated to
bridging these worlds; see \crossrefchaptert{processing}. The challenge that island effects posit to
any theory of grammar is central to linguistic theory and cognitive science: how to integrate
theoretical linguistics and psycholinguistic models of on-line language processing so that
fine-grained predictions about variability in acceptability judgments across nearly isomorphic
clauses can be explained.


 
%\section*{Abbreviations}
\section*{\acknowledgmentsUS}

Many thanks to Bob Borsley, Berthold Crysmann, Anne Abeillé, Jean-Pierre Koenig, and Stefan Müller
for detailed comments about an earlier draft, and to Stefan Müller for invaluable editorial
assistance.

{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this] 
}
\end{document}

%      <!-- Local IspellDict: en_US-w_accents -->
