%% -*- coding:utf-8 -*-
\documentclass[output=paper]{langsci/langscibook} 
\author{Frank Van Eynde\affiliation{University of Leuven}}
\title{Nominal structures}

% \chapterDOI{} %will be filled in at production

%\epigram{Change epigram in chapters/03.tex or remove it there }
\abstract{This chapter shows how nominal structures are treated in HPSG. The introduction puts the discussion in the broader framework of the NP vs. DP debate and differentiates three HPSG treatments, i.e. the specifier treatment, the DP treatment and the functor treatment. They are each presented in some detail and applied to the analysis of ordinary nominals, covering both syntactic and semantic aspects. A comparison reveals that the DP treatment does not mesh as well with the monostratal surface-oriented nature of the HPSG framework as the other treatments. Then it is shown how the specifier treatment and the functor treatment deal with nominals that have idiosyncratic properties, such as the Big Mess Construction, gerunds, nominalized infinitives and irregular P+NP combinations. Their analysis typically requires an appeal to phrasal constraints and a modicum of constructionalism.}
\maketitle

\begin{document}
\label{chap-np}

\section{Introduction}


We use the term `nominal' in a theory-neutral way as standing for a noun
and its phrasal projection. In this broad sense all of the bracketed 
strings in (\ref{rb}) are nominals. 

\begin{exe} 
\ex\label{rb}  [the [red [box]]] is empty
\end{exe} 
  
\noindent
For the analysis of nominals there are two main approaches in generative grammar. 
One treats the noun (N) as the head all the way through. In that analysis the 
largest bracketed string in (\ref{rb}) is an NP. 
The other makes a distinction between the nominal core, 
consisting of a noun with its complements and modifiers, if any,  
and a functional outer layer, comprising determiners, quantifiers and 
numerals. In that analysis the noun is the head of \emph{red box}, 
while the determiner is the head of \emph{the red box}, so that the category 
of the latter is DP. 

The first approach, henceforth called the {\bf NP approach}, prevailed in 
generative grammar up to and including the Government and Binding model
\citep{Chomsky81}. One of its modules, the categorial component, 
consists of phrase structure rules, such as those in (\ref{ps0}). 

\begin{exe} 
\ex\label{ps0}   
\begin{xlist}
\ex  VP ~ $\rightarrow$ ~ V ~ NP 
\ex  NP ~ $\rightarrow$ ~ Det ~ Nom
\end{xlist} 
\end{exe}

\noindent
They are required to ``meet some variety of X-bar theory'' \citep[5]{Chomsky81}. 
The original variety is that of \citet{Chomsky70}. It consists of the following
cross-categorial rule schemata:

\begin{exe} 
\ex\label{xbar} 
\begin{xlist} 
\ex\label{xbar1}   X$'$ ~ $\rightarrow$ ~ X ~  ... 
\ex\label{xbar2}   X$''$ ~ $\rightarrow$ ~ [Spec, X$'$] ~ X$'$ 
\end{xlist} 
\end{exe} 

\noindent
X$'$ stands for the combination of a head X and its complements,
where X is N, A or V, and X$''$ stands for the combination of 
X$'$ and its specifier ``where [Spec, N$'$] will be analyzed as the determiner'' 
\citep[52]{Chomsky70}. 
X-bar theory was further developed in \citet{Jackendoff77}, who added a
schema for the addition of adjuncts and who extended the range of 
X with P, the category of adpositions. Generalized Phrase Structure Grammar 
developed a monostratal version of it, exemplified by the analysis 
in Figure \ref{sis}, quoted from \citet[126]{GPSG85}. 
The top node is the double-bar category N$''$, which 
consists of the determiner and the single-bar category N$'$. 
The AP and the relative clause are adjoined to N$'$, and 
the lowest N$'$ consists of the noun and its PP complement.

\begin{figure}
\centering
\begin{forest}
sm edges
[ N$''$
  [Det [that]]
  [N$'$
    [AP [very tall,roof]]
    [N$'$
      [N$'$
        [N [sister]]
        [PP [of Leslie,roof]]]
      [{S[+R]} [who we met,roof]]]]]
\end{forest}
\caption{\label{sis}The NP approach} 
\end{figure}
 
The second approach, henceforth called the {\bf DP approach}, results from an
extension of the range of X in (\ref{xbar}) to the functional categories. 
This was motivated by the fact that some of the phrase structure rules, 
such as (\ref{ps1}), do not fit the X-bar mould. 

\begin{exe} 
\ex\label{ps1}   S ~ $\rightarrow$ ~ NP ~ Aux ~ VP
\end{exe}   

\noindent
To repair this, the category Aux, which contained both auxiliaries and 
inflectional verbal affixes \citep{Chomsky57}, was renamed as I(nfl) and treated as the head of S. 
More specifically, I(nfl) was claimed to combine with a VP complement, yielding I$'$, 
and the latter was claimed to combine with an NP specifier (the subject), yielding I$''$
(formerly S).
For the analysis of nominals such an overhaul did not at first seem necessary, 
since the relevant PS rules did fit the X-bar mould, but it took place nonetheless, 
mainly in order to capture similarities between nominal and clausal structures. 
These are especially conspicuous in gerunds, nominalized infinitives and nominals 
with a deverbal head, and were seen as evidence for the claim that determiners have their 
own phrasal projection, just like the members of I(nfl) \citep{Abney87}. 
More specifically, members of D were claimed to take an N$''$ complement (formerly Nom), 
yielding D$'$, and the latter was claimed to have a specifier sister, as in Figure \ref{abn}.
The DP approach was also taken on board in other frameworks, 
such as Word Grammar \citep{Hudson90} and Lexical Functional Grammar \citep[99]{Bresnan00}. 

\begin{figure}
	\centering
	\begin{forest}
sm edges
[D$''$ 
	[D$'$
		[D [that]]
		[N$''$
			[N$'$ 
				[N [sister]]
				[PP [of Leslie,roof]]]]]]
	\end{forest}
	\caption{\label{abn} The DP approach } 
\end{figure}
     
Turning now to Head-driven Phrase Structure Grammar we find three different treatments.  
The first and oldest can be characterized as a lexicalist version of the NP approach, more 
specifically of its monostratal formulation in GPSG.  
It is first proposed in \citet{ps} and further developed in \citet{ps2} and 
\citet{GS00}. We henceforth call it the {\bf specifier treatment}, 
after the role which it assigns to the determiner. 
The second is a lexicalist version of the DP approach.  
It is first proposed in \citet{Netter94} and further developed in \citet{Netter96a},
and \citet{NerbonneMullen00}. We will call it the {\bf DP treatment}. 
The third adopts the NP approach, but neutralizes the distinction between adjuncts and specifiers, 
treating them both as functors. It is first proposed in \citet{VanEynde98a} and 
\citet{Allegranza98} and further developed in \citet{VanEynde03}, \citet{VanEynde06} 
and \citet{Allegranza06}. It is also adopted in Sign-Based Construction Grammar \citep{Sag2012}. 
We will call it the {\bf functor treatment}. This chapter presents the three treatments and 
compares them wherever this seems appropriate.  

We first focus on ordinary nominals (Section 2) and then on nominals with idiosyncratic 
properties (Section 3). For exemplification we use English and a number of other Germanic 
and Romance languages, including Dutch, German, Italian and French.  
We assume familiarity with the typed feature structure notation and with such basic notions 
as unification, inheritance and token-identity. 
    

\section{Ordinary nominals} 


We use the term `ordinary nominal' for a nominal that contains a noun, 
any number of complements and/or adjuncts and at most one determiner. 
This section shows how such nominals are analyzed in respectively the 
specifier treatment (2.1), the functor treatment (2.2) and the DP treatment (2.3).  

    
\subsection{The specifier treatment} 


The specifier treatment adopts the same distinction between heads, complements, 
specifiers and adjuncts as X-bar syntax, but the integration of these notions 
in a monostratal lexicalist framework inevitably leads to various differences. 
The presentation is mainly based on \citet{ps2} and \citet{GS00}. 
We first discuss the syntactic structure (2.1.1) and the semantic composition (2.1.2) 
of nominals, and then turn to nominals with a phrasal specifier (2.1.3). 


\subsubsection{Syntactic structure}


Continuing with the same example as in Figures 1 and 2, 
a relational noun, such as \emph{sister}, selects a PP as its complement 
and a determiner as its specifier, as spelled out in its {\sc category} value
(\ref{n}). 

\begin{exe} 
\ex\label{n}
\begin{avm}
[\type{category}         \\ 
 head  & \type{noun}     \\
 spr   & <Det>            \\
 comps & <PP\[\type{of}\]> ]
\end{avm}
\end{exe}

\noindent
The combination with a matching PP, as in \emph{sister of Leslie},   
is subsumed by the \type{head-comp(lement)s-phr} type, as defined in \crossrefchaptert{properties}, 
and yields a nominal with an empty {\sc comps} list.  
Similarly, the combination of this nominal with a matching determiner, as in \emph{that sister of Leslie},    
is subsumed by the \type{head-sp(ecifie)r-phr} type, as defined in \crossrefchaptert{properties}. 
and yields a nominal with an empty {\sc spr} list. This is spelled out in Figure \ref{les}. 

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{noun}, {\sc spr} ~ \eliste, {\sc comps} ~ \eliste]}
		[\avmbox{2} [that]]
		[{[{\sc head} ~ \avmbox{1}, {\sc spr} ~ \liste{\avmbox{2}}, {\sc comps} ~ \eliste]} 
			[{[{\sc head} ~ \avmbox{1}, {\sc spr} ~ \liste{\avmbox{2} Det}, {\sc comps} ~ \liste{\avmbox{3} PP}]} [sister]]
			[\avmbox{3} [of Leslie,roof]]]]	
	\end{forest}
	\caption{\label{les} Adnominal Complements and Specifiers}
\end{figure}

Since the noun is the head of \emph{sister of Leslie} and since the latter is 
the head of \emph{that sister of Leslie}, the Head Feature Principle implies 
that the phrase as a whole shares the {\sc head} value of the noun ($\avmbox{1}$). 
The valence features, {\sc comps} and {\sc spr}, have a double role. 
On the one hand, they register the degree of saturation of the nominal; 
in this role they supersede the bar levels of X-bar theory. 
On the other hand, they capture co-occurrence restrictions, 
such as the fact that the complement of \emph{sister} be a PP, rather than an NP or a clause, 
and that its specifier be a singular determiner, rather than a plural one.\footnote{The 
value of the third valence feature ({\sc subj}) is invariably the empty list for nouns, 
except in the case of predicative nouns, whose {\sc subj} list contains an NP 
that is identified with the target of the predication, see \citet[409]{GS00}.}

In contrast to complements and specifiers, adjuncts are not included in valence lists, 
since their addition has no effect on the degree of saturation. At the same time, 
since the combination of an adjunct with its head is subject to 
co-occurrence restrictions, one needs a way to capture those. 
For that purpose \citet[55--57]{ps2} employs the feature {\sc mod(ified)}. 
It is part of the {\sc head} value of the substantive parts-of-speech, 
i.e. noun, verb, adjective and preposition. Its value is of type \type{synsem} 
in the case of modifying items and of type \type{none} otherwise.

\begin{exe} 
\ex   \type{substantive}: \begin{avm} 
                          [mod ~ \type{synsem} $\vee$ \type{none}]  
                          \end{avm} 
\end{exe} 

\noindent
As an example, let us take the number and gender agreement 
between nouns and adjectives in the Romance languages.\footnote{This is an 
instance of concord, as defined in \crossrefchaptert{agreement}.}  
The Italian \emph{grossa} `big', for instance, 
is compatible with singular feminine nouns, such as \emph{scatola} `box', 
but not with plural feminine nouns, such as \emph{scatole} `boxes', nor
with masculine nouns, such as \emph{libro} `book' or \emph{libri} `books'. 
This is made explicit in its {\sc mod} value, spelled out in (\ref{rd}). 

\begin{exe} 
\ex\label{rd}
\begin{avm}
[\type{category}                              \\
 head [\type{adjective}                         \\
       mod|loc|cat [head [\type{noun}          \\
                          number ~ \type{sing} \\
                          gender ~ \type{fem}] \\
                    spr ~ <Det> ]]]
\end{avm}
\end{exe}

\noindent
The token-identity of the {\sc mod(ified)} value of the adjective
with the {\sc synsem} value of its head sister is part of the 
definition of type \type{head-adj(unct)-phr}, as defined in \crossrefchaptert{properties}. 
Since the {\sc mod(ified)} feature is part of the {\sc head} value, it follows from the 
Head Feature Principle that it is shared between an adjective 
and the AP which it projects. As a consequence, the {\sc mod(ified)} value of 
\emph{molto grossa} `very big' is shared with that of \emph{grossa} `big'. 
 
Besides its role in modeling agreement, the {\sc mod} value is also instrumental 
in capturing constraints on linear order.  
The fact, for instance, that the AP in \emph{that very tall bridge} appears 
after the determiner and not before, as in *\emph{very tall that bridge},
is captured by the stipulation that the {\sc mod} value of the adjective 
is a nominal with a determiner on its {\sc spr} list. This blocks the 
combination with a fully saturated NP, as in 
*\emph{very tall that bridge}.\footnote{This order constraint is overruled in 
the Big Mess Construction, see Section 3.2.} Also here, the {\sc mod} value of 
the adjective is shared with that of the AP, as shown in Figure \ref{lea}. 

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{noun}, {\sc spr} ~ \eliste]}
	[\avmbox{2} [that]]
	[{[{\sc head} ~ \avmbox{1}, {\sc spr} ~ \liste{\avmbox{2}}]}
		[{[{\sc head} ~ \avmbox{4}]}
			[{[{\sc head} ~ \type{adv}]} [very]]
			[{[{\sc head} ~ \avmbox{4} [\type{adj} {\sc mod} ~ \avmbox{3}]]} [tall]]]
		[{\avmbox{3} [{\sc head} ~ \avmbox{1}, {\sc spr} ~ \liste{\avmbox{2} Det}]}, before drawing tree={x+=2em} [sister of Leslie, roof, before drawing tree={x+=2em}]]]]
	\end{forest}
	\caption{\label{lea} Adnominal Modifiers}
\end{figure}
 

\subsubsection{Semantic composition}


Given the monostratal nature of {\sc hpsg}, semantic representations 
do not constitute a separate level of representation, but take the form 
of attribute value pairs that are added to the syntactic representations.   
Phrase formation and semantic composition are, hence, modeled in tandem.  
Technically, the {\sc content} feature is declared for the same type of objects 
as the {\sc category} feature. 

\begin{exe} 
\ex  \type{local}: \begin{avm} 
                   [cat ~ \type{category} \\
                    content ~ \type{semantic-object}]
                   \end{avm} 
\end{exe} 

\noindent 
For nominals the value of the {\sc content} feature is of type \type{scope-obj(ect)} 
in \citet{GS00}. 
A scope-object is an index-restriction pair in which the index stands for 
entities and in which the restriction is a set of facts which constrain the 
denotation of the index, as in the {\sc content} value of the noun \emph{box}:    

\begin{exe} 
\ex\label{red} 
\begin{avm}
[\type{scope-obj}        \\
 index ~ @1 \type{index} \\
 restr ~ \{[\type{box}   \\
            arg ~ @1 ]\}]
\end{avm} 
\end{exe}

\noindent
This is comparable to the representations which are canonically used in 
Predicate Logic (PL), such as \{x$|$\type{box}(x)\}, where x stands for 
the entities that the predicate \emph{box} applies to. In contrast to 
PL variables, HPSG indices are sorted with respect to person, number 
and gender. This provides the means to model the type of agreement that 
is called index agreement in \crossrefchaptert{agreement}.

\begin{exe} 
\ex  \type{index}: \begin{avm}
                     [person ~ \type{person} \\
                      number ~ \type{number} \\
                      gender ~ \type{gender}] 
                     \end{avm} 
\end{exe} 

{\sc content} values of attributive adjectives are also of type \type{scope-object}. 
When combined with a noun, as in \emph{red box}, the resulting representation 
is one in which the indices of the adjective and the noun are identical, as in 
(\ref{redbox}).\footnote{This is an example of intersective modification. 
The semantic contribution of other types of adjectives, such as  
\emph{alleged} and \emph{fake}, are modeled differently \citep[330--331]{ps2}.}   

\begin{exe} 
\ex\label{redbox} 
\begin{avm}
[\type{scope-obj}         \\
 index ~ @1                \\
 restr ~ \{ [\type{red}   \\
             arg ~ @1 ],
            [\type{box}   \\
             arg ~ @1 ]\}]
\end{avm}
\end{exe}

\noindent
Also this is comparable to the PL practice of representing such 
combinations with one variable to which both predicates apply, as in 
\{x$|$\type{red}(x) \& \type{box}(x)\}. What triggers the index sharing is 
the {\sc mod(ified)} value of the adjective, as illustrated by the {\sc avm} of 
\emph{red} in (\ref{reddd}) \citep[55]{ps2}.\footnote{\avmbox{$\Sigma$} 
stands for an object of type \type{set}, as in \citet{GS00}.} 

\begin{exe} 
\ex\label{reddd}
\begin{avm}
[cat|head [\type{adjective}                               \\
           mod|loc|content [\type{scope-obj}            \\
                            index ~ @1                   \\
                            restr ~ \avmbox{$\Sigma$} ]] \\
 content [index ~ @1                                     \\
          restr ~ \{[\type{red}                         \\
                     arg ~ @1 ]\} ~ $\bigcup$ ~ \avmbox{$\Sigma$} ]]
\end{avm}
\end{exe}

\noindent
The adjective selects a scope-object, shares its index and adds its own 
restriction to those that are already present. The resulting {\sc content} 
value is then shared with the mother.

To model the semantic contribution of determiners, \citet{GS00} 
makes a distinction between scope-objects that contain a quantifier 
(\type{quant-rel}), and those that do not (\type{parameter}). 
In terms of this distinction, the 
addition of a determiner to a nominal, as in \emph{every red box}, 
triggers a shift from \type{parameter} to \type{quant-rel}. 
To capture this the specifier treatment employs the feature {\sc spec(ified)}. 
It is part of the {\sc head} value of the determiners, and its value is of type 
\type{semantic-obj(ect)}.\footnote{In \citet[45]{ps2} the {\sc spec(ified)}
feature was also assigned to other function words, such as complementizers, 
and its value was of type \type{synsem}.}   

\begin{exe} 
\ex   \type{determiner}: \begin{avm} [spec ~ \type{semantic-obj}] \end{avm}  
\end{exe} 

\noindent
In the case of \emph{every}, the {\sc spec} value is an object of 
type \type{parameter}, but its own {\sc content} value is a subtype of 
\type{quant-rel} and this quantifier is put in store, to be retrieved 
at the place where its scope is determined, as illustrated by the AVM
of \emph{every} in (\ref{every}) \citep[204]{GS00}.  

\begin{exe} 
\ex\label{every} 
\begin{avm}
[cat|head [\type{det}               \\
           spec [\type{parameter}   \\
                 index ~ @1          \\
                 restr ~ \avmbox{$\Sigma$} ]] \\
 content ~ @2 [\type{every-rel}     \\
               index ~ @1            \\
               restr ~ \avmbox{$\Sigma$} ] \\
 store ~ \{ @2 \}]
\end{avm}
\end{exe}

\noindent 
Notice that the addition of the {\sc spec} feature yields an analysis in which the determiner 
and the nominal select each other: The nominal selects 
its specifier by means of the valence feature {\sc spr} and the determiner selects the nominal 
by means of {\sc spec}.  


\subsubsection{Nominals with a phrasal specifier} 


Specifiers of nominals tend to be single words, but they can also take the form 
of a phrase. The bracketed phrase in [\emph{the Queen of England's}] \emph{sister},
for instance, is in complementary distribution with the possessive
determiner in \emph{her sister} and has a comparable semantic contribution.   
For this reason it is treated along the same lines. More specifically, the 
possessive marker \emph{'s} is treated as a determiner that takes an NP as its specifier, 
as shown in Figure \ref{cousin} \citep[51--54]{ps2} and \citep[193]{GS00}.\footnote{The treatment 
of the phonologically reduced \emph{'s} as the head of a phrase is comparable to 
the treatment of the homophonous word in \emph{he's ill} as the head of a VP.
Notice that the possessive \emph{'s} is not a genitive affix, for if it were, it 
would be affixed to the head noun \emph{Queen}, as in *\emph{the Queen's of England sister}, see  
\citet[199]{SagWasow03}.}

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{noun}, {\sc spr} ~ \eliste]}
	[{\avmbox{2} [{\sc head} ~ \avmbox{3} \type{det}, {\sc spr} ~ \eliste]}
		[{\avmbox{4} [{\sc head} ~ \type{noun}, {\sc spr} ~ \eliste]} [the Queen of England, roof]]
		[{[{\sc head} ~ \avmbox{3}, {\sc spr} ~ \liste{\avmbox{4}}]} ['s]]]
	[{[{\sc head} ~ \avmbox{1}, {\sc spr} ~ \liste{\avmbox{2}}]} [sister]]]
	\end{forest}
	\caption{\label{cousin} Phrasal Specifiers }  
\end{figure}

In this anaysis the specifier of \emph{sister} is a DetP that is headed by \emph{'s} 
and the latter takes the NP \emph{The Queen of England} as its specifier.\footnote{Since the 
specifier of \emph{'s} is an NP, it may in turn contain a specifier that is headed 
by \emph{'s}, as in \emph{John's uncle's car}.}
Semantically, \emph{'s} relates the index of its specifier (the possessor) to the index
of the nominal that it selects (the possessed), as spelled out in (\ref{poss}).\footnote{The
terms `possessor' and `possessed' are meant to be understood in a broad not-too-literal sense.}     

\begin{exe} 
\ex\label{poss}
\begin{avm}
[cat [head [\type{det}                       \\
            spec [\type{parameter}           \\
                  index ~ @1                  \\
                  restr ~ \avmbox{$\Sigma$} ]] \\
      spr ~ <[index ~ @3]> ]                  \\
 content ~ @2 [\type{the-rel}                \\
               index ~ @1                     \\
               restr ~ \{[\type{poss-rel}    \\
                          possessor ~ @3      \\
                          possessed ~ @1 ]\} ~ $\bigcup$ ~ \avmbox{$\Sigma$} ] \\
 store ~ \{ @2 \}]  
\end{avm}
\end{exe}

\noindent
The assignment of \type{the-rel} as the {\sc content} value captures 
the definiteness of the resulting NP. Notice that this analysis contains a DetP, 
but in spite of that, it is not an instance of the DP approach, since the 
determiner does not head the nominal, but only its specifier. 


\subsection{The functor treatment} 


The functor treatment also adopts the NP approach, but --in contrast to the 
specifier treatment-- it does not draw a distinction between specifiers and adjuncts, 
nor between lexical (or substantive) and functional categories.\footnote{The term \emph{functor} 
is also used in Categorial (Unification) Grammar, where it has 
a very broad meaning, subsuming the nonhead daughter in combinations of a 
head with a specifier or an adjunct, and the head daughter otherwise, 
see \citet{Bouma88}. This broad notion is also adopted in 
\citet{Reape94}. We adopt a more restrictive version in which functors 
are nonhead daughters which lexically select their head sister.}  
The presentation in this section is mainly based on \citet{VanEynde06} 
and \citet{Allegranza06}. We first introduce the basics (Section~\ref{sec-basics}) and then 
turn to nominals with a phrasal specifier (Section~\ref{sec-phrasal-spec}) and to nominals without specifier (Section~\ref{sec-without-spec}).    


\subsubsection{Basics} 
\label{sec-basics}


The distinction between specifiers and adjuncts is usually motivated by 
the assumption that the former are obligatory and non-stackable, while the latter  
are optional and stackable. In practice, though, this distinction 
is blurred by the fact that many nominals are well-formed without specifier.
To accommodate this one can employ phonetically empty determiners, optionality in 
the {\sc spr} list or a non-branching phrase type, as in 2.1.4,   
but the functor treatment provides a more radical response and abandons 
the distinction between specifiers and adjuncts. 
Technically, this implies that the {\sc spr} feature is dropped.\footnote{Intriguingly,
Chomsky has recently argued that there is no need for the notion of specifier in 
Transformational Grammar either, see \citet[43]{Chomsky13}.}  

Besides, the functor treatment abandons the distinction between lexical and 
functional categories. The words which are commonly treated as determiners
are not treated as members of a separate functional category `Det', but are 
claimed to belong to independently needed lexical categories, such as Adjective, 
Adverb, Pronoun and Noun. The argumentation is mainly --but not only-- based on 
matters of NP-internal agreement and inflectional variation \citep{VanEynde06}. 
In Dutch, for instance, prenominal adjectives show agreement with 
the nouns they modify: They take the affix \emph{-e} in combination with plural 
and singular non-neuter nominals, but not in combination with singular 
neuter nominals.\footnote{If the adjective is preceded by a definite determiner, 
it also takes the affix in singular neuter nominals. This phenomenon is treated 
in Section 2.2.2.} 

\begin{exe} 
\ex\label{wit} 
\gll  zwarte muren,        zwarte verf,              zwart zand \\
      black wall.{\sc pl}, black paint.{\sc sg.fem}, black sand.{\sc sg.neu} \\
\end{exe} 

\noindent
The same holds for several of the words that are claimed to be determiners, 
such as the possessive \emph{ons} `our' and the interrogative \emph{welk} `which'. 

\begin{exe} 
\ex\label{ons}
\gll onze ouders,         onze muur,             ons huis     \\
     our parent.{\sc pl}, our wall.{\sc sg.mas}, our house.{\sc sg.neu} \\
\ex\label{welk} 
\gll welke boeken,        welke man,              welk boek   \\
     which book.{\sc pl}, which man.{\sc sg.mas}, which book.{\sc sg.neu} \\
\end{exe} 

\noindent
By contrast, prenominal (pro)nouns do not show NP-internal agreement and  
never take the \emph{-e} affix. This not only holds for genitive nouns, 
such as \emph{Jans}, but also for several of the words that are claimed 
to be determiners, such as the interrogative \emph{wiens} `whose' 
and the quantifying \emph{wat} `some'. 

\begin{exe} 
\ex\label{jans}
\gll  Jans ouders,                   Jans fiets,                      Jans huis \\ 
      Jan.{\sc gen} parent.{\sc pl}, Jan.{\sc gen} bike.{\sc sg.mas}, Jan.{\sc gen} house.{\sc sg.neu} \\
\ex\label{wiens}
\gll  wiens ouders,          wiens muur,              wiens huis \\ 
      whose parent.{\sc pl}, whose wall.{\sc sg.mas}, whose house.{\sc sg.neu} \\
\ex\label{wat}
\gll  wat boeken,         wat verf,                wat zand  \\
      some book.{\sc pl}, some paint.{\sc sg.fem}, some sand.{\sc sg.neu} \\
\end{exe} 

\noindent
To model this the functor treatment assigns adjectival status 
to determiners like \emph{ons} `our' and \emph{welk} `which', 
and pronominal status to determiners like \emph{wiens} `whose' and
\emph{wat} `some'.  
This distinction is also relevant for other languages. The Italian 
possessives of the first and second person, for instance, 
show the same alternation for number and gender as the adjectives
and are subject to the same constraints on NP-internal agreement.

\begin{exe}  
\ex 
\gll  il nostro futuro, la nostra scuola, i nostri genitori, le nostre scatole \\
      the our future.{\sc sg.mas}, the our school.{\sc sg.fem}, the our parent.{\sc pl.mas}, the our box.{\sc pl.fem} \\ 
\end{exe} 

\noindent
The possessive of the third person plural, by contrast, does not show any 
inflectional variation and does not show agreement with the 
noun.\footnote{Confirming evidence for the pronominal status of \emph{loro} is 
provided by the fact that it is also used as a personal pronoun, as in 
\emph{l'ho dato a loro} `I gave it to them'.}

\begin{exe} 
\ex 
\gll  il loro futuro, la loro scuola, i loro genitori, le loro scatole \\   
      the their future.{\sc sg.mas}, the their school.{\sc sg.fem}, the their parent.{\sc pl.mas}, the their box.{\sc pl.fem} \\ 
\end{exe}

\noindent
An example from French is provided in \citet{Abeilleetal04}, 
which assigns adverbial status to the specifier in 
\emph{beaucoup de farine} `much flour'.  
Technically, the elimination of the distinction between lexical and 
functional categories implies that there is no longer any need for 
drawing a distinction between the selection features {\sc mod(ified)} and {\sc spec(ified)}. 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{headed-phr}
	[\type{head-arg-phr}
		[\type{head-subj-phr}]
		[\type{head-comps-phr}]
		[...]]
	[\type{head-nonarg-phr}
		[\type{head-functor-phr}]
		[\type{head-indep-phr}]]]	
	\end{forest}
	\caption{\label{typ} Hierarchy of Headed Phrases}
\end{figure}

To spell out the functor treatment we start from the 
hierarchy of headed phrases in Figure \ref{typ}. The basic distinction is
that between \type{head-argument-phr} and \type{head-nonargument-phr}. 
In the former the head daughter selects its non-head sister(s) by means of 
valence features, such as {\sc comps} and {\sc subj} (but not {\sc spr}!), 
and it is their values that register the degree of saturation of the phrase, 
as shown for {\sc comps} in Section 2.1.1.  
In head-nonargument phrases the degree of saturation is registered  
by the {\sc mark(ing)} feature. It is declared for objects of type \type{category}, 
along with the {\sc head} and valence features.\footnote{The {\sc marking} feature  
is introduced in \citet[46]{ps2} to model the combination of a complementizer 
and a clause.} Its value is shared with the head daughter in head-argument phrases
and with the non-head daughter in head-nonargument phrases, as spelled out in 
(\ref{mark1}) and (\ref{mark2}) respectively. 

\begin{exe}
\ex\label{mark1} 
\begin{avm}
[\type{head-arg-phr}                      \\
 synsem|loc|cat|mark ~ @1 \type{marking}  \\
 head-dtr|synsem|loc|cat|mark ~ @1] 
\end{avm}
\ex\label{mark2} 
\begin{avm}
[\type{head-nonarg-phr}                    \\
 synsem|loc|cat|mark ~ @1 \type{marking}   \\
 dtrs ~ <[synsem|loc|cat|mark ~ @1] ~, @2 > \\
 head-dtr ~ @2 \type{sign}]
\end{avm}
\end{exe}

\noindent
Besides, instead of {\sc mod} and {\sc spec}, the functor treatment employs 
one feature to model the selection of a head by its non-head sister. It is   
called {\sc sel(ect)}, and its value is an object of type \type{synsem} that is 
shared with the head daughter, as spelled out in (\ref{hefu}).  

\begin{exe}
\ex\label{hefu} 
\begin{avm}
[\type{head-functor-phr}                     \\
 dtrs ~ <[synsem|loc|cat|head|sel ~ @1]~, @2 \type{sign}> \\
 head-dtr ~ @2 [synsem ~ @1 \type{synsem}]]
\end{avm}
\end{exe} 

\noindent
This is a subtype of \type{head-nonargument-phr}. It subsumes 
the phrases in which the non-head daughter selects its head sister.
As such it contrasts with phrases of type  
\type{head-independent-phr}, whose defining characteristic is 
that the nonhead daughter does not select its head sister: Its {\sc sel} value 
is hence \type{none}.\footnote{This type is introduced in 
\citet[130]{VanEynde98a}. It will be used in Section 3 to deal with 
idiosyncratic nominals.}   

\begin{exe}
\ex\label{hein} 
\begin{avm}
[\type{head-independent-phr}                           \\
 dtrs ~ <[synsem|loc|cat|head|sel ~ \type{none}]~, @1> \\
 head-dtr ~ @1 \type{sign}]
\end{avm}
\end{exe}    

An illustration of the functor treatment is given in Figure \ref{markyy}. 
The combination of the noun with the adjective is an instance of \type{head-functor-phr}, 
in which the adjectival functor selects an unmarked nominal ($\avmbox{3}$),  
shares its {\sc marking} value ($\avmbox{5}$), and, being a non-argument, 
shares it with the mother as well. 
The combination of the resulting nominal with the demonstrative is also 
an instance of \type{head-functor-phr},
in which the demonstrative functor selects an unmarked nominal ($\avmbox{4}$), 
but --differently from the adjective-- its {\sc marking} value is of type 
\type{marked}, and this value is shared with the mother ($\avmbox{2}$).    
This accounts for the fact that adjectives can be stacked, while  
determiners cannot. It also accounts for the ill-formedness of combinations like 
*\emph{long the bridge}, since attributive adjectives are not compatible with a
marked nominal.  
Notice that the demonstrative does not belong to a functional part of speech, such as Det.
Instead, it is claimed to be a pronoun. The only difference between adnominal \emph{that} and nominal
\emph{that} concerns the {\sc sel(ect)} value: While the former selects an unmarked singular nominal, 
the latter does not select anything.      

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} [\type{noun} {\sc sel} ~ \type{none}], {\sc mark} ~ \avmbox{2} \type{marked}]}
	[{[{\sc head} ~ [\type{pron} {\sc sel} ~ \avmbox{4}], {\sc mark} ~ \avmbox{2}]} [that]]
	[{\avmbox{4} [{\sc head} ~ \avmbox{1}, {\sc mark} ~ \avmbox{5} \type{unmarked}]}
		[{[{\sc head} ~ [\type{adj} {\sc sel} ~ \avmbox{3}], {\sc mark} ~ \avmbox{5}]} [long]]
		[{\avmbox{3} [{\sc head} ~ \avmbox{1}, {\sc mark} ~ \avmbox{5}]} [bridge]]]]
	\end{forest}
	\caption{\label{markyy} Adnominal functors}
\end{figure}

Whether an adnominal functor is marked or unmarked is subject to cross-linguistic variation. 
The Italian possessives, for instance, are unmarked and can, hence, be preceded 
by an article, as in \emph{il suo cane} `the his dog' and \emph{un mio amico} `a my friend', 
while the French possessives are marked: \emph{(*le) mon chien} `(*the) my dog' 
and \emph{(*un) mon ami} `(*a) my friend'. 

The treatment of postnominal dependents is similar to that in the specifier treatment. 
A relational noun like \emph{sister}, for instance, takes an optional PP[\type{of}] complement.  
Similarly, a deverbal noun like \emph{description} takes two optional PP complements. 
If realized, as in \emph{a description of the Hungarian NP by Ivan}, 
their indices are identified with those of the second and the first argument respectively. 
There is a difference, though, in the treatment of nominals 
whose first argument is realized as a possessive, as in 
\emph{his description of the Hungarian NP}. In this case, the first argument is not put on the 
{\sc spr} list of the noun. Instead, the possessive selects an unmarked nominal with an optional 
PP[\type{by}] complement on its {\sc comps} list, and shares the index of that PP, 
as spelled out in Figure \ref{possy}. 

\begin{figure}
	\centering
\oneline{%
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} [\type{noun} {\sc sel} ~ \type{none}], {\sc mark} ~ \avmbox{2} \type{marked}]}
	[{[{\sc head} ~ [\type{pron} {\sc sel} ~ \avmbox{4}], {\sc mark} ~ \avmbox{2}]$_{i}$} [his]]
	[{\avmbox{4} [{\sc head} ~ \avmbox{1}, {\sc comps} ~ \liste{(PP[\type{by}]$_{i}$)}, {\sc mark} ~ \avmbox{5} \type{unmarked}]}
		[{[{\sc head} ~ \avmbox{1}, {\sc comps} ~ \liste{ (PP[\type{by}]$_{i}$), \avmbox{3}}, {\sc mark} ~ \avmbox{5}]} [description]]
		[{\avmbox{3} [{\sc head} ~ \type{prep}, {\sc comps} ~ \eliste]} [of the Hungarian NP, roof]]]]
	\end{forest}
}
	\caption{\label{possy} Deverbal nominals }
\end{figure}


        
\subsubsection{Nominals with a phrasal functor} 
\label{sec-phrasal-spec}


To illustrate how the functor treatment deals with phrasal specifiers we 
take the nominal \emph{a hundred pages}. It has a left branching structure in
which the indefinite article selects the unmarked singular noun \emph{hundred} --its plural 
counterpart is \emph{hundreds}--  
and in which the resulting NP selects the unmarked plural noun 
\emph{pages}, as spelled out in Figure \ref{glorie}. The  
{\sc marking} value of the article is shared with its mother ($\avmbox{1}$), 
yielding a marked NP, and since the latter is a functor it is also shared with 
the NP as a whole ($\avmbox{1}$). 
This accounts for the fact that it cannot be preceded by a determiner, 
as in *\emph{those a hundred pages}. By contrast, if the place of the article is taken 
by a numeral, as in \emph{two hundred pages}, the addition of a determiner is 
possible, as in \emph{those two hundred pages}, since the numeral is unmarked. 
Notice that the {\sc head} value of the entire NP is identified with that 
of \emph{pages}. This accounts for the fact that it is plural.  
 
\begin{figure}
	\centering
\oneline{%
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{5} [\type{noun} {\sc sel} ~ \type{none}], {\sc mark} ~ \avmbox{1}]}
	[{[{\sc head} ~ \avmbox{4} [\type{noun} {\sc sel} ~ \avmbox{2}], {\sc mark} ~ \avmbox{1}]}
		[{[{\sc head} ~ [\type{pron} {\sc sel} ~ \avmbox{3}], {\sc mark} ~ \avmbox{1} \type{marked}]} [a]]
		[{\avmbox{3} [{\sc head} ~ \avmbox{4}, {\sc mark} ~ \type{unmarked}]} [hundred]]]
	[{\avmbox{2} [{\sc head} ~ \avmbox{5}, {\sc mark} ~ \type{unmarked}]} [pages]]]
	\end{forest}
}
	\caption{\label{glorie} Phrasal functors }
\end{figure}


\subsubsection{The hierarchy of MARKING values} 
\label{sec-without-spec}

Nominals without specifier have a {\sc marking} value of type \type{unmarked}. 
Whether they need to be marked, depends on a multitude of factors and is 
largely language specific. 
Latin, for instance, does not require its nominals to contain a specifier, also 
if they are singular and count. Yet, there are cases in which an unmarked nominal 
is incomplete.    
To illustrate this let us take the attributive adjectives of Dutch again. 
As already pointed out in Section 2.2.1, they take the form without affix in  
singular neuter nominals, as in \emph{zwart zand} `black sand'. A complication, 
though, is that they canonically take the form with the affix if the nominal is  
introduced by a definite determiner, as in \emph{het zwarte zand} `the black sand'. 
This has consequences for the status of nominals with a singular neuter head: 
\emph{zwart zand} and *\emph{zwarte zand}, for instance, are both unmarked, 
but while the former is well-formed as it is, the latter is only 
well-formed if it is preceded by a definite determiner. 
To model this \citet{VanEynde06} differentiates between two types 
of \type{unmarked} nominals, as shown in Figure \ref{bare}. 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{marking}
	[\type{unmarked}
		[\type{incomplete}]
		[\type{bare}]]
	[\type{marked}]]		
	\end{forest}
	\caption{\label{bare} Hierarchy of Marking values} 
\end{figure}

Employing the more specific subtypes, the adjectives without affix which select a singular 
neuter nominal have the value \type{bare}, while their declined counterparts which select 
a singular neuter nominal have the value \type{incomplete}. 
Since this {\sc marking} value is shared with the mother, the {\sc marking} value 
of \emph{zwart zand} is \type{bare}, while that of *\emph{zwarte zand} is \type{incomplete}. 
The fact that the latter must be preceded by a definite determiner
is modeled in the {\sc sel(ect)} value of the determiner: 
While definite determiners select an unmarked nominal, which implies that 
they are compatible with both bare and incomplete nominals,
non-definite determiners select a bare nominal and are, hence, not compatible 
with an incomplete one, as in *\emph{een zwarte huis} `a black house'. 
The {\sc marking} feature is, hence, useful to differentiate bare 
nominals from incomplete nominals.  
In a similar way, one can make finer-grained distinctions in the hierarchy of  
\type{marked} values to capture co-occurrence restrictions between determiners and 
nominals, as in the functor treatment of the Italian determiner system of 
\citet{Allegranza06}. See also the treatment of nominals with idiosyncratic properties 
in Section 3. 


\subsection{The DP treatment} 


In contrast to the specifier treatment and the functor treatment, the DP 
treatment identifies the determiner as the head of an NP. The presentation in this section   
is mainly based on \citet{Netter94}. We first sketch its main characteristics (2.3.1)
and then list some problems for it (2.3.2). 


\subsubsection{Two types of complementation} 


\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{det}, {\sc comps} ~ \eliste]}
	[{[{\sc head} ~ \avmbox{1}, {\sc comps} ~ \liste{\avmbox{4} NP}]} [that]]
	[{\avmbox{4} [{\sc head} ~ \avmbox{2} \type{noun}, {\sc comps} ~ \eliste]}
		[{[{\sc head} ~ \avmbox{2}, {\sc comps} ~ \liste{\avmbox{3} PP}]} [sister]]
		[\avmbox{3} [of Leslie, roof]]]]
	\end{forest}
	\caption{\label{net} Lexicalist DP Treatment }
\end{figure}

As in the transformational DP treatment, nominal projections are divided in a nominal core, 
consisting of a noun and its complements and/or adjuncts, on the one hand, and an external 
functional layer, comprising the determiner, on the other hand. The former is analyzed in much the 
same way as in the specifier treatment. The relational noun   
\emph{sister} in Figure \ref{net}, for instance, is combined with 
its PP complement in the usual way. The addition of the determiner, by contrast, is 
modeled differently. It is not the nominal which selects its specifier by means of the valence 
feature {\sc spr}; instead, it is the determiner which selects a nominal by means of 
the valence feature {\sc comps}, yielding a DP with an empty {\sc comps} list. 
To account for the fact that this DP inherits many of its properties from its nominal 
non-head daughter, Netter makes a distinction between functional 
complementation and ordinary complementation, and differentiates between major and minor 
{\sc head} features: 

\begin{exe} 
\ex \type{part-of-speech}:\\
 \begin{avm}
                              [major [n ~ \type{boolean} \\
                                      v ~ \type{boolean} ] \\
                               minor [fcompl ~ \type{boolean}]]
                              \end{avm} 
\end{exe} 

\noindent
The {\sc major} attribute includes the boolean features N and V, where 
nouns are [+N, --V], adjectives [+N, +V], verbs [--N, +V] and prepositions [--N, --V]. 
Besides, [+N] categories also have the features {\sc case}, {\sc number} and {\sc gender}. 
Typical of functional complementation is that the functional head shares the 
{\sc major} value of its complement \citep[311--312]{Netter94}. 

\begin{exe} 
\ex\label{maj} In a lexical category of type \type{func-cat} the value of its {\sc major} 
      attribute is token identical with the {\sc major} value of its complement. 
\end{exe} 

\noindent
Since a determiner shares the {\sc major} value of its nominal complement and since this value is 
also shared with the DP (given the Head Feature Principle), it follows that the resulting 
DP is [+N,--V] and that its {\sc case}, {\sc number} and {\sc gender} values are 
identical to those of its nominal non-head daughter. 
This differentiates functional complementation from ordinary complementation, where 
a head and its complement do not share their {\sc major} value. The noun \emph{sister} 
in Figure \ref{net}, for instance, does not share the part-of-speech of its PP complement. 

The {\sc minor} attribute is used to model properties which a functional head does {\em not} share with its 
complement. It includes {\sc fcompl}, a feature which registers whether a projection is 
functionally complete or not. Its value is positive for determiners, negative for  
singular count nouns and underspecified for plurals and mass nouns.  
Determiners take a nominal complement with a negative {\sc fcompl} value, but their 
own {\sc fcompl} value is positive and since they are the head, they share this value with 
the mother, as in Figure \ref{netter}. 
In this analysis, a nominal is complete, if it is both saturated 
(empty {\sc comps} list) and functionally complete (positive {\sc fcompl}). 

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc major} ~ \avmbox{1} [--V, +N[\type{sing}]], {\sc minor$|$fcompl} ~ +]}
		[{[{\sc major} ~ \avmbox{1}, {\sc minor$|$fcompl} ~ +]} [that]]
		[{[{\sc major} ~ \avmbox{1}, {\sc minor$|$fcompl} ~ --]} [sister of Leslie, roof]]]
	\end{forest}
	\caption{\label{netter} Functional Completeness }
\end{figure}


\subsubsection{Two problems for the DP treatment}  


A problem for the DP treatment concerns the notion of 
functional complementation, as defined in (\ref{maj}). 
If determiners share the {\sc major} value of the nominals which they select, 
then it follows that they are nominal themselves, i.e. [+N, --V].
However, while this makes sense for determiners with (pro)nominal properties,
such as the English \emph{that}, the Dutch \emph{wat} `some' and the Italian 
\emph{loro} `their', it is rather implausible for determiners with adjectival properties,
such as the Italian \emph{mio} `my', the Dutch \emph{ons} `our' and the German
\emph{welch} `which'.   
The problem also affects the associated agreement features, i.e. {\sc case}, 
{\sc number} and {\sc gender}. If a determiner 
is required to share the values of these features with its nominal complement,
as spelled out in (\ref{maj}), then we get implausible results for nominals in 
which the determiner and the noun do not show agreement.    
In the Dutch \emph{wiens huizen} `whose houses' and 
\emph{'s lands hoogste bergen} `the highest mountains of the country', for instance,  
the selected nominals are plural and non-genitive, and so are the entire DPs, but 
the selecting determiner (phrase) is singular and genitive. 

Another problem concerns the assumption ``that all substantive categories will 
require the complement they combine with to be both saturated and 
functionally complete'' \citep[311]{Netter94}. Complements of verbs and 
prepositions must, hence, be positively specified for {\sc fcompl}. 
This, however, is contradicted by the existence of 
prepositions which require their complement to be functionally incomplete. 
The German prepositions \emph{am, im, vom, beim} and \emph{zum}, for instance, 
which diachronically result from the incorporation of a dative form of the 
definite article (\emph{dem}) in respectively \emph{an} `at', 
\emph{in} `in', \emph{von} `of', \emph{bei} `at' and \emph{zu} `to',
count as syntactic atoms in {\sc hpsg}, just like 
the forms without incorporated article, but in contrast to the 
latter they require their complement to lack a determiner: While  
\emph{in dem/diesem Zimmer} `in the/this room' and \emph{im Zimmer} `in.the room'
are both well-formed, *\emph{im dem/diesem Zimmer} `in.the the/this room' is 
not.\footnote{A similar phenomenon exists in Italian, where the prepositions 
\emph{a} `at', \emph{da} `from', \emph{su} `on', \emph{di} `of' and \emph{in} `in' 
have alternate forms with an incorporated form of the definite article, as in  
\emph{al, dal, sul, del} and \emph{nel}, each with 
feminine and plural counterparts (\emph{alla, agli, alle, dalla, dagli, dalle, ...}). 
The nominals which combine with such forms may not be introduced by a determiner:
\emph{di/*della questa scatola} `of/*of.the this box' and \emph{in/*nel ogni palazzo} 
`in/*in.the every palace'. The same holds m.m. for the French 
portemanteau forms \emph{du}, \emph{des}, \emph{au} and \emph{aux}.}   
Moreover, there are prepositions which do not contain an 
incorporated article, but nonetheless require a determinerless nominal. 
The Dutch \emph{te} and \emph{per}, for instance, are not compatible with nominals 
that contain a determiner, even if the nominal is singular and count, 
as in \emph{te (*het/een) paard} `on horse' and \emph{per (*de/een) trein} `by train'.   
This must be due to a lexical property of these prepositions, 
since most other prepositions require such nominals to have 
a determiner, as in \emph{ze viel van *(het/een) paard} `she fell from *(the/a) horse' 
and \emph{ze zit op *(de/een) trein naar Londen} `she is on *(the/a) train to London'.  
The requirement that complements of prepositions must be functionally complete is,
hence, untenable.  


\subsection{Conclusion} 


This section has presented the three main treatments of nominal structure in HPSG. 
Two adopt the NP approach and one the DP approach.
Overall, the former turn out to be more amenable to integration  
in a monostratal surface-oriented framework like HPSG than the latter. 
See also \citet{MuellerHeadless}. Of the two NP treatments
the specifier treatment is closer to early versions of \xbar syntax and to GPSG.   
The functor treatment is closer to versions of Categorial (Unification) Grammar, and 
has also been adopted in Sign-Based Construction Grammar \citep{Sag2012}.
 

\section{Idiosyncratic nominals}


This section focusses on the analysis of nominals with idiosyncratic properties. 
Since their analysis often requires 
a relaxation of the strictly lexicalist approach of early HPSG we first introduce some basic 
notions of constructional HPSG. 

The lexicalist approach of early HPSG can be characterized as one in which the 
properties of phrases are mainly determined by properties of the constituent words 
and only to a small extent by properties of the combinatory operations as such. 
\citet[391]{ps2}, for instance, employs no more than seven types 
of combinatory operations, including those which were exemplified in Section 2.1, 
i.e. head-specifier, head-complements and head-adjunct.\footnote{The remaining four 
are head-subject, head-subject-complements, head-marker and head-filler.}   
Over time, though, this radical lexicalism gave way to an 
approach in which the properties of the combinatory operations  
play a larger role. The small inventory of highly 
abstract phrase types got replaced by a more fine-grained hierarchy  
in which the types contain more specific and --if need be-- idiosyncratic 
constraints. This development started in \citet{Sag97}, was 
elaborated in \citet{GS00}, and gained momentum afterward,
leading to what is now known as constructional HPSG. 
Characteristic of this approach is the use of a bi-dimensional hierarchy 
of phrasal signs. In such a hierarchy the phrases are not only partitioned 
in terms of {\sc headedness}, but also in terms of a second dimension, called  
{\sc clausality}, as in Figure \ref{bidim}. 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{phrase}
	[{\sc headedness}
		[\type{headed-phr}
			[\type{head-subj-phr}, name=A]
			[...]]
		[\type{non-headed-phr}]]
	[{\sc clausality}
		[\type{clause}
			[\type{declarative-cl} [\type{decl-head-subj-cl}, name=B]]
			[...]]
		[\type{non-clause}]]]
\draw (A.south) -- (B.north);
	\end{forest}
	\caption{\label{bidim} Bi-dimensional Hierarchy of Clauses}  
\end{figure}

The types in the {\sc clausality} dimension are associated with constraints,
in much the same way as the types in the {\sc headedness} dimension.  
Clauses, for instance, are required to denote an object of type \type{message} 
\citep[41]{GS00}.

\begin{exe}
\ex 
\begin{avm} 
[\type{clause} \\
 synsem|loc|content ~ \type{message}] 
\end{avm}
\end{exe}

\noindent
At a finer-grained level, the clauses are partitioned into 
declarative, interrogative, imperative, exclamative and relative
clauses, each with their own constraints. 
Interrogative clauses, for instance, denote a question, 
while indicative declarative clauses denote a proposition.

Exploiting the possibilities of multiple inheritance one can 
define types which inherit properties from more than one supertype. 
The type \type{decl(arative)-head-subj(ect)-cl(ause)}, for instance, inherits 
the properties of \type{head-subject-phr}, on the one hand, and 
\type{declarative-cl}, on the other hand. Besides, it may 
have properties of its own, such as the fact that its head daughter 
is a finite non-inverted verb \citep[43]{GS00}. 
This combination of multiple inheritance and specific   
constraints on maximal phrase types is also useful for the analysis of 
nominals with idiosyncratic properties, as will be shown presently. 


\subsection{The Big Mess Construction} 


In ordinary nominals determiners precede attributive adjectives. Changing the order 
yields ill-formed combinations, such as *\emph{long that bridge} and *\emph{very tall every man}. 
However, this otherwise illegitimate order is precisely what we find in 
the Big Mess Construction (BMC), a term coined by \citet{Berman74}.  

\begin{exe}
\ex\label{bime}
\begin{xlist}
\ex   It's \emph{so good a bargain} I can't resist buying it.
\ex   \emph{How serious a problem} is this?
\end{xlist}
\end{exe} 

\noindent
The idiosyncratic order in (\ref{bime}) is required if the nominal is introduced 
by the indefinite article, and if the preceding AP is introduced by one of a small 
set of degree markers, including \emph{so, as, how, this, that} and \emph{too}. 

A specifier treatment of the BMC is provided in \citet[201]{GS00}. It adopts  
a left branching structure, as in  [[[\emph{so good}] \emph{a}] \emph{bargain}], 
in which \emph{so good} is the specifier of the indefinite article and in which 
\emph{so good a} is the specifier of \emph{bargain}. This is comparable to the 
treatment of the possessive in [[[\emph{the Queen of England}] \emph{'s}] \emph{sister}],
see Section 2.1.3.  
However, while there is evidence that \emph{the Queen of England's} is a constituent,
since it may occur independently as in \emph{this crown is the Queen of England's}, there is 
no evidence that \emph{so good a} is a constituent. Instead, there is evidence that 
the article forms a constituent with the following noun, since it also precedes the noun 
when the AP is in postnominal position, as in \emph{we never had a bargain so good as this one}.

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ $\avmbox{3}$ \type{noun}, {\sc mark} ~ $\avmbox{1}$ \type{a}]}
		[{[{\sc mark} ~ $\avmbox{1}$, {\sc sel} ~ $\avmbox{2}$]} [a]]
		[{$\avmbox{2}$ [{\sc head} ~ $\avmbox{3}$, {\sc mark} ~ \type{unmarked}]} [bargain]]]
	\end{forest}
	\caption{\label{aprob} The lower NP }
\end{figure}

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ $\avmbox{3}$ \type{adj}, {\sc mark} ~ $\avmbox{1}$ \type{marked}]}
		[{[{\sc mark} ~ $\avmbox{1}$, {\sc sel} ~ $\avmbox{2}$]} [so]]
		[{$\avmbox{2}$ [{\sc head} ~ $\avmbox{3}$, {\sc mark} ~ \type{unmarked}]} [good]]]
	\end{forest}
	\caption{\label{sohow} The marked AP }
\end{figure}

A structure in which the article sides with the noun, as in [[\emph{so good}] [\emph{a bargain}]], 
is adopted in \citet{VanEynde07}, \citet{KimSells11}, \citet{KaySag12}, 
\citet{ArnoldSadler14} and \citet{VanEynde18}, all of which are functor treatments. 
The structure of the lower NP is spelled out in Figure \ref{aprob}. 
The article has a {\sc marking} value of type \emph{a} which is a subtype of \type{marked} and which it
shares with the mother.\footnote{The {\sc marking} value of the article looks similar to its 
{\sc phonology} value, but it is not the same. The {\sc phonology} values of \emph{a} and \emph{an}, 
for instance, are different, but their {\sc marking} value is not.} 
The AP is also treated as an instance of the head-functor type 
in \citet{VanEynde07}, \citet{KimSells11} and \citet{VanEynde18}. 
The adverb has a {\sc marking} value of type \type{marked}, 
so that the AP is marked as well, as shown in Figure \ref{sohow}.   
In combination with the fact that the article selects an unmarked nominal, 
this accounts for the ill-formedness of (\ref{ster}). 

\begin{exe}
\ex\label{ster}
\begin{xlist}
\ex   [*] {It's a so good bargain I can't resist buying it.} 
\ex   [*] {A how serious problem is it?}   
\end{xlist}
\end{exe}

\noindent
By contrast, adverbs like \emph{very} and \emph{extremely} are unmarked,
so that the APs which they introduce are admissible in this position, as in (\ref{st}).  

\begin{exe}
\ex\label{st}
\begin{xlist}
\ex  This is a very serious problem. 
\ex  We struck an extremely good bargain. 
\end{xlist} 
\end{exe} 

To model the combination of the AP with the lower NP it may at first seem 
plausible to treat the AP as a functor which selects  
an NP that is introduced by the indefinite article. This, however, has 
unwanted consequences: Given that {\sc sel(ect)} is a {\sc head} feature, 
its value is shared between the AP and the adjective, so that the latter 
has the same {\sc sel(ect)} value as the AP, erroneously licensing such 
combinations as *\emph{good a bargain}. To avoid this 
\citet{VanEynde18} exploits the possibilities of 
the bi-dimensional hierarchy of phrase types in Figure \ref{prot}. 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{phrase}
	[{\sc headedness}
		[\type{headed-phr}
			[\type{hd-nonargument-phr}
				[\type{hd-functor-phr} [\type{regular-nominal-phr}, name=B2]]
				[\type{hd-independent-phr}, name=A1]]]]
	[{\sc clausality}
		[\type{non-clause}
			[\type{nominal-parameter}
			[\type{intersective-modification}, name=B1 [\type{big-mess-phr}, name=A2]]]]]]
\draw (A1.south) -- (A2.north);
\draw (B1.south) -- (B2.north);
	\end{forest}
	\caption{ \label{prot} Bi-dimensional Hierarchy of Nominals} 
\end{figure}

The types in the {\sc headedness} dimension are a subset of those in Figure \ref{typ}.  
The types in the {\sc clausality} dimension mainly capture semantic and 
category-specific properties, in analogy with the hierarchy of clausal phrases 
in \citet{GS00}. One of the non-clausal phrase types is \type{nominal-parameter}: 

\begin{exe}
\ex\label{param} 
\begin{avm}
[\type{nominal-parameter}                                             \\
 synsem|loc [cat|head ~ \type{noun}                                   \\
             content [\type{parameter}                                \\
                      index ~ @1 \type{index}                         \\
                      restr ~ \avmbox{$\Sigma_{1}$} ~ $\bigcup$ ~ \avmbox{$\Sigma_{2}$}]] \\
 dtrs ~ <[synsem|loc|content|restr ~ \avmbox{$\Sigma_{1}$} ] ~, @2>     \\
 head-dtr ~ @2 [synsem|loc|content [\type{parameter}                  \\
                                    index ~ @1                         \\
                                    restr ~ \avmbox{$\Sigma_{2}$} ]]]
\end{avm}
\end{exe}

\noindent
The mother shares its index with the head daughter ($\avmbox{1}$) and 
its {\sc restr(iction)} value is the union of the {\sc restr} values 
of the daughters (\avmbox{$\Sigma_{1}$} and \avmbox{$\Sigma_{2}$}). 
In the hierarchy of non-clausal phrases, this type contrasts amongst others with 
the quantified nominals, which have a {\sc content} value of type 
\type{quant-rel} \citep[203--205]{GS00}. A subtype of \type{nominal-parameter} is  
\type{intersective-modification}, as defined in (\ref{mononom}).  

\begin{exe}
\ex\label{mononom} 
\begin{avm}
[\type{intersective-modification}                \\
 synsem|loc|content|index ~ @1 \type{index}      \\
 dtrs ~ <[synsem|loc|content|index ~ @1 ] ~, @2 > \\
 head-dtr ~ @2 \type{sign} ]
\end{avm}
\end{exe}

\noindent 
This constraint requires the mother to share its index also with the 
non-head daughter. It captures the intuition that the 
noun and its non-head sister apply to the same entities, as in 
the case of \emph{red box}.\footnote{Another subtype of \type{nominal-parameter} 
is \type{inverted-predication}, which subsumes amongst others 
the binominal noun phrase construction and certain types of apposition
(see Section 3.4).}  

The maximally specific types inherit properties of one of the types of headed phrases,
on the one hand, and of one of the non-clausal phrase types, on the other hand.  
Regular nominal phrases, for instance, such as \emph{red box}, are subsumed 
by a type, called \type{regular-nominal-phr}, that inherits the 
constraints of \type{head-functor-phrase}, on the one hand, and 
\type{intersective-modification}, on the other hand.  
Another maximal type is \type{big-mess-phrase}. 
Its immediate supertype in the {\sc clausality} hierarchy is the same 
as for the regular nominal phrases, i.e. \type{intersective-modification}, 
but the one in the {\sc headedness} hierarchy is different: 
Being a subtype of \type{head-independent-phrase}, 
its non-head daughter does not select the head-daughter. 
Beside the inherited properties the BMC has some properties of its own.   
They are spelled out in (\ref{bigmess2}).

\begin{exe}
\ex\label{bigmess2} 
\begin{avm}
[\type{big-mess-phr}                                    \\     
 dtrs ~ <[\type{hd-functor-phr}                         \\
          synsem|loc|cat [head ~ \type{adjective}       \\
                          mark ~ \type{marked} ]], @1> \\
 head-dtr ~ @1 [\type{regular-nominal-phr}              \\
                synsem|loc|cat|mark ~ \type{a}]]
\end{avm}
\end{exe}

\noindent
The head daughter is required to be a regular nominal phrase 
whose {\sc marking} value is \type{a}, and the non-head daughter 
is required to be an adjectival head-functor phrase
with a {\sc marking} value of type \type{marked}. 
This licenses APs which are introduced by a marked adverb, 
as in \emph{so good a bargain} and \emph{how serious a problem}, 
while it excludes unmarked APs, as in 
*\emph{good a bargain} and *\emph{very big a house}.
Iterative application is not licensed, since (\ref{bigmess2}) requires the 
head daughter to be of type \type{regular-nominal-phr}, which is incompatible with the type 
\type{big-mess-phrase}. This accounts for the fact that a big
mess phrase cannot contain another big mess phrase, as in
*\emph{that splendid so good a bargain}.
A reviewer remarked that this analysis allows combinations like 
\emph{so big a red expensive house}, suggesting that it should not. 
We are not sure, though, that this combination is ill-formed, 
and are anyway not inclined to exclude the presence of adjectives in 
the lower NP, since that would erroneously block the formation of the 
well-formed combinations in (\ref{shrub}).

\begin{exe} 
\ex\label{shrub} 
\begin{xlist} 
\ex  How big a new shrub from France were you thinking of buying? 
\ex  That's as beautiful a little black dress as I've ever seen.  
\end{xlist} 
\end{exe} 

\noindent
These are quoted from \citet[116]{Zwicky95} and \citet[42]{Troseth09} respectively. 


\subsection{Nominals with a verbal core} 


Ordinary nominals have a nominal core, but there are also nominals  
with a verbal core, such as gerunds and nominalized infinitives. They are 
of special interest, since they figure prominently in the argumentation 
that triggered the shift from the NP approach to the DP approach in Transformational 
Grammar. We first present a specifier treatment of the English gerund and then 
a functor treatment of the Dutch nominalized infinitive. 


\subsubsection{The English gerund: a mixed category} 


\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{part-of-speech}
	[\type{noun}
		[\type{proper-noun}]
		[\type{common-noun}]
		[\type{gerund}, name=A2]]
	[\type{relational}, name=A1
		[\type{verb}]
		[\type{adjective}]]]
\draw (A1.south) -- (A2.north);
	\end{forest}
	\caption{ \label{ger} Gerunds as a Mixed Category }
\end{figure} 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{phrase}
	[{\sc headedness}
		[\type{headed-phr}
			[\type{head-subj-phr} [\type{nonfin-head-subj-cx}, name=A2]]
			[\type{head-spr-phr} [\type{noun-poss-cx}, name=B2]]
			[...]]]
	[{\sc clausality}
		[\type{clause}, name=A1]
		[\type{non-clause}, name=B1]]]
\draw (A1.south) -- (A2.north);
\draw (B1.south) -- (B2.north);
	\end{forest}
	\caption{\label{bido} Bi-dimensional Hierarchy of Gerundial Phrases } 
\end{figure}

Some examples of gerunds are given in (\ref{gen}--\ref{opt}), 
quoted from \citet[1290]{Quirketal85}. 

\begin{exe} 
\ex\label{gen}  [Brown's deftly painting his daughter] is a delight to watch. 
\ex\label{acc}  I dislike [Brown painting his daughter]. 
\ex\label{opt}  Brown is well known for [painting his daughter].
\end{exe}

\noindent
The bracketed phrases have the external distribution of an NP, 
taking the subject position in (\ref{gen}), 
the complement position of a transitive verb in (\ref{acc}) and 
the complement position of a preposition in (\ref{opt}). 
The internal structure of these phrases, though, shows a mixture of nominal and verbal 
characteristics. Typically nominal is the presence of the possessive in (\ref{gen}). 
Typically verbal are the presence of the accusative subject in (\ref{acc}), of the adverbial modifier in 
(\ref{gen}) and of the NP complements in (\ref{gen}--\ref{opt}). 
To model this Rob Malouf treats the gerund as a mixed category, introducing a separate 
part-of-speech for it, which is a subtype of \type{noun}, on the one hand, and 
\type{relational}, on the other hand, see Figure \ref{ger} \citep[65]{Malouf00}. 
The distinctive properties of this mixed category are spelled out in a lexical rule 
which derives gerunds from the homophonous present participles \citep[66]{Malouf00}.\footnote{Boxed 
capitals stand for objects of type \type{list}, as in \citet{GS00}.} 

\begin{exe}
\ex 
\begin{avm} 
[head  & [\type{verb}           \\
          vform ~ \type{prp}] \\
 subj  & <@1 np>               \\
 comps & @A                    \\
 spr   & < ~ > ]
\end{avm} ~ $\Rightarrow$ ~ \begin{avm} 
                            [head  & \type{gerund} \\
                             subj  & <@1>           \\
                             comps & @A             \\
                             spr   & <@1> ] 
                            \end{avm}
\end{exe}

\noindent
The gerunds are claimed to take the same kind of complements 
as the present participles from which they are derived ($\avmbox{A}$), and
the compatibility with adverbial modifiers follows from the 
fact that adverbs typically modify objects of type \type{relational}. 
The availability of different options for realizing the subject is 
captured by the inclusion of the subject requirement of the present 
participle in both the {\sc subj} list and the {\sc spr} list of the gerund
($\avmbox{1}$). To model the two options \citet[15]{Malouf00} employs the  
bi-dimensional hierarchy of phrase types in Figure \ref{bido}. 
The combination with an accusative subject
is subsumed by \type{nonfin-head-subj-cx}, which is a subtype of 
\type{head-subject-phr} and \type{clause}. Its defining properties are 
spelled out in (\ref{acccx}) \citep[16]{Malouf00}. 

\begin{exe}
\ex\label{acccx} 
\begin{avm} 
[\type{nonfin-head-subj-cx}                  \\
 synsem|loc|cat|head|root ~ --                \\
 dtrs ~ <[synsem|loc|cat|head [\type{noun}   \\
                               case ~ \type{acc}]], @1 > \\
 head-dtr ~ @1 ] 
\end{avm}
\end{exe} 

\noindent
This construction type subsumes combinations of a non-finite head with 
an accusative subject, as in (\ref{acc}). When the non-finite head is a gerund, 
the {\sc head} value of the resulting clause is \type{gerund} 
and since that is a subtype of \type{noun}, the clause is also a nominal phrase. 
This accounts for the fact that its external distribution is that of an NP.  
By contrast, the combination with a possessive subject
is subsumed by \type{noun-poss-cx}, which is a subtype of 
\type{head-spr-phr} and \type{non-clause} \citep[16]{Malouf00}.

\begin{exe} 
\ex\label{gencx} 
\begin{avm} 
[\type{noun-poss-cx}                                      \\
 synsem|loc [cat|head ~ \type{noun}                       \\
             content ~ \type{nom-obj}]                    \\
 dtrs ~ <[synsem|loc|cat|head [\type{noun}                \\
                               case ~ \type{gen}]], @1 > \\
 head-dtr ~ @1 ] 
\end{avm}
\end{exe}
 
\noindent
This construction subsumes combinations of a nominal and a 
possessive specifier, as in \emph{Brown's house}.\footnote{Malouf treats 
the English possessive as a genitive, differently from \citet{SagWasow03}.}   
It also subsumes combinations with the gerund, as in (\ref{gen}), since 
\type{gerund} is a subtype of \type{noun}.  

In sum, Malouf's analysis of the gerund involves the postulation of a
mixed type in the part-of-speech hierarchy, which allows the gerund to simultaneously
function as a nominal and a relational category. 


\subsubsection{The Dutch nominalized infinitive: phrasal conversion } 


The closest equivalent of the gerund in Dutch is the nominalized infinitive. 
Some examples are given in (\ref{schenk}--\ref{mobil}). 

\begin{exe} 
\ex\label{schenk} 
\gll   [geld wegschenken] maakt vrouwen gelukkig  \\
       [money donate.{\sc inf}] makes women happy \\
\trans `Donating money makes women happy' 
\ex\label{mobil}   
\gll   voor [het op diepte houden van de Vlaamse kusthavens] dient gebaggerd te worden  \\  
       for [the on depth keep.{\sc inf} of the Flemish coast.ports] needs dredged to be \\
\trans `Dredging is necessary to keep the Flemish coastal harbors accessible'  
\end{exe} 

\noindent
Also here the bracketed phrases have the external distribution of an NP, 
taking the subject position in (\ref{schenk}) and 
the complement position of a transitive verb in (\ref{mobil}). 
And also here, the internal structure shows a mix of nominal and verbal
characteristics. 
Typically nominal are the presence of the article and the postnominal PP
complement in (\ref{mobil}). 
Typically verbal are the presence of the direct object complement in (\ref{schenk})
and the predicative PP complement in (\ref{mobil}). 
To model this \citet{VanEynde19} makes a distinction between the verbal core 
and the nominal fringe of a nominalized infinitive, as in the structure of 
(\ref{mobil}), spelled out in Figure \ref{kust}.  

\begin{figure}
	\centering
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{noun}, {\sc comps} ~ \eliste]}
	[N [het]]
	[{[{\sc head} ~ \avmbox{1}, {\sc comps} ~ \eliste]}
		[{[{\sc head} ~ \avmbox{1}, {\sc comps} ~ \liste{\avmbox{2} PP$_{j}$}]}
			[{[{\sc head} ~ \avmbox{4} \type{verb}, {\sc comps} ~ \liste{NP$_{j}$}]}
			[\avmbox{3} [op diepte, roof]]
			[{[{\sc head} ~ \avmbox{4}, {\sc comps} ~ \liste{NP$_{j}$, \avmbox{3} PP$_{k}$}]} [houden]]]]
		[\avmbox{2}, before drawing tree={x+=3em} [van de kusthavens, roof, before drawing tree={x+=3em}]]]]
	\end{forest}
	\caption{\label{kust} Dutch nominalized infinitive}
\end{figure}   

The infinitive is treated as unambiguously verbal at the lexical level and 
remains verbal when combined with its predicative PP complement, but then it  
is converted into a nominal projection and combined with a postnominal PP[\type{van}]
complement and the definite article.\footnote{Since the article is a pronominal 
determiner, it is assigned the category N in the functor treatment.}     
The conversion is modelled in terms of a non-headed phrase type, 
since it does not comply with the Head Feature Principle: 
The {\sc head} value of the mother is not shared with the daughter.  
More specifically, there is a subtype of the non-headed phrases, called \type{convert-phr}, 
whose defining characteristic is that they have a single daughter. 
In that respect they differ from coordinate phrases, which have at least two daughters.  

\begin{exe} 
\ex\label{conv} 
\begin{avm} 
[\type{convert-phr} \\
 dtrs ~ <X> ]
\end{avm}
\end{exe}

\noindent
The conversion which we observe in the Dutch nominalized infinitive is modeled 
in terms of a phrase type that inherits properties of \type{convert-phr} and 
\type{non-clause}. Its properties are spelled out in (\ref{nominf}). 

\begin{figure}
	\centering
	\begin{forest}
%sm edges
[\type{phrase}
	[{\sc headedness}
		[\type{non-headed-phr}
			[\type{coord-phr}]
			[\type{convert-phr} [\type{nom-inf-phr}, name=A2]]
			[...]]]
	[{\sc clausality}
		[\type{clause}]
		[\type{non-clause}, name=A1]]]
\draw (A1.south) -- (A2.north);
	\end{forest}
	\caption{\label{bido6} Nominalized Infinitives in the Bi-dimensional Hierarchy  } 
\end{figure}

\begin{exe} 
\ex\label{nominf} 
\begin{avm} 
[\type{nom-inf-phr}                                      \\
 synsem|cat [head [\type{noun}                           \\
                   number ~ \type{singular}              \\
                   gender ~ \type{neuter}]               \\
             subj ~ < ~ >                                 \\
             comps ~ <(pp$_{j}$)> ~ $\oplus$ ~ <(pp$_{i}$)> ~ $\oplus$ ~ @A \\ 
             marking ~ \type{unmarked}]                     \\           
 dtrs ~ <[synsem|cat [head|vform ~ \type{inf}               \\
                      subj ~ <np$_{j}$>                         \\
                      comps ~ <(np$_{i}$)> ~ $\oplus$ ~ @A ]]>]
\end{avm}
\end{exe}

\noindent
The daughter is an infinitive with a non-empty {\sc subj} list and 
a possibly empty {\sc comps} list. $\avmbox{A}$  
The mother is an unmarked singular neuter nominal that inherits the 
unsaturated complement requirements of its daughter, 
albeit with the twist that NP complements become PP complements. 
The subject requirement of the infinitive is added to the {\sc comps} list 
of the nominal and becomes a PP too. It is made optional, since 
it is often left unexpressed, as in (\ref{schenk}--\ref{mobil}). If present,  
it is introduced by \emph{van} `of' or \emph{door} `by', as in 
(\ref{van}) and (\ref{door}) respectively. 

\begin{exe} 
\ex\label{van} 
\gll   het trage afsterven van de koraalriffen   \\
       the slow die.{\sc inf} of the coral.reefs \\
\trans `the slow dying of the coral reefs' 
\ex\label{door}
\gll   het uitschakelen van Chelsea door Real Madrid      \\
       the eliminate.{\sc inf} of Chelsea by Real Madrid  \\ 
\trans `the elimination of Chelsea by Real Madrid'  
\end{exe} 

For a full treatment that also covers the semantic type shift that 
accompanies the syntactic conversion, we refer to \citet{VanEynde19}.


\subsection{Idiosyncratic P+NP combinations} 


When a nominal combines with a preposition the result is usually a PP, but 
not always. The French \emph{de} `of', for instance, heads a PP in 
\emph{je viens de Roubaix} `I come from Roubaix', 
but not in \emph{beaucoup de farine} `much flour'. 
As argued in \citet{Abeilleetal04}, there are many differences 
between these two uses of \emph{de}, both syntactic and semantic ones.
To model them they treat the former as the ordinary head of a PP and 
the latter as a weak head. Typical of a weak head is that it shares 
nearly all properties of its complement, as spelled out in (\ref{de}).

\begin{exe} 
\ex\label{de} 
\begin{avm} 
[cat [head ~ @1                              \\
      subj ~ @A                              \\
      spr ~ @B                               \\
      comps ~ <[cat [head ~ @1               \\
                     subj ~ @A               \\
                     spr ~ @B                \\
                     comps ~ < ~ >]          \\
                     mark ~ \type{unmarked} \\
                content ~ @4 ]>              \\ 
      mark ~ \type{de}]                      \\
 content ~ @4 ] 
\end{avm}
\end{exe} 

\noindent
\emph{de} has the same values for {\sc head}, {\sc subj}, {\sc spr} and 
{\sc content} as its complement. 
When combined with \emph{farine} `flour', it is, hence, a noun 
that selects a specifier and that denotes a parameter.\footnote{The 
weak \emph{de} also combines with infinitival VPs, as in \emph{de sortir un peu
te ferait du bien} `getting out a bit would do you some good'. In that 
combination, \emph{de} is a verb that selects a subject and that denotes 
a state-of-affairs.} The only difference between the weak head and its complement  
concerns the {\sc marking} value: \emph{de} requires an unmarked complement, but 
its own {\sc marking} value is of type \emph{de}, and this value is shared with the mother.  
Specifiers that require the presence of \emph{de}, such as \emph{beaucoup} `much/many', 
select a nominal with that {\sc marking} value, 
as shown in (\ref{coup}).\footnote{In this AVM, quoted from \citet[18]{Abeilleetal04}, 
the value of {\sc spec} is of type \type{synsem}, as in \citet{ps2}, and not of type 
\type{semantic-object}, as in \citet{GS00}.}  

\begin{exe} 
\ex\label{coup} 
\begin{avm} 
[cat|head [\type{adverb}                                   \\
           spec|loc [cat [head ~ \type{noun}               \\
                          spr ~ <X>                         \\
                          mark ~ \type{de}]                \\
                     content [index ~ @2                    \\
                              restr ~ \avmbox{$\Sigma$} ]]] \\ 
 content ~ @1 [\type{beaucoup-rel}        \\
               index ~ @2                  \\
               restr ~ \avmbox{$\Sigma$} ] \\
 store ~ \{ @1 \} ]
\end{avm}
\end{exe} 

\noindent
Besides, the selected nominal is required to be unsaturated for {\sc spr} and 
the quantifier is required to share the index and restrictions of its head sister. 

The weak head analysis of \citet{Abeilleetal04} fits the mold of the specifier treatment. 
The same data can be analyzed in a way that fits the mold of the functor treatment,
as shown in Figure \ref{beau}. 
In this analysis the preposition in \emph{beaucoup de farine} `much flour'
is a functor that selects a nominal of type \type{bare} and that 
shares its own {\sc marking} value (\emph{de}) with the mother. 
The functor \emph{beaucoup} in its turn selects a nominal with the {\sc marking} value 
\type{de} and shares its own {\sc marking} value with the NP as a whole. 
In this analysis \emph{de} does not share the part-of-speech, valence and 
meaning of its nominal sister. Instead, those properties are shared directly between
\emph{farine} and \emph{de farine}. 

\begin{figure}
	\centering
\oneline{%
	\begin{forest}
sm edges
[{[{\sc head} ~ \avmbox{1} \type{noun}[{\sc sel} ~ \type{none}], {\sc mark} ~ \avmbox{2} \type{marked}]}
	[{[{\sc head} ~ [\type{adv} {\sc sel} ~ \avmbox{4}], {\sc mark} ~ \avmbox{2}]} [beaucoup]]
	[{\avmbox{4} [{\sc head} ~ \avmbox{1}, {\sc mark} ~ \avmbox{5} \type{de}]}
		[{[{\sc head} ~ [\type{prep} {\sc sel} ~ \avmbox{3}], {\sc mark} ~ \avmbox{5}]} [de]]
		[{\avmbox{3} [{\sc head} ~ \avmbox{1}, {\sc mark} ~ \type{bare}]} [farine]]]]
	\end{forest}
}
	\caption{\label{beau} A prepositional functor}
\end{figure}

While the differences between the weak head treatment and the functor treatment are small and subtle, 
they may have empirical consequences. This is shown in \citet{Maekawa15}, which offers 
an analysis of English nominals of the \emph{kind}/""\emph{type}/""\emph{sort} variety.  
To account for the fact that the determiner shows agreement with the 
rightmost noun in \emph{these sort of problems} and \emph{those kind of pitch changes}
Maekawa considers the option of treating \emph{of} and the immediately preceding noun as 
weak heads, but considers this unsatisfactory, since it has the unwanted effect of treating 
\emph{sort/kind/type} as plural. As an alternative, he develops an analysis in which \emph{of} and 
the preceding noun are functors \citep[149]{Maekawa15}. This avoids the unwanted side-effect, 
since functors do not share the {\sc head} value of their head sister. 
Further evidence is provided in \citet{Maekawa16}, which analyzes combinations of a   
singular numeral with a plural noun, as in \emph{these thousand pages}. If 
\emph{thousand} is treated as a weak head, it inadvertantly inherits the number value of
\emph{pages}, i.e. plural, whereas it is in fact singular, its plural counterpart being 
\emph{thousands}. This unwanted effect is avoided if \emph{thousand} is treated as a functor.


\subsection{Other nominals with idiosyncratic properties} 


A much studied nominal with idiosyncratic traits is the binominal noun phrase 
construction (BNPC), exemplified in (\ref{climb}). 

\begin{exe}
\ex\label{climb}
\begin{xlist}
\ex  She blames it on [her nitwit of a husband]. 
\ex  She had [a skullcracker of a headache]. 
\end{xlist}
\end{exe}

\noindent
In contrast to ordinary [NP--\emph{of}--NP] sequences, 
as in \emph{an employee of a Japanese company}, where the 
first nominal is the head of the entire NP, and where the second 
nominal is part of its PP adjunct, the {\sc bnpc} shows a pattern
in which the relation between the nominals is a predicative one: 
her husband is claimed to be a nitwit, and the headache is claimed to be 
like a skullcracker. HPSG treatments of the {\sc bnpc} are provided in
\citet{KimSells14} and \citet{VanEynde18}. The latter uses 
the phrase type hierarchy in (\ref{prot}), defining the {\sc bnpc} as 
a maximal type that inherits from \type{head-independent-phr} and 
\type{inverted-predication}. To capture the intuition that the second
nominal is the head of the entire NP, the preposition \emph{of} is 
not treated as the head of a PP, but as a functor that selects 
a nominal head. 

Comparable to NPs with a verbal core are NPs with an adjectival core,
such as \emph{the very poor} and \emph{the merely skeptical}. They are  
described and provided with an HPSG analysis in \citet{ArnoldSpencer2015}.
Interestingly, it employs a device for phrasal conversion that is similar to 
the one for the Dutch nominalized infinitive in Section 3.2.2. 

Idiosyncratic are also the nominals with an extracted determiner, as in 
the French (\ref{combi}) and the Dutch (\ref{gelu}). 

\begin{exe} 
\ex\label{combi}  
\gll   Combien as-tu lu [\_\_ de livres en latin] ?    \\
       how.many have-you read [\_\_ of books in Latin] \\ 
\trans `How many books have you read in Latin?' 
\ex\label{gelu}
\gll   Wat zijn dat [\_\_ voor vreemde geluiden] ? \\
       what are that [\_\_ for strange noises]     \\ 
\trans `What kind of strange noises are those?'  
\end{exe} 

\noindent
The French example is analyzed in \citet[20--21]{Abeilleetal04} and the Dutch one in 
\citet[47--50]{VanEynde04}. 

Another special case is apposition. It comes in (at least) two types, known as 
close apposition and loose apposition. Relevant examples are respectively 
\emph{my brother Richard is a soldier} and \emph{Sarajevo, the capital of Bosnia, is where WW I began}.
Both are compared and analyzed in \citet{Kim12} and \citet{Kim14}. 
\citet{VanEyndeKim16} provides an analysis of loose apposition in
Sign-Based Construction Grammar. 



\section{Conclusion} 


This chapter has provided a survey of how nominals are analyzed in HPSG. 
Over time three treatments have taken shape, i.e. the specifier treatment, 
the functor treatment and the DP treatment. They were presented and applied to ordinary 
nominals in Section 2. Nominals with idiosyncratic properties were discussed in Section 3.
Modeling them requires a relaxation of the strictly lexicalist stance of early HPSG.  
The more flexible tools of constructional HPSG have been put to use in the analysis 
of the Big Mess Construction and of nominals with a verbal core, such as gerunds and 
nominalized infinitives. For other nominals with idiosyncratic properties pointers 
have been given to relevant literature. 

In terms of the NP vs. DP debate, it is clear that the NP approach has been 
more successful in HPSG than the DP approach. This fits in with the tendency to 
avoid the postulation of functional categories with their respective projections. 
Clauses, for instance, are not analyzed as IPs either. The finer-grained 
differences between the specifier treatment and the functor treatment are 
a topic of ongoing debate. They both have their advocates and for the analysis of 
ordinary nominals there does not seem to be any evidence that decisively tilts the balance. 
For nominals with idiosyncratic properties, however, there is a clear tendency in the recent 
literature to adopt the functor treatment, usually in combination with phrasal constraints.    


 
%\section*{Abbreviations}
\section*{Acknowledgements}

{\sloppy
\printbibliography[heading=subbibliography,notkeyword=this] 
}
\end{document}
